### 2015-12-23 Schloss_PacBio16S_PeerJ_2016.Rmd
Copy over 2015 version of manuscript

---
title: "**Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: mbio.csl
fontsize: 11pt
geometry: margin=1.0in
---

```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = FALSE)

inline_hook <- function(x){
	print(x)
	if(is.numeric(x)){
		if(abs(x - round(x)) < .Machine$double.eps^0.5){
			paste(format(x,big.mark=',', digits=0, scientific=FALSE))
		} else {
			paste(format(x,big.mark=',', digits=1, nsmall=1, scientific=FALSE))
		}

	} else {
    	paste(x)      
	}
}
knitr::knit_hooks$set(inline=inline_hook)

```

\vspace{35mm}

**Running title:** 16S rRNA genes sequencing with PacBio

\vspace{15mm}

**Authors:** Patrick D. Schloss^1#^, Sarah L. Westcott^1^, Matthew L. Jenior^1^,
and Sarah K. Highlander^2^

\vspace{40mm}

$\dagger$ To whom correspondence should be addressed: pschloss@umich.edu

1\. Department of Microbiology and Immunology, 1500 W. Medical Center, University of Michigan, Ann Arbor, MI 48109

2\. Genomic Medicine, J. Craig Venter Institute, 4120 Capricorn Lane, La Jolla, CA 92307

\newpage
\linenumbers

## Abstract
Over the past 10 years, microbial ecologists have largely abandoned sequencing
16S rRNA genes by the Sanger sequencing method and have instead adopted highly
parallelized sequencing platforms. These new platforms, such as 454 and
Illumina's MiSeq, have allowed researchers to obtain millions of high quality,
but short sequences. These platforms have allowed researchers to significantly
improve the design of their experiments. The tradeoff has been the decline in
the number of full-length reference sequences that are deposited into databases.
To overcome this problem, we tested the ability of the PacBio Single Molecule,
Real-Time (SMRT) DNA sequencing platform to generate sequence reads from the 16S
rRNA gene. We generated sequencing data from the V4, V3-V5, V1-V3, V1-V6, and
V1-V9 variable regions from within the 16S rRNA gene from a synthetic mock
community and natural samples collected from human feces, mouse feces, and soil.
The synthetic mock community allowed us to assess the actual sequencing error
rate and how that error rate changed when different curation methods were
applied. We developed a simple method based on sequence characteristics and
quality scores to reduce the observed error rate for the V1-V9 region from 2.16%
to 0.32%. Unfortunately, this error rate was still 16-times higher than the
error rate that has been observed for the shorter reads generated by 454 and
Illumina's MiSeq sequencing platforms. Although the longer reads frequently
provided better classification, the wider adoption of this approach for 16S rRNA
gene sequencing is likely limited by its high sequencing error and low yield of
sequencing data relative to the other available platforms.

***Keywords:*** Microbial ecology, bioinformatics, sequencing error

## Introduction
Advances in sequencing technologies over the past 10 years have introduced
considerable advances to the field of microbial ecology. Clone-based Sanger
sequencing of the 16S rRNA gene has largely been replaced by various platforms
produced by 454/Roche {e.g. Sogin, 2006 #1510}, Illumina
{e.g. Gloor, 2010 #2611}, and IonTorrent {e.g. Junemann, 2012 #2630}. It was
once common to sequence fewer than 100 16S rRNA gene sequences from several
samples using the Sanger approach {e.g. McCaig, 1999 #592}. Now it is common to
generate thousands of sequences from each of several hundred samples
{The Human Microbiome Consortium, 2012 #2617}. The advance in throughput has
come at the cost of read length. Sanger sequencing regularly generated 800 nt
per read and because the DNA was cloned, it was possible to obtain multiple
reads per fragment to yield a full-length sequence from a representative single
molecule. At approximately $8 (US) per sequencing read, most
researchers have effectively decided that full-length sequences are not worth
the increased cost relative to the cost of more recently developed approaches.
There is still a clear need to generate high-throughput full-length sequence
reads that are of sufficient quality that they can be used as references for
analyses based on obtaining short sequence reads.

Historically, all sequencing platforms were created to primarily perform genome
sequencing. When sequencing a genome, it is assumed that the same base of DNA
will be sequenced multiple times and the consensus of multiple sequence reads is
used to generate contigs. Thus, although an individual base call may have a high
error rate, the consensus sequence will have a low error rate. To sequence the
16S rRNA gene researchers use conserved primers to amplify a sub-region from
within the gene that is isolated from many organisms. Because the fragments
are not cloned, it is not possible to obtain high sequence coverage from the
same DNA molecule using these platforms. Thus, to reduce sequencing error rates
it has become imperative to develop stringent sequence curation and denoising
algorithms {Kozich, 2013 #2719; Schloss, 2011 #2466}.
There has been a tradeoff between read length, number of reads per sample, and
the error rate. For instance, we recently demonstrated that using the Illumina
MiSeq and the 454 Titanium platforms the raw error rate varies between 1
and 2% {Kozich, 2013 #2719; Schloss, 2011 #2466}. Yet, it was possible to obtain
error rates below 0.02% by adopting various denoising algorithms. However, the
resulting fragments were
only 250-nt long. In the case of 454 Titanium, extending the length of the
fragment introduces length-based errors and in the case of the Illumina MiSeq,
increasing the length of the fragment reduces the overlap between the read pairs
reducing the ability of each read to mutually reduce the sequencing error.
Inadequate denoising of sequencing reads can have many negative effects
including limited ability to identify chimeras
{Haas, 2011 #2404;Edgar, 2011 #2406} and inflation of alpha- and beta-diversity
metrics {Huse, 2010 #2034; Kunin, 2010 #2259; Kozich, 2013 #2719; Schloss, 2011 #2466}.
Although MiSeq and 454 enjoy widespread use in the field, the MiSeq
platform is emerging as the leader because of the ability to sequence 15-20
million fragments that can be distributed across hundreds of samples for less
than $5000 (US).

As these sequencing platforms have grown in popularity, there has been a decline
in the number of full-length 16S rRNA genes being deposited into GenBank that
could serve as references. This is particularly frustrating since the
technologies have significantly improved our ability to detect and identify
novel populations for which we lack full-length reference sequences. A related
problem is the perceived limitation that the short reads generated by the 454
and Illumina platforms cannot be reliably classified to the genus or
species level. Previous investigators have utilized simulations to
demonstrate that increased read lengths usually increase the accuracy and
sensitivity of classification against reference databases
{Liu, 2008 #2188;Werner, 2012 #2474;Wang, 2007 #1809}. There is
clearly a need to develop sequencing technologies that will allow researchers to
generate high quality full-length 16S rRNA gene sequences in a high throughput
manner.

New advances in single molecule sequencing technologies, such as the platform
produced by Pacific Biosciences (PacBio), offer the opportunity to once again
obtain full-length sequence reads with a high depth of coverage from a large
number of samples. To this point, the PacBio Single Molecule, Real-Time (SMRT)
DNA Sequencing System has received limited application in the microbial ecology
research domain {Mosher, 2014 #3376; Fichot, 2013 #3453; Mosher, 2013 #3454}.
The SMRT system ligates hairpin adapters (i.e. SMRTbells)
to the ends of double-stranded DNA. Although the DNA molecule is linear, it is
effectively circularized allowing the sequencing polymerase to process around
the molecule multiple times {Au, 2012 #3455}. According to Pacific Biosciences
the platform is able to generate median read lengths longer than 8 kb with the
P4-C2 chemistry; however, the single pass error rate is approximately 15%. Given
the circular nature of the DNA fragment, the full read length can be used to
cover the DNA fragment multiple times resulting in a reduced error rate.
Therefore, one should be able to obtain multiple coverage of the full 16S rRNA
gene at a reduced error rate.

Despite the opportunity to potentially generate high-quality full-length
sequences, the Pacific Biosciences platform has not been widely adopted for
sequencing 16S rRNA genes {Fichot, 2013 #3453; Mosher, 2014 #3376; Mosher, 2014 #3376}.
Previous studies utilizing the technology have removed reads with mismatched
primers and barcodes, ambiguous base  calls, and low quality scores
{Fichot, 2013 #3453}. Others have utilized the platform without describing the
bioinformatic pipeline that was utilized {Mosher, 2014 #3376; Mosher, 2014 #3376}.
Regardless of the curation methods, the error rates associated with sequencing
the 16S rRNA gene on the platform have never been reported. In the current
study, we assessed the quality of data generated by the Pacific Biosciences
sequencer and whether it could fill the need for generating high-quality,
full-length sequence data. We hypothesized that by modulating the 16S rRNA gene
fragment length we could alter the read depth and obtain reads longer than are
currently available by the 454 and Illumina platforms but with the same quality.
To test this hypothesis, we developed a sequence curation pipeline that was
optimized by reducing the sequencing error rate of a mock bacterial community
with known composition. The resulting pipeline was then applied to 16S rRNA gene
fragments that were isolated from soil and human and mouse feces.

## Materials and Methods
***Community DNA.*** We utilized genomic DNA isolated from four communities.
These same DNA extracts were previously used to develop an Illumina MiSeq-based
sequencing strategy {Kozich, 2013 #2719}.  Briefly, we used a “Mock Community” composed of
genomic DNA from 21 bacterial strains: *Acinetobacter baumannii* ATCC 17978,
*Actinomyces odontolyticus* ATCC 17982, *Bacillus cereus* ATCC 10987,
*Bacteroides vulgatus* ATCC 8482, *Clostridium beijerinckii* ATCC 51743,
*Deinococcus radiodurans* ATCC 13939, *Enterococcus faecalis* ATCC 47077,
*Escherichia coli* ATCC 70096, *Helicobacter pylori* ATCC 700392, *Lactobacillus
gasseri* ATCC 33323, *Listeria monocytogenes* ATCC BAA-679, *Neisseria
meningitidis* ATCC BAA-335, *Porphyromonas gingivalis* ATCC 33277,
*Propionibacterium acnes* DSM 16379, *Pseudomonas aeruginosa* ATCC 47085,
*Rhodobacter sphaeroides* ATCC 17023, *Staphylococcus aureus* ATCC BAA-1718,
*Staphylococcus epidermidis* ATCC 12228, *Streptococcus agalactiae* ATCC
BAA-611, *Streptococcus mutans* ATCC 700610, *Streptococcus pneumoniae* ATCC
BAA-334. The mock community DNA is available through BEI resources
(v3.1, HM-278D). Genomic DNAs from the three other communities were obtained
using the MO BIO PowerSoil DNA extraction kit. The human and mouse fecal samples
were obtained using protocols that were reviewed and approved by the University
Committee on Use and Care of Animals (Protocol #PRO00004877) and the Institutional Review Board at the
University of Michigan (Protocol #HUM00057066). The human stool donor provided informed consent.

***Library generation and sequencing.***
The DNAs were each amplified in triplicate using barcoded primers targeting the
V4, V1-V3, V3-V5, V1-V5, V1-V6, and V1-V9 variable regions (Table 1). The
primers were synthesized so that the 5’ end of the forward and reverse primers
were each tagged with a 5-nt barcode sequence to allow multiplexing of samples
within a single sequencing run. Methods describing PCR, amplicon cleanup, and
pooling were described previously {Kozich, 2013 #2719}. The SMRTbell adapters
were ligated onto  the PCR products and the libraries were sequenced at the
University of Michigan DNA Sequencing Core using the P4-C2 chemistry on a PacBio
RS II SMRT DNA Sequencing System.

***Data analysis.***
All sequencing data were curated using mothur {Schloss, 2009 #1816} and
analyzed using the R programming language {R Core Team, 2014 #3456}. The raw
data can be obtained from the Sequence Read Archive at NCBI under accession
SRP051686, which are associated with BioProject PRJNA271568. Several specific
features were incorporated into mothur to
facilitate the analysis of PacBio  sequence data.  First, because non-ambiguous
base calls are assigned to Phred quality scores of zero, the consensus fastq
files were parsed so that scores of zero were interpreted as corresponding to an
ambiguous base call (i.e. N) in the  fastq.info command using the pacbio=T
option. Second, because the consensus  sequence can be generated in the forward
and reverse complement orientations, a checkorient option was added to the
trim.seqs command in order to identify the  proper orientation. These features
were incorporated into mothur v.1.30. Because chimeric molecules can be
generated during PCR and would artificially inflate the sequencing error, it was
necessary to remove these data prior to assessing the error rate. Because we
knew the true sequences for the strains in the mock community we could calculate
all possible chimeras between strains in the mock community (*in silico*
chimeras). If a sequence read was 3 or more nucleotides more similar to an *in
silico* chimera than it was to a non-chimeric reference sequence, it was
classified as a chimera and removed from further consideration. Identification
of *in silico* chimeras and calculation of sequencing error rates
was performed using the seq.error command in mothur {Schloss, 2011 #2466}. *De
novo* chimera detection was also performed on the mock and other sequence data
using the abundance-based algorithm implemented in UCHIME {Edgar, 2011 #2406}.
Sequences sequences were aligned against a SILVA-based reference alignment
{Pruesse, 2007 #1735} using a profile-based aligner {Schloss, 2009 #1825} and
were classified against the SILVA {Pruesse, 2007 #1735}, RDP {Cole, 2014 #3460},
and greengenes {Werner, 2012 #2474} reference taxonomies  using a
naive Bayesian classifier {Wang, 2007 #1809}. Sequences were assigned to
operational taxonomic units using the average neighbor clustering algorithm with
a 3% distance threshold {Schloss, 2011 #3156}.  Detailed methods including this
paper as an R markdown file are available as a public online repository
(http://github.com/SchlossLab/ Schloss_PacBio16S_PeerJ_2015).

```{r setup, echo=FALSE, eval=TRUE, tidy=TRUE}
rm(list=ls())
library("knitr")
#pdf.options(encoding="ISOLatin2.enc")
regions <- c("v4", "v13", "v35", "v15", "v16", "v19")
clrs <- rainbow(length(regions))
pretty.region <- c("v13"="V1-V3", "v15"="V1-V5", "v16"="V1-V6", "v19"="V1-V9", "v35"="V3-V5", "v4"="V4")
names(clrs) <- regions

opts_chunk$set("fig.path"="figures/")
opts_chunk$set("fig.align"="center")
opts_chunk$set("dev" = c("png", "cairo_pdf"))

opts_chunk$set("tidy" = TRUE)
opts_chunk$set("echo" = FALSE)
opts_chunk$set("eval" = TRUE)
opts_chunk$set("results" = 'hide')
opts_chunk$set("warning" = FALSE)
opts_chunk$set("cache" = TRUE)
```

## Results and Discussion

```{r error_profile}
sub.matrix <- read.table(file=paste("pipeline_dev/", regions[1], "/", regions[1], ".mock.filter.error.matrix", sep=""), header=T, row.names=1)[-7,]
for(i in 2:length(regions)){
	sub.matrix <- sub.matrix + read.table(file=paste("pipeline_dev/", regions[i], "/", regions[i], ".mock.filter.error.matrix", sep=""), header=T, row.names=1)[-7,]
}

matches <- c(sub.matrix[1,1], sub.matrix[2,2], sub.matrix[3,3], sub.matrix[4,4])
total <- sum(sub.matrix)
errors <- sub.matrix
diag(errors[1:4,1:4]) <- 0
init.error.rate <- 100 * sum(errors) / total				#1.925717

subst.bias <- apply(errors[1:4,1:4], 2, sum)/apply(sub.matrix[1:4,1:4], 2, sum)
subst.bias <- 100 * subst.bias/sum(subst.bias)
#       rA        rT        rG        rC
# 22.43918 27.21620 24.06557 26.27904
#these are more or less equal to each other

insert.bias <- (sub.matrix[,"rGap"]/apply(sub.matrix, 1, sum))[1:4]
insert.bias <- 100 * insert.bias/sum(insert.bias)
#       qA       qT       qG       qC
# 23.46112 24.36667 24.33064 27.84157
#these are more or less equal to each other

delete.bias <- (sub.matrix["qGap",]/apply(sub.matrix, 2, sum))[1:4]
delete.bias <- 100 * delete.bias/sum(delete.bias)
#           rA      rT       rG       rC
#qGap 11.43234 9.46907 44.57506 34.52353
#interestingly the G's and C's are more likely to be deleted than the As or Ts

substitutions <- 100*sum(errors[1:4,1:4])/sum(errors)	#35.80084
insertions <- 100*sum(errors[,5])/sum(errors)			#45.27368
deletions <- 100*sum(errors[5,])/sum(errors)			#17.27259
ambiguous <- 100*sum(errors[6,])/sum(errors)			#2.101505

# Need to connect error types with the quality scores
error.quality <- read.table(file=paste("pipeline_dev/", regions[1], "/", regions[1], ".mock.filter.error.quality", sep=""), header=T, row.names=1)[-7,]
for(i in 2:length(regions)){
	error.quality <- error.quality + read.table(file=paste("pipeline_dev/", regions[i], "/", regions[i], ".mock.filter.error.quality", sep=""), header=T, row.names=1)[-7,]
}
error.quality["72",] <- error.quality["72",] + error.quality["80",]
error.quality <- error.quality[1:72,]

#get quantiles
matches <- rep(as.numeric(rownames(error.quality)), error.quality$matches)
m.quant <- quantile(matches, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))

subs <- rep(as.numeric(rownames(error.quality)), error.quality$substitutions)
s.quant <- quantile(subs, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))

ins <- rep(as.numeric(rownames(error.quality)), error.quality$insertions)
i.quant <- quantile(ins, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))

ambig <- rep(as.numeric(rownames(error.quality)), error.quality$ambiguous)
a.quant <- quantile(ambig, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
```
***The PacBio error profile.***
To build a sequence curation pipeline, we first needed to characterize the error
rate associated with sequencing the 16S rRNA gene. We observed an average
sequencing error rate of `r format(init.error.rate, digits=3, nsmall=2)`%. Insertions, deletions, substitutions, and ambiguous base calls
accounted for `r format(insertions, digits=3, nsmall=1)`,
`r format(deletions, digits=3, nsmall=1)`, `r format(substitutions, digits=3, nsmall=1)`,
and `r format(ambiguous, digits=2, nsmall=1)`% of
the errors, respectively. The substitution errors were equally likely and all four bases
were equally likely to cause insertion errors. Interestingly, guanines
(`r format(delete.bias["rG"], digits=3, nsmall=1)`%) and cytosines
(`r format(delete.bias["rC"], digits=3, nsmall=1)`%) were more likely to be deleted
than adenines (`r format(delete.bias["rA"], digits=3, nsmall=1)`%) or thymidines
(`r format(delete.bias["rT"], digits=2, nsmall=1)`%). When we considered the Phred
quality score of each base call, we observed a median quality score of `r m.quant["50%"]`
for correct base calls and scores of `r s.quant["50%"]` and `r i.quant["50%"]` for
substitutions and insertions, respectively (Figure 1A). Although there
was a broad distribution of quality scores with each type of base call, the errors could
largely be distinguished from the correct base calls.

```{r initialError}
getInitError <- function(folder){
	error <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".mock.filter.error.summary", sep=""), header=T, row.names=1)
	error <- error[error$numparents==1,]
	return(sum(error$mismatches)/sum(error$total))
}

init.rates <- unlist(lapply(regions, getInitError))
names(init.rates) <- regions

getReasonsLost <- function(region){
	composite <- read.table(file=paste("pipeline_dev/", region, "/", region, ".composite", sep=""), header=T, row.names=1)
	reasons <- as.character(composite$reason)

	nseqs <- length(reasons)
	good <- sum(reasons=="g")	/ nseqs
	start.stop <- sum(grepl("s", reasons) | grepl("e", reasons)) / nseqs
	homop <- length(grep("h", reasons)) / nseqs
	ambig <- length(grep("n", reasons)) / nseqs
	return(c(good=good, start.stop=start.stop, homop=homop, ambig=ambig, ngood=nseqs))
}

reasons.lost <- matrix(unlist(lapply(regions, getReasonsLost)), ncol=5, byrow=T)
colnames(reasons.lost) <- c("good", "start.stop", "homop", "ambig", "nseqs")
rownames(reasons.lost) <- regions

orig.nseqs <- reasons.lost[,"nseqs"]
basic.nseqs <- orig.nseqs * reasons.lost[,1]
reasons.lost <- reasons.lost[,-5]

getBasicError <- function(folder){
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
	good <- composite[composite$reason == "g",]
	return(sum(good$mismatches)/sum(good$total))
}

basic.error <- unlist(lapply(regions, getBasicError))
names(basic.error) <- regions
#         v4         v13         v35         v15         v16         v19
#0.005373722 0.014743697 0.013730112 0.009027326 0.014249660 0.013027623
composite.basic.error <- 100 * sum(basic.error*basic.nseqs)/(sum(basic.nseqs))
```

***A basic sequence curation procedure.***
To establish a simple curation procedure, we culled any sequence that contained
an ambiguous base call, had a string of the same base repeated 9 or more times,
did not start and end at the expected alignment coordinates for that region
of the 16S rRNA gene, or that was chimeric. This reduced the experiment-wide
error rate from `r format(init.error.rate, digits=3, nsmall=2)` to
`r format(composite.basic.error, digits=2, nsmall=2)`%. This basic procedure
resulted in the removal of between
`r format(round(min(100*(1-reasons.lost[,"good"])), digits=1), nsmall=1)`
(`r pretty.region[names(which.min(100*(1-reasons.lost[,"good"])))]`) and
`r round(max(100*(1-reasons.lost[,"good"])), digits=1)`
(`r pretty.region[names(which.max(100*(1-reasons.lost[,"good"])))]`)% of the
reads. The percentage of reads removed increased with the length of the
fragment (Figure 2). The number of reads removed
because of the presence of ambiguous base calls was similar to the number of
reads that were removed for not fully aligning to the correct region within the
16S rRNA gene (Table 2). The latter class of errors was
generally due to sequence truncations that could not be explained.

```{r barcodePrimerAnalysis}
getFragment_BPErrorRates <- function(folder){
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]
	bc.primer <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs

	return(cbind(composite$error, bc.primer))
}

errors <- lapply(regions, getFragment_BPErrorRates)

errors.table <- matrix(errors[[1]], ncol=2, byrow=F)
err.bcprimer <- aggregate(100*errors.table[,1], by=list(errors.table[,2]),
													function(x){c(mean(x), length(x))})

getBCPrimerError <- function(folder){
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
																		 ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]
	mismatch <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs

	good0 <- mismatch == 0
	composite.good0 <- composite[good0,]

	good1 <- mismatch <= 1
	composite.good1 <- composite[good1,]

	return(c(mean(composite.good0[,"error"]), nrow(composite.good0)/nrow(composite),
	mean(composite.good1[,"error"]), nrow(composite.good1)/nrow(composite) ))
}

bcprimer.error <- matrix(unlist(lapply(regions, getBCPrimerError)), nrow=6, byrow=T)
rownames(bcprimer.error) <- regions
colnames(bcprimer.error) <- c("0.error", "0.frac", "1.error", "1.frac")

#         0.error    0.frac     1.error    1.frac
# v4  0.002616919 0.6417469 0.003711755 0.9081720
# v13 0.011102712 0.5799452 0.012493713 0.8675194
# v35 0.010838504 0.6464069 0.011901910 0.8934573
# v15 0.005375053 0.5079055 0.007029568 0.8338150
# v16 0.010374438 0.5456238 0.011893323 0.8609559
# v19 0.008543379 0.5072207 0.010086351 0.8237548
```
```{r coverageAnalysis}
errorAtCoverageCutoff <- function(folder){
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]

	total.seqs <- nrow(composite)

	deep <- composite$freq >= 10
	composite <- composite[deep,]

	covered.seqs <- nrow(composite)

	error <- sum(composite$mismatches)/sum(composite$total)
	frac.kept <- covered.seqs / total.seqs
	return(c(error=error, kept=frac.kept))
}

coverage.error <- unlist(lapply(regions, errorAtCoverageCutoff))
coverage.error <- matrix(coverage.error, ncol=2, byrow=T)
rownames(coverage.error) <- regions
colnames(coverage.error) <- c("10.error", "10.frac")

#       10.error   10.frac
#v4  0.002524566 0.7897874
#v13 0.010359113 0.6592051
#v35 0.010084961 0.6642832
#v15 0.006534609 0.6717103
#v16 0.010039601 0.4605835
#v19 0.009506114 0.5620395

coverage.reduction <- (1-coverage.error[,"10.error"]/basic.error)*100
#      v4      v13      v35      v15      v16      v19
#53.02015 29.73870 26.54859 27.61302 29.54498 27.03110
```
```{r qualityScoreAnalysis}
getErrorRateFromAveQ <- function(folder, threshold=60){
	write(folder, "")
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]

	rates <- composite[composite$aveQ >= threshold,]

	remaining.error <- mean(rates[,"error"])
	fraction.kept <- nrow(rates)/nrow(composite)

	return(c(remaining.error, fraction.kept))
}

aveq.error <- matrix(unlist(lapply(regions, getErrorRateFromAveQ)), ncol=2, byrow=T)
rownames(aveq.error) <- regions
colnames(aveq.error) <- c("60.error", "60.frac")

#       60.error   60.frac
#v4  0.002623098 0.8692667
#v13 0.009914371 0.7608497
#v35 0.009035060 0.7697533
#v15 0.004553350 0.7005270
#v16 0.008921902 0.6561142
#v19 0.005714256 0.5010315

quality.reduction <- (1-aveq.error[,"60.error"]/basic.error)*100
```
```{r oligosCoverageAnalysis}
oligosCoverage <- function(folder){
	write(folder, "")
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]

	mismatch <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs
	good0 <- mismatch == 0 & composite$freq >= 10
	composite.good0 <- composite[good0,]

	good1 <- mismatch <= 1 & composite$freq >= 10
	composite.good1 <- composite[good1,]

	return(c(mean(composite.good0[,"error"]), nrow(composite.good0)/nrow(composite),
	mean(composite.good1[,"error"]), nrow(composite.good1)/nrow(composite) ))
}

oligosCoverage.error <- matrix(unlist(lapply(regions, oligosCoverage)), ncol=4, byrow=T)
rownames(oligosCoverage.error) <- regions
colnames(oligosCoverage.error) <- c("0.10.error", "0.10.frac", "1.10.error", "1.10.frac")

#     0.10.error 0.10.frac  1.10.error 1.10.frac
#v4  0.001565052 0.5567846 0.002068184 0.7513111
#v13 0.009371691 0.4353586 0.009899256 0.6180905
#v35 0.009096211 0.4907043 0.009503092 0.6321058
#v15 0.004050098 0.3803978 0.005253236 0.5899354
#v16 0.007945951 0.2960894 0.008956478 0.4301676
#v19 0.006706336 0.3188918 0.007903900 0.4927793
```
```{r oligosQScoreAnalysis}
oligosQScore <- function(folder){
	write(folder, "")
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]

	mismatch <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs

	good <- mismatch == 0 & composite$aveQ >= 60
	composite.good0 <- composite[good,]

	good1 <- mismatch <= 1 & composite$aveQ >= 60
	composite.good1 <- composite[good1,]

	return(c(mean(composite.good0[,"error"]), nrow(composite.good0)/nrow(composite),
	mean(composite.good1[,"error"]), nrow(composite.good1)/nrow(composite) ))
}

oligosQScore.error <- matrix(unlist(lapply(regions, oligosQScore)), ncol=4, byrow=T)
rownames(oligosQScore.error) <- regions
colnames(oligosQScore.error) <- c("0.60.error", "0.60.frac", "1.60.error", "1.60.frac")

#     0.60.error 0.60.frac  1.60.error 1.60.frac
#v4  0.001715135 0.6015066 0.002215499 0.8232097
#v13 0.009153895 0.5013705 0.009700915 0.7108269
#v35 0.008365403 0.5557740 0.008690839 0.7281015
#v15 0.003217812 0.4102346 0.003993206 0.6282727
#v16 0.007589961 0.4115456 0.008358922 0.6101800
#v19 0.004870705 0.3209549 0.005294429 0.4653699
```
```{r coverageQScoreAnalysis}
coverageQScore <- function(folder){
	write(folder, "")
		composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
		composite <- composite[composite$reason == "g",]

	good <- composite$freq >= 10 & composite$aveQ>= 60
	composite.good <- composite[good,]

	return(c(mean(composite.good[,"error"]), nrow(composite.good)/nrow(composite) ))
}

coverageQScore.error <- matrix(unlist(lapply(regions, coverageQScore)), ncol=2, byrow=T)
rownames(coverageQScore.error) <- regions
colnames(coverageQScore.error) <- c("10.60.error", "10.60.frac")

#    10.60.error 10.60.frac
#v4  0.002361882  0.7874988
#v13 0.010016818  0.6541800
#v35 0.009296289  0.6564176
#v15 0.004481083  0.5824549
#v16 0.009612120  0.4587213
#v19 0.005692918  0.3884468
```
```{r allFiltersAnalysis}
allFilters <- function(folder){
write(folder, "")
		composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
		composite <- composite[composite$reason == "g",]
	mismatch <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs

good0 <- mismatch == 0 & composite$aveQ>= 60 & composite$freq >= 10
composite.good0 <- composite[good0,]

good1 <- mismatch <= 1 & composite$aveQ >= 60 & composite$freq >= 10
composite.good1 <- composite[good1,]

return(c(mean(composite.good0[,"error"]), nrow(composite.good0)/nrow(composite),
mean(composite.good1[,"error"]), nrow(composite.good1)/nrow(composite) ))
}

allFilters.error <- matrix(unlist(lapply(regions, allFilters)), ncol=4, byrow=T)
rownames(allFilters.error) <- regions
colnames(allFilters.error) <- c("0.10.60.error", "0.10.60.frac", "1.10.60.error", "1.10.60.frac")

#    0.10.60.error 0.10.60.frac 1.10.60.error 1.10.60.frac
#v4    0.001515500    0.5562125   0.001993325    0.7499762
#v13   0.009299146    0.4337597   0.009827404    0.6155779
#v35   0.008613200    0.4862353   0.008979207    0.6253128
#v15   0.003078476    0.3488609   0.003905562    0.5268616
#v16   0.007738039    0.2954687   0.008814923    0.4295469
#v19   0.004797594    0.2496316   0.005266467    0.3631005
```

***Identifying correlates of increased sequencing error.***
In contrast to the 454 and Illumina-based platforms where the sequencing
quality decays with length, the consensus sequencing approach employed by the
PacBio sequencer is thought to generate a uniform distribution of errors. This
makes it impossible to simply trim sequences to high quality regions. Therefore,
we sought to identify characteristics within sequences that would allow us to
identify and remove those sequences with errors using three different
approaches. First, we hypothesized that errors in the barcode and primer would
be correlated with the error rate for the entire sequence. We observed a strong
relationship between the number of mismatches to the barcodes and primers and the
error rate of the rest of the sequence fragment (Figure 1B). Although
allowing no mismatches to the barcodes and primers yielded the lowest error rate,
that stringent criterion removed a large fraction of the reads from the dataset and
allowing at most one mismatch marginally increased the error rate while preserving
more sequences in the dataset (Figure 2). Second, we hypothesized
that increased sequencing coverage should yield lower error rates. We found that
once we had obtained 10-fold coverage of the fragments, the error rate did not
change appreciably (Figure 1C). When we compared the error
rates of reads with at least 10-fold coverage to those with less coverage, we
reduced the error rate by
`r paste(round(sort(coverage.reduction)[c(1,5)], digits=1), collapse=" to ")`%
for each region except the V4 region for which the error rate was reduced by
`r round(coverage.reduction["v4"], digits=1)`%. Third, based on the earlier
analysis associating errors with quality scores, we used two quality score-based
approaches for identifying reads with errors (Figure 3). We calculated the
minimum average quality score across all 50-nt windows within each sequence and
we also calculated the average quality score across each sequence. We then
associated both methods of calculating the average quality score with the error
rate of the reads and the fraction of sequences that would be retained if each
threshold were selected. Using the sliding window approach we did not observe
any clear break points indicating that one quality score would be better than
another (Figure 3AB). In contrast, using the whole sequence quality score
average we observed a decrease in the error rate and the fraction of sequences
retained when the threshold was increased above 60 (Figure 3CD). When we used
this threshold, we were able to reduce the error rate by
`r paste(round(range(quality.reduction), digits=1), collapse=" to ")`%
(Figure 2A). We noted that the fraction of reads retained
decreased as the length of the fragment increased with retention of
`r round(100*aveq.error["v4", "60.frac"],digits=1)`% of the V4 reads and
`r round(100*aveq.error["v19", "60.frac"],digits=1)`% of the V1-V9 reads (Figure
2B). Next, we asked whether which combinations of culling reads with mismatches
to the expected barcodes and primers, less than 10-fold sequencing coverage, and
an average quality score less than 60 made the most meaningful reductions in the
error rate while preserving the most reads when implemented with the basic
curation pipeline (Figure 2B). We observed similar error rates when we required
one or fewer mismatches to the barcodes and primers and an average quality score
above 60 as when we also required a minimum 10-fold coverage. Culling sequences
that had more than one mismatch to the barcodes and primers and those with an
average quality score less than 60 reduced the error rate to between
`r paste(round(100*range(oligosQScore.error[,"1.60.error"]), digits=2), collapse=" and ")`.
This procedure resulted in the removal of
`r paste(rev(round(100*(1-range(oligosQScore.error[,"1.60.frac"])), digits=0)), collapse=" and ")`%
of the reads (Figure 2). The remainder of this paper uses this sequence curation
approach.

```{r precluster}
summarizeError <- function(folder, method){
  file1 <- paste("analysis/", folder, "/", folder, ".mock1.", method, ".error.summary", sep="")
  file2 <- paste("analysis/", folder, "/", folder, ".mock2.", method, ".error.summary", sep="")
  file3 <- paste("analysis/", folder, "/", folder, ".mock3.", method, ".error.summary", sep="")

  summary <- rbind(read.table(file=file1, header=T, row.names=1), read.table(file=file2, header=T, row.names=1), read.table(file=file3, header=T, row.names=1))
  nochim <- summary[summary$numparents==1,]
  error <- sum(nochim$weight * nochim$mismatches) / sum(nochim$weight * nochim$total)
  return(error)
}

unique.error <- unlist(lapply(regions, summarizeError, "unique"))
precluster.error <- unlist(lapply(regions, summarizeError, "precluster"))
final.error <- cbind(unique.error, precluster.error)
rownames(final.error) <- regions

#    unique.error precluster.error
#v4   0.002277391      0.001418871
#v13  0.009081866      0.007722886
#v35  0.009719159      0.008257968
#v15  0.004740243      0.002645747
#v16  0.007130228      0.005676505
#v19  0.005158478      0.003163881

reduction <- 100*(1-final.error[,"precluster.error"]/final.error[,"unique.error"])
rnd.reduction <- round(reduction, 0)
min.rnd.red <- min(rnd.reduction)
max.rnd.red <- max(rnd.reduction)

min.region <- paste(pretty.region[names(rnd.reduction[rnd.reduction==min.rnd.red])],collapse=" and ")

max.region <- pretty.region[[names(rnd.reduction[rnd.reduction==max.rnd.red])]]
```

***Pre-clustering sequences to further reduce sequencing noise.***
Previously, we implemented a pre-clustering algorithm where sequences were sorted
by their abundance in decreasing order and rare sequences are clustered with a
more abundant sequence if the rare sequences have fewer mismatches than a
defined threshold when compared to the more abundant sequence. The recommended
threshold was a 1-nt difference per 100-nt of sequence data. For example, the
threshold for 250 bp fragment from the V4 region would be 2 nt or 14 for the
1458 bp V1-V9 fragments. This approach removes residual PCR and sequencing
errors while not overwhelming the resolution needed to identify OTUs that are
based on a 3% distance threshold. The tradeoff of this approach is that one would
unable to differentiate V1-V9 sequences that truly differed by less than 14 nt.
When we applied
this approach to our PacBio data, we observed a reduction in the error rate
between `r min.rnd.red` (`r min.region`) and `r max.rnd.red`% (`r max.region`).
The final error rates varied between
`r round(min(100*final.error[,"precluster.error"]), digits=2)`
(`r pretty.region[names(which.min(final.error[,"precluster.error"]))]`)
and `r round(max(100*final.error[,"precluster.error"]), digits=2)`%
(`r pretty.region[names(which.max(final.error[,"precluster.error"]))]`); the
full-length, V1-V9, fragments had an error rate of
`r round(100* final.error["v19","precluster.error"], digits=2)`%
(Figure 2B). These error rates are 7-40
times higher than what we have previously observed using the 454 and
Illumina MiSeq platforms (0.02%){Schloss, 2011 #2466; Kozich, 2013 #2719}

```{r}
getOTUResults <- function(region){
  perfect <- read.table(file=paste0("analysis/", region, "/HMP_MOCK.filter.pick.phylip.an.summary"), header=T)
  nochims <- read.table(file=paste0("analysis/", region, "/", region, ".mock.precluster.perfect.pick.an.ave-std.summary"), header=T)
  observed <- read.table(file=paste0("analysis/",region, "/", region, ".trim.unique.good.filter.unique.precluster.pick.an.merge.groups.ave-std.summary"), header=T)
	obs.mock <- observed[observed$group=="mock"   & observed$method=="ave", "sobs"]
	obs.soil <- observed[observed$group=="soil"   & observed$method=="ave", "sobs"]
	obs.mouse <- observed[observed$group=="mouse" & observed$method=="ave", "sobs"]
	obs.human <- observed[observed$group=="human" & observed$method=="ave", "sobs"]
	nseqs <- nochims[nochims$method=="ave", "nseqs"]

	return(c(perfect[1,2], nochims[1,3], obs.mock, obs.soil, obs.mouse, obs.human, nseqs))
}

otu.table <- t(sapply(regions, getOTUResults))
nseqs <- otu.table[1,7]
perfect.otu <- otu.table[1,1]
otu.table <- otu.table[,-c(1,7)]

otu.table <- cbind(100 * final.error[rownames(otu.table),"precluster.error"], otu.table)
colnames(otu.table) <- c("Error Rate (%)", "Mock with no chimeras", "Observed mock", "Soil", "Mouse", "Human")
rownames(otu.table) <- pretty.region[rownames(otu.table)]

#kable(otu.table, digits=c(2, rep(1,5)))

#|      | Error Rate (%)| Mock with no chimeras| Observed mock|  Soil| Mouse| Human|
#|:-----|--------------:|---------------------:|-------------:|-----:|-----:|-----:|
#|V4    |           0.14|                  26.0|          27.4| 335.8|  78.5|  58.1|
#|V1-V3 |           0.77|                  53.6|          77.0| 436.6| 159.7| 116.8|
#|V3-V5 |           0.83|                  83.1|         106.8| 388.9| 259.6|  91.6|
#|V1-V5 |           0.26|                  34.6|          40.7| 403.2|  88.8|  74.7|
#|V1-V6 |           0.57|                  57.4|          89.7| 431.9| 138.3|  95.9|
#|V1-V9 |           0.32|                  29.5|          41.8| 390.9|  81.9| 102.0|

getCoverage <- function(region){
  observed <- read.table(file=paste0("analysis/",region, "/", region, ".trim.unique.good.filter.unique.precluster.pick.an.merge.groups.ave-std.summary"), header=T)
	obs.mock <- observed[observed$group=="mock"   & observed$method=="ave", "coverage"]
	obs.soil <- observed[observed$group=="soil"   & observed$method=="ave", "coverage"]
	obs.mouse <- observed[observed$group=="mouse" & observed$method=="ave", "coverage"]
	obs.human <- observed[observed$group=="human" & observed$method=="ave", "coverage"]

	return(c(obs.mock, obs.soil, obs.mouse, obs.human))
}

coverage.table <- t(sapply(regions, getCoverage))
colnames(coverage.table) <- c("Mock", "Soil", "Mouse", "Human")
rownames(coverage.table) <- pretty.region[rownames(coverage.table)]
```

***Effects of error rates on OTU assignments.***
The sequencing error rate is known to affect the number of OTUs that are
observed {Schloss, 2011 #2466}. For each region, we determined that if there were no chimeras or
PCR or sequencing errors, then we would expect to find `r perfect.otu[[1]]` OTUs. When
achieved perfect chimera removal, but allowed for PCR and sequencing errors, we
observed between `r diff<-otu.table[,2]-perfect.otu;
round(min(diff), digits=1)` (`r names(which.min(diff))`) and `r round(max(diff), digits=1)`
(`r names(which.max(diff))`) extra OTUs. The range in the number of extra OTUs was largely
explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, final.error[,"precluster.error"])$estimate, 2)`). Next, we determined
the number of OTUs that were observed when we used UCHIME to identify chimeric sequence.
Under these more realistic conditions, we observed between `r diff<-otu.table[,3]-perfect.otu;
round(min(diff), digits=1)` (`r names(which.min(diff))`) and `r round(max(diff), digits=1)`
(`r names(which.max(diff))`) extra OTUs. Finally, we calculated the number of OTUs in the
soil, mouse, and human samples using the same pipeline with chimera detection and removal
based on the UCHIME algorithm. Again, we found that there was a strong correlation between
the number of observed OTUs and the error rate for the
soil (R=`r format(round(cor.test(otu.table[,"Soil"], final.error[,"precluster.error"])$estimate, 2), nsmall=2)`),
mouse (R=`r format(round(cor.test(otu.table[,"Mouse"], final.error[,"precluster.error"])$estimate, 2), nsmall=2)`),
and human samples (R=`r format(round(cor.test(otu.table[,"Human"], final.error[,"precluster.error"])$estimate, 2), nsmall=2)`).
These results underscore the effect of sequencing error on the inflation of the number of
observed OTUs.

```{r, classificationSummary}
data <- read.table(file="taxonomy.depth.analysis", header=T)

rdp <- cbind("Mock"=data[data$database=="rdp" & data$sample=="mock", "total"],
						 "Human"=data[data$database=="rdp" & data$sample=="human", "total"],
						 "Mouse"=data[data$database=="rdp" & data$sample=="mouse", "total"],
						 "Soil"=data[data$database=="rdp" & data$sample=="soil", "total"])
rownames(rdp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")

gg <- cbind("Mock"=data[data$database=="gg" & data$sample=="mock", "total"],
						"Human"=data[data$database=="gg" & data$sample=="human", "total"],
						"Mouse"=data[data$database=="gg" & data$sample=="mouse", "total"],
						"Soil"=data[data$database=="gg" & data$sample=="soil", "total"])
rownames(gg) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")

silva <- cbind("Mock"=data[data$database=="silva" & data$sample=="mock", "total"],
							 "Human"=data[data$database=="silva" & data$sample=="human", "total"],
							 "Mouse"=data[data$database=="silva" & data$sample=="mouse", "total"],
							 "Soil"=data[data$database=="silva" & data$sample=="soil", "total"])
rownames(silva) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")

gg.sp <- cbind("Mock"=data[data$database=="gg" & data$sample=="mock", "X7"],
							 "Human"=data[data$database=="gg" & data$sample=="human", "X7"],
							 "Mouse"=data[data$database=="gg" & data$sample=="mouse", "X7"],
							 "Soil"=data[data$database=="gg" & data$sample=="soil", "X7"])
rownames(gg.sp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
```

***Increasing sequence length improves classification.***
We classified all of the sequence data we generated using the naïve Bayesian
classifier using the RDP, SILVA, and greengenes reference taxonomies (Figure
4). In general, increasing the length of the region improved the
ability to assign the sequence to a genus or species. Interestingly, each of the
samples we analyzed varied in the ability to assign its sequences to the depth of
genus or species. Furthermore, the reference database that did the best job of
classifying the sequences varied by sample type. For example, the SILVA
reference did the best for the human feces and soil samples and the RDP did the
best for the mouse feces samples. An advantage of the greengenes database is
that it contains
information for 2,514 species-level lineages for 11% of the reference sequences;
the other databases only provided taxonomic data to the genus level. There was a
modest association between the length of the fragment and the
ability to classify sequences to the species-level for the human samples; there
was no such association for the mouse and soil samples. In fact, at most
`r format(round(max(gg.sp[,"Soil"]), digits=1), nsmall=1)`% of
the soil sequences and `r format(round(max(gg.sp[,"Mouse"]), digits=1), nsmall=1)`%
of the mouse sequences could be classified to a
species. These results indicate that the ability to classify sequences to the
genus or species level is a function of read length, sample type, and the
reference database.

```{r}
error <- read.table(file="analysis/v4/v4.trim.unique.good.filter.unique.pick.error.summary", header=T, row.names=1)
error.nochim <- error[error$numparents==1,]
one.off <- error.nochim[error.nochim$mismatches==1,]
one.off.sum <- sum(one.off$weight)
v4.table <- table(one.off$weight)
v4.total <- sum(error$weight)

error <- read.table(file="analysis/v35/v35.trim.unique.good.filter.unique.pick.error.summary", header=T, row.names=1)
error.nochim <- error[error$numparents==1,]
one.off <- error.nochim[error.nochim$mismatches==1,]
one.off.sum <- sum(one.off$weight)
v35.table <- table(one.off$weight)
v35.total <- sum(error$weight)

error <- read.table(file="analysis/v15/v15.trim.unique.good.filter.unique.pick.error.summary", header=T, row.names=1)
error.nochim <- error[error$numparents==1,]
one.off <- error.nochim[error.nochim$mismatches==1,]
one.off.sum <- sum(one.off$weight)
v15.table <- table(one.off$weight)
v15.total <- sum(error$weight)
```

***Sequencing errors are not random.***
Above, we described that although there was no obvious bias in the substitution
or insertion rate, we did observe that guanines and cytosines were more likely
to be deleted than adenines and thymidines. This lack of randomness in the error
profile suggested that there might be a systematic non-random distribution of
the errors across the sequences. This would manifest itself by the creation of
duplicate sequences with the same error. Because we were able to obtain a large
number of reads from the mock communities where we sequenced the V4
(N=`r v4.total`), V1-V5 (N=`r v15.total` sequences), and V3-V5 (N=`r v35.total`)
regions, we investigated the mock community data from these regions further. We
identified all of the sequences that had a 1-nt difference to the true sequence.
For these three regions, a majority of the sequences with 1-nt errors were only
observed once (V4: `r round(100*v4.table[1]/sum(v4.table), 1)`%, V1-V5:
`r round(100*v15.table[1]/sum(v15.table), 1)`%, V3-V5:
`r round(100*v35.table[1]/sum(v35.table), 1)`%). We found that the frequency of the
most abundant 1-nt error paralleled the number of sequences. There were two sequences
in the V4 dataset that occurred `r names(v4.table[length(v4.table)])` times, one
sequence in the V1-V5 dataset that occurred `r names(v15.table[length(v15.table)])`
times, and one sequence in the V3-V5 dataset that occurred
`r names(v35.table[length(v35.table)])` times. Contrary to previous reports
{Carneiro, 2012 #3458;Koren, 2012 #3459}, these results indicate that
reproducible errors occur with the PacBio sequencing platform and that they can be
quite frequent.

## Conclusions
The various sequencing platforms that are available to microbial ecologists are
able to fill unique needs and have their own strengths and weaknesses. For
sequencing the 16S rRNA gene, the 454 platform is able to generate a moderate
number of high-quality 500-nt sequence fragments (error rates below 0.02%) {Schloss, 2011 #2466} and
the MiSeq platform is able to generate a large number of high-quality 250-nt
sequence fragments (error rates below 0.02%) {Kozich, 2013 #2719}. The promise of the PacBio
sequencing platform was the generation of high-quality near full-length sequence
fragments. As we have shown in this study, it is possible to generate near
full-length sequences; however, the error rate associated with those reads is
considerable (i.e. `r round(100*final.error["v19","precluster.error"],2)`%)
and requires a level of sequencing coverage that is not commonly observed in a
typical sequencing run. This results in the generation of a small number of low
quality full-length sequences. When we considered the shorter V4 region, which is
similar in length to what is sequenced by the MiSeq platform, the error rates
we observed with the PacBio platform were nearly 5-fold higher than what has
previously been reported. It appears that the promise offered by the PacBio
platform has not been realized.

The widespread adoption of the 454 and MiSeq platforms and decrease in the use
of Sanger
sequencing for the 16S rRNA gene has resulted in a decrease in the
generation of the full-length reference sequences that are needed for performing
phylogenetic analyses and designing lineage specific PCR primers and fluorescent
*in situ* hybridization (FISH) probes. It remains to be determined whether the
elevated error rates we observed for full-length sequences are prohibitive for
these applications. We can estimate the distribution of errors assuming that the
errors follow a binomial distribution along the length of the 1,500 nt
gene with the error rate that we achieved from the V1-V9 mock community data
prior to pre-clustering the sequences, which was
`r round(100*final.error["v19","unique.error"],2)`% (Figure 5). Under
these conditions one would only expect
`r round(100*dbinom(0, 1500, final.error["v19","unique.error"]), 2)`% of the
sequences to have no errors. In fact, 95% of the reads would have at least
`r qbinom(0.05, 1500, final.error["v19","unique.error"])` errors and 50% of the
reads would have at least
`r qbinom(0.50, 1500, final.error["v19","unique.error"])` errors. If the error
rate could be dropped to 0.25%, then 95% of
the reads would have at least
`r qbinom(0.05, 1500, 0.0025)` error and 50% of the
reads would have at least
`r qbinom(0.50, 1500, 0.0025)` errors. If it were possible to replicate the low
error rates we have previously observed using the 454 and Illumina MiSeq
platforms, which was 0.02%, then we would expect
`r round(100*dbinom(0, 1500, 0.0002), 1)`% of the sequences to have no errors.
In fact, 95% of the reads would have `r qbinom(0.95, 1500, 0.0002)` or fewer
errors. Although full-length sequence data is highly desired, at this point, it
does not appear that the PacBio platform can provide the data of sufficient
quality to fill the niche of generating reference sequences.

Full-length sequences are frequently seen as a panacea to overcome the
limitations of taxonomic classifications. The ability to classify each of our
sample types benefited from the generation of full-length sequences. It was
interesting that the benefit varied by sample type and database. For example,
using the mouse libraries, the ability to classify each of the regions differed
by less than 5%
when classifying against the SILVA and greengenes databases. The effect of the
database that was used was also interesting. The RDP database outperformed the
other databases for the mouse samples and the SILVA database outperformed the
others for the human and soil samples. The three databases were equally
effective for classifying the mock community. Finally, since only the
greengenes database provided species-level information for its reference
sequences it was the only database that allowed for resolution of species-level
classification. The sequences from the mouse and soil libraries were not
effectively classified to the species level (all less than 10%). In contrast,
classification of the human libraries resulted in more than 40% of the sequences
being classified to a genus, regardless of the region. That the variation in
species-level classification for the human libraries was less than 10% suggests
that the benefit of added length is minimal considering the lower sequencing
yield.

The development of newer sequencing technologies continue to advance
and there is justifiable excitement to apply these technologies to sequence the
16S rRNA gene. Although it is clearly possible to generate sequencing data from
these various platforms, it is critical that we assess the platforms for their
ability to generate high quality data and the particular niche that the new
approach will fill. With this in mind, it is essential that researchers utilize
mock communities as part of their experimental design so that they can quantify
their error rates. The ability to generate large numbers of near full-length 16S
rRNA gene sequences is an exciting advance. At this point, the excitement must
be tempered by the appreciation that the error rates limit the application of
the approach.

## Acknowledgements
The Genomic DNA from Microbial Mock Community A (Even, Low Concentration, v3.1,
HM-278D) was obtained through the NIH Biodefense and Emerging Infections
Research Resources Repository, NIAID, NIH as part of the Human Microbiome
Project.

## Funding statement
This study was supported by grants from the NIH (R01HG005975, R01GM099514 and
P30DK034933 to PDS and U54HG004973 to SKH).

## References

## Tables

**Table 1. Summary of the primer pairs used to generate the 16S rRNA gene fragment fragments and the characteristics of each region.**

```{r Table_1, warning=TRUE, results='asis'}

getSloppyRevComp <- function(sequence){
	rev.sequence <- paste(rev(unlist(strsplit(sequence, ""))), collapse="")
	revcomp.sequence <- chartr("ATGCRYSWKMBDHVN", "TACG...........", rev.sequence)
	return(revcomp.sequence)
}

getPos <- function(primer, sequence=ecoli, forward=TRUE){
	if(forward==FALSE){
		primer <- getSloppyRevComp(primer)
	}
	primer <- gsub("[^ATGC]", ".", primer)

	pos <- regexpr(primer, sequence)

	if(forward==FALSE){
		pos <- pos + nchar(primer) -1
	}
	return(pos)

}

#taken from... http://www.ncbi.nlm.nih.gov/nuccore/NC_022648.1?report=fasta&from=4564120&to=4565661
ecoli <- "AAATTGAAGAGTTTGATCATGGCTCAGATTGAACGCTGGCGGCAGGCCTAACACATGCAAGTCGAACGGTAACAGGAAGCAGCTTGCTGCTTTGCTGACGAGTGGCGGACGGGTGAGTAATGTCTGGGAAACTGCCTGATGGAGGGGGATAACTACTGGAAACGGTAGCTAATACCGCATAACGTCGCAAGACCAAAGAGGGGGACCTTCGGGCCTCTTGCCATCGGATGTGCCCAGATGGGATTAGCTAGTAGGTGGGGTAACGGCTCACCTAGGCGACGATCCCTAGCTGGTCTGAGAGGATGACCAGCCACACTGGAACTGAGACACGGTCCAGACTCCTACGGGAGGCAGCAGTGGGGAATATTGCACAATGGGCGCAAGCCTGATGCAGCCATGCCGCGTGTATGAAGAAGGCCTTCGGGTTGTAAAGTACTTTCAGCGGGGAGGAAGGGAGTAAAGTTAATACCTTTGCTCATTGACGTTACCCGCAGAAGAAGCACCGGCTAACTCCGTGCCAGCAGCCGCGGTAATACGGAGGGTGCAAGCGTTAATCGGAATTACTGGGCGTAAAGCGCACGCAGGCGGTTTGTTAAGTCAGATGTGAAATCCCCGGGCTCAACCTGGGAACTGCATCTGATACTGGCAAGCTTGAGTCTCGTAGAGGGGGGTAGAATTCCAGGTGTAGCGGTGAAATGCGTAGAGATCTGGAGGAATACCGGTGGCGAAGGCGGCCCCCTGGACGAAGACTGACGCTCAGGTGCGAAAGCGTGGGGAGCAAACAGGATTAGATACCCTGGTAGTCCACGCCGTAAACGATGTCGACTTGGAGGTTGTGCCCTTGAGGCGTGGCTTCCGGAGCTAACGCGTTAAGTCGACCGCCTGGGGAGTACGGCCGCAAGGTTAAAACTCAAATGAATTGACGGGGGCCCGCACAAGCGGTGGAGCATGTGGTTTAATTCGATGCAACGCGAAGAACCTTACCTGGTCTTGACATCCACGGAAGTTTTCAGAGATGAGAATGTGCCTTCGGGAACCGTGAGACAGGTGCTGCATGGCTGTCGTCAGCTCGTGTTGTGAAATGTTGGGTTAAGTCCCGCAACGAGCGCAACCCTTATCCTTTGTTGCCAGCGGTCCGGCCGGGAACTCAAAGGAGACTGCCAGTGATAAACTGGAGGAAGGTGGGGATGACGTCAAGTCATCATGGCCCTTACGACCAGGGCTACACACGTGCTACAATGGCGCATACAAAGAGAAGCGACCTCGCGAGAGCAAGCGGACCTCATAAAGTGCGTCGTAGTCCGGATTGGAGTCTGCAACTCGACTCCATGAAGTCGGAATCGCTAGTAATCGTGGATCAGAATGCCACGGTGAATACGTTCCCGGGCCTTGTACACACCGCCCGTCACACCATGGGAGTGGGTTGCAAAAGAAGTAGGTAGCTTAACCTTCGGGAGGGCGCTTACCACTTTGTGATTCATGACTGGGGTGAAGTCGTAACAAGGTAACCGTAGGGGAACCTGCGGTTGGATCACCTCCTTA"

oligos <- read.table(file="pacbio.oligos", stringsAsFactors=F)[1:6,]
rownames(oligos) <- pretty.region[oligos$V4]
oligos <- oligos[,c("V2", "V3")]
colnames(oligos) <- c("Forward", "Reverse")

start <- sapply(oligos[,1], getPos)
end <- sapply(oligos[,2], getPos, forward=FALSE)

coordinates <- paste(start, end, sep="-")
length <- end - start - nchar(oligos[,1]) - nchar(oligos[,2]) + 1
t1 <- cbind(oligos, coordinates, length)
knitr::kable(t1, digits=1, row.names=TRUE, col.names=c("Forward", "Reverse", "*E. coli* coordinates^a^", "Length (bp)^b^"), align=c("l", "l", "c", "c"))
```

^a^	The coordinates where the start and end of the forward and reverse primers anneal, respectively.

^b^	The number of bases between the primers.

**Table 2. Summary of the reasons that sequences were excluded because of the
basic sequence curation steps**

```{r Table_2, warning=TRUE, results='asis'}
rownames(reasons.lost) <- pretty.region[rownames(reasons.lost)]
reasons.lost <- cbind(orig.nseqs, 100*reasons.lost, basic.nseqs)
knitr::kable(reasons.lost, digits=1, row.names=TRUE,
						 col.names=c("Initial sequences (N)", "Good reads (%)",
						 						"Wrong start/end position (%)",
						 						"Excessively long homopolymers (%)",
						 						"Ambiguous base calls (%)",
						 						"Sequences remaining (N\\)"),
						 align="c")
```

## Figures

**Figure 1. Summary of errors in data generated using PacBio sequencing platform
to sequence various regions within the 16S rRNA gene.** Quality scores varied
with error types (A). The sequencing error rate of the amplified gene fragments
increased with mismatches to the barcodes and primers (B). The sequencing error
rate declined with increasd sequencing coverage; however, increasing the
sequencing depth beyond 10-fold coverage had no meaningful effect on the
sequencing error rate (C).

**Figure 2. Change in error rate (A) and the percentage of sequences that were
retained (B) when using various sequence curation methods.** The condition that
was used for downstream analyses is indicated by the star. The plotted numbers
represent the region that was sequenced. For example "15" represents the data
for the V1-V5 region.

**Figure 3. The relationship between the error rate of each region and the
composite quality scores for the sequences.** The error rates (A and C) and
percentage of sequences (B and D) were calculated for the reads that had a
composite quality score above the plotted value. The composite quality scores
were calculated by either determining the minimum value of the average quality
score wihin all 50-nt windows within each region (A and B) or by calculating
the average quality score across the entire sequence read (C and D).

**Figure 4. Percentage of unique sequences that could be classified.**
Classifications were performed using taxonomy references curated from the RDP,
SILVA, or greengenes databases for the four types of samples that were sequenced
across the six regions from the 16S rRNA gene. Only the greengenes taxonomy
reference provided species-level information.

**Figure 5. The percentage of V1-V9 sequences that were predicted to have
between 0 and 20 errors as a function of the error rate of the sequences.** The
highest error rate, `r round(100*final.error["v19","unique.error"], 2)`%,
corresponds to what was observed before the pre-clustering step.
The smallest error rate (0.02%) corresponds to our previous observations using
the 454 and MiSeq sequencing platforms. The predicted number of errors was
assumed to follow a binomial distribution.

### Figure 1
```{r Figure_1, fig.width=6.5, fig.height=3.0}
layout(matrix(c(1,2,3), nrow=1), heights=1, widths=c(0.75,1,1))

#Panel A depicting range of quality scores for different errors
par(mar=c(6, 5, 0.5, 0.5))
plot(1, xlim=c(0.5, 4.5), ylim=c(0,75), type="n", yaxt="n", xaxt="n", xlab="",
		 ylab="")
polygon(x=c(0.75, 1.25, 1.25, 0.75), y=c(m.quant["25%"], m.quant["25%"],
																				 m.quant["75%"], m.quant["75%"]))
polygon(x=c(1.75, 2.25, 2.25, 1.75), y=c(s.quant["25%"], s.quant["25%"],
																				 s.quant["75%"], s.quant["75%"]))
polygon(x=c(2.75, 3.25, 3.25, 2.75), y=c(i.quant["25%"], i.quant["25%"],
																				 i.quant["75%"], i.quant["75%"]))
polygon(x=c(3.75, 4.25, 4.25, 3.75), y=c(a.quant["25%"], a.quant["25%"],
																				 a.quant["75%"], a.quant["75%"]))

segments(x0=seq(0.75, 3.75, 1), x1=seq(1.25,4.25,1), y0=c(m.quant["50%"],
											s.quant["50%"], i.quant["50%"], a.quant["50%"]), lwd=4)

arrows(x0=1:4, y0=c(m.quant["75%"], s.quant["75%"], i.quant["75%"],
										a.quant["75%"]), y1=c(m.quant["97.5%"], s.quant["97.5%"],
																					i.quant["97.5%"], a.quant["97.5%"]),
			 angle=90, length=0.05)

arrows(x0=1:4, y0=c(m.quant["25%"], s.quant["25%"], i.quant["25%"],
										a.quant["25%"]), y1=c(m.quant["2.5%"], s.quant["2.5%"],
																					i.quant["2.5%"], a.quant["2.5%"]),
			 angle=90, length=0.05)

axis(1, at=c(1,2,3,4), label=c("Matches", "Substitutions", "Insertions",
															 "Ambiguous"), las=2, cex.axis=1)
axis(2, las=2, at=c(0,25,50,75), cex.axis=1.5)
title(ylab="Quality score", cex.lab=1.5)
mtext(side=2, line=2.5, at=75, text="A", cex=2, font=2, las=2)

#Panel B depicting relationship between errors in barcodes/primers and the
#rest of the sequence.
par(mar=c(5,5,0.5,0.5))
plot(err.bcprimer$x[1:6,1]~err.bcprimer$Group.1[1:6], pch=19,
		 ylab="Error rate (%)", xlab="", axes=F, ylim=c(0,5), cex.lab=1.5)
mtext(side=1, line=3.75, text="Total mismatches to\nbarcodes and primers",
			cex=0.9)
axis(1, cex.axis=1.5)
axis(2, las=2, cex.axis=1.5)
box()
mtext(side=2, line=2.5, at=5, text="B", cex=2, font=2, las=2)

#Panel C depicting relationship between sequencing coverage and error rate for
#the different regions that were sequence
plotLines <- function(folder){
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
																		 ".composite", sep=""),
													header=T,row.names=1)

	composite <- composite[composite$reason=="g",]
	error.by.depth <- aggregate(composite$error, by=list(composite$freq),
															function(x){c(mean=mean(x),
																						quantile(x, prob=c(0.025, 0.975)),
																						n=length(x))})

	#exclude any coverage values where we don't have at least 50 observations
	error.by.depth <- error.by.depth[error.by.depth$x[,"n"]>=50,]
	fold.coverage <- error.by.depth[, "Group.1"]
	error.by.depth <- error.by.depth$x

	mean <- error.by.depth[,"mean"]
	points(mean~fold.coverage, type="l", lwd=2, col=clrs[folder])
}

par(mar=c(5,2,0.5, 0.5))
plot(1, xlim=c(2,30), ylim=c(0,0.05), type="n", xlab="", ylab="", axes=F)
axis(2, at=seq(0,0.05, 0.01), label=FALSE, tick=T, las=2)
#axis(2, at=seq(0,0.05, 0.01), label=seq(0,5, 1), las=2)
axis(1, cex.axis=1.5)
lapply(regions, plotLines)
legend(x=18, y=0.05, legend=pretty.region[regions], lty=1, lwd=2, col=clrs[regions])
mtext(side=2, line=0.5, at=0.05, text="C", cex=2, font=2, las=2)
title(xlab="Coverage", ylab="", cex.lab=1.5)
box()

layout(1)
```

### Figure 2
```{r Figure_2, width=6.5, height=6}
error.fraction <- cbind(initial.error=init.rates, basic.error=basic.error,
												bcprimer.error, coverage.error, aveq.error,
												oligosCoverage.error, oligosQScore.error,
												coverageQScore.error, allFilters.error, final.error)
rownames(error.fraction) <- gsub("v(\\d*)", "\\1", rownames(error.fraction))

error <- 100 * error.fraction[,grepl("error", colnames(error.fraction))]
fraction <- 100 * error.fraction[,grepl("frac", colnames(error.fraction))]

getJitter <- function(x){
	jitter <- 0.15
	runif(6, x-jitter, x+jitter)
}

l <- layout(matrix(c(1,2,3), nrow=3), heights=c(1,1,0.6))

par(mar=c(0.5, 7, 0.5, 0.5))
plot(1, type="n", xlim=c(1, 14), ylim=c(0,2.5), axes=F, xlab="", ylab="",
		 cex.lab=1.5)
text(x=getJitter(1.0), y=error[,1], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(2.0), y=error[,2], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(3.0), y=error[,3], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(4.0), y=error[,4], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(5.0), y=error[,5], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(6.0), y=error[,6], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(7.0), y=error[,7], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(8.0), y=error[,8], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(9.0), y=error[,9], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(10.0), y=error[,10], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(11.0), y=error[,11], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(12.0), y=error[,12], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(13.0), y=error[,13], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(14.0), y=error[,15], label=rownames(error.fraction), cex=1.25)
text(x=10, y=2, label="*", cex=3, font=2)

mtext(text="Error rate (%)", side=2, line=4, cex.lab=1.2)
box()
axis(2, las=2, at=seq(0,2.5, 0.5), label=format(seq(0, 2.5, 0.5),nsmall=1),
		 cex.axis=1.5)

abline(v=seq(3.5,13.5, 1), col="grey")
abline(v=c(1.5, 2.5, 13.5), col="black", lwd=2)

mtext(side=2, line=4.5, at=2.44, text="A", cex=2, font=2, las=1)

par(mar=c(0.5, 7, 0.5, 0.5))
plot(1, type="n", xlim=c(1, 14), ylim=c(0,100), axes=F,
		 ylab="Reads remaining\nfrom basic criteria (%)", xlab="", cex.lab=1.5)
text(x=getJitter(3.0), y=fraction[,1], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(4.0), y=fraction[,2], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(5.0), y=fraction[,3], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(6.0), y=fraction[,4], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(7.0), y=fraction[,5], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(8.0), y=fraction[,6], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(9.0), y=fraction[,7], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(10.0), y=fraction[,8], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(11.0), y=fraction[,9], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(12.0), y=fraction[,10], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(13.0), y=fraction[,11], label=rownames(error.fraction), cex=1.25)
text(x=getJitter(14.0), y=fraction[,8], label=rownames(error.fraction), cex=1.25)

box()
abline(v=seq(3.5,13.5, 1), col="grey")
abline(v=c(1.5, 2.5, 13.5), col="black", lwd=2)

mtext(side=2, line=4.5, at=98, text="B", cex=2, font=2, las=1)
axis(2, las=2, at=seq(0,100, 25), label=seq(0, 100, 25), cex.axis=1.5)
text(1,0, label="NA")
text(2,0, label="NA")

plot(1, type="n", xlim=c(1, 14), ylim=c(0,1), axes=F, ylab="", xlab="")

text(x=c(1:14)+0.2, y=rep(1, 11), label=c("Initial data", "Basic criteria",
		"No Mismatches", "\u22641 Mismatch", "Coverage", "Quality Score",
		"No Mismatches\n& Coverage", "\u22641 Mismatch\n& Coverage",
		"No Mismatches\n& Quality", "\u22641 Mismatch\n& Quality",
		"Coverage & Quality", "No Mismatches,\nCoverage & Quality",
		"\u22641 Mismatch,\nCoverage & Quality", "After Pre-clustering"),
		srt=90, pos=2, cex=1)

layout(1)
```

### Figure 3
```{r Figure_3, fig.height=6.5, fig.width=6.5}
aggregateErrorMinQualityScore <- function(folder){
	write(folder, "")
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
																		 ".composite", sep=""), header=T,
													row.names=1)
	composite <- composite[composite$reason == "g",]

	remaining.error <- rep(0, max(composite$minQ)+1)
	qscores <- 0:max(composite$minQ)

	remaining.error <- unlist(lapply(qscores,
														function(x){mean(composite[composite$minQ >= x,"error"])}))
	names(remaining.error) <- qscores
	points(x=10:70, remaining.error[10:70], type="l", col=clrs[folder], lwd=2)
}

fractionMinQualityScore <- function(folder){
	write(folder, "")

	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
																		 ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]

	binned <- aggregate(composite$error, by=list(round(composite$minQ)), length)
	remaining <- 1-cumsum(binned$x)/sum(binned$x)
	points(binned$Group.1, remaining, type="l", col=clrs[folder], lwd=2)
}

aggregateErrorAveQualityScore <- function(folder){
	write(folder, "")
	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
																		 ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]

	remaining.error <- rep(0, max(composite$aveQ)+1)
	qscores <- 0:max(composite$aveQ)
	remaining.error <- unlist(lapply(qscores,
												function(x){mean(composite[composite$aveQ >= x,"error"])}))
	names(remaining.error) <- qscores
	points(x=10:70, remaining.error[10:70], type="l", col=clrs[folder], lwd=2)
}

fractionAveQualityScore <- function(folder){
write(folder, "")

	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
																		 ".composite", sep=""), header=T, row.names=1)
	composite <- composite[composite$reason == "g",]

	binned <- aggregate(composite$error, by=list(round(composite$aveQ)), length)

	remaining <- 1-cumsum(binned$x)/sum(binned$x)
	points(binned$Group.1, remaining, type="l", col=clrs[folder], lwd=2)
}

layout(1)
layout(matrix(c(1,2,3,4), nrow=2, byrow=F), width=c(1.1,1), height=c(1, 1.3))

#A
par(mar=c(0.5,5,0.5, 0.5))
plot(1, xlim=c(10,75), ylim=c(0,0.015), type="n", axes=F, xlab="", ylab="")
lapply(regions, aggregateErrorMinQualityScore)
mtext(side=2, line=2.5, text="Error rate of sequences\nabove threshold (%)", cex.lab=0.9)
axis(1, at=seq(10,70,10), label=F, tick=T)
axis(2, las=2, at=seq(0,0.015,0.005), label=format(seq(0,1.5,0.5), nsmall=1), cex.axis=1.25)
box()
mtext(side=2, line=2.6, at=0.015, text="A", las=2, cex=2, font=2)

#B
par(mar=c(6,5,0.5, 0.5))
plot(1, xlim=c(10,75), ylim=c(0,1), type="n", yaxt="n", xlab="", ylab="", cex.axis=1.25)
lapply(regions, fractionMinQualityScore)
mtext(side=1, line=4.25, text="Minimum average quality\nscore within a 50-nt window\nacross the full sequence", cex=0.9)
mtext(side=2, line=2.5, text="Sequences above\nthreshold (%)", cex.lab=0.9)

axis(2, las=2, at=seq(0,1,0.2), label=seq(0,100,20), cex.axis=1.25)
mtext(side=2, line=2.6, at=1, text="B", las=2, cex=2, font=2)

#C
par(mar=c(0.5,2,0.5, 0.5))
plot(1, xlim=c(10,75), ylim=c(0,0.015), type="n", axes=F, xlab="", ylab="")
lapply(regions, aggregateErrorAveQualityScore)
axis(1, at=seq(10,70,10), label=F, tick=T)
axis(2, at=seq(0,0.02,0.005), label=F, tick=T)
box()
mtext(side=2, line=0.6, at=0.015, text="C", las=2, cex=2, font=2)

#D
par(mar=c(6,2,0.5, 0.5))
plot(1, xlim=c(10,75), ylim=c(0,1), type="n", yaxt="n", xlab="", ylab="", cex.axis=1.25)
lapply(regions, fractionAveQualityScore)
mtext(side=1, text="Average quality score", cex=0.9, line=3)
title(ylab="Sequences above threshold (%)")
axis(2, at=seq(0,1,0.2), label=FALSE, tick=TRUE)
mtext(side=2, line=0.6, at=1, text="D", las=2, cex=2, font=2)

legend(x=10, y=0.50, legend=pretty.region[regions], lty=1, lwd=2, col=clrs[regions], cex=1.0)
```

### Figure 4

```{r, Figure_4, fig.width=7, fig.height=7}
par(mar=c(5, 5, 0.5, 1))
dotchart(gg, xlim=c(0,100), col="black", xlab="Unique reads that classified\nto genus or species level (%)", pch=19)

points(x=rdp[,"Mock"], y=25:30, pch=15, col="black")
points(x=rdp[,"Human"], y=17:22, pch=15, col="black")
points(x=rdp[,"Mouse"], y=9:14, pch=15, col="black")
points(x=rdp[,"Soil"], y=1:6, pch=15, col="black")

points(x=silva[,"Mock"], y=25:30, pch=17, col="black")
points(x=silva[,"Human"], y=17:22, pch=17, col="black")
points(x=silva[,"Mouse"], y=9:14, pch=17, col="black")
points(x=silva[,"Soil"], y=1:6, pch=17, col="black")

points(x=gg.sp[,"Mock"], y=25:30, pch=21, bg="gray")
points(x=gg.sp[,"Human"], y=17:22, pch=21, bg="gray")
points(x=gg.sp[,"Mouse"], y=9:14, pch=21, bg="gray")
points(x=gg.sp[,"Soil"], y=1:6, pch=21, bg="gray")

legend(x=63, y=12, legend=c("RDP (gen.)", "SILVA (gen.)", "greengenes (gen.+sp.)", "greengenes (sp.)"), pch=c(15, 17, 19, 21), col="black", pt.bg=c("black", "black", "black", "gray"), bg="white")
```

### Figure 5

```{r Figure_5, fig.width=7, fig.height=4}
x <- 0:20
e1 <- final.error["v19","unique.error"]
e2 <- 0.0025
e3 <- 0.0010
e4 <- 0.0002

par(mar=c(5,6,0.5, 0.5))
plot(x, dbinom(x, 1458, e1), type="l", ylim=c(0,0.35), xlab="Number of errors",
		 ylab="Full-length 16S rRNA\ngene sequences (%)", yaxt="n", lwd=3)
points(x, dbinom(x, 1458, e2), type="l", col="red", lwd=3)
points(x, dbinom(x, 1458, e3), type="l", col="blue", lwd=3)
points(x, dbinom(x, 1458, e4), type="l", col="darkgreen", lwd=3)
axis(2, at=seq(0,0.35,0.05), label=seq(0,35,5), las=1)
legend(x=12, y=0.30, legend=paste0(format(round(100*c(e1, e2, e3, e4), 2), nsmall=2), "%"),
			 lwd=3, col=c("black", "red", "blue", "darkgreen"))
```

```{r, versions}
sessionInfo()
```

### 2015-12-23 Schloss_PacBio16S_PeerJ_2016.Rmd
Fix formatting

@@
 it has become imperative to develop stringent sequence curation and denoising
-algorithms {Kozich, 2013 #2719; Schloss, 2011 #2466}.
-There has been a tradeoff between read length, number of reads per sample, and
-the error rate. For instance, we recently demonstrated that using the Illumina
-MiSeq and the 454 Titanium platforms the raw error rate varies between 1
-and 2% {Kozich, 2013 #2719; Schloss, 2011 #2466}. Yet, it was possible to obtain
-error rates below 0.02% by adopting various denoising algorithms. However, the
-resulting fragments were
-only 250-nt long. In the case of 454 Titanium, extending the length of the
+algorithms {Kozich, 2013 #2719; Schloss, 2011 #2466}. There has been a tradeoff
+between read length, number of reads per sample, and the error rate. For
+instance, we recently demonstrated that using the Illumina MiSeq and the 454
+Titanium platforms the raw error rate varies between 1 and 2% {Kozich, 2013
+#2719; Schloss, 2011 #2466}. Yet, it was possible to obtain error rates below
+0.02% by adopting various denoising algorithms. However, the resulting fragments
+were only 250-nt long. In the case of 454 Titanium, extending the length of the
 fragment introduces length-based errors and in the case of the Illumina MiSeq,
@@
 Inadequate denoising of sequencing reads can have many negative effects
-including limited ability to identify chimeras
-{Haas, 2011 #2404;Edgar, 2011 #2406} and inflation of alpha- and beta-diversity
-metrics {Huse, 2010 #2034; Kunin, 2010 #2259; Kozich, 2013 #2719; Schloss, 2011 #2466}.
+including limited ability to identify chimeras {Haas, 2011 #2404;Edgar, 2011 #2406}
+and inflation of alpha- and beta-diversity metrics
+{Huse, 2010 #2034; Kunin, 2010 #2259; Kozich, 2013 #2719; Schloss, 2011 #2466}.
 Although MiSeq and 454 enjoy widespread use in the field, the MiSeq
@@
 research domain {Mosher, 2014 #3376; Fichot, 2013 #3453; Mosher, 2013 #3454}.
-The SMRT system ligates hairpin adapters (i.e. SMRTbells)
-to the ends of double-stranded DNA. Although the DNA molecule is linear, it is
-effectively circularized allowing the sequencing polymerase to process around
-the molecule multiple times {Au, 2012 #3455}. According to Pacific Biosciences
-the platform is able to generate median read lengths longer than 8 kb with the
-P4-C2 chemistry; however, the single pass error rate is approximately 15%. Given
-the circular nature of the DNA fragment, the full read length can be used to
-cover the DNA fragment multiple times resulting in a reduced error rate.
-Therefore, one should be able to obtain multiple coverage of the full 16S rRNA
-gene at a reduced error rate.
+The SMRT system ligates hairpin adapters (i.e. SMRTbells) to the ends of
+double-stranded DNA. Although the DNA molecule is linear, it is effectively
+circularized allowing the sequencing polymerase to process around the molecule
+multiple times {Au, 2012 #3455}. According to Pacific Biosciences the platform
+is able to generate median read lengths longer than 8 kb with the P4-C2
+chemistry; however, the single pass error rate is approximately 15%. Given the
+circular nature of the DNA fragment, the full read length can be used to cover
+the DNA fragment multiple times resulting in a reduced error rate. Therefore,
+one should be able to obtain multiple coverage of the full 16S rRNA gene at a
+reduced error rate.
 
@@
 Previous studies utilizing the technology have removed reads with mismatched
-primers and barcodes, ambiguous base  calls, and low quality scores
-{Fichot, 2013 #3453}. Others have utilized the platform without describing the
-bioinformatic pipeline that was utilized {Mosher, 2014 #3376; Mosher, 2014 #3376}.
-Regardless of the curation methods, the error rates associated with sequencing
-the 16S rRNA gene on the platform have never been reported. In the current
-study, we assessed the quality of data generated by the Pacific Biosciences
-sequencer and whether it could fill the need for generating high-quality,
-full-length sequence data. We hypothesized that by modulating the 16S rRNA gene
-fragment length we could alter the read depth and obtain reads longer than are
-currently available by the 454 and Illumina platforms but with the same quality.
-To test this hypothesis, we developed a sequence curation pipeline that was
-optimized by reducing the sequencing error rate of a mock bacterial community
-with known composition. The resulting pipeline was then applied to 16S rRNA gene
-fragments that were isolated from soil and human and mouse feces.
+primers and barcodes, ambiguous base  calls, and low quality scores {Fichot, 2013 #3453}.
+Others have utilized the platform without describing the bioinformatic pipeline
+that was utilized {Mosher, 2014 #3376; Mosher, 2014 #3376}. Regardless of the
+curation methods, the error rates associated with sequencing the 16S rRNA gene
+on the platform have never been reported. In the current study, we assessed the
+quality of data generated by the Pacific Biosciences sequencer and whether it
+could fill the need for generating high-quality, full-length sequence data. We
+hypothesized that by modulating the 16S rRNA gene fragment length we could alter
+the read depth and obtain reads longer than are currently available by the 454
+and Illumina platforms but with the same quality. To test this hypothesis, we
+developed a sequence curation pipeline that was optimized by reducing the
+sequencing error rate of a mock bacterial community with known composition. The
+resulting pipeline was then applied to 16S rRNA gene fragments that were
+isolated from soil and human and mouse feces.
 
@@
 These same DNA extracts were previously used to develop an Illumina MiSeq-based
-sequencing strategy {Kozich, 2013 #2719}.  Briefly, we used a “Mock Community” composed of
-genomic DNA from 21 bacterial strains: *Acinetobacter baumannii* ATCC 17978,
-*Actinomyces odontolyticus* ATCC 17982, *Bacillus cereus* ATCC 10987,
-*Bacteroides vulgatus* ATCC 8482, *Clostridium beijerinckii* ATCC 51743,
+sequencing strategy {Kozich, 2013 #2719}.  Briefly, we used a “Mock Community”
+composed of genomic DNA from 21 bacterial strains: *Acinetobacter baumannii*
+ATCC 17978, *Actinomyces odontolyticus* ATCC 17982, *Bacillus cereus* ATCC
+10987, *Bacteroides vulgatus* ATCC 8482, *Clostridium beijerinckii* ATCC 51743,
 *Deinococcus radiodurans* ATCC 13939, *Enterococcus faecalis* ATCC 47077,
@@
 BAA-611, *Streptococcus mutans* ATCC 700610, *Streptococcus pneumoniae* ATCC
-BAA-334. The mock community DNA is available through BEI resources
-(v3.1, HM-278D). Genomic DNAs from the three other communities were obtained
-using the MO BIO PowerSoil DNA extraction kit. The human and mouse fecal samples
-were obtained using protocols that were reviewed and approved by the University
-Committee on Use and Care of Animals (Protocol #PRO00004877) and the Institutional Review Board at the
-University of Michigan (Protocol #HUM00057066). The human stool donor provided informed consent.
+BAA-334. The mock community DNA is available through BEI resources (v3.1,
+HM-278D). Genomic DNAs from the three other communities were obtained using the
+MO BIO PowerSoil DNA extraction kit. The human and mouse fecal samples were
+obtained using protocols that were reviewed and approved by the University
+Committee on Use and Care of Animals (Protocol #PRO00004877) and the
+Institutional Review Board at the University of Michigan (Protocol
+#HUM00057066). The human stool donor provided informed consent.
 
@@
 pooling were described previously {Kozich, 2013 #2719}. The SMRTbell adapters
-were ligated onto  the PCR products and the libraries were sequenced at the
-University of Michigan DNA Sequencing Core using the P4-C2 chemistry on a PacBio
-RS II SMRT DNA Sequencing System.
+were ligated onto  the PCR products and the libraries were sequenced by Pacific
+Biosciences using the P6-C24 chemistry on a PacBio RS II SMRT DNA Sequencing
+System.
 

### 2015-12-23 Schloss_PacBio16S_PeerJ_2016.Rmd
Fix formatting

@@
 ***Data analysis.***
-All sequencing data were curated using mothur {Schloss, 2009 #1816} and
-analyzed using the R programming language {R Core Team, 2014 #3456}. The raw
-data can be obtained from the Sequence Read Archive at NCBI under accession
-SRP051686, which are associated with BioProject PRJNA271568. Several specific
-features were incorporated into mothur to
-facilitate the analysis of PacBio  sequence data.  First, because non-ambiguous
-base calls are assigned to Phred quality scores of zero, the consensus fastq
-files were parsed so that scores of zero were interpreted as corresponding to an
-ambiguous base call (i.e. N) in the  fastq.info command using the pacbio=T
-option. Second, because the consensus  sequence can be generated in the forward
-and reverse complement orientations, a checkorient option was added to the
-trim.seqs command in order to identify the  proper orientation. These features
-were incorporated into mothur v.1.30. Because chimeric molecules can be
-generated during PCR and would artificially inflate the sequencing error, it was
-necessary to remove these data prior to assessing the error rate. Because we
-knew the true sequences for the strains in the mock community we could calculate
-all possible chimeras between strains in the mock community (*in silico*
-chimeras). If a sequence read was 3 or more nucleotides more similar to an *in
-silico* chimera than it was to a non-chimeric reference sequence, it was
-classified as a chimera and removed from further consideration. Identification
-of *in silico* chimeras and calculation of sequencing error rates
+All sequencing data were curated using mothur {Schloss, 2009 #1816} and analyzed
+using the R programming language {R Core Team, 2014 #3456}. The raw data can be
+obtained from the Sequence Read Archive at NCBI under accession SRP051686, which
+are associated with BioProject PRJNA271568. Several specific features were
+incorporated into mothur to facilitate the analysis of PacBio  sequence data.
+First, because non-ambiguous base calls are assigned to Phred quality scores of
+zero, the consensus fastq files were parsed so that scores of zero were
+interpreted as corresponding to an ambiguous base call (i.e. N) in the
+fastq.info command using the pacbio=T option. Second, because the consensus
+sequence can be generated in the forward and reverse complement orientations, a
+checkorient option was added to the trim.seqs command in order to identify the
+proper orientation. These features were incorporated into mothur v.1.30. Because
+chimeric molecules can be generated during PCR and would artificially inflate
+the sequencing error, it was necessary to remove these data prior to assessing
+the error rate. Because we knew the true sequences for the strains in the mock
+community we could calculate all possible chimeras between strains in the mock
+community (*in silico* chimeras). If a sequence read was 3 or more nucleotides
+more similar to an *in silico* chimera than it was to a non-chimeric reference
+sequence, it was classified as a chimera and removed from further consideration.
+Identification of *in silico* chimeras and calculation of sequencing error rates
 was performed using the seq.error command in mothur {Schloss, 2011 #2466}. *De
@@
 were classified against the SILVA {Pruesse, 2007 #1735}, RDP {Cole, 2014 #3460},
-and greengenes {Werner, 2012 #2474} reference taxonomies  using a
-naive Bayesian classifier {Wang, 2007 #1809}. Sequences were assigned to
-operational taxonomic units using the average neighbor clustering algorithm with
-a 3% distance threshold {Schloss, 2011 #3156}.  Detailed methods including this
-paper as an R markdown file are available as a public online repository
+and greengenes {Werner, 2012 #2474} reference taxonomies  using a naive Bayesian
+classifier {Wang, 2007 #1809}. Sequences were assigned to operational taxonomic
+units using the average neighbor clustering algorithm with a 3% distance
+threshold {Schloss, 2011 #3156}.  Detailed methods including this paper as an R
+markdown file are available as a public online repository
 (http://github.com/SchlossLab/ Schloss_PacBio16S_PeerJ_2015).
@@
 ```{r setup, echo=FALSE, eval=TRUE, tidy=TRUE}
-rm(list=ls())
-library("knitr")
-#pdf.options(encoding="ISOLatin2.enc")
 regions <- c("v4", "v13", "v35", "v15", "v16", "v19")
@@
 names(clrs) <- regions
-
-opts_chunk$set("fig.path"="figures/")
-opts_chunk$set("fig.align"="center")
-opts_chunk$set("dev" = c("png", "cairo_pdf"))
-
-opts_chunk$set("tidy" = TRUE)
-opts_chunk$set("echo" = FALSE)
-opts_chunk$set("eval" = TRUE)
-opts_chunk$set("results" = 'hide')
-opts_chunk$set("warning" = FALSE)
-opts_chunk$set("cache" = TRUE)
 ```

### 2015-12-23 code/consolidate_data.R
Pool report files by region and output to data/process

@@ -188,12 +188,16 @@ generate_report <- function(report_file) {
 	stopifnot(sum(rownames(error_summary) != rownames(alignment_quality)) == 0)
 	stopifnot(sum(rownames(error_summary) != rownames(quality_score_report)) == 0)
 
+	batch <- rep(gsub("data/mothur_(.*)/.*", "\\1", report_file), nrow(error_summary))
+	
+
 	report_data <- cbind(
 		error_summary,
 		coverage_pred_error,
 		mismatches,
 		alignment_quality,
-		quality_score_report
+		quality_score_report,
+		batch
 	)
 
 	write.table(report_data, file=report_file, quote=F)

### 2015-12-24 code/consolidate_data.R
Create a single mock.report file in data/process

@@ -189,7 +189,7 @@ generate_report <- function(report_file) {
 	stopifnot(sum(rownames(error_summary) != rownames(quality_score_report)) == 0)
 
 	batch <- rep(gsub("data/mothur_(.*)/.*", "\\1", report_file), nrow(error_summary))
-	
+	region <- rep(gsub("data/mothur_.*/(V.*).mock.*", "\\1", report_file), nrow(error_summary))
 
 	report_data <- cbind(
 		error_summary,
@@ -197,7 +197,8 @@ generate_report <- function(report_file) {
 		mismatches,
 		alignment_quality,
 		quality_score_report,
-		batch
+		batch,
+		region
 	)
 
 	write.table(report_data, file=report_file, quote=F)

### 2015-12-24 code/pool_error_quality.R
Compress mock.error.report and pool data/mothur*/*error.quality

@@ -1,0 +1,15 @@
+pool <- function(file_string){
+	files <- unlist(strsplit(file_string, " "))
+
+	pooled_data <- read.table(file=files[1], header=T)
+	files <- files[-1]
+
+	qscores <- pooled_data$qscore
+
+	for(f in files){
+		temp_data <- read.table(file=files[1], header=T)
+		stopifnot(temp_data$qscore == qscores)
+	}
+	pooled_data$qscore <- qscores
+	write.table(pooled_data, "data/process/mock.quality.report", quote=F, row.names=F, sep='\t')
+}

### 2016-01-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Add flags for what needs revision

@@
 
+regions <- c("v4", "v13", "v35", "v15", "v16", "v19")
+clrs <- rainbow(length(regions))
+pretty.region <- c("v13"="V1-V3", "v15"="V1-V5", "v16"="V1-V6", "v19"="V1-V9", "v35"="V3-V5", "v4"="V4")
+names(clrs) <- regions
 ```
@@
 
-## Abstract
+## Abstract [ Revise ]
 Over the past 10 years, microbial ecologists have largely abandoned sequencing
@@
 
-## Introduction
+## Introduction [ Revise ]
 Advances in sequencing technologies over the past 10 years have introduced
@@
 
-***Library generation and sequencing.***
+***Library generation and sequencing.*** [ Revise ]
 The DNAs were each amplified in triplicate using barcoded primers targeting the
@@
 
-***Data analysis.***
+***Data analysis.***  [ Revise ]
 All sequencing data were curated using mothur {Schloss, 2009 #1816} and analyzed
@@
 (http://github.com/SchlossLab/ Schloss_PacBio16S_PeerJ_2015).
-
-```{r setup, echo=FALSE, eval=TRUE, tidy=TRUE}
-regions <- c("v4", "v13", "v35", "v15", "v16", "v19")
-clrs <- rainbow(length(regions))
-pretty.region <- c("v13"="V1-V3", "v15"="V1-V5", "v16"="V1-V6", "v19"="V1-V9", "v35"="V3-V5", "v4"="V4")
-names(clrs) <- regions
-```
 

### 2016-01-05 Schloss_PacBio16S_PeerJ_2016.Rmd

@@
       in_header: header.tex
-csl: mbio.csl
+csl: peerj.csl
 fontsize: 11pt

### 2016-01-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Update code and text for The PacBio error profile section

@@
 names(clrs) <- regions
+
+mock_error <- read.table(file="data/process/mock.error.report")
+mock_error_nochim <- mock_error[mock_error$numparents==1,]
+mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.10,]
 ```
@@
 ```{r error_profile}
-sub.matrix <- read.table(file=paste("pipeline_dev/", regions[1], "/", regions[1], ".mock.filter.error.matrix", sep=""), header=T, row.names=1)[-7,]
-for(i in 2:length(regions)){
-	sub.matrix <- sub.matrix + read.table(file=paste("pipeline_dev/", regions[i], "/", regions[i], ".mock.filter.error.matrix", sep=""), header=T, row.names=1)[-7,]
-}
-
-matches <- c(sub.matrix[1,1], sub.matrix[2,2], sub.matrix[3,3], sub.matrix[4,4])
-total <- sum(sub.matrix)
-errors <- sub.matrix
-diag(errors[1:4,1:4]) <- 0
-init.error.rate <- 100 * sum(errors) / total				#1.925717
-
-subst.bias <- apply(errors[1:4,1:4], 2, sum)/apply(sub.matrix[1:4,1:4], 2, sum)
-subst.bias <- 100 * subst.bias/sum(subst.bias)
-#       rA        rT        rG        rC
-# 22.43918 27.21620 24.06557 26.27904
-#these are more or less equal to each other
-
-insert.bias <- (sub.matrix[,"rGap"]/apply(sub.matrix, 1, sum))[1:4]
-insert.bias <- 100 * insert.bias/sum(insert.bias)
-#       qA       qT       qG       qC
-# 23.46112 24.36667 24.33064 27.84157
-#these are more or less equal to each other
-
-delete.bias <- (sub.matrix["qGap",]/apply(sub.matrix, 2, sum))[1:4]
-delete.bias <- 100 * delete.bias/sum(delete.bias)
-#           rA      rT       rG       rC
-#qGap 11.43234 9.46907 44.57506 34.52353
-#interestingly the G's and C's are more likely to be deleted than the As or Ts
-
-substitutions <- 100*sum(errors[1:4,1:4])/sum(errors)	#35.80084
-insertions <- 100*sum(errors[,5])/sum(errors)			#45.27368
-deletions <- 100*sum(errors[5,])/sum(errors)			#17.27259
-ambiguous <- 100*sum(errors[6,])/sum(errors)			#2.101505
+mutations <- c("AA", "AT", "AG", "AC", "TA", "TT", "TG", "TC", "GA", "GT", "GG", "GC", "CA", "CT", "CG", "CC", "Ai", "Ti", "Gi", "Ci", "dA", "dT", "dG", "dC")
+
+mutation_table <- mock_error_good[, which(colnames(mock_error_good)%in%mutations)]
+mutation_count <- apply(mutation_table, 2, sum)
+
+mutation_total <- sum(mutation_count)
+matches <- sum(mutation_count[c("AA", "TT", "GG", "CC")])
+mis_matches <- mutation_total - matches
+init_error_rate <- 100 * mis_matches / mutation_total		#0.6514488
+
+#substitutions
+substitutions <- mutations[grep("[ATGC].", mutations)]
+substitutions <- substitutions[grep(".[ATGC]", substitutions)]
+substitutions <- substitutions[! substitutions %in% c("AA", "TT", "GG", "CC")]
+substitution_counts <- mutation_count[substitutions]
+substitution_total <- sum(substitution_counts)
+substitution_bias <- substitution_counts / substitution_total
+
+reference <- gsub(".(.)", "\\1", names(substitution_bias))
+# aggregate(substitution_bias, by=list(reference), mean)
+#
+# Group.1          x
+# 1       A 0.10192188
+# 2       C 0.05944286
+# 3       G 0.07434043
+# 4       T 0.09762816
+
+substitution_rate <- 100 * substitution_total / mis_matches
+
+# [1] 50.90346
+
+#deletions
+deletion_counts <- mutation_count[grep("d.", mutations)]
+deletion_total <- sum(deletion_counts)
+deletion_bias <- 100 * deletion_counts / deletion_total
+
+#dA        dT        dG        dC
+#24.25770 17.99132 39.44415 18.30683
+
+deletion_rate <- 100 * deletion_total / mis_matches
+
+#[1] 17.93361
+
+#insertions
+insertion_counts <- mutation_count[grep(".i", mutations)]
+insertion_total <- sum(insertion_counts)
+insertion_bias <- 100 * insertion_counts / insertion_total
+
+#Ai        Ti        Gi        Ci
+#23.13509 19.63951 30.05906 27.16633
+
+insertion_rate <- 100 * insertion_total / mis_matches
+
+#[1] 31.16293
 
 # Need to connect error types with the quality scores
-error.quality <- read.table(file=paste("pipeline_dev/", regions[1], "/", regions[1], ".mock.filter.error.quality", sep=""), header=T, row.names=1)[-7,]
-for(i in 2:length(regions)){
-	error.quality <- error.quality + read.table(file=paste("pipeline_dev/", regions[i], "/", regions[i], ".mock.filter.error.quality", sep=""), header=T, row.names=1)[-7,]
-}
-error.quality["72",] <- error.quality["72",] + error.quality["80",]
-error.quality <- error.quality[1:72,]
+error_quality <- read.table(file="data/process/mock.quality.report", header=T, row.names=1)
 
 #get quantiles
-matches <- rep(as.numeric(rownames(error.quality)), error.quality$matches)
+matches <- rep(as.numeric(rownames(error_quality)), error_quality$matches)
 m.quant <- quantile(matches, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
-
-subs <- rep(as.numeric(rownames(error.quality)), error.quality$substitutions)
+m.94 <- 100 * sum(matches >= 94)/length(matches)
+# [1] 80.52649
+
+subs <- rep(as.numeric(rownames(error_quality)), error_quality$substitutions)
 s.quant <- quantile(subs, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
-
-ins <- rep(as.numeric(rownames(error.quality)), error.quality$insertions)
+s.94 <- 100 * sum(subs >= 94)/length(subs)
+# [1] 79.96451
+
+ins <- rep(as.numeric(rownames(error_quality)), error_quality$insertions)
 i.quant <- quantile(ins, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
-
-ambig <- rep(as.numeric(rownames(error.quality)), error.quality$ambiguous)
-a.quant <- quantile(ambig, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
-```
+i.94 <- 100 * sum(ins >= 94)/length(ins)
+# [1] 80.35283
+```
+
 ***The PacBio error profile.***
@@
 rate associated with sequencing the 16S rRNA gene. We observed an average
-sequencing error rate of `r format(init.error.rate, digits=3, nsmall=2)`%. Insertions, deletions, substitutions, and ambiguous base calls
-accounted for `r format(insertions, digits=3, nsmall=1)`,
-`r format(deletions, digits=3, nsmall=1)`, `r format(substitutions, digits=3, nsmall=1)`,
-and `r format(ambiguous, digits=2, nsmall=1)`% of
-the errors, respectively. The substitution errors were equally likely and all four bases
+sequencing error rate of `r format(100*init_error_rate, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
+accounted for `r format(insertion_rate, digits=3, nsmall=1)`,
+`r format(deletion_rate, digits=3, nsmall=1)`, and `r format(substitution_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
+likely and all four bases
 were equally likely to cause insertion errors. Interestingly, guanines
-(`r format(delete.bias["rG"], digits=3, nsmall=1)`%) and cytosines
-(`r format(delete.bias["rC"], digits=3, nsmall=1)`%) were more likely to be deleted
-than adenines (`r format(delete.bias["rA"], digits=3, nsmall=1)`%) or thymidines
-(`r format(delete.bias["rT"], digits=2, nsmall=1)`%). When we considered the Phred
-quality score of each base call, we observed a median quality score of `r m.quant["50%"]`
-for correct base calls and scores of `r s.quant["50%"]` and `r i.quant["50%"]` for
-substitutions and insertions, respectively (Figure 1A). Although there
-was a broad distribution of quality scores with each type of base call, the errors could
-largely be distinguished from the correct base calls.
+(`r format(deletion_bias["dG"], digits=3, nsmall=1)`%) and adenines
+(`r format(deletion_bias["dA"], digits=3, nsmall=1)`%) were more likely to be deleted
+than cytosines (`r format(deletion_bias["dC"], digits=3, nsmall=1)`%) or thymidines
+(`r format(deletion_bias["dT"], digits=2, nsmall=1)`%). The PacBio quality values varied between
+`r paste(range(c(matches, subs, ins)), collapse=' and ')`.
+Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(m.94, digits=3, nsmall=1)`%), substitutions (`r format(s.94, digits=3, nsmall=1)`%), and insertions (`r format(i.94, digits=3, nsmall=1)`%).
+Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(cor(mock_error_good$error, mock_error_good$pred_error), digits=2, nsmall=2)`; Figure 1A).
 
@@
 of the 16S rRNA gene, or that was chimeric. This reduced the experiment-wide
-error rate from `r format(init.error.rate, digits=3, nsmall=2)` to
+error rate from `r format(init_error_rate, digits=3, nsmall=2)` to
 `r format(composite.basic.error, digits=2, nsmall=2)`%. This basic procedure
@@
 
+[revise - Fig 1A plot predicted error rate vs observed error rate]
 **Figure 1. Summary of errors in data generated using PacBio sequencing platform

### 2016-01-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Update code and text for basic error correction section

@@
 
-regions <- c("v4", "v13", "v35", "v15", "v16", "v19")
+regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
 clrs <- rainbow(length(regions))
-pretty.region <- c("v13"="V1-V3", "v15"="V1-V5", "v16"="V1-V6", "v19"="V1-V9", "v35"="V3-V5", "v4"="V4")
+pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")
 names(clrs) <- regions
@@
 # [1] 80.35283
-```
-
-***The PacBio error profile.***
+
+## Basic error correction analysis
+get_raw_error <- function(region){
+	region_data <- mock_error_good[mock_error_good$region == region,]
+	return(100*sum(region_data$mismatches)/sum(region_data$total))
+}
+
+raw_rates <- sapply(regions, get_raw_error)
+
+get_basic_kept <- function(region){
+	region_data <- mock_error_good[mock_error_good$region == region,]
+	100 * sum(region_data$good_coords & region_data$good_homop)/nrow(region_data)
+}
+
+basic_kept <- sapply(regions, get_basic_kept)
+
+get_basic_error <- function(region){
+	region_data <- mock_error_good[mock_error_good$region == region & mock_error_good$good_coords & mock_error_good$good_homop,]
+	return(100*sum(region_data$mismatches)/sum(region_data$total))
+}
+
+basic_rates <- sapply(regions, get_basic_error)
+
+mock_error_basic <- mock_error_good[mock_error_good$good_coords & mock_error_good$good_homop,]
+
+basic_error_rate <- 100 * sum(mock_error_basic$mismatches) / sum(mock_error_basic$total)		#[1] 0.6448736
+```
+
+***The PacBio error profile and a basic sequence curation procedure.***
 To build a sequence curation pipeline, we first needed to characterize the error
@@
 
+We established a simple curation procedure by culling any sequence that had a
+string of the same base repeated more than 8 times  or did not start and end at
+the expected alignment coordinates for that region of the 16S rRNA gene. This
+reduced the experiment-wide error rate from `r format(init_error_rate, digits=3, nsmall=2)` to
+`r format(basic_error_rate, digits=2, nsmall=2)`%. This basic procedure
+resulted in the removal of between
+`r format(min(100-basic_kept), digits=3, nsmall=1)`
+(`r pretty_region[names(which.min(100-basic_kept))]`) and
+`r format(max(100-basic_kept), digits=3, nsmall=1)`
+(`r pretty_region[names(which.max(100-basic_kept))]`)% of the reads (Table 2). Although
+the percentage of reads removed increased with the length of the fragment, there
+was no obvious relationship between fragment length and error rate (Figure 2).
+
 ```{r initialError}
-getInitError <- function(folder){
-	error <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".mock.filter.error.summary", sep=""), header=T, row.names=1)
-	error <- error[error$numparents==1,]
-	return(sum(error$mismatches)/sum(error$total))
-}
-
-init.rates <- unlist(lapply(regions, getInitError))
-names(init.rates) <- regions
-
-getReasonsLost <- function(region){
-	composite <- read.table(file=paste("pipeline_dev/", region, "/", region, ".composite", sep=""), header=T, row.names=1)
-	reasons <- as.character(composite$reason)
-
-	nseqs <- length(reasons)
-	good <- sum(reasons=="g")	/ nseqs
-	start.stop <- sum(grepl("s", reasons) | grepl("e", reasons)) / nseqs
-	homop <- length(grep("h", reasons)) / nseqs
-	ambig <- length(grep("n", reasons)) / nseqs
-	return(c(good=good, start.stop=start.stop, homop=homop, ambig=ambig, ngood=nseqs))
-}
-
-reasons.lost <- matrix(unlist(lapply(regions, getReasonsLost)), ncol=5, byrow=T)
-colnames(reasons.lost) <- c("good", "start.stop", "homop", "ambig", "nseqs")
-rownames(reasons.lost) <- regions
-
-orig.nseqs <- reasons.lost[,"nseqs"]
-basic.nseqs <- orig.nseqs * reasons.lost[,1]
-reasons.lost <- reasons.lost[,-5]
-
-getBasicError <- function(folder){
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
-	good <- composite[composite$reason == "g",]
-	return(sum(good$mismatches)/sum(good$total))
-}
-
-basic.error <- unlist(lapply(regions, getBasicError))
-names(basic.error) <- regions
-#         v4         v13         v35         v15         v16         v19
-#0.005373722 0.014743697 0.013730112 0.009027326 0.014249660 0.013027623
-composite.basic.error <- 100 * sum(basic.error*basic.nseqs)/(sum(basic.nseqs))
-```
-
-***A basic sequence curation procedure.***
-To establish a simple curation procedure, we culled any sequence that contained
-an ambiguous base call, had a string of the same base repeated 9 or more times,
-did not start and end at the expected alignment coordinates for that region
-of the 16S rRNA gene, or that was chimeric. This reduced the experiment-wide
-error rate from `r format(init_error_rate, digits=3, nsmall=2)` to
-`r format(composite.basic.error, digits=2, nsmall=2)`%. This basic procedure
-resulted in the removal of between
-`r format(round(min(100*(1-reasons.lost[,"good"])), digits=1), nsmall=1)`
-(`r pretty.region[names(which.min(100*(1-reasons.lost[,"good"])))]`) and
-`r round(max(100*(1-reasons.lost[,"good"])), digits=1)`
-(`r pretty.region[names(which.max(100*(1-reasons.lost[,"good"])))]`)% of the
-reads. The percentage of reads removed increased with the length of the
-fragment (Figure 2). The number of reads removed
-because of the presence of ambiguous base calls was similar to the number of
-reads that were removed for not fully aligning to the correct region within the
-16S rRNA gene (Table 2). The latter class of errors was
-generally due to sequence truncations that could not be explained.
+
+```
 

### 2016-01-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Update code and text for Identifying correlates of increased sequencing error section

@@
 mock_error_basic <- mock_error_good[mock_error_good$good_coords & mock_error_good$good_homop,]
+basic_n <- aggregate(mock_error_basic$error, by=list(mock_error_basic$region), length)$x
+names(basic_n) <- names(basic_rates)
 
@@
 
-```{r initialError}
-
-```
-
-```{r barcodePrimerAnalysis}
-getFragment_BPErrorRates <- function(folder){
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-	bc.primer <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs
-
-	return(cbind(composite$error, bc.primer))
-}
-
-errors <- lapply(regions, getFragment_BPErrorRates)
-
-errors.table <- matrix(errors[[1]], ncol=2, byrow=F)
-err.bcprimer <- aggregate(100*errors.table[,1], by=list(errors.table[,2]),
-													function(x){c(mean(x), length(x))})
-
-getBCPrimerError <- function(folder){
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
-																		 ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-	mismatch <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs
-
-	good0 <- mismatch == 0
-	composite.good0 <- composite[good0,]
-
-	good1 <- mismatch <= 1
-	composite.good1 <- composite[good1,]
-
-	return(c(mean(composite.good0[,"error"]), nrow(composite.good0)/nrow(composite),
-	mean(composite.good1[,"error"]), nrow(composite.good1)/nrow(composite) ))
-}
-
-bcprimer.error <- matrix(unlist(lapply(regions, getBCPrimerError)), nrow=6, byrow=T)
-rownames(bcprimer.error) <- regions
-colnames(bcprimer.error) <- c("0.error", "0.frac", "1.error", "1.frac")
-
-#         0.error    0.frac     1.error    1.frac
-# v4  0.002616919 0.6417469 0.003711755 0.9081720
-# v13 0.011102712 0.5799452 0.012493713 0.8675194
-# v35 0.010838504 0.6464069 0.011901910 0.8934573
-# v15 0.005375053 0.5079055 0.007029568 0.8338150
-# v16 0.010374438 0.5456238 0.011893323 0.8609559
-# v19 0.008543379 0.5072207 0.010086351 0.8237548
-```
-```{r coverageAnalysis}
-errorAtCoverageCutoff <- function(folder){
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-
-	total.seqs <- nrow(composite)
-
-	deep <- composite$freq >= 10
-	composite <- composite[deep,]
-
-	covered.seqs <- nrow(composite)
-
-	error <- sum(composite$mismatches)/sum(composite$total)
-	frac.kept <- covered.seqs / total.seqs
-	return(c(error=error, kept=frac.kept))
-}
-
-coverage.error <- unlist(lapply(regions, errorAtCoverageCutoff))
-coverage.error <- matrix(coverage.error, ncol=2, byrow=T)
-rownames(coverage.error) <- regions
-colnames(coverage.error) <- c("10.error", "10.frac")
-
-#       10.error   10.frac
-#v4  0.002524566 0.7897874
-#v13 0.010359113 0.6592051
-#v35 0.010084961 0.6642832
-#v15 0.006534609 0.6717103
-#v16 0.010039601 0.4605835
-#v19 0.009506114 0.5620395
-
-coverage.reduction <- (1-coverage.error[,"10.error"]/basic.error)*100
-#      v4      v13      v35      v15      v16      v19
-#53.02015 29.73870 26.54859 27.61302 29.54498 27.03110
-```
-```{r qualityScoreAnalysis}
-getErrorRateFromAveQ <- function(folder, threshold=60){
-	write(folder, "")
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-
-	rates <- composite[composite$aveQ >= threshold,]
-
-	remaining.error <- mean(rates[,"error"])
-	fraction.kept <- nrow(rates)/nrow(composite)
-
-	return(c(remaining.error, fraction.kept))
-}
-
-aveq.error <- matrix(unlist(lapply(regions, getErrorRateFromAveQ)), ncol=2, byrow=T)
-rownames(aveq.error) <- regions
-colnames(aveq.error) <- c("60.error", "60.frac")
-
-#       60.error   60.frac
-#v4  0.002623098 0.8692667
-#v13 0.009914371 0.7608497
-#v35 0.009035060 0.7697533
-#v15 0.004553350 0.7005270
-#v16 0.008921902 0.6561142
-#v19 0.005714256 0.5010315
-
-quality.reduction <- (1-aveq.error[,"60.error"]/basic.error)*100
-```
-```{r oligosCoverageAnalysis}
-oligosCoverage <- function(folder){
-	write(folder, "")
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-
-	mismatch <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs
-	good0 <- mismatch == 0 & composite$freq >= 10
-	composite.good0 <- composite[good0,]
-
-	good1 <- mismatch <= 1 & composite$freq >= 10
-	composite.good1 <- composite[good1,]
-
-	return(c(mean(composite.good0[,"error"]), nrow(composite.good0)/nrow(composite),
-	mean(composite.good1[,"error"]), nrow(composite.good1)/nrow(composite) ))
-}
-
-oligosCoverage.error <- matrix(unlist(lapply(regions, oligosCoverage)), ncol=4, byrow=T)
-rownames(oligosCoverage.error) <- regions
-colnames(oligosCoverage.error) <- c("0.10.error", "0.10.frac", "1.10.error", "1.10.frac")
+```{r barcode_primer_analysis}
+mock_error_basic$pbdiffs <- mock_error_basic$fbdiffs + mock_error_basic$rbdiffs + mock_error_basic$fpdiffs + mock_error_basic$rpdiffs
+
+#This is for the Figure 2B plot...
+pb_overall <- aggregate(mock_error_basic$error, by=list(mock_error_basic$pbdiffs), function(x){c(mean=mean(100*x))})
+
+# want to get the error rate and fraction of sequences for each region with a
+# threshold of mismatches or fewer
+
+pb_by_region_mm <- aggregate(mock_error_basic$error, by=list(mock_error_basic$pbdiffs, mock_error_basic$region), function(x){c(mean=mean(100*x), nseqs=length(x))})
+
+aggregate_pb_data <- function(region){
+	pb_region <- pb_by_region_mm[pb_by_region_mm$Group.2 == region,]
+
+	product <- pb_region$x[,"mean"] * pb_region$x[,"nseqs"]
+	pb_region_nseqs <- cumsum(pb_region$x[,"nseqs"])
+	pb_region_error <- cumsum(product) / pb_region_nseqs
+	pb_region_frac <- pb_region_nseqs / sum(pb_region$x[,"nseqs"])
+	c(error=pb_region_error, fraction=pb_region_frac)
+}
+
+pb_data <- sapply(regions, aggregate_pb_data)
+
+pb_error <- pb_data[grep("error", rownames(pb_data)),]
+rownames(pb_error) <- 0:6
+
+pb_fraction <- pb_data[grep("fraction", rownames(pb_data)),]
+rownames(pb_fraction) <- 0:6
+
+pb_error_reduction_0 <- 100 * (1 - pb_error["0",]/basic_rates)
+pb_error_reduction_1 <- 100 * (1 - pb_error["1",]/basic_rates)
+
+pb_nseqs_reduction_0 <- 100 * (1 - pb_fraction["0",])
+pb_nseqs_reduction_1 <- 100 * (1 - pb_fraction["1",])
+```
+
+```{r coverage_analysis}
+#This is for the Figure 2C plot...
+coverage_overall_agg <- aggregate(100 * mock_error_basic$error, by=list(mock_error_basic$coverage), mean)
+coverage_overall <- coverage_overall_agg$x
+names(coverage_overall) <- coverage_overall_agg$Group.1
+coverage_overall <- coverage_overall[as.character(3:40)]
+
+# need to get the error rate for coverage values between 3 and 40 for each region
+
+error_by_coverage <- function(region){
+	coverage_region <- mock_error_basic[mock_error_basic$region == region,]
+	coverage_error_agg <- aggregate(100 * coverage_region$error, by=list(coverage_region$coverage), mean)
+	coverage_error <- coverage_error_agg[coverage_error_agg$Group.1 <= 40,"x"]
+	names(coverage_error) <- coverage_error_agg[coverage_error_agg$Group.1 <= 40,"Group.1"]
+	coverage_error
+}
+
+coverage_error <- sapply(regions, error_by_coverage)
+#seems to plateau right around 10
+
+error_at_below_coverage_cutoff <- function(region, cutoff=10){
+	region_data <- mock_error_basic[mock_error_basic$region == region,]
+
+	deep <- region_data$coverage >= 10
+	coverage_data <- region_data[deep,]
+
+	error <- 100 * sum(coverage_data$mismatches)/sum(coverage_data$total)
+	frac_kept <- nrow(coverage_data) / nrow(region_data)
+	c(error=error, fraction=frac_kept)
+}
+
+coverage_data <- sapply(regions, error_at_below_coverage_cutoff)
+
+coverage_error <- coverage_data["error",]
+coverage_fraction <- coverage_data["fraction",]
+
+coverage_error_reduction <- 100 * (1 - coverage_error/basic_rates)
+coverage_nseqs_reduction <- 100 * (1 - coverage_fraction)
+```
+
+```{r predicted_error_analysis }
+#This is for the Figure 2D plot...
+pred_overall <- aggregate(100 * mock_error_basic$error, by=list(round(mock_error_basic$pred_error, 3)), mean)
+
+error_at_above_pred_error_cutoff <- function(region, threshold=0.999){
+	region_data <- mock_error_basic[mock_error_basic$region == region,]
+
+	above_threshold <- region_data[region_data$pred_error >= threshold,]
+
+	c(error=100*sum(above_threshold$mismatches)/sum(above_threshold$total), nseqs=nrow(above_threshold)/nrow(region_data))
+}
+
+pred_error_data <- sapply(regions, error_at_above_pred_error_cutoff, threshold=0.999)
+
+pred_error_reduction <- 100 * (1 - pred_error_data["error",]/basic_rates)
+pred_nseqs_reduction <- 100 * (1 - pred_error_data["nseqs",])
+```
+
+```{r oligos_coverage_analysis}
+get_oligos_coverage <- function(region){
+	region_data <- mock_error_basic[mock_error_basic$region == region,]
+
+	good0 <- region_data$pbdiffs <= 0 & region_data$coverage >= 10
+	region_data_good0 <- region_data[good0,]
+
+	good1 <- region_data$pbdiffs <= 1 & region_data$coverage >= 10
+	region_data_good1 <- region_data[good1,]
+
+	c(	100*mean(region_data_good0[,"error"]), nrow(region_data_good0)/nrow(region_data),
+		100*mean(region_data_good1[,"error"]), nrow(region_data_good1)/nrow(region_data))
+}
+
+oligos_coverage_error <- matrix(unlist(lapply(regions, get_oligos_coverage)), ncol=4, byrow=T)
+rownames(oligos_coverage_error) <- regions
+colnames(oligos_coverage_error) <- c("0.10.error", "0.10.frac", "1.10.error", "1.10.frac")
 
 #     0.10.error 0.10.frac 1.10.error 1.10.frac
-#v4  0.001565052 0.5567846 0.002068184 0.7513111
-#v13 0.009371691 0.4353586 0.009899256 0.6180905
-#v35 0.009096211 0.4907043 0.009503092 0.6321058
-#v15 0.004050098 0.3803978 0.005253236 0.5899354
-#v16 0.007945951 0.2960894 0.008956478 0.4301676
-#v19 0.006706336 0.3188918 0.007903900 0.4927793
-```
+#V1V3  0.4515344 0.5898447  0.4632910 0.7968639
+#V1V5  0.4187311 0.5557715  0.4270938 0.7384810
+#V1V6  0.4399351 0.5410259  0.4460726 0.7167325
+#V1V9  0.4317629 0.4444776  0.4420337 0.5752849
+#V3V5  0.6421004 0.6854427  0.6555240 0.8102832
+#V4    0.6825316 0.5512332  0.7023460 0.8245533
+
+oligos_coverage_error_reduction <- 100 * (1 - oligos_coverage_error[,"1.10.error"]/basic_rates)
+oligos_coverage_nseqs_reduction <- 100 * (1 - oligos_coverage_error[,"1.10.frac"])
+```
+
 ```{r oligosQScoreAnalysis}
-oligosQScore <- function(folder){
-	write(folder, "")
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-
-	mismatch <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs
-
-	good <- mismatch == 0 & composite$aveQ >= 60
-	composite.good0 <- composite[good,]
-
-	good1 <- mismatch <= 1 & composite$aveQ >= 60
-	composite.good1 <- composite[good1,]
-
-	return(c(mean(composite.good0[,"error"]), nrow(composite.good0)/nrow(composite),
-	mean(composite.good1[,"error"]), nrow(composite.good1)/nrow(composite) ))
-}
-
-oligosQScore.error <- matrix(unlist(lapply(regions, oligosQScore)), ncol=4, byrow=T)
-rownames(oligosQScore.error) <- regions
-colnames(oligosQScore.error) <- c("0.60.error", "0.60.frac", "1.60.error", "1.60.frac")
-
-#     0.60.error 0.60.frac  1.60.error 1.60.frac
-#v4  0.001715135 0.6015066 0.002215499 0.8232097
-#v13 0.009153895 0.5013705 0.009700915 0.7108269
-#v35 0.008365403 0.5557740 0.008690839 0.7281015
-#v15 0.003217812 0.4102346 0.003993206 0.6282727
-#v16 0.007589961 0.4115456 0.008358922 0.6101800
-#v19 0.004870705 0.3209549 0.005294429 0.4653699
-```
-```{r coverageQScoreAnalysis}
-coverageQScore <- function(folder){
-	write(folder, "")
-		composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
-		composite <- composite[composite$reason == "g",]
-
-	good <- composite$freq >= 10 & composite$aveQ>= 60
-	composite.good <- composite[good,]
-
-	return(c(mean(composite.good[,"error"]), nrow(composite.good)/nrow(composite) ))
-}
-
-coverageQScore.error <- matrix(unlist(lapply(regions, coverageQScore)), ncol=2, byrow=T)
-rownames(coverageQScore.error) <- regions
-colnames(coverageQScore.error) <- c("10.60.error", "10.60.frac")
-
-#    10.60.error 10.60.frac
-#v4  0.002361882  0.7874988
-#v13 0.010016818  0.6541800
-#v35 0.009296289  0.6564176
-#v15 0.004481083  0.5824549
-#v16 0.009612120  0.4587213
-#v19 0.005692918  0.3884468
-```
-```{r allFiltersAnalysis}
-allFilters <- function(folder){
-write(folder, "")
-		composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder, ".composite", sep=""), header=T, row.names=1)
-		composite <- composite[composite$reason == "g",]
-	mismatch <- composite$fbdiffs + composite$rbdiffs + composite$fpdiffs + composite$rpdiffs
-
-good0 <- mismatch == 0 & composite$aveQ>= 60 & composite$freq >= 10
-composite.good0 <- composite[good0,]
-
-good1 <- mismatch <= 1 & composite$aveQ >= 60 & composite$freq >= 10
-composite.good1 <- composite[good1,]
-
-return(c(mean(composite.good0[,"error"]), nrow(composite.good0)/nrow(composite),
-mean(composite.good1[,"error"]), nrow(composite.good1)/nrow(composite) ))
-}
-
-allFilters.error <- matrix(unlist(lapply(regions, allFilters)), ncol=4, byrow=T)
-rownames(allFilters.error) <- regions
-colnames(allFilters.error) <- c("0.10.60.error", "0.10.60.frac", "1.10.60.error", "1.10.60.frac")
+get_oligos_pred_error <- function(region){
+	region_data <- mock_error_basic[mock_error_basic$region == region,]
+
+	good0 <- region_data$pbdiffs <= 0 & region_data$pred_error >= 0.999
+	region_data_good0 <- region_data[good0,]
+
+	good1 <- region_data$pbdiffs <= 1 & region_data$pred_error >= 0.999
+	region_data_good1 <- region_data[good1,]
+
+	c(	100 * mean(region_data_good0[,"error"]), nrow(region_data_good0)/nrow(region_data),
+		100 * mean(region_data_good0[,"error"]), nrow(region_data_good0)/nrow(region_data) )
+}
+
+oligos_pred_error <- matrix(unlist(lapply(regions, get_oligos_pred_error)), ncol=4, byrow=T)
+rownames(oligos_pred_error) <- regions
+colnames(oligos_pred_error) <- c("0.999.error", "0.999.frac", "1.999.error", "1.999.frac")
+
+#     0.999.error 0.999.frac 1.999.error 1.999.frac
+#V1V3   0.2806049  0.4749822   0.2806049  0.4749822
+#V1V5   0.2177965  0.4266279   0.2177965  0.4266279
+#V1V6   0.2168503  0.4296550   0.2168503  0.4296550
+#V1V9   0.2177073  0.3438720   0.2177073  0.3438720
+#V3V5   0.2229757  0.5131411   0.2229757  0.5131411
+#V4     0.3485654  0.3961328   0.3485654  0.3961328
+
+oligos_pred_error_reduction <- 100 * (1 - oligos_pred_error[,"1.999.error"]/basic_rates)
+oligos_pred_nseqs_reduction <- 100 * (1 - oligos_pred_error[,"1.999.frac"])
+```
+
+```{r coverage_pred_error_analysis}
+get_coverage_pred <- function(region){
+	region_data <- mock_error_basic[mock_error_basic$region == region,]
+
+	good <- region_data$coverage >= 10 & region_data$pred_error >= 0.999
+	region_data_good <- region_data[good,]
+
+	c(100 * mean(region_data_good[,"error"]), nrow(region_data_good)/nrow(region_data) )
+}
+
+coverage_pred_error <- matrix(unlist(lapply(regions, get_coverage_pred)), ncol=2, byrow=T)
+rownames(coverage_pred_error) <- regions
+colnames(coverage_pred_error) <- c("10.999.error", "10.999.frac")
+
+#    10.999.error 10.999.frac
+#V1V3   0.2843963    65.86436
+#V1V5   0.2188792    56.55108
+#V1V6   0.2178326    57.33756
+#V1V9   0.2191284    42.66675
+#V3V5   0.2236951    60.74254
+#V4     0.3527900    61.79213
+
+coverage_pred_error_reduction <- 100 * (1 - coverage_pred_error[,"10.999.error"]/basic_rates)
+coverage_pred_nseqs_reduction <- 100 * (1 - coverage_pred_error[,"10.999.frac"])
+```
+
+```{r all_filters_analysis}
+get_all_filters <- function(region){
+	region_data <- mock_error_basic[mock_error_basic$region == region,]
+
+	good0 <- region_data$pbdiffs == 0 & region_data$pred_error >= 0.999 & region_data$coverage >= 10
+	region_data_good0 <- region_data[good0,]
+
+	good1 <- region_data$pbdiffs <= 1 & region_data$pred_error >= 0.999 & region_data$coverage >= 10
+	region_data_good1 <- region_data[good1,]
+
+	c(	100 * mean(region_data_good0[,"error"]), nrow(region_data_good0)/nrow(region_data),
+		100 * mean(region_data_good1[,"error"]), nrow(region_data_good1)/nrow(region_data) )
+}
+
+all_filters_error <- matrix(unlist(lapply(regions, get_all_filters)), ncol=4, byrow=T)
+rownames(all_filters_error) <- regions
+colnames(all_filters_error) <- c("0.10.999.error", "0.10.999.frac", "1.10.999.error", "1.10.999.frac")
 
@@
 #v19   0.004797594    0.2496316   0.005266467    0.3631005
+
+#     0.10.999.error 0.10.999.frac 1.10.999.error 1.10.999.frac
+#V1V3      0.2808218     0.4696170      0.2840387     0.6183602
+#V1V5      0.2184805     0.4192252      0.2182989     0.5374066
+#V1V6      0.2170989     0.4182253      0.2174090     0.5421862
+#V1V9      0.2176989     0.3213802      0.2191334     0.4016057
+#V3V5      0.2241187     0.5038275      0.2240779     0.5841414
+#V4        0.3492870     0.3932393      0.3526755     0.5665274
+
+all_filters_error_reduction <- 100 * (1 - all_filters_error[,"1.10.999.error"]/basic_rates)
+all_filters_nseqs_reduction <- 100 * (1 - all_filters_error[,"1.10.999.frac"])
+```
+
+```{r summary}
+base_error_reduction <- rbind(	pb_error_reduction_0,
+								pb_error_reduction_1,
+								pred_error_reduction,
+								coverage_error_reduction,
+								coverage_pred_error_reduction,
+								oligos_coverage_error_reduction,
+								oligos_pred_error_reduction,
+								all_filters_error_reduction)
+
+base_nseqs_reduction <- rbind(	pb_nseqs_reduction_0,
+								pb_nseqs_reduction_1,
+								pred_nseqs_reduction,
+								coverage_nseqs_reduction,
+								coverage_pred_nseqs_reduction,
+								oligos_coverage_nseqs_reduction,
+								oligos_pred_nseqs_reduction,
+								all_filters_nseqs_reduction)
 ```
@@
 allowing no mismatches to the barcodes and primers yielded the lowest error rate,
-that stringent criterion removed a large fraction of the reads from the dataset and
-allowing at most one mismatch marginally increased the error rate while preserving
-more sequences in the dataset (Figure 2). Second, we hypothesized
-that increased sequencing coverage should yield lower error rates. We found that
-once we had obtained 10-fold coverage of the fragments, the error rate did not
-change appreciably (Figure 1C). When we compared the error
-rates of reads with at least 10-fold coverage to those with less coverage, we
-reduced the error rate by
-`r paste(round(sort(coverage.reduction)[c(1,5)], digits=1), collapse=" to ")`%
-for each region except the V4 region for which the error rate was reduced by
-`r round(coverage.reduction["v4"], digits=1)`%. Third, based on the earlier
-analysis associating errors with quality scores, we used two quality score-based
-approaches for identifying reads with errors (Figure 3). We calculated the
-minimum average quality score across all 50-nt windows within each sequence and
-we also calculated the average quality score across each sequence. We then
-associated both methods of calculating the average quality score with the error
-rate of the reads and the fraction of sequences that would be retained if each
-threshold were selected. Using the sliding window approach we did not observe
-any clear break points indicating that one quality score would be better than
-another (Figure 3AB). In contrast, using the whole sequence quality score
-average we observed a decrease in the error rate and the fraction of sequences
-retained when the threshold was increased above 60 (Figure 3CD). When we used
-this threshold, we were able to reduce the error rate by
-`r paste(round(range(quality.reduction), digits=1), collapse=" to ")`%
-(Figure 2A). We noted that the fraction of reads retained
-decreased as the length of the fragment increased with retention of
-`r round(100*aveq.error["v4", "60.frac"],digits=1)`% of the V4 reads and
-`r round(100*aveq.error["v19", "60.frac"],digits=1)`% of the V1-V9 reads (Figure
-2B). Next, we asked whether which combinations of culling reads with mismatches
-to the expected barcodes and primers, less than 10-fold sequencing coverage, and
-an average quality score less than 60 made the most meaningful reductions in the
-error rate while preserving the most reads when implemented with the basic
-curation pipeline (Figure 2B). We observed similar error rates when we required
-one or fewer mismatches to the barcodes and primers and an average quality score
-above 60 as when we also required a minimum 10-fold coverage. Culling sequences
-that had more than one mismatch to the barcodes and primers and those with an
-average quality score less than 60 reduced the error rate to between
-`r paste(round(100*range(oligosQScore.error[,"1.60.error"]), digits=2), collapse=" and ")`.
-This procedure resulted in the removal of
-`r paste(rev(round(100*(1-range(oligosQScore.error[,"1.60.frac"])), digits=0)), collapse=" and ")`%
-of the reads (Figure 2). The remainder of this paper uses this sequence curation
-approach.
+that stringent criterion removed a large fraction of the reads from the dataset.
+Allowing at most one mismatch only marginally increased the error rate while
+retaining more sequences in the dataset (Figure 2). Second, we hypothesized that
+increased sequencing coverage should yield lower error rates. We found that once
+we had obtained 10-fold coverage of the fragments, the error rate did not change
+appreciably (Figure 1C). When we compared the error rates of reads with at least
+10-fold coverage to those with less coverage, we reduced the error rate by
+`r paste(format(range(coverage_reduction), digits=3, nsmall=1), collapse=" to ")`%
+(Figure 2). Third, based on the earlier analysis finding a relationship between the
+predicted error rate and the observed error rate, we sought a minimum predicted
+error rate that would allow us to reduce the observed error rate. The average
+observed error rate for sequences with predicted error rates between 0.01 and
+0.10% was linear (Figure 1D). We decided to use a threshold of 0.01% because a
+large number of sequence reads were lost when we used a smaller threshold. When
+we used this threshold, we were able to reduce the error rate by
+`r paste(format(range(pred_error_reduction), digits=3, nsmall=1), collapse=" to ")`%
+(Figure 2A). Finally, we quantified the effect of combining filters. We found that
+any combination of filters that included the predicted error rate threshold had
+the most significant impact on reducing the observed error rate. Furthermore,
+the inclusion of the mismatch and coverage filters had a negligible impact on
+error rates, but had a significant impact on the number of sequences included in
+the analysis. For instance, among the V1-V9 data, requiring sequences to have a
+predicted error rate less than 0.01% resulted in a `r format(base_error_reduction["pred_error_reduction","V1V9"], digits=3, nsmall=1)`% and resulted in the removal of `r format(base_nseqs_reduction["pred_error_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences.
+Adding the mismatch or coverage filter
+had no effect on the reduction of error, but resulted in the removal of
+`r format(base_nseqs_reduction["oligos_pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)` and
+`r format(base_nseqs_reduction["coverage_pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences, respectively. Use of all filters had no impact on the reduction in the observed
+error rate, but resulted in the removal of `r format(base_nseqs_reduction["all_filters_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences. The remainder of this paper only uses sequences
+with a predicted error rate less than 0.01%.
 

### 2016-01-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Some light editing

@@
 `r paste(format(range(coverage_reduction), digits=3, nsmall=1), collapse=" to ")`%
-(Figure 2). Third, based on the earlier analysis finding a relationship between the
-predicted error rate and the observed error rate, we sought a minimum predicted
+(Figure 2). Third, based on the observed correlation between the
+predicted and observed error rates, we sought to identify a minimum predicted
 error rate that would allow us to reduce the observed error rate. The average

### 2016-01-06 Schloss_PacBio16S_PeerJ_2016.Rmd

@@
 
+[ This figure removed ]
 **Figure 3. The relationship between the error rate of each region and the
@@
 were calculated by either determining the minimum value of the average quality
-score wihin all 50-nt windows within each region (A and B) or by calculating
+score within all 50-nt windows within each region (A and B) or by calculating
 the average quality score across the entire sequence read (C and D).

### 2016-01-06 code/get_unique_pc_fasta.sh
Process all samples/regions through pre.cluster step

@@ -1,0 +1,77 @@
+FASTA=$1
+
+STUB=$(echo $FASTA | sed -e 's/.fasta//')
+echo $STUB
+
+REGION=$(echo $FASTA | sed -re 's/.*\/(V[^.]*)\..*/\1/')
+echo $REGION
+
+START=0
+END=0
+
+
+# | Region | Start | End   | Length | diffs |
+# |--------|-------|-------|--------|-------|
+# | v13    | 1046  | 13125 | 489    |    4  | 
+# | v15    | 1046  | 27659 | 879    |    8  | 
+# | v16    | 1046  | 34113 | 1028   |   10  |
+# | v19    | 1046  | 43116 | 1458   |   14  |
+# | v35    | 6428  | 27659 | 545    |    5  |
+# | v4     | 13862 | 23444 | 253    |    2  |
+# |--------|-------|-------|--------|-------|
+
+if [ $REGION == V1V9 ]
+then
+	START=1046
+	END=43116
+	DIFFS=14
+elif [ $REGION == V1V3 ]
+then
+	START=1046
+	END=13125
+	DIFFS=4
+elif [ $REGION == V1V5 ]
+then
+    	START=1046
+        END=27659
+	DIFFS=8
+elif [ $REGION == V1V6 ]
+then
+    	START=1046
+        END=34113
+	DIFFS=10
+elif [ $REGION == V3V5 ]
+then
+    	START=6248
+        END=27659
+	DIFFS=5
+elif [ $REGION == V4 ]
+then
+    	START=13862
+        END=23444
+	DIFFS=2
+fi
+
+mothur "#unique.seqs(fasta=$FASTA);
+	align.seqs(fasta=current, reference=data/references/silva.seed.align, processors=8);
+	screen.seqs(fasta=current, name=current, start=$START, end=$END, maxhomop=8);
+	filter.seqs(vertical=T, trump=.);
+	unique.seqs(fasta=current, name=current);
+	pre.cluster(fasta=current, name=current, diffs=$DIFFS)"
+
+#$STUB.unique.good.filter.names
+#$STUB.unique.good.filter.unique.fasta
+#$STUB.unique.good.filter.unique.precluster.names
+#$STUB.unique.good.filter.unique.precluster.fasta
+
+rm $STUB.unique.good.filter.unique.precluster.map
+rm $STUB.unique.good.filter.fasta
+rm $STUB.unique.bad.accnos
+rm $STUB.good.names
[…]

### 2016-01-06 code/get_unique_pc_fasta.sh
Remove call to delete *.filter file

@@ -74,4 +74,3 @@ rm $STUB.unique.align
 rm $STUB.unique.flip.accnos
 rm $STUB.names
 rm $STUB.unique.fasta
-rm *.filter

### 2016-01-06 code/get_unique_pc_fasta.sh
Fix V3V5 coordinates

@@ -12,12 +12,12 @@ END=0
 
 # | Region | Start | End   | Length | diffs |
 # |--------|-------|-------|--------|-------|
-# | v13    | 1046  | 13125 | 489    |    4  | 
-# | v15    | 1046  | 27659 | 879    |    8  | 
-# | v16    | 1046  | 34113 | 1028   |   10  |
-# | v19    | 1046  | 43116 | 1458   |   14  |
-# | v35    | 6428  | 27659 | 545    |    5  |
-# | v4     | 13862 | 23444 | 253    |    2  |
+#x| v13    | 1046  | 13125 | 489    |    4  | 
+#x| v15    | 1046  | 27659 | 879    |    8  | 
+#x| v16    | 1046  | 34113 | 1028   |   10  |
+#x| v19    | 1046  | 43116 | 1458   |   14  |
+#x| v35    | 6428  | 27659 | 545    |    5  |
+#x| v4     | 13862 | 23444 | 253    |    2  |
 # |--------|-------|-------|--------|-------|
 
 if [ $REGION == V1V9 ]
@@ -42,7 +42,7 @@ then
 	DIFFS=10
 elif [ $REGION == V3V5 ]
 then
-    	START=6248
+    	START=6428
         END=27659
 	DIFFS=5
 elif [ $REGION == V4 ]

### 2016-01-07 consolidate_taxonomy.R
Synthesize taxonomy data

@@ -1,0 +1,29 @@
+#	read in a bunch of taxonomy files and figure out how deeply each dataset
+#	was classified. outputs a summary file into data/process/
+
+count_levels <- function(tax_string){
+	length(unlist(strsplit(tax_string, split=";")))
+}
+
+count_tax_levels <- function(combo){
+	taxonomy_file_name <- paste0("data/mothur_pool/",combo["region"],".",combo["dataset"],".screen.unique.good.filter.unique.precluster.pick.", combo["db"], ".wang.taxonomy")
+	taxonomy <- read.table(file=taxonomy_file_name, stringsAsFactors=FALSE)
+	classified <- gsub("unclassified;", "", taxonomy$V2)
+	depth_count <- table(factor(sapply(classified, count_levels), levels=0:7))
+	depth_count / sum(depth_count)
+}
+
+tax_levels <- 0:7
+regions <- c("V4", "V1V3", "V3V5", "V1V5", "V1V6", "V1V9")
+databases <- c("pds", "gg", "nr")
+datasets <- c("mock", "mouse", "human", "soil")
+
+combos <- cbind(region=rep(regions, each=length(databases)*length(datasets)),
+				dataset=rep(datasets, length(regions)*length(databases)),
+				db=rep(rep(databases, each=length(datasets)), length(regions)))
+
+tax_level_counts <- sapply(1:nrow(combos), function(x){count_tax_levels(combos[x,])})
+genus_species <- tax_level_counts[,"6"] + tax_level_counts[,"7"]
+composite <- cbind(combos, tax_level_counts, genus_species)
+
+write.table(file="taxonomy.depth.analysis", composite, quote=F, row.names=F)

### 2016-01-07 code/consolidate_taxonomy.R
Fix path for output

@@ -26,4 +26,4 @@ tax_level_counts <- sapply(1:nrow(combos), function(x){count_tax_levels(combos[x
 genus_species <- tax_level_counts[,"6"] + tax_level_counts[,"7"]
 composite <- cbind(combos, tax_level_counts, genus_species)
 
-write.table(file="taxonomy.depth.analysis", composite, quote=F, row.names=F)
+write.table(file="data/process/taxonomy.depth.analysis", composite, quote=F, row.names=F)

### 2016-01-07 code/consolidate_taxonomy.R
Transpose output from sapply

@@ -22,7 +22,7 @@ combos <- cbind(region=rep(regions, each=length(databases)*length(datasets)),
 				dataset=rep(datasets, length(regions)*length(databases)),
 				db=rep(rep(databases, each=length(datasets)), length(regions)))
 
-tax_level_counts <- sapply(1:nrow(combos), function(x){count_tax_levels(combos[x,])})
+tax_level_counts <- t(sapply(1:nrow(combos), function(x){count_tax_levels(combos[x,])}))
 genus_species <- tax_level_counts[,"6"] + tax_level_counts[,"7"]
 composite <- cbind(combos, tax_level_counts, genus_species)
 

### 2016-01-07 Schloss_PacBio16S_PeerJ_2016.Rmd
Update unique/precluster error analysis code

@@
 `r format(base_nseqs_reduction["coverage_pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences, respectively. Use of all filters had no impact on the reduction in the observed
-error rate, but resulted in the removal of `r format(base_nseqs_reduction["all_filters_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences. The remainder of this paper only uses sequences
-with a predicted error rate less than 0.01%.
+error rate, but resulted in the removal of `r format(base_nseqs_reduction["all_filters_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences.
+The remainder of this paper only uses sequences with a predicted error rate less than 0.01%.
 
 ```{r precluster}
-summarizeError <- function(folder, method){
-  file1 <- paste("analysis/", folder, "/", folder, ".mock1.", method, ".error.summary", sep="")
-  file2 <- paste("analysis/", folder, "/", folder, ".mock2.", method, ".error.summary", sep="")
-  file3 <- paste("analysis/", folder, "/", folder, ".mock3.", method, ".error.summary", sep="")
-
-  summary <- rbind(read.table(file=file1, header=T, row.names=1), read.table(file=file2, header=T, row.names=1), read.table(file=file3, header=T, row.names=1))
-  nochim <- summary[summary$numparents==1,]
-  error <- sum(nochim$weight * nochim$mismatches) / sum(nochim$weight * nochim$total)
-  return(error)
-}
-
-unique.error <- unlist(lapply(regions, summarizeError, "unique"))
-precluster.error <- unlist(lapply(regions, summarizeError, "precluster"))
-final.error <- cbind(unique.error, precluster.error)
-rownames(final.error) <- regions
-
-#    unique.error precluster.error
-#v4   0.002277391      0.001418871
-#v13  0.009081866      0.007722886
-#v35  0.009719159      0.008257968
-#v15  0.004740243      0.002645747
-#v16  0.007130228      0.005676505
-#v19  0.005158478      0.003163881
-
-reduction <- 100*(1-final.error[,"precluster.error"]/final.error[,"unique.error"])
-rnd.reduction <- round(reduction, 0)
-min.rnd.red <- min(rnd.reduction)
-max.rnd.red <- max(rnd.reduction)
+get_error_rate <- function(file_name){
+	error_summary <- read.table(file=file_name, header=T, stringsAsFactors=FALSE)
+	nochim <- error_summary[error_summary$numparents==1,]
+    sum(nochim$weight * nochim$mismatches) / sum(nochim$weight * nochim$total)
+}
+
+regions <- c("V4", "V1V3", "V1V5", "V1V6", "V1V9", "V3V5")
+
+unique_file_names <- paste0("data/mothur_pool/", regions, ".mock.screen.unique.good.filter.unique.error.summary")
+
+unique_error <- sapply(unique_file_names, get_error_rate)
+
+precluster_file_names <- paste0("data/mothur_pool/", regions,".mock.screen.unique.good.filter.unique.precluster.error.summary")
+
+precluster_error <- sapply(precluster_file_names, get_error_rate)
+
+pooled_error <- cbind(unique = unique_error, precluster = precluster_error)
+rownames(pooled_error) <- regions
+
+#          unique   precluster
+#V4   0.003205151 0.0015805612
+#V1V3 0.002529819 0.0006841053
+#V1V5 0.001987322 0.0001938975
+#V1V6 0.001910158 0.0003086235
+#V1V9 0.001944654 0.0002743820
+#V3V5 0.002076498 0.0003532300
+
+reduction <- 100*(1-pooled_error[,"precluster"]/pooled_error[,"unique"])
 
 min.region <- paste(pretty.region[names(rnd.reduction[rnd.reduction==min.rnd.red])],collapse=" and ")
-
 max.region <- pretty.region[[names(rnd.reduction[rnd.reduction==max.rnd.red])]]
@@
 more abundant sequence if the rare sequences have fewer mismatches than a
-defined threshold when compared to the more abundant sequence. The recommended
+defined threshold when compared to the more abundant sequence [refs]. The recommended
 threshold was a 1-nt difference per 100-nt of sequence data. For example, the
@@
 unable to differentiate V1-V9 sequences that truly differed by less than 14 nt.
-When we applied
-this approach to our PacBio data, we observed a reduction in the error rate
+When we applied this approach to our PacBio data, we observed a reduction in the error rate
 between `r min.rnd.red` (`r min.region`) and `r max.rnd.red`% (`r max.region`).
 The final error rates varied between
-`r round(min(100*final.error[,"precluster.error"]), digits=2)`
-(`r pretty.region[names(which.min(final.error[,"precluster.error"]))]`)
-and `r round(max(100*final.error[,"precluster.error"]), digits=2)`%
-(`r pretty.region[names(which.max(final.error[,"precluster.error"]))]`); the
-full-length, V1-V9, fragments had an error rate of
-`r round(100* final.error["v19","precluster.error"], digits=2)`%
-(Figure 2B). These error rates are 7-40
-times higher than what we have previously observed using the 454 and
+`r round(100*min(pooled_error[,"precluster"]), digits=2)` (`r names(which.min(pooled_error[,"precluster"]))`) and `r round(100*max(pooled_error[,"precluster"]), digits=2)`% (`r names(which.max(pooled_error[,"precluster"]))`).
+ The full-length (i.e. V1V9) fragments had an error rate of
+`r round(100* pooled_error["V1V9","precluster"], digits=2)`%
+(Figure 2B), this is similar to what we have previously observed using the 454 and
 Illumina MiSeq platforms (0.02%){Schloss, 2011 #2466; Kozich, 2013 #2719}

### 2016-01-27 Schloss_PacBio16S_PeerJ_2016.Rmd
Complete code edits for Results section

@@
 
-min.region <- paste(pretty.region[names(rnd.reduction[rnd.reduction==min.rnd.red])],collapse=" and ")
-max.region <- pretty.region[[names(rnd.reduction[rnd.reduction==max.rnd.red])]]
+min.region <- pretty_region[names(reduction[which.min(reduction)])]
+max.region <- pretty_region[names(reduction[which.max(reduction)])]
 ```
@@
 more abundant sequence if the rare sequences have fewer mismatches than a
-defined threshold when compared to the more abundant sequence [refs]. The recommended
+defined threshold when compared to the more abundant sequence [@refs]. The recommended
 threshold was a 1-nt difference per 100-nt of sequence data. For example, the
@@
 When we applied this approach to our PacBio data, we observed a reduction in the error rate
-between `r min.rnd.red` (`r min.region`) and `r max.rnd.red`% (`r max.region`).
+between `r min(reduction)` (`r min.region`) and `r max(reduction)`% (`r max.region`).
 The final error rates varied between
@@
 ```{r}
+getAveSobs <- function(region, sample){
+	file_name <- paste0("data/mothur_pool/", region, ".", sample, ".screen.unique.good.filter.unique.precluster.pick.an.ave-std.summary")
+
+	data_sobs <- NA
+
+	if(file.exists(file_name)){
+		data <- read.table(file=file_name, header=T)
+		data_sobs <- data[data$method=="ave", "sobs"]
+	}
+
+	data_sobs
+}
+
 getOTUResults <- function(region){
-  perfect <- read.table(file=paste0("analysis/", region, "/HMP_MOCK.filter.pick.phylip.an.summary"), header=T)
-  nochims <- read.table(file=paste0("analysis/", region, "/", region, ".mock.precluster.perfect.pick.an.ave-std.summary"), header=T)
-  observed <- read.table(file=paste0("analysis/",region, "/", region, ".trim.unique.good.filter.unique.precluster.pick.an.merge.groups.ave-std.summary"), header=T)
-	obs.mock <- observed[observed$group=="mock"   & observed$method=="ave", "sobs"]
-	obs.soil <- observed[observed$group=="soil"   & observed$method=="ave", "sobs"]
-	obs.mouse <- observed[observed$group=="mouse" & observed$method=="ave", "sobs"]
-	obs.human <- observed[observed$group=="human" & observed$method=="ave", "sobs"]
+ 	perfect <- read.table(file=paste0("data/mothur_pool/HMP_MOCK.", region, ".pick.phylip.an.summary"), header=T)
+ 	nochims <- read.table(file=paste0("data/mothur_pool/", region, ".mock.screen.unique.good.filter.unique.precluster.perfect.an.ave-std.summary"), header=T)
+
+ 	obs.mock <- getAveSobs(region, "mock")
+ 	obs.mouse <- getAveSobs(region, "mouse")
+ 	obs.human <- getAveSobs(region, "human")
+ 	obs.soil <- getAveSobs(region, "soil")
+
 	nseqs <- nochims[nochims$method=="ave", "nseqs"]
 
-	return(c(perfect[1,2], nochims[1,3], obs.mock, obs.soil, obs.mouse, obs.human, nseqs))
+	return(c(perfect[1,2], nochims[1,"sobs"], obs.mock, obs.soil, obs.mouse, obs.human, nseqs))
 }
@@
 otu.table <- t(sapply(regions, getOTUResults))
-nseqs <- otu.table[1,7]
+nseqs <- otu.table[[1,7]]
 perfect.otu <- otu.table[1,1]
@@
 
-otu.table <- cbind(100 * final.error[rownames(otu.table),"precluster.error"], otu.table)
+otu.table <- cbind(100 * pooled_error[rownames(otu.table),"precluster"], otu.table)
 colnames(otu.table) <- c("Error Rate (%)", "Mock with no chimeras", "Observed mock", "Soil", "Mouse", "Human")
-rownames(otu.table) <- pretty.region[rownames(otu.table)]
+rownames(otu.table) <- pretty_region[rownames(otu.table)]
 
@@
 #|:-----|--------------:|---------------------:|-------------:|-----:|-----:|-----:|
-#|V4    |           0.14|                  26.0|          27.4| 335.8|  78.5|  58.1|
-#|V1-V3 |           0.77|                  53.6|          77.0| 436.6| 159.7| 116.8|
-#|V3-V5 |           0.83|                  83.1|         106.8| 388.9| 259.6|  91.6|
-#|V1-V5 |           0.26|                  34.6|          40.7| 403.2|  88.8|  74.7|
-#|V1-V6 |           0.57|                  57.4|          89.7| 431.9| 138.3|  95.9|
-#|V1-V9 |           0.32|                  29.5|          41.8| 390.9|  81.9| 102.0|
-
-getCoverage <- function(region){
-  observed <- read.table(file=paste0("analysis/",region, "/", region, ".trim.unique.good.filter.unique.precluster.pick.an.merge.groups.ave-std.summary"), header=T)
-	obs.mock <- observed[observed$group=="mock"   & observed$method=="ave", "coverage"]
-	obs.soil <- observed[observed$group=="soil"   & observed$method=="ave", "coverage"]
-	obs.mouse <- observed[observed$group=="mouse" & observed$method=="ave", "coverage"]
-	obs.human <- observed[observed$group=="human" & observed$method=="ave", "coverage"]
-
-	return(c(obs.mock, obs.soil, obs.mouse, obs.human))
-}
-
-coverage.table <- t(sapply(regions, getCoverage))
-colnames(coverage.table) <- c("Mock", "Soil", "Mouse", "Human")
-rownames(coverage.table) <- pretty.region[rownames(coverage.table)]
+#|V4    |           0.16|                  33.4|          48.7| 630.1| 212.4| 117.0|
+#|V1-V3 |           0.07|                  20.5|          27.7| 726.3| 107.5| 105.4|
+#|V3-V5 |           0.04|                  21.1|          31.2|    NA| 112.3|  83.7|
+#|V1-V5 |           0.02|                  19.8|          27.2| 694.4|  66.0|  85.3|
+#|V1-V6 |           0.03|                  20.0|          46.4| 693.3|  75.5|  89.2|
+#|V1-V9 |           0.03|                  19.5|          40.3|    NA| 102.7| 117.6|
+
 ```
@@
 (`r names(which.max(diff))`) extra OTUs. The range in the number of extra OTUs was largely
-explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, final.error[,"precluster.error"])$estimate, 2)`). Next, we determined
+explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, pooled_error[,"precluster"])$estimate, 2)`). Next, we determined
 the number of OTUs that were observed when we used UCHIME to identify chimeric sequence.
@@
 soil, mouse, and human samples using the same pipeline with chimera detection and removal
-based on the UCHIME algorithm. Again, we found that there was a strong correlation between
+based on the UCHIME algorithm. Suprisingly, there was not a clear relationship across sample type and region. Again, we found that there was a strong correlation between
 the number of observed OTUs and the error rate for the
-soil (R=`r format(round(cor.test(otu.table[,"Soil"], final.error[,"precluster.error"])$estimate, 2), nsmall=2)`),
-mouse (R=`r format(round(cor.test(otu.table[,"Mouse"], final.error[,"precluster.error"])$estimate, 2), nsmall=2)`),
-and human samples (R=`r format(round(cor.test(otu.table[,"Human"], final.error[,"precluster.error"])$estimate, 2), nsmall=2)`).
+mouse (R=`r format(round(cor.test(otu.table[,"Mouse"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`)
+and human samples (R=`r format(round(cor.test(otu.table[,"Human"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`).
 These results underscore the effect of sequencing error on the inflation of the number of
@@
 ```{r, classificationSummary}
-data <- read.table(file="taxonomy.depth.analysis", header=T)
-
-rdp <- cbind("Mock"=data[data$database=="rdp" & data$sample=="mock", "total"],
-						 "Human"=data[data$database=="rdp" & data$sample=="human", "total"],
-						 "Mouse"=data[data$database=="rdp" & data$sample=="mouse", "total"],
-						 "Soil"=data[data$database=="rdp" & data$sample=="soil", "total"])
+data <- read.table(file="data/process/taxonomy.depth.analysis", header=T)
+
+rdp <- 100 * cbind("Mock"=data[data$db=="pds" & data$dataset=="mock", "genus_species"],
+				 "Human"=data[data$db=="pds" & data$dataset=="human", "genus_species"],
+				 "Mouse"=data[data$db=="pds" & data$dataset=="mouse", "genus_species"],
+				 "Soil"=data[data$db=="pds" & data$dataset=="soil", "genus_species"])
 rownames(rdp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
 
-gg <- cbind("Mock"=data[data$database=="gg" & data$sample=="mock", "total"],
-						"Human"=data[data$database=="gg" & data$sample=="human", "total"],
-						"Mouse"=data[data$database=="gg" & data$sample=="mouse", "total"],
-						"Soil"=data[data$database=="gg" & data$sample=="soil", "total"])
+gg <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "genus_species"],
+				"Human"=data[data$db=="gg" & data$dataset=="human", "genus_species"],
+				"Mouse"=data[data$db=="gg" & data$dataset=="mouse", "genus_species"],
+				"Soil"=data[data$db=="gg" & data$dataset=="soil", "genus_species"])
 rownames(gg) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
 
-silva <- cbind("Mock"=data[data$database=="silva" & data$sample=="mock", "total"],
-							 "Human"=data[data$database=="silva" & data$sample=="human", "total"],
-							 "Mouse"=data[data$database=="silva" & data$sample=="mouse", "total"],
-							 "Soil"=data[data$database=="silva" & data$sample=="soil", "total"])
+silva <- 100 * cbind("Mock"=data[data$db=="nr" & data$dataset=="mock", "genus_species"],
+					 "Human"=data[data$db=="nr" & data$dataset=="human", "genus_species"],
+					 "Mouse"=data[data$db=="nr" & data$dataset=="mouse", "genus_species"],
+					 "Soil"=data[data$db=="nr" & data$dataset=="soil", "genus_species"])
 rownames(silva) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
 
-gg.sp <- cbind("Mock"=data[data$database=="gg" & data$sample=="mock", "X7"],
-							 "Human"=data[data$database=="gg" & data$sample=="human", "X7"],
-							 "Mouse"=data[data$database=="gg" & data$sample=="mouse", "X7"],
-							 "Soil"=data[data$database=="gg" & data$sample=="soil", "X7"])
+gg.sp <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "X7"],
+					 "Human"=data[data$db=="gg" & data$dataset=="human", "X7"],
+					 "Mouse"=data[data$db=="gg" & data$dataset=="mouse", "X7"],
+					 "Soil"=data[data$db=="gg" & data$dataset=="soil", "X7"])
 rownames(gg.sp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
@@
 ```{r}
-error <- read.table(file="analysis/v4/v4.trim.unique.good.filter.unique.pick.error.summary", header=T, row.names=1)
+
+getRegionNonRandom <- function(region){
+	file_name <- paste0("data/mothur_pool/", region, ".mock.screen.unique.good.filter.unique.error.summary")
+
+	error <- read.table(file=file_name, header=T, row.names=1)
 	error.nochim <- error[error$numparents==1,]
+
 	one.off <- error.nochim[error.nochim$mismatches==1,]
-one.off.sum <- sum(one.off$weight)
-v4.table <- table(one.off$weight)
-v4.total <- sum(error$weight)
-
-error <- read.table(file="analysis/v35/v35.trim.unique.good.filter.unique.pick.error.summary", header=T, row.names=1)
-error.nochim <- error[error$numparents==1,]
-one.off <- error.nochim[error.nochim$mismatches==1,]
-one.off.sum <- sum(one.off$weight)
-v35.table <- table(one.off$weight)
-v35.total <- sum(error$weight)
-
-error <- read.table(file="analysis/v15/v15.trim.unique.good.filter.unique.pick.error.summary", header=T, row.names=1)
-error.nochim <- error[error$numparents==1,]
-one.off <- error.nochim[error.nochim$mismatches==1,]
-one.off.sum <- sum(one.off$weight)
-v15.table <- table(one.off$weight)
-v15.total <- sum(error$weight)
+
+	one.off.n <- sum(one.off$weight)
+	one.off.table <- table(one.off$weight)
+	nseqs <- sum(error.nochim$weight)
+	one.off.table
+}
+
+non_random <- lapply(regions, getRegionNonRandom)
+names(non_random) <- regions
+
+percent_one_off <- 100 * sapply(non_random, function(x){x[1]/sum(x)})
+
 ```
@@
 Above, we described that although there was no obvious bias in the substitution
-or insertion rate, we did observe that guanines and cytosines were more likely
-to be deleted than adenines and thymidines. This lack of randomness in the error
+or insertion rate, we did observe that guanines and adenines were more likely to
+be deleted than cytosines or thymidines. This lack of randomness in the error
 profile suggested that there might be a systematic non-random distribution of
 the errors across the sequences. This would manifest itself by the creation of
-duplicate sequences with the same error. Because we were able to obtain a large
-number of reads from the mock communities where we sequenced the V4
-(N=`r v4.total`), V1-V5 (N=`r v15.total` sequences), and V3-V5 (N=`r v35.total`)
-regions, we investigated the mock community data from these regions further. We
-identified all of the sequences that had a 1-nt difference to the true sequence.
-For these three regions, a majority of the sequences with 1-nt errors were only
-observed once (V4: `r round(100*v4.table[1]/sum(v4.table), 1)`%, V1-V5:
-`r round(100*v15.table[1]/sum(v15.table), 1)`%, V3-V5:
-`r round(100*v35.table[1]/sum(v35.table), 1)`%). We found that the frequency of the
-most abundant 1-nt error paralleled the number of sequences. There were two sequences
-in the V4 dataset that occurred `r names(v4.table[length(v4.table)])` times, one
-sequence in the V1-V5 dataset that occurred `r names(v15.table[length(v15.table)])`
-times, and one sequence in the V3-V5 dataset that occurred
-`r names(v35.table[length(v35.table)])` times. Contrary to previous reports
+duplicate sequences with the same error. We identified all of the mock community
+sequences that had a 1-nt difference to the true sequence. For these three
+regions, between `r min(percent_one_off)` and `r max(percent_one_off)` of the
+sequences with 1-nt errors were only observed once. We found that the frequency
+of the most abundant 1-nt error paralleled the number of sequences.
+Surprisingly, the same 1-nt error appeared
+`r as.numeric(names(non_random$V1V6[length(non_random$V1V6)]))` times in the
+V1-V6 mock data and another 1-nt error appeared
+`r as.numeric(names(non_random$V1V9[length(non_random$V1V9)]))` times in the
+V1-V9 mock data. Contrary to previous reports
 {Carneiro, 2012 #3458;Koren, 2012 #3459}, these results indicate that
-reproducible errors occur with the PacBio sequencing platform and that they can be
-quite frequent.
+reproducible errors occur with the PacBio sequencing platform and that they can
+be quite abundant. Through the use of the pre-clustering step described above
+these 1-nt errors would be ameliorated; however, this result indicates that
+caution should be used when attempting to use fine-scale OTU definitions.
 
@@
 full-length sequences; however, the error rate associated with those reads is
-considerable (i.e. `r round(100*final.error["v19","precluster.error"],2)`%)
+considerable (i.e. `r round(100*pooled_error["v19","precluster.error"],2)`%)
 and requires a level of sequencing coverage that is not commonly observed in a
@@
 oligos <- read.table(file="pacbio.oligos", stringsAsFactors=F)[1:6,]
-rownames(oligos) <- pretty.region[oligos$V4]
+rownames(oligos) <- pretty_region[oligos$V4]
 oligos <- oligos[,c("V2", "V3")]
@@
 ```{r Table_2, warning=TRUE, results='asis'}
-rownames(reasons.lost) <- pretty.region[rownames(reasons.lost)]
+rownames(reasons.lost) <- pretty_region[rownames(reasons.lost)]
 reasons.lost <- cbind(orig.nseqs, 100*reasons.lost, basic.nseqs)
@@
 lapply(regions, plotLines)
-legend(x=18, y=0.05, legend=pretty.region[regions], lty=1, lwd=2, col=clrs[regions])
+legend(x=18, y=0.05, legend=pretty_region[regions], lty=1, lwd=2, col=clrs[regions])
 mtext(side=2, line=0.5, at=0.05, text="C", cex=2, font=2, las=2)
@@
 
-legend(x=10, y=0.50, legend=pretty.region[regions], lty=1, lwd=2, col=clrs[regions], cex=1.0)
+legend(x=10, y=0.50, legend=pretty_region[regions], lty=1, lwd=2, col=clrs[regions], cex=1.0)
 ```

### 2016-01-28 Schloss_PacBio16S_PeerJ_2016.Rmd
Edit Conclusion section

@@
 ```{r}
-
 getRegionNonRandom <- function(region){
@@
 percent_one_off <- 100 * sapply(non_random, function(x){x[1]/sum(x)})
-
 ```
@@
 the MiSeq platform is able to generate a large number of high-quality 250-nt
-sequence fragments (error rates below 0.02%) {Kozich, 2013 #2719}. The promise of the PacBio
-sequencing platform was the generation of high-quality near full-length sequence
-fragments. As we have shown in this study, it is possible to generate near
-full-length sequences; however, the error rate associated with those reads is
-considerable (i.e. `r round(100*pooled_error["v19","precluster.error"],2)`%)
-and requires a level of sequencing coverage that is not commonly observed in a
-typical sequencing run. This results in the generation of a small number of low
-quality full-length sequences. When we considered the shorter V4 region, which is
-similar in length to what is sequenced by the MiSeq platform, the error rates
-we observed with the PacBio platform were nearly 5-fold higher than what has
-previously been reported. It appears that the promise offered by the PacBio
-platform has not been realized.
+sequence fragments (error rates below 0.02%) {Kozich, 2013 #2719}. The promise
+of the PacBio sequencing platform was the generation of high-quality near
+full-length sequence fragments. As we have shown in this study, it is possible
+to generate near full-length sequences with error rates that are slightly
+higher, but comparable to the other platforms (i.e.
+`r round(100*pooled_error["V1V9","precluster"],2)`%).
+With the exception of the V4 region, the error rates were less than
+`r 100*max(pooled_error[!(rownames(pooled_error) %in% "V4"),"precluster"])`%.
+When we considered the shorter V4 region, which is similar in length to what is
+sequenced by the MiSeq platform, the error rates we observed with the PacBio
+platform were nearly 8-fold higher than what has previously been reported. It
+was unclear why these shorter reads had such a high error rate relative to the
+other regions. At this point, the primary limitation of generating full-length
+sequences on the PacBio platform is the cost of generating the data and
+accessibility to the sequencers.
 
 The widespread adoption of the 454 and MiSeq platforms and decrease in the use
-of Sanger
-sequencing for the 16S rRNA gene has resulted in a decrease in the
+of Sanger sequencing for the 16S rRNA gene has resulted in a decrease in the
 generation of the full-length reference sequences that are needed for performing
@@
 *in situ* hybridization (FISH) probes. It remains to be determined whether the
-elevated error rates we observed for full-length sequences are prohibitive for
+error rates we observed for full-length sequences are prohibitive for
 these applications. We can estimate the distribution of errors assuming that the
@@
 prior to pre-clustering the sequences, which was
-`r round(100*final.error["v19","unique.error"],2)`% (Figure 5). Under
+`r round(100*pooled_error["V1V9","unique"],2)`% (Figure 5). Under
 these conditions one would only expect
-`r round(100*dbinom(0, 1500, final.error["v19","unique.error"]), 2)`% of the
-sequences to have no errors. In fact, 95% of the reads would have at least
-`r qbinom(0.05, 1500, final.error["v19","unique.error"])` errors and 50% of the
-reads would have at least
-`r qbinom(0.50, 1500, final.error["v19","unique.error"])` errors. If the error
-rate could be dropped to 0.25%, then 95% of
-the reads would have at least
-`r qbinom(0.05, 1500, 0.0025)` error and 50% of the
-reads would have at least
-`r qbinom(0.50, 1500, 0.0025)` errors. If it were possible to replicate the low
-error rates we have previously observed using the 454 and Illumina MiSeq
-platforms, which was 0.02%, then we would expect
-`r round(100*dbinom(0, 1500, 0.0002), 1)`% of the sequences to have no errors.
-In fact, 95% of the reads would have `r qbinom(0.95, 1500, 0.0002)` or fewer
-errors. Although full-length sequence data is highly desired, at this point, it
-does not appear that the PacBio platform can provide the data of sufficient
-quality to fill the niche of generating reference sequences.
+`r round(100*dbinom(0, 1500, pooled_error["V1V9","unique"]), 2)`% of the
+sequences to have no errors and 50% of the sequences would have at least
+`r qbinom(0.50, 1500, pooled_error["V1V9","unique"])` errors.
+After applying the pre-clustering denoising step, the erro rate drops by `r round(pooled_error["V1V9","unique"]/pooled_error["V1V9","precluster"])`-fold to
+`r 100*pooled_error["V1V9","precluster"]`%. With this error rate, we would
+expect `r round(100*dbinom(0, 1500, pooled_error["V1V9","precluster"]), 2)`%
+of the sequences to have no sequencing errors. Of course, the tradeoff is the
+loss of resolution among closely related sequences.
 
@@
 using the mouse libraries, the ability to classify each of the regions differed
-by less than 5%
-when classifying against the SILVA and greengenes databases. The effect of the
-database that was used was also interesting. The RDP database outperformed the
-other databases for the mouse samples and the SILVA database outperformed the
-others for the human and soil samples. The three databases were equally
-effective for classifying the mock community. Finally, since only the
+by less than 5% when classifying against the SILVA and greengenes databases. The
+effect of the database that was used was also interesting. The RDP database
+outperformed the other databases for the mouse samples and the SILVA database
+outperformed the others for the human and soil samples. The three databases were
+equally effective for classifying the mock community. Finally, since only the
 greengenes database provided species-level information for its reference
@@
 mock communities as part of their experimental design so that they can quantify
-their error rates. The ability to generate large numbers of near full-length 16S
-rRNA gene sequences is an exciting advance. At this point, the excitement must
-be tempered by the appreciation that the error rates limit the application of
-the approach.
+their error rates. The ability to generate near full-length 16S rRNA gene
+sequences is an exciting advance that will hopefully expand our ability to
+improve the characterization of microbial communities.
 
@@
 reference provided species-level information.
-
-**Figure 5. The percentage of V1-V9 sequences that were predicted to have
-between 0 and 20 errors as a function of the error rate of the sequences.** The
-highest error rate, `r round(100*final.error["v19","unique.error"], 2)`%,
-corresponds to what was observed before the pre-clustering step.
-The smallest error rate (0.02%) corresponds to our previous observations using
-the 454 and MiSeq sequencing platforms. The predicted number of errors was
-assumed to follow a binomial distribution.
 
@@
 ```
-
-### Figure 5
-
-```{r Figure_5, fig.width=7, fig.height=4}
-x <- 0:20
-e1 <- final.error["v19","unique.error"]
-e2 <- 0.0025
-e3 <- 0.0010
-e4 <- 0.0002
-
-par(mar=c(5,6,0.5, 0.5))
-plot(x, dbinom(x, 1458, e1), type="l", ylim=c(0,0.35), xlab="Number of errors",
-		 ylab="Full-length 16S rRNA\ngene sequences (%)", yaxt="n", lwd=3)
-points(x, dbinom(x, 1458, e2), type="l", col="red", lwd=3)
-points(x, dbinom(x, 1458, e3), type="l", col="blue", lwd=3)
-points(x, dbinom(x, 1458, e4), type="l", col="darkgreen", lwd=3)
-axis(2, at=seq(0,0.35,0.05), label=seq(0,35,5), las=1)
-legend(x=12, y=0.30, legend=paste0(format(round(100*c(e1, e2, e3, e4), 2), nsmall=2), "%"),
-			 lwd=3, col=c("black", "red", "blue", "darkgreen"))
-```
-
-```{r, versions}
-sessionInfo()
-```

### 2016-01-28 Schloss_PacBio16S_PeerJ_2016.Rmd
Trying to get write.paper to run

@@
 10-fold coverage to those with less coverage, we reduced the error rate by
-`r paste(format(range(coverage_reduction), digits=3, nsmall=1), collapse=" to ")`%
+`r paste(format(range(base_error_reduction), digits=3, nsmall=1), collapse=" to ")`%
 (Figure 2). Third, based on the observed correlation between the
@@
 
-## References
-
 ## Tables
@@
 par(mar=c(6, 5, 0.5, 0.5))
-plot(1, xlim=c(0.5, 4.5), ylim=c(0,75), type="n", yaxt="n", xaxt="n", xlab="",
-		 ylab="")
+plot(1, xlim=c(0.5, 4.5), ylim=c(0,75), type="n", yaxt="n", xaxt="n", xlab="", ylab="")
 polygon(x=c(0.75, 1.25, 1.25, 0.75), y=c(m.quant["25%"], m.quant["25%"],
@@
 					a.quant["75%"]), y1=c(m.quant["97.5%"], s.quant["97.5%"],
-																					i.quant["97.5%"], a.quant["97.5%"]),
-			 angle=90, length=0.05)
+					i.quant["97.5%"], a.quant["97.5%"]), angle=90, length=0.05)
 
@@
 					a.quant["25%"]), y1=c(m.quant["2.5%"], s.quant["2.5%"],
-																					i.quant["2.5%"], a.quant["2.5%"]),
-			 angle=90, length=0.05)
-
-axis(1, at=c(1,2,3,4), label=c("Matches", "Substitutions", "Insertions",
-															 "Ambiguous"), las=2, cex.axis=1)
+					i.quant["2.5%"], a.quant["2.5%"]), angle=90, length=0.05)
+
+axis(1, at=c(1,2,3,4), label=c("Matches", "Substitutions", "Insertions", "Ambiguous"), las=2, cex.axis=1)
 axis(2, las=2, at=c(0,25,50,75), cex.axis=1.5)
@@
 	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
-																		 ".composite", sep=""),
-													header=T,row.names=1)
+								 ".composite", sep=""), header=T,row.names=1)
 
@@
 	error.by.depth <- aggregate(composite$error, by=list(composite$freq),
-															function(x){c(mean=mean(x),
-																						quantile(x, prob=c(0.025, 0.975)),
-																						n=length(x))})
+						function(x){c(mean=mean(x), quantile(x, prob=c(0.025, 0.975)), n=length(x))})
 
@@
 	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
-																		 ".composite", sep=""), header=T,
-													row.names=1)
+								 ".composite", sep=""), header=T, row.names=1)
 	composite <- composite[composite$reason == "g",]
@@
 
-	remaining.error <- unlist(lapply(qscores,
-														function(x){mean(composite[composite$minQ >= x,"error"])}))
+	remaining.error <- unlist(lapply(qscores, function(x){mean(composite[composite$minQ >= x,"error"])}))
 	names(remaining.error) <- qscores
@@
 ```
+
+## References

### 2016-01-28 Schloss_PacBio16S_PeerJ_2016.Rmd
Trying to get write.paper to work

@@
 the analysis. For instance, among the V1-V9 data, requiring sequences to have a
-predicted error rate less than 0.01% resulted in a `r format(base_error_reduction["pred_error_reduction","V1V9"], digits=3, nsmall=1)`% and resulted in the removal of `r format(base_nseqs_reduction["pred_error_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences.
+predicted error rate less than 0.01% resulted in a `r format(base_error_reduction["pred_error_reduction","V1V9"], digits=3, nsmall=1)`% and resulted in the removal of `r format(base_nseqs_reduction["pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences.
 Adding the mismatch or coverage filter
@@
 
-oligos <- read.table(file="pacbio.oligos", stringsAsFactors=F)[1:6,]
+oligos <- read.table(file="data/references/pacbio.oligos", stringsAsFactors=F)[1:6,]
 rownames(oligos) <- pretty_region[oligos$V4]

### 2016-01-29 Schloss_PacBio16S_PeerJ_2016.Rmd
Extract the analysis of the error profile

@@
 
-mock_error <- read.table(file="data/process/mock.error.report")
-mock_error_nochim <- mock_error[mock_error$numparents==1,]
-mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.10,]
+library(json)
 ```
@@
 ```{r error_profile}
-mutations <- c("AA", "AT", "AG", "AC", "TA", "TT", "TG", "TC", "GA", "GT", "GG", "GC", "CA", "CT", "CG", "CC", "Ai", "Ti", "Gi", "Ci", "dA", "dT", "dG", "dC")
-
-mutation_table <- mock_error_good[, which(colnames(mock_error_good)%in%mutations)]
-mutation_count <- apply(mutation_table, 2, sum)
-
-mutation_total <- sum(mutation_count)
-matches <- sum(mutation_count[c("AA", "TT", "GG", "CC")])
-mis_matches <- mutation_total - matches
-init_error_rate <- 100 * mis_matches / mutation_total		#0.6514488
-
-#substitutions
-substitutions <- mutations[grep("[ATGC].", mutations)]
-substitutions <- substitutions[grep(".[ATGC]", substitutions)]
-substitutions <- substitutions[! substitutions %in% c("AA", "TT", "GG", "CC")]
-substitution_counts <- mutation_count[substitutions]
-substitution_total <- sum(substitution_counts)
-substitution_bias <- substitution_counts / substitution_total
-
-reference <- gsub(".(.)", "\\1", names(substitution_bias))
-# aggregate(substitution_bias, by=list(reference), mean)
-#
-# Group.1          x
-# 1       A 0.10192188
-# 2       C 0.05944286
-# 3       G 0.07434043
-# 4       T 0.09762816
-
-substitution_rate <- 100 * substitution_total / mis_matches
-
-# [1] 50.90346
-
-#deletions
-deletion_counts <- mutation_count[grep("d.", mutations)]
-deletion_total <- sum(deletion_counts)
-deletion_bias <- 100 * deletion_counts / deletion_total
-
-#dA        dT        dG        dC
-#24.25770 17.99132 39.44415 18.30683
-
-deletion_rate <- 100 * deletion_total / mis_matches
-
-#[1] 17.93361
-
-#insertions
-insertion_counts <- mutation_count[grep(".i", mutations)]
-insertion_total <- sum(insertion_counts)
-insertion_bias <- 100 * insertion_counts / insertion_total
-
-#Ai        Ti        Gi        Ci
-#23.13509 19.63951 30.05906 27.16633
-
-insertion_rate <- 100 * insertion_total / mis_matches
-
-#[1] 31.16293
-
-# Need to connect error types with the quality scores
-error_quality <- read.table(file="data/process/mock.quality.report", header=T, row.names=1)
-
-#get quantiles
-matches <- rep(as.numeric(rownames(error_quality)), error_quality$matches)
-m.quant <- quantile(matches, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
-m.94 <- 100 * sum(matches >= 94)/length(matches)
-# [1] 80.52649
-
-subs <- rep(as.numeric(rownames(error_quality)), error_quality$substitutions)
-s.quant <- quantile(subs, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
-s.94 <- 100 * sum(subs >= 94)/length(subs)
-# [1] 79.96451
-
-ins <- rep(as.numeric(rownames(error_quality)), error_quality$insertions)
-i.quant <- quantile(ins, prob=c(0.025, 0.25, 0.5, 0.75, 0.975))
-i.94 <- 100 * sum(ins >= 94)/length(ins)
-# [1] 80.35283
+error_profile <- fromJSON(scan("data/process/error_profile.json", quiet=T, what="character"))
 
@@
 basic_error_rate <- 100 * sum(mock_error_basic$mismatches) / sum(mock_error_basic$total)		#[1] 0.6448736
+
 ```
@@
 rate associated with sequencing the 16S rRNA gene. We observed an average
-sequencing error rate of `r format(100*init_error_rate, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
-accounted for `r format(insertion_rate, digits=3, nsmall=1)`,
-`r format(deletion_rate, digits=3, nsmall=1)`, and `r format(substitution_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
+sequencing error rate of `r format(error_profile$overall_error, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
+accounted for `r format(error_profile$ins_rate, digits=3, nsmall=1)`,
+`r format(error_profile$del_rate, digits=3, nsmall=1)`, and `r format(error_profile$subs_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
 likely and all four bases
 were equally likely to cause insertion errors. Interestingly, guanines
-(`r format(deletion_bias["dG"], digits=3, nsmall=1)`%) and adenines
-(`r format(deletion_bias["dA"], digits=3, nsmall=1)`%) were more likely to be deleted
-than cytosines (`r format(deletion_bias["dC"], digits=3, nsmall=1)`%) or thymidines
-(`r format(deletion_bias["dT"], digits=2, nsmall=1)`%). The PacBio quality values varied between
-`r paste(range(c(matches, subs, ins)), collapse=' and ')`.
-Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(m.94, digits=3, nsmall=1)`%), substitutions (`r format(s.94, digits=3, nsmall=1)`%), and insertions (`r format(i.94, digits=3, nsmall=1)`%).
-Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(cor(mock_error_good$error, mock_error_good$pred_error), digits=2, nsmall=2)`; Figure 1A).
+(`r format(error_profile$del_bias$dG, digits=3, nsmall=1)`%) and adenines
+(`r format(error_profile$del_bias$dA, digits=3, nsmall=1)`%) were more likely to be deleted
+than cytosines (`r error_profile$del_bias$dC, digits=3, nsmall=1)`%) or thymidines
+(`r error_profile$del_bias$dT, digits=2, nsmall=1)`%). The PacBio quality values varied between
+`r paste(error_profile$qual_range, collapse=' and ')`.
+Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(error_profile$percent_over_93$m, digits=3, nsmall=1)`%), substitutions (`r format(error_profile$percent_over_93$s, digits=3, nsmall=1)`%), and insertions (`r format(error_profile$percent_over_93$i, digits=3, nsmall=1)`%).
+Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(error_profile$qual_error_corr, digits=2, nsmall=2)`; Figure 1A).
 
@@
 	coverage_error_agg <- aggregate(100 * coverage_region$error, by=list(coverage_region$coverage), mean)
-	coverage_error <- coverage_error_agg[coverage_error_agg$Group.1 <= 40,"x"]
-	names(coverage_error) <- coverage_error_agg[coverage_error_agg$Group.1 <= 40,"Group.1"]
+
+	coverage_error <- rep(NA, max(mock_error_basic$coverage))
+	coverage_error[coverage_error_agg$Group.1] <- coverage_error_agg[,"x"]
 	coverage_error
@@
 
-	deep <- region_data$coverage >= 10
+	deep <- region_data$coverage >= cutoff
 	coverage_data <- region_data[deep,]

### 2016-01-29 code/get_error_profile.R

@@ -1,0 +1,96 @@
+library(rjson)
+
+mock_error <- read.table(file="data/process/mock.error.report")
+mock_error_nochim <- mock_error[mock_error$numparents==1,]
+mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.10,]
+
+mutations <- c("AA", "AT", "AG", "AC", "TA", "TT", "TG", "TC", "GA", "GT", "GG", "GC", "CA", "CT", "CG", "CC", "Ai", "Ti", "Gi", "Ci", "dA", "dT", "dG", "dC")
+
+mutation_table <- mock_error_good[, which(colnames(mock_error_good)%in%mutations)]
+mutation_count <- apply(mutation_table, 2, sum)
+
+mutation_total <- sum(mutation_count)
+matches <- sum(mutation_count[c("AA", "TT", "GG", "CC")])
+mis_matches <- mutation_total - matches
+init_error_rate <- 100 * mis_matches / mutation_total		#0.6514488
+
+
+#substitutions
+substitutions <- mutations[grep("[ATGC].", mutations)]
+substitutions <- substitutions[grep(".[ATGC]", substitutions)]
+substitutions <- substitutions[! substitutions %in% c("AA", "TT", "GG", "CC")]
+substitution_counts <- mutation_count[substitutions]
+substitution_total <- sum(substitution_counts)
+substitution_bias <- substitution_counts / substitution_total
+
+reference <- gsub(".(.)", "\\1", names(substitution_bias))
+# aggregate(substitution_bias, by=list(reference), mean)
+#
+# Group.1          x
+# 1       A 0.10192188
+# 2       C 0.05944286
+# 3       G 0.07434043
+# 4       T 0.09762816
+
+substitution_rate <- 100 * substitution_total / mis_matches
+
+# [1] 50.90346
+
+
+#deletions
+deletion_counts <- mutation_count[grep("d.", mutations)]
+deletion_total <- sum(deletion_counts)
+deletion_bias <- 100 * deletion_counts / deletion_total
+
+#dA        dT        dG        dC
+#24.25770 17.99132 39.44415 18.30683
+
+deletion_rate <- 100 * deletion_total / mis_matches
+
+#[1] 17.93361
+
+#insertions
+insertion_counts <- mutation_count[grep(".i", mutations)]
+insertion_total <- sum(insertion_counts)
[…]

### 2016-01-29 Schloss_PacBio16S_PeerJ_2016.Rmd
Extract the analysis of the error rates by each method

@@
 
-## Basic error correction analysis
-get_raw_error <- function(region){
-	region_data <- mock_error_good[mock_error_good$region == region,]
-	return(100*sum(region_data$mismatches)/sum(region_data$total))
-}
-
-raw_rates <- sapply(regions, get_raw_error)
-
-get_basic_kept <- function(region){
-	region_data <- mock_error_good[mock_error_good$region == region,]
-	100 * sum(region_data$good_coords & region_data$good_homop)/nrow(region_data)
-}
-
-basic_kept <- sapply(regions, get_basic_kept)
-
-get_basic_error <- function(region){
-	region_data <- mock_error_good[mock_error_good$region == region & mock_error_good$good_coords & mock_error_good$good_homop,]
-	return(100*sum(region_data$mismatches)/sum(region_data$total))
-}
-
-basic_rates <- sapply(regions, get_basic_error)
-
-mock_error_basic <- mock_error_good[mock_error_good$good_coords & mock_error_good$good_homop,]
-basic_n <- aggregate(mock_error_basic$error, by=list(mock_error_basic$region), length)$x
-names(basic_n) <- names(basic_rates)
-
-basic_error_rate <- 100 * sum(mock_error_basic$mismatches) / sum(mock_error_basic$total)		#[1] 0.6448736
-
 ```
@@
 
-```{r barcode_primer_analysis}
-mock_error_basic$pbdiffs <- mock_error_basic$fbdiffs + mock_error_basic$rbdiffs + mock_error_basic$fpdiffs + mock_error_basic$rpdiffs
+```{r random_stuff }
 
@@
 
-# want to get the error rate and fraction of sequences for each region with a
-# threshold of mismatches or fewer
-
-pb_by_region_mm <- aggregate(mock_error_basic$error, by=list(mock_error_basic$pbdiffs, mock_error_basic$region), function(x){c(mean=mean(100*x), nseqs=length(x))})
-
-aggregate_pb_data <- function(region){
-	pb_region <- pb_by_region_mm[pb_by_region_mm$Group.2 == region,]
-
-	product <- pb_region$x[,"mean"] * pb_region$x[,"nseqs"]
-	pb_region_nseqs <- cumsum(pb_region$x[,"nseqs"])
-	pb_region_error <- cumsum(product) / pb_region_nseqs
-	pb_region_frac <- pb_region_nseqs / sum(pb_region$x[,"nseqs"])
-	c(error=pb_region_error, fraction=pb_region_frac)
-}
-
-pb_data <- sapply(regions, aggregate_pb_data)
-
-pb_error <- pb_data[grep("error", rownames(pb_data)),]
-rownames(pb_error) <- 0:6
-
-pb_fraction <- pb_data[grep("fraction", rownames(pb_data)),]
-rownames(pb_fraction) <- 0:6
-
-pb_error_reduction_0 <- 100 * (1 - pb_error["0",]/basic_rates)
-pb_error_reduction_1 <- 100 * (1 - pb_error["1",]/basic_rates)
-
-pb_nseqs_reduction_0 <- 100 * (1 - pb_fraction["0",])
-pb_nseqs_reduction_1 <- 100 * (1 - pb_fraction["1",])
-```
-
-```{r coverage_analysis}
 #This is for the Figure 2C plot...
@@
 
-# need to get the error rate for coverage values between 3 and 40 for each region
-
-error_by_coverage <- function(region){
-	coverage_region <- mock_error_basic[mock_error_basic$region == region,]
-	coverage_error_agg <- aggregate(100 * coverage_region$error, by=list(coverage_region$coverage), mean)
-
-	coverage_error <- rep(NA, max(mock_error_basic$coverage))
-	coverage_error[coverage_error_agg$Group.1] <- coverage_error_agg[,"x"]
-	coverage_error
-}
-
-coverage_error <- sapply(regions, error_by_coverage)
-#seems to plateau right around 10
-
-error_at_below_coverage_cutoff <- function(region, cutoff=10){
-	region_data <- mock_error_basic[mock_error_basic$region == region,]
-
-	deep <- region_data$coverage >= cutoff
-	coverage_data <- region_data[deep,]
-
-	error <- 100 * sum(coverage_data$mismatches)/sum(coverage_data$total)
-	frac_kept <- nrow(coverage_data) / nrow(region_data)
-	c(error=error, fraction=frac_kept)
-}
-
-coverage_data <- sapply(regions, error_at_below_coverage_cutoff)
-
-coverage_error <- coverage_data["error",]
-coverage_fraction <- coverage_data["fraction",]
-
-coverage_error_reduction <- 100 * (1 - coverage_error/basic_rates)
-coverage_nseqs_reduction <- 100 * (1 - coverage_fraction)
-```
-
-```{r predicted_error_analysis }
 #This is for the Figure 2D plot...
@@
 
-error_at_above_pred_error_cutoff <- function(region, threshold=0.999){
-	region_data <- mock_error_basic[mock_error_basic$region == region,]
-
-	above_threshold <- region_data[region_data$pred_error >= threshold,]
-
-	c(error=100*sum(above_threshold$mismatches)/sum(above_threshold$total), nseqs=nrow(above_threshold)/nrow(region_data))
-}
-
-pred_error_data <- sapply(regions, error_at_above_pred_error_cutoff, threshold=0.999)
-
-pred_error_reduction <- 100 * (1 - pred_error_data["error",]/basic_rates)
-pred_nseqs_reduction <- 100 * (1 - pred_error_data["nseqs",])
-```
-
-```{r oligos_coverage_analysis}
-get_oligos_coverage <- function(region){
-	region_data <- mock_error_basic[mock_error_basic$region == region,]
-
-	good0 <- region_data$pbdiffs <= 0 & region_data$coverage >= 10
-	region_data_good0 <- region_data[good0,]
-
-	good1 <- region_data$pbdiffs <= 1 & region_data$coverage >= 10
-	region_data_good1 <- region_data[good1,]
-
-	c(	100*mean(region_data_good0[,"error"]), nrow(region_data_good0)/nrow(region_data),
-		100*mean(region_data_good1[,"error"]), nrow(region_data_good1)/nrow(region_data))
-}
-
-oligos_coverage_error <- matrix(unlist(lapply(regions, get_oligos_coverage)), ncol=4, byrow=T)
-rownames(oligos_coverage_error) <- regions
-colnames(oligos_coverage_error) <- c("0.10.error", "0.10.frac", "1.10.error", "1.10.frac")
-
-#     0.10.error 0.10.frac 1.10.error 1.10.frac
-#V1V3  0.4515344 0.5898447  0.4632910 0.7968639
-#V1V5  0.4187311 0.5557715  0.4270938 0.7384810
-#V1V6  0.4399351 0.5410259  0.4460726 0.7167325
-#V1V9  0.4317629 0.4444776  0.4420337 0.5752849
-#V3V5  0.6421004 0.6854427  0.6555240 0.8102832
-#V4    0.6825316 0.5512332  0.7023460 0.8245533
-
-oligos_coverage_error_reduction <- 100 * (1 - oligos_coverage_error[,"1.10.error"]/basic_rates)
-oligos_coverage_nseqs_reduction <- 100 * (1 - oligos_coverage_error[,"1.10.frac"])
-```
-
-```{r oligosQScoreAnalysis}
-get_oligos_pred_error <- function(region){
-	region_data <- mock_error_basic[mock_error_basic$region == region,]
-
-	good0 <- region_data$pbdiffs <= 0 & region_data$pred_error >= 0.999
-	region_data_good0 <- region_data[good0,]
-
-	good1 <- region_data$pbdiffs <= 1 & region_data$pred_error >= 0.999
-	region_data_good1 <- region_data[good1,]
-
-	c(	100 * mean(region_data_good0[,"error"]), nrow(region_data_good0)/nrow(region_data),
-		100 * mean(region_data_good0[,"error"]), nrow(region_data_good0)/nrow(region_data) )
-}
-
-oligos_pred_error <- matrix(unlist(lapply(regions, get_oligos_pred_error)), ncol=4, byrow=T)
-rownames(oligos_pred_error) <- regions
-colnames(oligos_pred_error) <- c("0.999.error", "0.999.frac", "1.999.error", "1.999.frac")
-
-#     0.999.error 0.999.frac 1.999.error 1.999.frac
-#V1V3   0.2806049  0.4749822   0.2806049  0.4749822
-#V1V5   0.2177965  0.4266279   0.2177965  0.4266279
-#V1V6   0.2168503  0.4296550   0.2168503  0.4296550
-#V1V9   0.2177073  0.3438720   0.2177073  0.3438720
-#V3V5   0.2229757  0.5131411   0.2229757  0.5131411
-#V4     0.3485654  0.3961328   0.3485654  0.3961328
-
-oligos_pred_error_reduction <- 100 * (1 - oligos_pred_error[,"1.999.error"]/basic_rates)
-oligos_pred_nseqs_reduction <- 100 * (1 - oligos_pred_error[,"1.999.frac"])
-```
-
-```{r coverage_pred_error_analysis}
-get_coverage_pred <- function(region){
-	region_data <- mock_error_basic[mock_error_basic$region == region,]
-
-	good <- region_data$coverage >= 10 & region_data$pred_error >= 0.999
-	region_data_good <- region_data[good,]
-
-	c(100 * mean(region_data_good[,"error"]), nrow(region_data_good)/nrow(region_data) )
-}
-
-coverage_pred_error <- matrix(unlist(lapply(regions, get_coverage_pred)), ncol=2, byrow=T)
-rownames(coverage_pred_error) <- regions
-colnames(coverage_pred_error) <- c("10.999.error", "10.999.frac")
-
-#    10.999.error 10.999.frac
-#V1V3   0.2843963    65.86436
-#V1V5   0.2188792    56.55108
-#V1V6   0.2178326    57.33756
-#V1V9   0.2191284    42.66675
-#V3V5   0.2236951    60.74254
-#V4     0.3527900    61.79213
-
-coverage_pred_error_reduction <- 100 * (1 - coverage_pred_error[,"10.999.error"]/basic_rates)
-coverage_pred_nseqs_reduction <- 100 * (1 - coverage_pred_error[,"10.999.frac"])
-```
-
-```{r all_filters_analysis}
-get_all_filters <- function(region){
-	region_data <- mock_error_basic[mock_error_basic$region == region,]
-
-	good0 <- region_data$pbdiffs == 0 & region_data$pred_error >= 0.999 & region_data$coverage >= 10
-	region_data_good0 <- region_data[good0,]
-
-	good1 <- region_data$pbdiffs <= 1 & region_data$pred_error >= 0.999 & region_data$coverage >= 10
-	region_data_good1 <- region_data[good1,]
-
-	c(	100 * mean(region_data_good0[,"error"]), nrow(region_data_good0)/nrow(region_data),
-		100 * mean(region_data_good1[,"error"]), nrow(region_data_good1)/nrow(region_data) )
-}
-
-all_filters_error <- matrix(unlist(lapply(regions, get_all_filters)), ncol=4, byrow=T)
-rownames(all_filters_error) <- regions
-colnames(all_filters_error) <- c("0.10.999.error", "0.10.999.frac", "1.10.999.error", "1.10.999.frac")
-
-#    0.10.60.error 0.10.60.frac 1.10.60.error 1.10.60.frac
-#v4    0.001515500    0.5562125   0.001993325    0.7499762
-#v13   0.009299146    0.4337597   0.009827404    0.6155779
-#v35   0.008613200    0.4862353   0.008979207    0.6253128
-#v15   0.003078476    0.3488609   0.003905562    0.5268616
-#v16   0.007738039    0.2954687   0.008814923    0.4295469
-#v19   0.004797594    0.2496316   0.005266467    0.3631005
-
-#     0.10.999.error 0.10.999.frac 1.10.999.error 1.10.999.frac
-#V1V3      0.2808218     0.4696170      0.2840387     0.6183602
-#V1V5      0.2184805     0.4192252      0.2182989     0.5374066
-#V1V6      0.2170989     0.4182253      0.2174090     0.5421862
-#V1V9      0.2176989     0.3213802      0.2191334     0.4016057
-#V3V5      0.2241187     0.5038275      0.2240779     0.5841414
-#V4        0.3492870     0.3932393      0.3526755     0.5665274
-
-all_filters_error_reduction <- 100 * (1 - all_filters_error[,"1.10.999.error"]/basic_rates)
-all_filters_nseqs_reduction <- 100 * (1 - all_filters_error[,"1.10.999.frac"])
-```
-
-```{r summary}
-base_error_reduction <- rbind(	pb_error_reduction_0,
-								pb_error_reduction_1,
-								pred_error_reduction,
-								coverage_error_reduction,
-								coverage_pred_error_reduction,
-								oligos_coverage_error_reduction,
-								oligos_pred_error_reduction,
-								all_filters_error_reduction)
-
-base_nseqs_reduction <- rbind(	pb_nseqs_reduction_0,
-								pb_nseqs_reduction_1,
-								pred_nseqs_reduction,
-								coverage_nseqs_reduction,
-								coverage_pred_nseqs_reduction,
-								oligos_coverage_nseqs_reduction,
-								oligos_pred_nseqs_reduction,
-								all_filters_nseqs_reduction)
 ```

### 2016-01-29 code/get_error_rate_table.R

@@ -1,0 +1,314 @@
+regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
+
+mock_error <- read.table(file="data/process/mock.error.report")
+mock_error_nochim <- mock_error[mock_error$numparents==1,]
+mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.10,]
+
+
+
+
+## Basic error correction analysis
+get_raw_error <- function(region){
+	region_data <- mock_error_good[mock_error_good$region == region,]
+	return(100*sum(region_data$mismatches)/sum(region_data$total))
+}
+
+raw_rates <- sapply(regions, get_raw_error)
+
+get_basic_kept <- function(region){
+	region_data <- mock_error_good[mock_error_good$region == region,]
+	100 * sum(region_data$good_coords & region_data$good_homop)/nrow(region_data)
+}
+
+percent_basic_kept <- sapply(regions, get_basic_kept)
+
+get_basic_error <- function(region){
+	region_data <- mock_error_good[mock_error_good$region == region & mock_error_good$good_coords & mock_error_good$good_homop,]
+	return(100*sum(region_data$mismatches)/sum(region_data$total))
+}
+
+basic_rates <- sapply(regions, get_basic_error)
+
+mock_error_basic <- mock_error_good[mock_error_good$good_coords & mock_error_good$good_homop,]
+
+
+#number of sequences per region that survive the basic step
+basic_n <- aggregate(mock_error_basic$error, by=list(mock_error_basic$region), length)$x
+names(basic_n) <- names(basic_rates)
+
+
+
+
+#primer and barcode analysis
+mock_error_basic$pbdiffs <- mock_error_basic$fbdiffs + mock_error_basic$rbdiffs + mock_error_basic$fpdiffs + mock_error_basic$rpdiffs
+
+
+
+# want to get the error rate and fraction of sequences for each region with a
+# threshold of mismatches or fewer
+
+pb_by_region_mm <- aggregate(mock_error_basic$error, by=list(mock_error_basic$pbdiffs, mock_error_basic$region), function(x){c(mean=mean(100*x), nseqs=length(x))})
+
+aggregate_pb_data <- function(region){
+	pb_region <- pb_by_region_mm[pb_by_region_mm$Group.2 == region,]
+
[…]

### 2016-01-29 code/consolidate_taxonomy.R
Change output file name

@@ -26,4 +26,4 @@ tax_level_counts <- t(sapply(1:nrow(combos), function(x){count_tax_levels(combos
 genus_species <- tax_level_counts[,"6"] + tax_level_counts[,"7"]
 composite <- cbind(combos, tax_level_counts, genus_species)
 
-write.table(file="data/process/taxonomy.depth.analysis", composite, quote=F, row.names=F)
+write.table(file="data/process/taxonomy_depth_analysis.tsv", composite, quote=F, row.names=F, sep="\t")

### 2016-01-29 code/get_error_rate_table.R
Add sequence counts

@@ -10,31 +10,24 @@ mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.10,]
 ## Basic error correction analysis
 get_raw_error <- function(region){
 	region_data <- mock_error_good[mock_error_good$region == region,]
-	return(100*sum(region_data$mismatches)/sum(region_data$total))
+	return(c(100*sum(region_data$mismatches)/sum(region_data$total), nrow(region_data)))
 }
-
 raw_rates <- sapply(regions, get_raw_error)
+rownames(raw_rates) <- c("error", "nseqs")
 
-get_basic_kept <- function(region){
-	region_data <- mock_error_good[mock_error_good$region == region,]
-	100 * sum(region_data$good_coords & region_data$good_homop)/nrow(region_data)
-}
-
-percent_basic_kept <- sapply(regions, get_basic_kept)
 
 get_basic_error <- function(region){
 	region_data <- mock_error_good[mock_error_good$region == region & mock_error_good$good_coords & mock_error_good$good_homop,]
-	return(100*sum(region_data$mismatches)/sum(region_data$total))
+	return(c(100*sum(region_data$mismatches)/sum(region_data$total), nrow(region_data)))
 }
-
 basic_rates <- sapply(regions, get_basic_error)
+rownames(basic_rates) <- c("error", "nseqs")
+
+
 
 mock_error_basic <- mock_error_good[mock_error_good$good_coords & mock_error_good$good_homop,]
 
 
-#number of sequences per region that survive the basic step
-basic_n <- aggregate(mock_error_basic$error, by=list(mock_error_basic$region), length)$x
-names(basic_n) <- names(basic_rates)
 
 
 
@@ -51,24 +44,21 @@ pb_by_region_mm <- aggregate(mock_error_basic$error, by=list(mock_error_basic$pb
 
 aggregate_pb_data <- function(region){
 	pb_region <- pb_by_region_mm[pb_by_region_mm$Group.2 == region,]
-
 	product <- pb_region$x[,"mean"] * pb_region$x[,"nseqs"]
 	pb_region_nseqs <- cumsum(pb_region$x[,"nseqs"])
 	pb_region_error <- cumsum(product) / pb_region_nseqs
-	pb_region_frac <- pb_region_nseqs / sum(pb_region$x[,"nseqs"])
[…]

### 2016-01-29 code/get_error_rate_table.R
Fix random typos to get it to run

@@ -237,7 +237,7 @@ base_error_reduction <- rbind(	raw = NA,
 								coverage_pred = coverage_pred_error_reduction,
 								pb_coverage = oligos_coverage_error_reduction,
 								pb_pred = oligos_pred_error_reduction,
-								all = all_filters_error_reduction_1,
+								all = all_filters_error_reduction,
 								precluster = precluster_error_reduction
 							)
 
@@ -249,7 +249,7 @@ base_nseqs_reduction <- rbind(	raw = NA,
 								coverage_pred = coverage_pred_nseqs_reduction,
 								pb_coverage = oligos_coverage_nseqs_reduction,
 								pb_pred = oligos_pred_nseqs_reduction,
-								all = all_filters_nseqs_reduction_1,
+								all = all_filters_nseqs_reduction,
 								precluster = precluster_nseqs_reduction
 							)
 
@@ -257,7 +257,7 @@ error <- rbind(	raw = raw_rates["error",],
 				basic = basic_rates["error",],
 				pb = pb_error["1",],
 				pred = pred_error_data["error",],
-				coverage = coverage_data["error",],
+				coverage = coverage_error["error",],
 				coverage_pred = coverage_pred_error[,"error"],
 				pb_coverage = oligos_coverage_error[,"error"],
 				pb_pred = oligos_pred_error[,"error"],
@@ -269,7 +269,7 @@ nseqs <- rbind(	raw = raw_rates["nseqs",],
 				basic = basic_rates["nseqs",],
 				pb = pb_nseqs["1",],
 				pred = pred_error_data["nseqs",],
-				coverage = coverage_data["nseqs",],
+				coverage = coverage_error["nseqs",],
 				coverage_pred = coverage_pred_error[,"nseqs"],
 				pb_coverage = oligos_coverage_error[,"nseqs"],
 				pb_pred = oligos_pred_error[,"nseqs"],
@@ -279,11 +279,11 @@ nseqs <- rbind(	raw = raw_rates["nseqs",],
 
 summary_data <- data.frame(
 				cbind(	error = as.vector(error),
-						nseqs = as.vector(nseqs)
+						nseqs = as.vector(nseqs),
 						error_red = as.vector(base_error_reduction),
 						nseqs_red = as.vector(base_nseqs_reduction)
 					))
-summary_data$region <- rep(colnames(error_summary), each=nrow(error_summary))
-summary_data$method <- rep(rownames(error_summary), ncol(error_summary))
[…]

### 2016-01-29 Schloss_PacBio16S_PeerJ_2016.Rmd
Use external error_summary file through pre-cluster analysis

@@
 regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
+pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")
+
 clrs <- rainbow(length(regions))
-pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")
 names(clrs) <- regions
 
-library(json)
+library(rjson)
 ```
@@
 
+error_summary <- read.table(file="data/process/error_summary.tsv", header=T)
+raw_to_basic_lost <- 100 * (1 - error_summary[error_summary$method == "basic","nseqs"]/error_summary[error_summary$method == "raw","nseqs"])
+
 ```
@@
 the expected alignment coordinates for that region of the 16S rRNA gene. This
-reduced the experiment-wide error rate from `r format(init_error_rate, digits=3, nsmall=2)` to
-`r format(basic_error_rate, digits=2, nsmall=2)`%. This basic procedure
-resulted in the removal of between
-`r format(min(100-basic_kept), digits=3, nsmall=1)`
-(`r pretty_region[names(which.min(100-basic_kept))]`) and
-`r format(max(100-basic_kept), digits=3, nsmall=1)`
-(`r pretty_region[names(which.max(100-basic_kept))]`)% of the reads (Table 2). Although
+reduced the experiment-wide error rate from `r format(mean(error_summary[error_summary$method=="raw","error"]), digits=3, nsmall=2)` to
+`r format(mean(error_summary[error_summary$method=="basic","error"]), digits=2, nsmall=2)`%. This basic procedure resulted in the removal of between
+`r format(min(raw_to_basic_lost), digits=3, nsmall=1)`
+(`r pretty_region[names(which.min(raw_to_basic_lost))]`) and
+`r format(max(raw_to_basic_lost), digits=3, nsmall=1)`
+(`r pretty_region[names(which.max(raw_to_basic_lost))]`)% of the reads (Table 2). Although
 the percentage of reads removed increased with the length of the fragment, there
@@
 pred_overall <- aggregate(100 * mock_error_basic$error, by=list(round(mock_error_basic$pred_error, 3)), mean)
-
 ```
@@
 10-fold coverage to those with less coverage, we reduced the error rate by
-`r paste(format(range(base_error_reduction), digits=3, nsmall=1), collapse=" to ")`%
+`r paste(format(range(error_summary[error_summary$method == "coverage","error_red"]), digits=3, nsmall=1), collapse=" to ")`%
 (Figure 2). Third, based on the observed correlation between the
@@
 we used this threshold, we were able to reduce the error rate by
-`r paste(format(range(pred_error_reduction), digits=3, nsmall=1), collapse=" to ")`%
+`r paste(format(range(error_summary[error_summary$method == "pred","error_red"]), digits=3, nsmall=1), collapse=" to ")`%
 (Figure 2A). Finally, we quantified the effect of combining filters. We found that
@@
 the analysis. For instance, among the V1-V9 data, requiring sequences to have a
-predicted error rate less than 0.01% resulted in a `r format(base_error_reduction["pred_error_reduction","V1V9"], digits=3, nsmall=1)`% and resulted in the removal of `r format(base_nseqs_reduction["pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences.
-Adding the mismatch or coverage filter
+predicted error rate less than 0.01% resulted in a
+`r format(error_summary[error_summary$method == "pred" & error_summary$region == "V1V9","error_red"], digits=3, nsmall=1)`% reduction in error and resulted in the removal of
+`r format(error_summary[error_summary$method == "pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`% of the sequences. Adding the mismatch or coverage filter
 had no effect on the reduction of error, but resulted in the removal of
-`r format(base_nseqs_reduction["oligos_pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)` and
-`r format(base_nseqs_reduction["coverage_pred_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences, respectively. Use of all filters had no impact on the reduction in the observed
-error rate, but resulted in the removal of `r format(base_nseqs_reduction["all_filters_nseqs_reduction","V1V9"], digits=3, nsmall=1)`% of the sequences.
+`r format(error_summary[error_summary$method == "pb_pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)` and
+`r format(error_summary[error_summary$method == "coverage_pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`
+% of the sequences, respectively. Use of all filters had no impact on the reduction in the observed
+error rate, but resulted in the removal of `r format(error_summary[error_summary$method == "all" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`% of the sequences.
 The remainder of this paper only uses sequences with a predicted error rate less than 0.01%.
-
-```{r precluster}
-get_error_rate <- function(file_name){
-	error_summary <- read.table(file=file_name, header=T, stringsAsFactors=FALSE)
-	nochim <- error_summary[error_summary$numparents==1,]
-    sum(nochim$weight * nochim$mismatches) / sum(nochim$weight * nochim$total)
-}
-
-regions <- c("V4", "V1V3", "V1V5", "V1V6", "V1V9", "V3V5")
-
-unique_file_names <- paste0("data/mothur_pool/", regions, ".mock.screen.unique.good.filter.unique.error.summary")
-
-unique_error <- sapply(unique_file_names, get_error_rate)
-
-precluster_file_names <- paste0("data/mothur_pool/", regions,".mock.screen.unique.good.filter.unique.precluster.error.summary")
-
-precluster_error <- sapply(precluster_file_names, get_error_rate)
-
-pooled_error <- cbind(unique = unique_error, precluster = precluster_error)
-rownames(pooled_error) <- regions
-
-#          unique   precluster
-#V4   0.003205151 0.0015805612
-#V1V3 0.002529819 0.0006841053
-#V1V5 0.001987322 0.0001938975
-#V1V6 0.001910158 0.0003086235
-#V1V9 0.001944654 0.0002743820
-#V3V5 0.002076498 0.0003532300
-
-reduction <- 100*(1-pooled_error[,"precluster"]/pooled_error[,"unique"])
-
-min.region <- pretty_region[names(reduction[which.min(reduction)])]
-max.region <- pretty_region[names(reduction[which.max(reduction)])]
-```
 
@@
 When we applied this approach to our PacBio data, we observed a reduction in the error rate
-between `r min(reduction)` (`r min.region`) and `r max(reduction)`% (`r max.region`).
+between
+`r format(min(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)`
+(`r pretty_region[which.min(error_summary[error_summary$method == "precluster","nseqs_red"])]`)
+and `r format(max(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)`%
+(`r pretty_region[which.max(error_summary[error_summary$method == "precluster","nseqs_red"])]`).
 The final error rates varied between
-`r round(100*min(pooled_error[,"precluster"]), digits=2)` (`r names(which.min(pooled_error[,"precluster"]))`) and `r round(100*max(pooled_error[,"precluster"]), digits=2)`% (`r names(which.max(pooled_error[,"precluster"]))`).
- The full-length (i.e. V1V9) fragments had an error rate of
-`r round(100* pooled_error["V1V9","precluster"], digits=2)`%
+`r round(min(error_summary[error_summary$method == "precluster","error"]), digits=2)`
+(`r pretty_region[which.min(error_summary[error_summary$method == "precluster","error"])]`)
+and
+`r round(max(error_summary[error_summary$method == "precluster","error"]), digits=2)`%
+(`r pretty_region[which.max(error_summary[error_summary$method == "precluster","error"])]`).
+The full-length (i.e. V1-V9) fragments had an error rate of
+`r round(error_summary[error_summary$method == "precluster" & error_summary$region=="V1V9","error"], digits=2)`%
 (Figure 2B), this is similar to what we have previously observed using the 454 and

### 2016-01-30 Schloss_PacBio16S_PeerJ_2016.Rmd
Extract taxonomy analysis

@@
 error_profile <- fromJSON(scan("data/process/error_profile.json", quiet=T, what="character"))
-
-error_summary <- read.table(file="data/process/error_summary.tsv", header=T)
-raw_to_basic_lost <- 100 * (1 - error_summary[error_summary$method == "basic","nseqs"]/error_summary[error_summary$method == "raw","nseqs"])
-
 ```
@@
 ***The PacBio error profile and a basic sequence curation procedure.***
-To build a sequence curation pipeline, we first needed to characterize the error
-rate associated with sequencing the 16S rRNA gene. We observed an average
-sequencing error rate of `r format(error_profile$overall_error, digits=2, nsmall=2)`%. Insertions, deletions, and substitutions
-accounted for `r format(error_profile$ins_rate, digits=3, nsmall=1)`,
-`r format(error_profile$del_rate, digits=3, nsmall=1)`, and `r format(error_profile$subs_rate, digits=3, nsmall=1)`% of the errors, respectively. The substitution errors were equally
-likely and all four bases
-were equally likely to cause insertion errors. Interestingly, guanines
-(`r format(error_profile$del_bias$dG, digits=3, nsmall=1)`%) and adenines
-(`r format(error_profile$del_bias$dA, digits=3, nsmall=1)`%) were more likely to be deleted
-than cytosines (`r error_profile$del_bias$dC, digits=3, nsmall=1)`%) or thymidines
-(`r error_profile$del_bias$dT, digits=2, nsmall=1)`%). The PacBio quality values varied between
-`r paste(error_profile$qual_range, collapse=' and ')`.
+To build a sequence curation pipeline, we first needed to characterize the error rate associated with sequencing the 16S rRNA gene.
+We observed an average sequencing error rate of `r format(error_profile$overall_error, digits=2, nsmall=2)`%.
+Insertions, deletions, and substitutions accounted for `r format(error_profile$ins_rate, digits=3, nsmall=1)`, `r format(error_profile$del_rate, digits=3, nsmall=1)`, and `r format(error_profile$subs_rate, digits=3, nsmall=1)`% of the errors, respectively.
+The substitution errors were equally likely and all four bases were equally likely to cause insertion errors.
+Interestingly, guanines (`r format(error_profile$del_bias$dG, digits=3, nsmall=1)`%) and adenines (`r format(error_profile$del_bias$dA, digits=3, nsmall=1)`%) were more likely to be deleted than cytosines (`r error_profile$del_bias$dC, digits=3, nsmall=1)`%) or thymidines (`r error_profile$del_bias$dT, digits=2, nsmall=1)`%).
+The PacBio quality values varied between `r paste(error_profile$qual_range, collapse=' and ')`.
 Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(error_profile$percent_over_93$m, digits=3, nsmall=1)`%), substitutions (`r format(error_profile$percent_over_93$s, digits=3, nsmall=1)`%), and insertions (`r format(error_profile$percent_over_93$i, digits=3, nsmall=1)`%).
@@
 
-We established a simple curation procedure by culling any sequence that had a
-string of the same base repeated more than 8 times  or did not start and end at
-the expected alignment coordinates for that region of the 16S rRNA gene. This
-reduced the experiment-wide error rate from `r format(mean(error_summary[error_summary$method=="raw","error"]), digits=3, nsmall=2)` to
-`r format(mean(error_summary[error_summary$method=="basic","error"]), digits=2, nsmall=2)`%. This basic procedure resulted in the removal of between
-`r format(min(raw_to_basic_lost), digits=3, nsmall=1)`
-(`r pretty_region[names(which.min(raw_to_basic_lost))]`) and
-`r format(max(raw_to_basic_lost), digits=3, nsmall=1)`
-(`r pretty_region[names(which.max(raw_to_basic_lost))]`)% of the reads (Table 2). Although
-the percentage of reads removed increased with the length of the fragment, there
-was no obvious relationship between fragment length and error rate (Figure 2).
+```{r basic}
+error_summary <- read.table(file="data/process/error_summary.tsv", header=T)
 
-```{r random_stuff }
-
-#This is for the Figure 2B plot...
-pb_overall <- aggregate(mock_error_basic$error, by=list(mock_error_basic$pbdiffs), function(x){c(mean=mean(100*x))})
-
-#This is for the Figure 2C plot...
-coverage_overall_agg <- aggregate(100 * mock_error_basic$error, by=list(mock_error_basic$coverage), mean)
-coverage_overall <- coverage_overall_agg$x
-names(coverage_overall) <- coverage_overall_agg$Group.1
-coverage_overall <- coverage_overall[as.character(3:40)]
-
-#This is for the Figure 2D plot...
-pred_overall <- aggregate(100 * mock_error_basic$error, by=list(round(mock_error_basic$pred_error, 3)), mean)
+raw_to_basic_lost <- 100 * (1 - error_summary[error_summary$method == "basic","nseqs"]/error_summary[error_summary$method == "raw","nseqs"])
 ```
 
+We established a simple curation procedure by culling any sequence that had a string of the same base repeated more than 8 times  or did not start and end at the expected alignment coordinates for that region of the 16S rRNA gene.
+This reduced the experiment-wide error rate from `r format(mean(error_summary[error_summary$method=="raw","error"]), digits=3, nsmall=2)` to `r format(mean(error_summary[error_summary$method=="basic","error"]), digits=2, nsmall=2)`%.
+This basic procedure resulted in the removal of between `r format(min(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(which.min(raw_to_basic_lost))]`) and `r format(max(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(which.max(raw_to_basic_lost))]`)% of the reads (Table 2).
+Although the percentage of reads removed increased with the length of the fragment, there was no obvious relationship between fragment length and error rate (Figure 2).
+
 ***Identifying correlates of increased sequencing error.***
-In contrast to the 454 and Illumina-based platforms where the sequencing
-quality decays with length, the consensus sequencing approach employed by the
-PacBio sequencer is thought to generate a uniform distribution of errors. This
-makes it impossible to simply trim sequences to high quality regions. Therefore,
-we sought to identify characteristics within sequences that would allow us to
-identify and remove those sequences with errors using three different
-approaches. First, we hypothesized that errors in the barcode and primer would
-be correlated with the error rate for the entire sequence. We observed a strong
-relationship between the number of mismatches to the barcodes and primers and the
-error rate of the rest of the sequence fragment (Figure 1B). Although
-allowing no mismatches to the barcodes and primers yielded the lowest error rate,
-that stringent criterion removed a large fraction of the reads from the dataset.
-Allowing at most one mismatch only marginally increased the error rate while
-retaining more sequences in the dataset (Figure 2). Second, we hypothesized that
-increased sequencing coverage should yield lower error rates. We found that once
-we had obtained 10-fold coverage of the fragments, the error rate did not change
-appreciably (Figure 1C). When we compared the error rates of reads with at least
-10-fold coverage to those with less coverage, we reduced the error rate by
-`r paste(format(range(error_summary[error_summary$method == "coverage","error_red"]), digits=3, nsmall=1), collapse=" to ")`%
-(Figure 2). Third, based on the observed correlation between the
-predicted and observed error rates, we sought to identify a minimum predicted
-error rate that would allow us to reduce the observed error rate. The average
-observed error rate for sequences with predicted error rates between 0.01 and
-0.10% was linear (Figure 1D). We decided to use a threshold of 0.01% because a
-large number of sequence reads were lost when we used a smaller threshold. When
-we used this threshold, we were able to reduce the error rate by
-`r paste(format(range(error_summary[error_summary$method == "pred","error_red"]), digits=3, nsmall=1), collapse=" to ")`%
-(Figure 2A). Finally, we quantified the effect of combining filters. We found that
-any combination of filters that included the predicted error rate threshold had
-the most significant impact on reducing the observed error rate. Furthermore,
-the inclusion of the mismatch and coverage filters had a negligible impact on
-error rates, but had a significant impact on the number of sequences included in
-the analysis. For instance, among the V1-V9 data, requiring sequences to have a
-predicted error rate less than 0.01% resulted in a
-`r format(error_summary[error_summary$method == "pred" & error_summary$region == "V1V9","error_red"], digits=3, nsmall=1)`% reduction in error and resulted in the removal of
-`r format(error_summary[error_summary$method == "pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`% of the sequences. Adding the mismatch or coverage filter
-had no effect on the reduction of error, but resulted in the removal of
-`r format(error_summary[error_summary$method == "pb_pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)` and
-`r format(error_summary[error_summary$method == "coverage_pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`
-% of the sequences, respectively. Use of all filters had no impact on the reduction in the observed
-error rate, but resulted in the removal of `r format(error_summary[error_summary$method == "all" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`% of the sequences.
+In contrast to the 454 and Illumina-based platforms where the sequencing quality decays with length, the consensus sequencing approach employed by the PacBio sequencer is thought to generate a uniform distribution of errors.
+This makes it impossible to simply trim sequences to high quality regions.
+Therefore, we sought to identify characteristics within sequences that would allow us to identify and remove those sequences with errors using three different approaches.
+First, we hypothesized that errors in the barcode and primer would be correlated with the error rate for the entire sequence.
+We observed a strong relationship between the number of mismatches to the barcodes and primers and the error rate of the rest of the sequence fragment (Figure 1B).
+Although allowing no mismatches to the barcodes and primers yielded the lowest error rate, that stringent criterion removed a large fraction of the reads from the dataset.
+Allowing at most one mismatch only marginally increased the error rate while retaining more sequences in the dataset (Figure 2).
+Second, we hypothesized that increased sequencing coverage should yield lower error rates.
+We found that once we had obtained 10-fold coverage of the fragments, the error rate did not change appreciably (Figure 1C).
+When we compared the error rates of reads with at least 10-fold coverage to those with less coverage, we reduced the error rate by `r paste(format(range(error_summary[error_summary$method == "coverage","error_red"]), digits=3, nsmall=1), collapse=" to ")`% (Figure 2).
+Third, based on the observed correlation between the predicted and observed error rates, we sought to identify a minimum predicted error rate that would allow us to reduce the observed error rate.
+The average observed error rate for sequences with predicted error rates between 0.01 and 0.10% was linear (Figure 1D).
+We decided to use a threshold of 0.01% because a large number of sequence reads were lost when we used a smaller threshold.
+When we used this threshold, we were able to reduce the error rate by `r paste(format(range(error_summary[error_summary$method == "pred","error_red"]), digits=3, nsmall=1), collapse=" to ")`% (Figure 2A).
+Finally, we quantified the effect of combining filters.
+We found that any combination of filters that included the predicted error rate threshold had the most significant impact on reducing the observed error rate.
+Furthermore, the inclusion of the mismatch and coverage filters had a negligible impact on error rates, but had a significant impact on the number of sequences included in the analysis.
+For instance, among the V1-V9 data, requiring sequences to have a predicted error rate less than 0.01% resulted in a `r format(error_summary[error_summary$method == "pred" & error_summary$region == "V1V9","error_red"], digits=3, nsmall=1)`% reduction in error and resulted in the removal of `r format(error_summary[error_summary$method == "pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`% of the sequences.
+Adding the mismatch or coverage filter had no effect on the reduction of error, but resulted in the removal of `r format(error_summary[error_summary$method == "pb_pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)` and `r format(error_summary[error_summary$method == "coverage_pred" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)` % of the sequences, respectively.
+Use of all filters had no impact on the reduction in the observed error rate, but resulted in the removal of `r format(error_summary[error_summary$method == "all" & error_summary$region == "V1V9","nseqs_red"], digits=3, nsmall=1)`% of the sequences.
 The remainder of this paper only uses sequences with a predicted error rate less than 0.01%.
@@
 ***Pre-clustering sequences to further reduce sequencing noise.***
-Previously, we implemented a pre-clustering algorithm where sequences were sorted
-by their abundance in decreasing order and rare sequences are clustered with a
-more abundant sequence if the rare sequences have fewer mismatches than a
-defined threshold when compared to the more abundant sequence [@refs]. The recommended
-threshold was a 1-nt difference per 100-nt of sequence data. For example, the
-threshold for 250 bp fragment from the V4 region would be 2 nt or 14 for the
-1458 bp V1-V9 fragments. This approach removes residual PCR and sequencing
-errors while not overwhelming the resolution needed to identify OTUs that are
-based on a 3% distance threshold. The tradeoff of this approach is that one would
-unable to differentiate V1-V9 sequences that truly differed by less than 14 nt.
-When we applied this approach to our PacBio data, we observed a reduction in the error rate
-between
-`r format(min(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)`
-(`r pretty_region[which.min(error_summary[error_summary$method == "precluster","nseqs_red"])]`)
-and `r format(max(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)`%
-(`r pretty_region[which.max(error_summary[error_summary$method == "precluster","nseqs_red"])]`).
-The final error rates varied between
-`r round(min(error_summary[error_summary$method == "precluster","error"]), digits=2)`
-(`r pretty_region[which.min(error_summary[error_summary$method == "precluster","error"])]`)
-and
-`r round(max(error_summary[error_summary$method == "precluster","error"]), digits=2)`%
-(`r pretty_region[which.max(error_summary[error_summary$method == "precluster","error"])]`).
-The full-length (i.e. V1-V9) fragments had an error rate of
-`r round(error_summary[error_summary$method == "precluster" & error_summary$region=="V1V9","error"], digits=2)`%
-(Figure 2B), this is similar to what we have previously observed using the 454 and
-Illumina MiSeq platforms (0.02%){Schloss, 2011 #2466; Kozich, 2013 #2719}
+Previously, we implemented a pre-clustering algorithm where sequences were sorted by their abundance in decreasing order and rare sequences are clustered with a more abundant sequence if the rare sequences have fewer mismatches than a defined threshold when compared to the more abundant sequence [@refs].
+The recommended threshold was a 1-nt difference per 100-nt of sequence data.
+For example, the threshold for 250 bp fragment from the V4 region would be 2 nt or 14 for the 1458 bp V1-V9 fragments.
+This approach removes residual PCR and sequencing errors while not overwhelming the resolution needed to identify OTUs that are based on a 3% distance threshold.
+The tradeoff of this approach is that one would unable to differentiate V1-V9 sequences that truly differed by less than 14 nt.
+When we applied this approach to our PacBio data, we observed a reduction in the error rate between `r format(min(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)` (`r pretty_region[which.min(error_summary[error_summary$method == "precluster","nseqs_red"])]`) and `r format(max(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)`% (`r pretty_region[which.max(error_summary[error_summary$method == "precluster","nseqs_red"])]`).
+The final error rates varied between `r round(min(error_summary[error_summary$method == "precluster","error"]), digits=2)` (`r pretty_region[which.min(error_summary[error_summary$method == "precluster","error"])]`) and `r round(max(error_summary[error_summary$method == "precluster","error"]), digits=2)`% (`r pretty_region[which.max(error_summary[error_summary$method == "precluster","error"])]`).
+The full-length (i.e.
+V1-V9) fragments had an error rate of `r round(error_summary[error_summary$method == "precluster" & error_summary$region=="V1V9","error"], digits=2)`% (Figure 2B), this is similar to what we have previously observed using the 454 and Illumina MiSeq platforms (0.02%){Schloss, 2011 #2466; Kozich, 2013 #2719}
 
@@
 ***Effects of error rates on OTU assignments.***
-The sequencing error rate is known to affect the number of OTUs that are
-observed {Schloss, 2011 #2466}. For each region, we determined that if there were no chimeras or
-PCR or sequencing errors, then we would expect to find `r perfect.otu[[1]]` OTUs. When
-achieved perfect chimera removal, but allowed for PCR and sequencing errors, we
-observed between `r diff<-otu.table[,2]-perfect.otu;
-round(min(diff), digits=1)` (`r names(which.min(diff))`) and `r round(max(diff), digits=1)`
-(`r names(which.max(diff))`) extra OTUs. The range in the number of extra OTUs was largely
-explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, pooled_error[,"precluster"])$estimate, 2)`). Next, we determined
-the number of OTUs that were observed when we used UCHIME to identify chimeric sequence.
-Under these more realistic conditions, we observed between `r diff<-otu.table[,3]-perfect.otu;
-round(min(diff), digits=1)` (`r names(which.min(diff))`) and `r round(max(diff), digits=1)`
-(`r names(which.max(diff))`) extra OTUs. Finally, we calculated the number of OTUs in the
-soil, mouse, and human samples using the same pipeline with chimera detection and removal
-based on the UCHIME algorithm. Suprisingly, there was not a clear relationship across sample type and region. Again, we found that there was a strong correlation between
-the number of observed OTUs and the error rate for the
-mouse (R=`r format(round(cor.test(otu.table[,"Mouse"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`)
-and human samples (R=`r format(round(cor.test(otu.table[,"Human"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`).
-These results underscore the effect of sequencing error on the inflation of the number of
-observed OTUs.
+The sequencing error rate is known to affect the number of OTUs that are observed {Schloss, 2011 #2466}.
+For each region, we determined that if there were no chimeras or PCR or sequencing errors, then we would expect to find `r perfect.otu[[1]]` OTUs.
+When achieved perfect chimera removal, but allowed for PCR and sequencing errors, we observed between `r diff<-otu.table[,2]-perfect.otu; round(min(diff), digits=1)` (`r names(which.min(diff))`) and `r round(max(diff), digits=1)` (`r names(which.max(diff))`) extra OTUs.
+The range in the number of extra OTUs was largely explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, pooled_error[,"precluster"])$estimate, 2)`).
+Next, we determined the number of OTUs that were observed when we used UCHIME to identify chimeric sequence.
+Under these more realistic conditions, we observed between `r diff<-otu.table[,3]-perfect.otu; round(min(diff), digits=1)` (`r names(which.min(diff))`) and `r round(max(diff), digits=1)` (`r names(which.max(diff))`) extra OTUs.
+Finally, we calculated the number of OTUs in the soil, mouse, and human samples using the same pipeline with chimera detection and removal based on the UCHIME algorithm.
+Suprisingly, there was not a clear relationship across sample type and region.
+Again, we found that there was a strong correlation between the number of observed OTUs and the error rate for the mouse (R=`r format(round(cor.test(otu.table[,"Mouse"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`) and human samples (R=`r format(round(cor.test(otu.table[,"Human"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`).
+These results underscore the effect of sequencing error on the inflation of the number of observed OTUs.
 
 ```{r, classificationSummary}
-data <- read.table(file="data/process/taxonomy.depth.analysis", header=T)
-
-rdp <- 100 * cbind("Mock"=data[data$db=="pds" & data$dataset=="mock", "genus_species"],
-				 "Human"=data[data$db=="pds" & data$dataset=="human", "genus_species"],
-				 "Mouse"=data[data$db=="pds" & data$dataset=="mouse", "genus_species"],
-				 "Soil"=data[data$db=="pds" & data$dataset=="soil", "genus_species"])
-rownames(rdp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
-
-gg <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "genus_species"],
-				"Human"=data[data$db=="gg" & data$dataset=="human", "genus_species"],
-				"Mouse"=data[data$db=="gg" & data$dataset=="mouse", "genus_species"],
-				"Soil"=data[data$db=="gg" & data$dataset=="soil", "genus_species"])
-rownames(gg) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
-
-silva <- 100 * cbind("Mock"=data[data$db=="nr" & data$dataset=="mock", "genus_species"],
-					 "Human"=data[data$db=="nr" & data$dataset=="human", "genus_species"],
-					 "Mouse"=data[data$db=="nr" & data$dataset=="mouse", "genus_species"],
-					 "Soil"=data[data$db=="nr" & data$dataset=="soil", "genus_species"])
-rownames(silva) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
-
-gg.sp <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "X7"],
-					 "Human"=data[data$db=="gg" & data$dataset=="human", "X7"],
-					 "Mouse"=data[data$db=="gg" & data$dataset=="mouse", "X7"],
-					 "Soil"=data[data$db=="gg" & data$dataset=="soil", "X7"])
-rownames(gg.sp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+data <- read.table(file="data/process/taxonomy_depth_analysis.tsv", header=T)
 ```
@@
 ***Increasing sequence length improves classification.***
-We classified all of the sequence data we generated using the naïve Bayesian
-classifier using the RDP, SILVA, and greengenes reference taxonomies (Figure
-4). In general, increasing the length of the region improved the
-ability to assign the sequence to a genus or species. Interestingly, each of the
-samples we analyzed varied in the ability to assign its sequences to the depth of
-genus or species. Furthermore, the reference database that did the best job of
-classifying the sequences varied by sample type. For example, the SILVA
-reference did the best for the human feces and soil samples and the RDP did the
-best for the mouse feces samples. An advantage of the greengenes database is
-that it contains
-information for 2,514 species-level lineages for 11% of the reference sequences;
-the other databases only provided taxonomic data to the genus level. There was a
-modest association between the length of the fragment and the
-ability to classify sequences to the species-level for the human samples; there
-was no such association for the mouse and soil samples. In fact, at most
-`r format(round(max(gg.sp[,"Soil"]), digits=1), nsmall=1)`% of
-the soil sequences and `r format(round(max(gg.sp[,"Mouse"]), digits=1), nsmall=1)`%
-of the mouse sequences could be classified to a
-species. These results indicate that the ability to classify sequences to the
-genus or species level is a function of read length, sample type, and the
-reference database.
+We classified all of the sequence data we generated using the naïve Bayesian classifier using the RDP, SILVA, and greengenes reference taxonomies (Figure 3).
+In general, increasing the length of the region improved the ability to assign the sequence to a genus or species.
+Interestingly, each of the samples we analyzed varied in the ability to assign its sequences to the depth of genus or species.
+Furthermore, the reference database that did the best job of classifying the sequences varied by sample type.
+For example, the SILVA reference did the best for the human feces and soil samples and the RDP did the best for the mouse feces samples.
+An advantage of the greengenes database is that it contains information for 2,514 species-level lineages for 11% of the reference sequences; the other databases only provided taxonomic data to the genus level.
+There was a modest association between the length of the fragment and the ability to classify sequences to the species-level for the human samples; there was no such association for the mouse and soil samples.
+In fact, at most `r format(round(100*max(data[data$db=="gg" & data$dataset=="soil","X7"]), digits=1), nsmall=1)`% of the soil sequences and `r format(round(100*max(data[data$db=="gg" & data$dataset=="mouse","X7"]), digits=1), nsmall=1)`% of the mouse sequences could be classified to a species.
+These results indicate that the ability to classify sequences to the genus or species level is a function of read length, sample type, and the reference database.
 
@@
 ***Sequencing errors are not random.***
-Above, we described that although there was no obvious bias in the substitution
-or insertion rate, we did observe that guanines and adenines were more likely to
-be deleted than cytosines or thymidines. This lack of randomness in the error
-profile suggested that there might be a systematic non-random distribution of
-the errors across the sequences. This would manifest itself by the creation of
-duplicate sequences with the same error. We identified all of the mock community
-sequences that had a 1-nt difference to the true sequence. For these three
-regions, between `r min(percent_one_off)` and `r max(percent_one_off)` of the
-sequences with 1-nt errors were only observed once. We found that the frequency
-of the most abundant 1-nt error paralleled the number of sequences.
-Surprisingly, the same 1-nt error appeared
-`r as.numeric(names(non_random$V1V6[length(non_random$V1V6)]))` times in the
-V1-V6 mock data and another 1-nt error appeared
-`r as.numeric(names(non_random$V1V9[length(non_random$V1V9)]))` times in the
-V1-V9 mock data. Contrary to previous reports
-{Carneiro, 2012 #3458;Koren, 2012 #3459}, these results indicate that
-reproducible errors occur with the PacBio sequencing platform and that they can
-be quite abundant. Through the use of the pre-clustering step described above
-these 1-nt errors would be ameliorated; however, this result indicates that
-caution should be used when attempting to use fine-scale OTU definitions.
+Above, we described that although there was no obvious bias in the substitution or insertion rate, we did observe that guanines and adenines were more likely to be deleted than cytosines or thymidines.
+This lack of randomness in the error profile suggested that there might be a systematic non-random distribution of the errors across the sequences.
+This would manifest itself by the creation of duplicate sequences with the same error.
+We identified all of the mock community sequences that had a 1-nt difference to the true sequence.
+For these three regions, between `r min(percent_one_off)` and `r max(percent_one_off)` of the sequences with 1-nt errors were only observed once.
+We found that the frequency of the most abundant 1-nt error paralleled the number of sequences.
+Surprisingly, the same 1-nt error appeared `r as.numeric(names(non_random$V1V6[length(non_random$V1V6)]))` times in the V1-V6 mock data and another 1-nt error appeared `r as.numeric(names(non_random$V1V9[length(non_random$V1V9)]))` times in the V1-V9 mock data.
+Contrary to previous reports {Carneiro, 2012 #3458;Koren, 2012 #3459}, these results indicate that reproducible errors occur with the PacBio sequencing platform and that they can be quite abundant.
+Through the use of the pre-clustering step described above these 1-nt errors would be ameliorated; however, this result indicates that caution should be used when attempting to use fine-scale OTU definitions.
 
 ## Conclusions
-The various sequencing platforms that are available to microbial ecologists are
-able to fill unique needs and have their own strengths and weaknesses. For
-sequencing the 16S rRNA gene, the 454 platform is able to generate a moderate
-number of high-quality 500-nt sequence fragments (error rates below 0.02%) {Schloss, 2011 #2466} and
-the MiSeq platform is able to generate a large number of high-quality 250-nt
-sequence fragments (error rates below 0.02%) {Kozich, 2013 #2719}. The promise
-of the PacBio sequencing platform was the generation of high-quality near
-full-length sequence fragments. As we have shown in this study, it is possible
-to generate near full-length sequences with error rates that are slightly
-higher, but comparable to the other platforms (i.e.
+The various sequencing platforms that are available to microbial ecologists are able to fill unique needs and have their own strengths and weaknesses.
+For sequencing the 16S rRNA gene, the 454 platform is able to generate a moderate number of high-quality 500-nt sequence fragments (error rates below 0.02%) {Schloss, 2011 #2466} and the MiSeq platform is able to generate a large number of high-quality 250-nt sequence fragments (error rates below 0.02%) {Kozich, 2013 #2719}.
+The promise of the PacBio sequencing platform was the generation of high-quality near full-length sequence fragments.
+As we have shown in this study, it is possible to generate near full-length sequences with error rates that are slightly higher, but comparable to the other platforms (i.e.
 `r round(100*pooled_error["V1V9","precluster"],2)`%).
-With the exception of the V4 region, the error rates were less than
-`r 100*max(pooled_error[!(rownames(pooled_error) %in% "V4"),"precluster"])`%.
-When we considered the shorter V4 region, which is similar in length to what is
-sequenced by the MiSeq platform, the error rates we observed with the PacBio
-platform were nearly 8-fold higher than what has previously been reported. It
-was unclear why these shorter reads had such a high error rate relative to the
-other regions. At this point, the primary limitation of generating full-length
-sequences on the PacBio platform is the cost of generating the data and
-accessibility to the sequencers.
+With the exception of the V4 region, the error rates were less than `r 100*max(pooled_error[!(rownames(pooled_error) %in% "V4"),"precluster"])`%.
+When we considered the shorter V4 region, which is similar in length to what is sequenced by the MiSeq platform, the error rates we observed with the PacBio platform were nearly 8-fold higher than what has previously been reported.
+It was unclear why these shorter reads had such a high error rate relative to the other regions.
+At this point, the primary limitation of generating full-length sequences on the PacBio platform is the cost of generating the data and accessibility to the sequencers.
 
-The widespread adoption of the 454 and MiSeq platforms and decrease in the use
-of Sanger sequencing for the 16S rRNA gene has resulted in a decrease in the
-generation of the full-length reference sequences that are needed for performing
-phylogenetic analyses and designing lineage specific PCR primers and fluorescent
-*in situ* hybridization (FISH) probes. It remains to be determined whether the
-error rates we observed for full-length sequences are prohibitive for
-these applications. We can estimate the distribution of errors assuming that the
-errors follow a binomial distribution along the length of the 1,500 nt
-gene with the error rate that we achieved from the V1-V9 mock community data
-prior to pre-clustering the sequences, which was
-`r round(100*pooled_error["V1V9","unique"],2)`% (Figure 5). Under
-these conditions one would only expect
-`r round(100*dbinom(0, 1500, pooled_error["V1V9","unique"]), 2)`% of the
-sequences to have no errors and 50% of the sequences would have at least
-`r qbinom(0.50, 1500, pooled_error["V1V9","unique"])` errors.
-After applying the pre-clustering denoising step, the erro rate drops by `r round(pooled_error["V1V9","unique"]/pooled_error["V1V9","precluster"])`-fold to
-`r 100*pooled_error["V1V9","precluster"]`%. With this error rate, we would
-expect `r round(100*dbinom(0, 1500, pooled_error["V1V9","precluster"]), 2)`%
-of the sequences to have no sequencing errors. Of course, the tradeoff is the
-loss of resolution among closely related sequences.
+The widespread adoption of the 454 and MiSeq platforms and decrease in the use of Sanger sequencing for the 16S rRNA gene has resulted in a decrease in the generation of the full-length reference sequences that are needed for performing phylogenetic analyses and designing lineage specific PCR primers and fluorescent *in situ* hybridization (FISH) probes.
+It remains to be determined whether the error rates we observed for full-length sequences are prohibitive for these applications.
+We can estimate the distribution of errors assuming that the errors follow a binomial distribution along the length of the 1,500 nt gene with the error rate that we achieved from the V1-V9 mock community data prior to pre-clustering the sequences, which was `r round(100*pooled_error["V1V9","unique"],2)`% (Figure 5).
+Under these conditions one would only expect `r round(100*dbinom(0, 1500, pooled_error["V1V9","unique"]), 2)`% of the sequences to have no errors and 50% of the sequences would have at least `r qbinom(0.50, 1500, pooled_error["V1V9","unique"])` errors.
+After applying the pre-clustering denoising step, the erro rate drops by `r round(pooled_error["V1V9","unique"]/pooled_error["V1V9","precluster"])`-fold to `r 100*pooled_error["V1V9","precluster"]`%.
+With this error rate, we would expect `r round(100*dbinom(0, 1500, pooled_error["V1V9","precluster"]), 2)`% of the sequences to have no sequencing errors.
+Of course, the tradeoff is the loss of resolution among closely related sequences.
 
-Full-length sequences are frequently seen as a panacea to overcome the
-limitations of taxonomic classifications. The ability to classify each of our
-sample types benefited from the generation of full-length sequences. It was
-interesting that the benefit varied by sample type and database. For example,
-using the mouse libraries, the ability to classify each of the regions differed
-by less than 5% when classifying against the SILVA and greengenes databases. The
-effect of the database that was used was also interesting. The RDP database
-outperformed the other databases for the mouse samples and the SILVA database
-outperformed the others for the human and soil samples. The three databases were
-equally effective for classifying the mock community. Finally, since only the
-greengenes database provided species-level information for its reference
-sequences it was the only database that allowed for resolution of species-level
-classification. The sequences from the mouse and soil libraries were not
-effectively classified to the species level (all less than 10%). In contrast,
-classification of the human libraries resulted in more than 40% of the sequences
-being classified to a genus, regardless of the region. That the variation in
-species-level classification for the human libraries was less than 10% suggests
-that the benefit of added length is minimal considering the lower sequencing
-yield.
+Full-length sequences are frequently seen as a panacea to overcome the limitations of taxonomic classifications.
+The ability to classify each of our sample types benefited from the generation of full-length sequences.
+It was interesting that the benefit varied by sample type and database.
+For example, using the mouse libraries, the ability to classify each of the regions differed by less than 5% when classifying against the SILVA and greengenes databases.
+The effect of the database that was used was also interesting.
+The RDP database outperformed the other databases for the mouse samples and the SILVA database outperformed the others for the human and soil samples.
+The three databases were equally effective for classifying the mock community.
+Finally, since only the greengenes database provided species-level information for its reference sequences it was the only database that allowed for resolution of species-level classification.
+The sequences from the mouse and soil libraries were not effectively classified to the species level (all less than 10%).
+In contrast, classification of the human libraries resulted in more than 40% of the sequences being classified to a genus, regardless of the region.
+That the variation in species-level classification for the human libraries was less than 10% suggests that the benefit of added length is minimal considering the lower sequencing yield.
 
-The development of newer sequencing technologies continue to advance
-and there is justifiable excitement to apply these technologies to sequence the
-16S rRNA gene. Although it is clearly possible to generate sequencing data from
-these various platforms, it is critical that we assess the platforms for their
-ability to generate high quality data and the particular niche that the new
-approach will fill. With this in mind, it is essential that researchers utilize
-mock communities as part of their experimental design so that they can quantify
-their error rates. The ability to generate near full-length 16S rRNA gene
-sequences is an exciting advance that will hopefully expand our ability to
-improve the characterization of microbial communities.
+The development of newer sequencing technologies continue to advance and there is justifiable excitement to apply these technologies to sequence the 16S rRNA gene.
+Although it is clearly possible to generate sequencing data from these various platforms, it is critical that we assess the platforms for their ability to generate high quality data and the particular niche that the new approach will fill.
+With this in mind, it is essential that researchers utilize mock communities as part of their experimental design so that they can quantify their error rates.
+The ability to generate near full-length 16S rRNA gene sequences is an exciting advance that will hopefully expand our ability to improve the characterization of microbial communities.
 
 ## Acknowledgements
-The Genomic DNA from Microbial Mock Community A (Even, Low Concentration, v3.1,
-HM-278D) was obtained through the NIH Biodefense and Emerging Infections
-Research Resources Repository, NIAID, NIH as part of the Human Microbiome
-Project.
+The Genomic DNA from Microbial Mock Community A (Even, Low Concentration, v3.1, HM-278D) was obtained through the NIH Biodefense and Emerging Infections Research Resources Repository, NIAID, NIH as part of the Human Microbiome Project.
 
 ## Funding statement
-This study was supported by grants from the NIH (R01HG005975, R01GM099514 and
-P30DK034933 to PDS and U54HG004973 to SKH).
+This study was supported by grants from the NIH (R01HG005975, R01GM099514 and P30DK034933 to PDS and U54HG004973 to SKH).
 
@@
 
-**Table 2. Summary of the reasons that sequences were excluded because of the
-basic sequence curation steps**
+**Table 2. Summary of the reasons that sequences were excluded because of the basic sequence curation steps**
 
@@
 
-[ This figure removed ]
-**Figure 3. The relationship between the error rate of each region and the
-composite quality scores for the sequences.** The error rates (A and C) and
-percentage of sequences (B and D) were calculated for the reads that had a
-composite quality score above the plotted value. The composite quality scores
-were calculated by either determining the minimum value of the average quality
-score within all 50-nt windows within each region (A and B) or by calculating
-the average quality score across the entire sequence read (C and D).
-
-**Figure 4. Percentage of unique sequences that could be classified.**
+**Figure 3. Percentage of unique sequences that could be classified.**
 Classifications were performed using taxonomy references curated from the RDP,
@@
 ### Figure 2
+```{r random_stuff }
+
+#This is for the Figure 2B plot...
+pb_overall <- aggregate(mock_error_basic$error, by=list(mock_error_basic$pbdiffs), function(x){c(mean=mean(100*x))})
+
+#This is for the Figure 2C plot...
+coverage_overall_agg <- aggregate(100 * mock_error_basic$error, by=list(mock_error_basic$coverage), mean)
+coverage_overall <- coverage_overall_agg$x
+names(coverage_overall) <- coverage_overall_agg$Group.1
+coverage_overall <- coverage_overall[as.character(3:40)]
+
+#This is for the Figure 2D plot...
+pred_overall <- aggregate(100 * mock_error_basic$error, by=list(round(mock_error_basic$pred_error, 3)), mean)
+```
+
 ```{r Figure_2, width=6.5, height=6}
@@
 ### Figure 3
-```{r Figure_3, fig.height=6.5, fig.width=6.5}
-aggregateErrorMinQualityScore <- function(folder){
-	write(folder, "")
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
-								 ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-
-	remaining.error <- rep(0, max(composite$minQ)+1)
-	qscores <- 0:max(composite$minQ)
-
-	remaining.error <- unlist(lapply(qscores, function(x){mean(composite[composite$minQ >= x,"error"])}))
-	names(remaining.error) <- qscores
-	points(x=10:70, remaining.error[10:70], type="l", col=clrs[folder], lwd=2)
-}
-
-fractionMinQualityScore <- function(folder){
-	write(folder, "")
-
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
-					 				".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-
-	binned <- aggregate(composite$error, by=list(round(composite$minQ)), length)
-	remaining <- 1-cumsum(binned$x)/sum(binned$x)
-	points(binned$Group.1, remaining, type="l", col=clrs[folder], lwd=2)
-}
-
-aggregateErrorAveQualityScore <- function(folder){
-	write(folder, "")
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
-								 ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-
-	remaining.error <- rep(0, max(composite$aveQ)+1)
-	qscores <- 0:max(composite$aveQ)
-	remaining.error <- unlist(lapply(qscores,
-					function(x){mean(composite[composite$aveQ >= x,"error"])}))
-	names(remaining.error) <- qscores
-	points(x=10:70, remaining.error[10:70], type="l", col=clrs[folder], lwd=2)
-}
-
-fractionAveQualityScore <- function(folder){
-write(folder, "")
-
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
-								 ".composite", sep=""), header=T, row.names=1)
-	composite <- composite[composite$reason == "g",]
-
-	binned <- aggregate(composite$error, by=list(round(composite$aveQ)), length)
-
-	remaining <- 1-cumsum(binned$x)/sum(binned$x)
-	points(binned$Group.1, remaining, type="l", col=clrs[folder], lwd=2)
-}
-
-layout(1)
-layout(matrix(c(1,2,3,4), nrow=2, byrow=F), width=c(1.1,1), height=c(1, 1.3))
-
-#A
-par(mar=c(0.5,5,0.5, 0.5))
-plot(1, xlim=c(10,75), ylim=c(0,0.015), type="n", axes=F, xlab="", ylab="")
-lapply(regions, aggregateErrorMinQualityScore)
-mtext(side=2, line=2.5, text="Error rate of sequences\nabove threshold (%)", cex.lab=0.9)
-axis(1, at=seq(10,70,10), label=F, tick=T)
-axis(2, las=2, at=seq(0,0.015,0.005), label=format(seq(0,1.5,0.5), nsmall=1), cex.axis=1.25)
-box()
-mtext(side=2, line=2.6, at=0.015, text="A", las=2, cex=2, font=2)
-
-#B
-par(mar=c(6,5,0.5, 0.5))
-plot(1, xlim=c(10,75), ylim=c(0,1), type="n", yaxt="n", xlab="", ylab="", cex.axis=1.25)
-lapply(regions, fractionMinQualityScore)
-mtext(side=1, line=4.25, text="Minimum average quality\nscore within a 50-nt window\nacross the full sequence", cex=0.9)
-mtext(side=2, line=2.5, text="Sequences above\nthreshold (%)", cex.lab=0.9)
-
-axis(2, las=2, at=seq(0,1,0.2), label=seq(0,100,20), cex.axis=1.25)
-mtext(side=2, line=2.6, at=1, text="B", las=2, cex=2, font=2)
-
-#C
-par(mar=c(0.5,2,0.5, 0.5))
-plot(1, xlim=c(10,75), ylim=c(0,0.015), type="n", axes=F, xlab="", ylab="")
-lapply(regions, aggregateErrorAveQualityScore)
-axis(1, at=seq(10,70,10), label=F, tick=T)
-axis(2, at=seq(0,0.02,0.005), label=F, tick=T)
-box()
-mtext(side=2, line=0.6, at=0.015, text="C", las=2, cex=2, font=2)
-
-#D
-par(mar=c(6,2,0.5, 0.5))
-plot(1, xlim=c(10,75), ylim=c(0,1), type="n", yaxt="n", xlab="", ylab="", cex.axis=1.25)
-lapply(regions, fractionAveQualityScore)
-mtext(side=1, text="Average quality score", cex=0.9, line=3)
-title(ylab="Sequences above threshold (%)")
-axis(2, at=seq(0,1,0.2), label=FALSE, tick=TRUE)
-mtext(side=2, line=0.6, at=1, text="D", las=2, cex=2, font=2)
-
-legend(x=10, y=0.50, legend=pretty_region[regions], lty=1, lwd=2, col=clrs[regions], cex=1.0)
-```
-
-### Figure 4
 
 ```{r, Figure_4, fig.width=7, fig.height=7}
+rdp <- 100 * cbind("Mock"=data[data$db=="pds" & data$dataset=="mock", "genus_species"],
+				 "Human"=data[data$db=="pds" & data$dataset=="human", "genus_species"],
+				 "Mouse"=data[data$db=="pds" & data$dataset=="mouse", "genus_species"],
+				 "Soil"=data[data$db=="pds" & data$dataset=="soil", "genus_species"])
+rownames(rdp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+
+gg <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "genus_species"],
+				"Human"=data[data$db=="gg" & data$dataset=="human", "genus_species"],
+				"Mouse"=data[data$db=="gg" & data$dataset=="mouse", "genus_species"],
+				"Soil"=data[data$db=="gg" & data$dataset=="soil", "genus_species"])
+rownames(gg) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+
+silva <- 100 * cbind("Mock"=data[data$db=="nr" & data$dataset=="mock", "genus_species"],
+					 "Human"=data[data$db=="nr" & data$dataset=="human", "genus_species"],
+					 "Mouse"=data[data$db=="nr" & data$dataset=="mouse", "genus_species"],
+					 "Soil"=data[data$db=="nr" & data$dataset=="soil", "genus_species"])
+rownames(silva) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+
+gg.sp <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "X7"],
+					 "Human"=data[data$db=="gg" & data$dataset=="human", "X7"],
+					 "Mouse"=data[data$db=="gg" & data$dataset=="mouse", "X7"],
+					 "Soil"=data[data$db=="gg" & data$dataset=="soil", "X7"])
+rownames(gg.sp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+
 par(mar=c(5, 5, 0.5, 1))

### 2016-01-30 Schloss_PacBio16S_PeerJ_2016.Rmd
Extract Sobs analysis

@@
 Finally, we calculated the number of OTUs in the soil, mouse, and human samples using the same pipeline with chimera detection and removal based on the UCHIME algorithm.
-Suprisingly, there was not a clear relationship across sample type and region.
+Surpisingly, there was not a clear relationship across sample type and region.
 Again, we found that there was a strong correlation between the number of observed OTUs and the error rate for the mouse (R=`r format(round(cor.test(otu.table[,"Mouse"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`) and human samples (R=`r format(round(cor.test(otu.table[,"Human"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`).
@@
 ```{r, classificationSummary}
-data <- read.table(file="data/process/taxonomy_depth_analysis.tsv", header=T)
+taxonomy_data <- read.table(file="data/process/taxonomy_depth_analysis.tsv", header=T)
 ```
@@
 There was a modest association between the length of the fragment and the ability to classify sequences to the species-level for the human samples; there was no such association for the mouse and soil samples.
-In fact, at most `r format(round(100*max(data[data$db=="gg" & data$dataset=="soil","X7"]), digits=1), nsmall=1)`% of the soil sequences and `r format(round(100*max(data[data$db=="gg" & data$dataset=="mouse","X7"]), digits=1), nsmall=1)`% of the mouse sequences could be classified to a species.
+In fact, at most `r format(round(100*max(taxonomy_data[taxonomy_data$db=="gg" & taxonomy_data$dataset=="soil","X7"]), digits=1), nsmall=1)`% of the soil sequences and `r format(round(100*max(taxonomy_data[taxonomy_data$db=="gg" & taxonomy_data$dataset=="mouse","X7"]), digits=1), nsmall=1)`% of the mouse sequences could be classified to a species.
 These results indicate that the ability to classify sequences to the genus or species level is a function of read length, sample type, and the reference database.

### 2016-01-30 code/get_sobs_table.R

@@ -1,0 +1,36 @@
+regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
+pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")
+
+getAveSobs <- function(region, sample){
+	file_name <- paste0("data/mothur_pool/", region, ".", sample, ".screen.unique.good.filter.unique.precluster.pick.an.ave-std.summary")
+
+	data_sobs <- NA
+
+	if(file.exists(file_name)){
+		data <- read.table(file=file_name, header=T)
+		data_sobs <- data[data$method=="ave", "sobs"]
+	}
+
+	data_sobs
+}
+
+getOTUResults <- function(region){
+ 	perfect <- read.table(file=paste0("data/mothur_pool/HMP_MOCK.", region, ".pick.phylip.an.summary"), header=T)
+ 	nochims <- read.table(file=paste0("data/mothur_pool/", region, ".mock.screen.unique.good.filter.unique.precluster.perfect.an.ave-std.summary"), header=T)
+
+ 	obs.mock <- getAveSobs(region, "mock")
+ 	obs.mouse <- getAveSobs(region, "mouse")
+ 	obs.human <- getAveSobs(region, "human")
+ 	obs.soil <- getAveSobs(region, "soil")
+
+	nseqs <- nochims[nochims$method=="ave", "nseqs"]
+
+	return(c(perfect[1,2], nochims[1,"sobs"], obs.mock, obs.soil, obs.mouse, obs.human, nseqs))
+}
+
+otu_table <- t(sapply(regions, getOTUResults))
+nseqs <- otu_table[[1,7]]
+
+colnames(otu_table) <- c("no_error","no_chimeras", "mock", "soil", "mouse", "human", "nseqs")
+
+write.table(file="data/process/sobs_table.tsv")

### 2016-01-30 code/get_one_off_table.R
Extract one-off analysis

@@ -1,0 +1,31 @@
+getRegionNonRandom <- function(region){
+	file_name <- paste0("data/mothur_pool/", region, ".mock.screen.unique.good.filter.unique.error.summary")
+
+	error <- read.table(file=file_name, header=T, row.names=1)
+	error.nochim <- error[error$numparents==1,]
+
+	one.off <- error.nochim[error.nochim$mismatches==1,]
+
+	one.off.n <- sum(one.off$weight)
+	one.off.table <- table(one.off$weight)
+	nseqs <- sum(error.nochim$weight)
+	one.off.table
+}
+
+regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
+non_random <- lapply(regions, getRegionNonRandom)
+names(non_random) <- regions
+
+max_freq <- max(sapply(non_random, function(x){max(as.numeric(names(x)))}))
+one_offs <- matrix(0, nrow=max_freq, ncol=length(regions))
+rownames(one_offs) <- 1:max_freq
+colnames(one_offs) <- regions
+
+for(r in regions){
+	data <- unlist(unname(non_random[r]))
+	one_offs[as.numeric(names(data)),r] <- data
+	one_offs[,r]
+}
+
+write.table(one_offs, "data/process/non_random_analysis.tsv", sep="\t", quote=F)
+

### 2016-01-30 Schloss_PacBio16S_PeerJ_2016.Rmd
Extract one-off analysis

@@
 ```{r}
-getAveSobs <- function(region, sample){
-	file_name <- paste0("data/mothur_pool/", region, ".", sample, ".screen.unique.good.filter.unique.precluster.pick.an.ave-std.summary")
-
-	data_sobs <- NA
-
-	if(file.exists(file_name)){
-		data <- read.table(file=file_name, header=T)
-		data_sobs <- data[data$method=="ave", "sobs"]
-	}
-
-	data_sobs
-}
-
-getOTUResults <- function(region){
- 	perfect <- read.table(file=paste0("data/mothur_pool/HMP_MOCK.", region, ".pick.phylip.an.summary"), header=T)
- 	nochims <- read.table(file=paste0("data/mothur_pool/", region, ".mock.screen.unique.good.filter.unique.precluster.perfect.an.ave-std.summary"), header=T)
-
- 	obs.mock <- getAveSobs(region, "mock")
- 	obs.mouse <- getAveSobs(region, "mouse")
- 	obs.human <- getAveSobs(region, "human")
- 	obs.soil <- getAveSobs(region, "soil")
-
-	nseqs <- nochims[nochims$method=="ave", "nseqs"]
-
-	return(c(perfect[1,2], nochims[1,"sobs"], obs.mock, obs.soil, obs.mouse, obs.human, nseqs))
-}
-
-otu.table <- t(sapply(regions, getOTUResults))
-nseqs <- otu.table[[1,7]]
-perfect.otu <- otu.table[1,1]
-otu.table <- otu.table[,-c(1,7)]
-
-otu.table <- cbind(100 * pooled_error[rownames(otu.table),"precluster"], otu.table)
-colnames(otu.table) <- c("Error Rate (%)", "Mock with no chimeras", "Observed mock", "Soil", "Mouse", "Human")
-rownames(otu.table) <- pretty_region[rownames(otu.table)]
-
-#kable(otu.table, digits=c(2, rep(1,5)))
-
-#|      | Error Rate (%)| Mock with no chimeras| Observed mock|  Soil| Mouse| Human|
-#|:-----|--------------:|---------------------:|-------------:|-----:|-----:|-----:|
-#|V4    |           0.16|                  33.4|          48.7| 630.1| 212.4| 117.0|
-#|V1-V3 |           0.07|                  20.5|          27.7| 726.3| 107.5| 105.4|
-#|V3-V5 |           0.04|                  21.1|          31.2|    NA| 112.3|  83.7|
-#|V1-V5 |           0.02|                  19.8|          27.2| 694.4|  66.0|  85.3|
-#|V1-V6 |           0.03|                  20.0|          46.4| 693.3|  75.5|  89.2|
-#|V1-V9 |           0.03|                  19.5|          40.3|    NA| 102.7| 117.6|
-
+sobs_data <- read.table(file="data/process/sobs_table.tsv", header=T)
 ```
@@
 ```{r}
-getRegionNonRandom <- function(region){
-	file_name <- paste0("data/mothur_pool/", region, ".mock.screen.unique.good.filter.unique.error.summary")
+one_offs <- read.table(file="data/process/non_random_analysis.tsv", header=T)
+percent_one_off <- 100 * prop.table(as.matrix(one_offs), 2)
 
-	error <- read.table(file=file_name, header=T, row.names=1)
-	error.nochim <- error[error$numparents==1,]
+get_max_one_off <- function(region){
+	max(as.numeric(rownames(one_offs[one_offs[,region]==1,])))
+}
+max_one_off <- sapply(regions, get_max_one_off)
 
-	one.off <- error.nochim[error.nochim$mismatches==1,]
-
-	one.off.n <- sum(one.off$weight)
-	one.off.table <- table(one.off$weight)
-	nseqs <- sum(error.nochim$weight)
-	one.off.table
-}
-
-non_random <- lapply(regions, getRegionNonRandom)
-names(non_random) <- regions
-
-percent_one_off <- 100 * sapply(non_random, function(x){x[1]/sum(x)})
 ```
@@
 We identified all of the mock community sequences that had a 1-nt difference to the true sequence.
-For these three regions, between `r min(percent_one_off)` and `r max(percent_one_off)` of the sequences with 1-nt errors were only observed once.
+For these three regions, between `r paste0(format(round(range(percent_one_off[1,]), digits=1), nsmall=1), collapse=" and ")` of the sequences with 1-nt errors were only observed once.
 We found that the frequency of the most abundant 1-nt error paralleled the number of sequences.
-Surprisingly, the same 1-nt error appeared `r as.numeric(names(non_random$V1V6[length(non_random$V1V6)]))` times in the V1-V6 mock data and another 1-nt error appeared `r as.numeric(names(non_random$V1V9[length(non_random$V1V9)]))` times in the V1-V9 mock data.
+Surprisingly, the same 1-nt error appeared `r max_one_off["V1V6"]` times (`r format(round(percent_one_off[max_one_off["V1V6"], "V1V6"], digits=2), nsmall=2)`%) in the V1-V6 mock data and another 1-nt error appeared `r max_one_off["V1V9"]` times (`r format(round(percent_one_off[max_one_off["V1V9"], "V1V9"], digits=2), nsmall=2)`%) in the V1-V9 mock data.
 Contrary to previous reports {Carneiro, 2012 #3458;Koren, 2012 #3459}, these results indicate that reproducible errors occur with the PacBio sequencing platform and that they can be quite abundant.
@@
 
+#plot one_offs
+```{r}
+one_offs <- read.table(file="data/process/non_random_analysis.tsv", header=T)
+
+regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
+
+z <- prop.table(as.matrix(one_offs), 2)[1:10,]
+plot(NA, xlim=c(1,10), ylim=c(0,1), xlab="", ylab="", axes=F)
+for(r in regions){
+	points(1:10, z[,r], type="l")
+}
+axis(1, at=1:10, label=1:10)
+axis(2)
+```
+
 ## References

### 2016-01-30 Schloss_PacBio16S_PeerJ_2016.Rmd
Fix up R code in Conclusion

@@
 max_one_off <- sapply(regions, get_max_one_off)
-
 ```
@@
 As we have shown in this study, it is possible to generate near full-length sequences with error rates that are slightly higher, but comparable to the other platforms (i.e.
-`r round(100*pooled_error["V1V9","precluster"],2)`%).
-With the exception of the V4 region, the error rates were less than `r 100*max(pooled_error[!(rownames(pooled_error) %in% "V4"),"precluster"])`%.
-When we considered the shorter V4 region, which is similar in length to what is sequenced by the MiSeq platform, the error rates we observed with the PacBio platform were nearly 8-fold higher than what has previously been reported.
+`r round(error_summary[error_summary$region=="V1V9" & error_summary$method=="precluster","error"],2)`%).
+With the exception of the V4 region (`r round(error_summary[error_summary$region=="V4" & error_summary$method=="precluster","error"], 2)`%), the error rates were less than `r round(max(error_summary[error_summary$region!="V4" & error_summary$method=="precluster","error"]), 2)`%.
+When we considered the shorter V4 region, which is similar in length to what is sequenced by the MiSeq platform, the error rates we observed with the PacBio platform were nearly `r round(error_summary[error_summary$region=="V4" & error_summary$method=="precluster","error"]/0.02)`-fold higher than what has previously been reported on the other platforms.
 It was unclear why these shorter reads had such a high error rate relative to the other regions.

### 2016-01-30 code/get_sobs_table.R
Get R code to output data correctly

@@ -6,7 +6,7 @@ getAveSobs <- function(region, sample){
 
 	data_sobs <- NA
 
-	if(file.exists(file_name)){
+	if(file.size(file_name) != 0){
 		data <- read.table(file=file_name, header=T)
 		data_sobs <- data[data$method=="ave", "sobs"]
 	}
@@ -33,4 +33,4 @@ nseqs <- otu_table[[1,7]]
 
 colnames(otu_table) <- c("no_error","no_chimeras", "mock", "soil", "mouse", "human", "nseqs")
 
-write.table(file="data/process/sobs_table.tsv")
+write.table(otu_table, file="data/process/sobs_table.tsv")

### 2016-01-30 Schloss_PacBio16S_PeerJ_2016.Rmd
Fix up R code in Conclusion

@@
 
+```{r binomial_analysis}
+unique <- error_summary[error_summary$region=="V1V9" & error_summary$method=="coverage_pred","error"]/100
+
+precluster <- error_summary[error_summary$region=="V1V9" & error_summary$method=="precluster","error"] / 100
+unique_to_precluster <- unique / precluster
+```
+
 The widespread adoption of the 454 and MiSeq platforms and decrease in the use of Sanger sequencing for the 16S rRNA gene has resulted in a decrease in the generation of the full-length reference sequences that are needed for performing phylogenetic analyses and designing lineage specific PCR primers and fluorescent *in situ* hybridization (FISH) probes.
 It remains to be determined whether the error rates we observed for full-length sequences are prohibitive for these applications.
-We can estimate the distribution of errors assuming that the errors follow a binomial distribution along the length of the 1,500 nt gene with the error rate that we achieved from the V1-V9 mock community data prior to pre-clustering the sequences, which was `r round(100*pooled_error["V1V9","unique"],2)`% (Figure 5).
-Under these conditions one would only expect `r round(100*dbinom(0, 1500, pooled_error["V1V9","unique"]), 2)`% of the sequences to have no errors and 50% of the sequences would have at least `r qbinom(0.50, 1500, pooled_error["V1V9","unique"])` errors.
-After applying the pre-clustering denoising step, the erro rate drops by `r round(pooled_error["V1V9","unique"]/pooled_error["V1V9","precluster"])`-fold to `r 100*pooled_error["V1V9","precluster"]`%.
-With this error rate, we would expect `r round(100*dbinom(0, 1500, pooled_error["V1V9","precluster"]), 2)`% of the sequences to have no sequencing errors.
-Of course, the tradeoff is the loss of resolution among closely related sequences.
+We can estimate the distribution of errors assuming that the errors follow a binomial distribution along the length of the 1,500 nt gene with the error rate that we achieved from the V1-V9 mock community data prior to pre-clustering the sequences, which was `r round(100*unique,2)`% (Figure 5).
+Under these conditions one would expect `r round(100*dbinom(0, 1500, unique), 2)`% of the sequences to have no errors and 50% of the sequences would have at least `r qbinom(0.50, 1500, unique)` errors.
+After applying the pre-clustering denoising step, the error rate drops by `r round(unique_to_precluster, 1)`-fold to `r 100*precluster`%.
+With this error rate, we would expect `r round(100*dbinom(0, 1500, precluster), 2)`% of the sequences to have no sequencing errors.
+The cost of the reduced error rate is the loss of resolution among closely related sequences.
 

### 2016-02-03 code/get_error_profile.R
Fix variable name to reflect the predicted rate rather than quality score

@@ -89,7 +89,7 @@ json <- toJSON(
 		ins_bias=insertion_bias,
 		percent_over_93 = c(m=m_93, s=s_93, i=i_93),
 		qual_range = range(matches, subs, ins),
-		qual_error_corr = cor(mock_error_good$error, mock_error_good$pred_error)
+		pred_error_corr = cor(mock_error_good$error, mock_error_good$pred_error)
 	)
 )
 

### 2016-02-03 Schloss_PacBio16S_PeerJ_2016.Rmd
Build Figure 1

@@
 Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(error_profile$percent_over_93$m, digits=3, nsmall=1)`%), substitutions (`r format(error_profile$percent_over_93$s, digits=3, nsmall=1)`%), and insertions (`r format(error_profile$percent_over_93$i, digits=3, nsmall=1)`%).
-Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(error_profile$qual_error_corr, digits=2, nsmall=2)`; Figure 1A).
+Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(error_profile$pred_error_corr, digits=2, nsmall=2)`; Figure 1A).
 
@@
 
-[revise - Fig 1A plot predicted error rate vs observed error rate]
 **Figure 1. Summary of errors in data generated using PacBio sequencing platform
-to sequence various regions within the 16S rRNA gene.** Quality scores varied
-with error types (A). The sequencing error rate of the amplified gene fragments
+to sequence various regions within the 16S rRNA gene.** The predicted error rate using PacBio's sequence analysis algorithm correlated well with the observed error rate (Pearson's R: `r format(error_profile$pred_error_corr, digits=2, nsmall=2)`; A).
+Because of the large number of sequences, we randomly selected 5% of the data to show in panel A.
+The sequencing error rate of the amplified gene fragments
 increased with mismatches to the barcodes and primers (B). The sequencing error
-rate declined with increasd sequencing coverage; however, increasing the
+rate declined with increased sequencing coverage; however, increasing the
 sequencing depth beyond 10-fold coverage had no meaningful effect on the
@@
 reference provided species-level information.
-
-### Figure 1
-```{r Figure_1, fig.width=6.5, fig.height=3.0}
-layout(matrix(c(1,2,3), nrow=1), heights=1, widths=c(0.75,1,1))
-
-#Panel A depicting range of quality scores for different errors
-par(mar=c(6, 5, 0.5, 0.5))
-plot(1, xlim=c(0.5, 4.5), ylim=c(0,75), type="n", yaxt="n", xaxt="n", xlab="", ylab="")
-polygon(x=c(0.75, 1.25, 1.25, 0.75), y=c(m.quant["25%"], m.quant["25%"],
-											 m.quant["75%"], m.quant["75%"]))
-polygon(x=c(1.75, 2.25, 2.25, 1.75), y=c(s.quant["25%"], s.quant["25%"],
-											 s.quant["75%"], s.quant["75%"]))
-polygon(x=c(2.75, 3.25, 3.25, 2.75), y=c(i.quant["25%"], i.quant["25%"],
-											 i.quant["75%"], i.quant["75%"]))
-polygon(x=c(3.75, 4.25, 4.25, 3.75), y=c(a.quant["25%"], a.quant["25%"],
-											 a.quant["75%"], a.quant["75%"]))
-
-segments(x0=seq(0.75, 3.75, 1), x1=seq(1.25,4.25,1), y0=c(m.quant["50%"],
-											s.quant["50%"], i.quant["50%"], a.quant["50%"]), lwd=4)
-
-arrows(x0=1:4, y0=c(m.quant["75%"], s.quant["75%"], i.quant["75%"],
-					a.quant["75%"]), y1=c(m.quant["97.5%"], s.quant["97.5%"],
-					i.quant["97.5%"], a.quant["97.5%"]), angle=90, length=0.05)
-
-arrows(x0=1:4, y0=c(m.quant["25%"], s.quant["25%"], i.quant["25%"],
-					a.quant["25%"]), y1=c(m.quant["2.5%"], s.quant["2.5%"],
-					i.quant["2.5%"], a.quant["2.5%"]), angle=90, length=0.05)
-
-axis(1, at=c(1,2,3,4), label=c("Matches", "Substitutions", "Insertions", "Ambiguous"), las=2, cex.axis=1)
-axis(2, las=2, at=c(0,25,50,75), cex.axis=1.5)
-title(ylab="Quality score", cex.lab=1.5)
-mtext(side=2, line=2.5, at=75, text="A", cex=2, font=2, las=2)
-
-#Panel B depicting relationship between errors in barcodes/primers and the
-#rest of the sequence.
-par(mar=c(5,5,0.5,0.5))
-plot(err.bcprimer$x[1:6,1]~err.bcprimer$Group.1[1:6], pch=19,
-		 ylab="Error rate (%)", xlab="", axes=F, ylim=c(0,5), cex.lab=1.5)
-mtext(side=1, line=3.75, text="Total mismatches to\nbarcodes and primers",
-			cex=0.9)
-axis(1, cex.axis=1.5)
-axis(2, las=2, cex.axis=1.5)
-box()
-mtext(side=2, line=2.5, at=5, text="B", cex=2, font=2, las=2)
-
-#Panel C depicting relationship between sequencing coverage and error rate for
-#the different regions that were sequence
-plotLines <- function(folder){
-	composite <- read.table(file=paste("pipeline_dev/", folder, "/", folder,
-								 ".composite", sep=""), header=T,row.names=1)
-
-	composite <- composite[composite$reason=="g",]
-	error.by.depth <- aggregate(composite$error, by=list(composite$freq),
-						function(x){c(mean=mean(x), quantile(x, prob=c(0.025, 0.975)), n=length(x))})
-
-	#exclude any coverage values where we don't have at least 50 observations
-	error.by.depth <- error.by.depth[error.by.depth$x[,"n"]>=50,]
-	fold.coverage <- error.by.depth[, "Group.1"]
-	error.by.depth <- error.by.depth$x
-
-	mean <- error.by.depth[,"mean"]
-	points(mean~fold.coverage, type="l", lwd=2, col=clrs[folder])
-}
-
-par(mar=c(5,2,0.5, 0.5))
-plot(1, xlim=c(2,30), ylim=c(0,0.05), type="n", xlab="", ylab="", axes=F)
-axis(2, at=seq(0,0.05, 0.01), label=FALSE, tick=T, las=2)
-#axis(2, at=seq(0,0.05, 0.01), label=seq(0,5, 1), las=2)
-axis(1, cex.axis=1.5)
-lapply(regions, plotLines)
-legend(x=18, y=0.05, legend=pretty_region[regions], lty=1, lwd=2, col=clrs[regions])
-mtext(side=2, line=0.5, at=0.05, text="C", cex=2, font=2, las=2)
-title(xlab="Coverage", ylab="", cex.lab=1.5)
-box()
-
-layout(1)
-```
 

### 2016-02-03 code/build_figure1.R

@@ -1,0 +1,84 @@
+regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
+pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")
+
+clrs <- rainbow(length(regions))
+names(clrs) <- regions
+
+
+
+mock_error <- read.table(file="data/process/mock.error.report")
+mock_error_nochim <- mock_error[mock_error$numparents==1,]
+mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.1,]
+mock_error_basic <- mock_error_good[mock_error_good$good_coords & mock_error_good$good_homop,]
+
+
+subset <- mock_error_basic[sample(1:nrow(mock_error_good), 0.05*nrow(mock_error_good)),]
+
+
+pdf(file="results/figures/figure_1.pdf", width=6.5, height=3.0)
+layout(matrix(c(1,2,3), nrow=1), heights=1, widths=c(1,1,1))
+
+#Panel A depicting relationship between predicted and observed error rates
+par(mar=c(5, 4, 0.5, 0.5))
+
+plot(subset$error, 1-subset$pred_error, pch=19, cex=0.15, xlim=c(0,0.1), ylim=c(0,0.1), xlab="", ylab="", axes=F)
+axis(1, at=seq(0,0.1,0.02), label=seq(0,10,2))
+axis(2, at=seq(0,0.1,0.02), label=seq(0,10,2), las=2)
+box()
+text(x=0.0050, y=0.0975, label="A", cex=2, font=2)
+mtext(side=2, line=2, text="Predicted Error (%)", cex=0.7)
+mtext(side=1, line=2.5, text="Observed Error (%)", cex=0.7)
+
+
+#Panel B depicting relationship between errors in barcodes/primers and the
+#rest of the sequence.
+par(mar=c(5,3,0.5,0))
+
+bcprimer <- apply(mock_error_basic[, c("fbdiffs", "rbdiffs", "fpdiffs", "rpdiffs")], 1, sum)
+error_bcprimer <- aggregate(mock_error_basic[,"error"], by=list(bcprimer), function(x){c(mean(x), quantile(x, probs=c(0.025, 0.975)))})
+
+plot(error_bcprimer$x[1:6,1]~error_bcprimer$Group.1[1:6], pch=19,
[…]

### 2016-02-03 code/get_error_rate_table.R

@@ -65,19 +65,6 @@ pb_nseqs_reduction_1 <- 100 * (1 - pb_nseqs["1",]/basic_rates["nseqs",])
 
 #coverage_analysis - need to get the error rate for coverage values between 3 and 40 for each region
 
-#error_by_coverage <- function(region){
-#	coverage_region <- mock_error_basic[mock_error_basic$region == region,]
-#	coverage_error_agg <- aggregate(100 * coverage_region$error, by=list(coverage_region$coverage), mean)
-#
-#	coverage_error <- rep(NA, max(mock_error_basic$coverage))
-#	coverage_error[coverage_error_agg$Group.1] <- coverage_error_agg[,"x"]
-#	coverage_error
-#}
-#
-#coverage_error <- sapply(regions, error_by_coverage)
-#seems to plateau right around 10
-
-
 error_at_below_coverage_cutoff <- function(region, cutoff=10){
 	region_data <- mock_error_basic[mock_error_basic$region == region,]
 	deep <- region_data$coverage >= cutoff

### 2016-02-03 Schloss_PacBio16S_PeerJ_2016.Rmd
Build Figure 2

@@
 
-### Figure 2
-```{r random_stuff }
-
-#This is for the Figure 2B plot...
-pb_overall <- aggregate(mock_error_basic$error, by=list(mock_error_basic$pbdiffs), function(x){c(mean=mean(100*x))})
-
-#This is for the Figure 2C plot...
-coverage_overall_agg <- aggregate(100 * mock_error_basic$error, by=list(mock_error_basic$coverage), mean)
-coverage_overall <- coverage_overall_agg$x
-names(coverage_overall) <- coverage_overall_agg$Group.1
-coverage_overall <- coverage_overall[as.character(3:40)]
-
-#This is for the Figure 2D plot...
-pred_overall <- aggregate(100 * mock_error_basic$error, by=list(round(mock_error_basic$pred_error, 3)), mean)
-```
-
-```{r Figure_2, width=6.5, height=6}
-error.fraction <- cbind(initial.error=init.rates, basic.error=basic.error,
-							bcprimer.error, coverage.error, aveq.error,
-							oligosCoverage.error, oligosQScore.error,
-							coverageQScore.error, allFilters.error, final.error)
-rownames(error.fraction) <- gsub("v(\\d*)", "\\1", rownames(error.fraction))
-
-error <- 100 * error.fraction[,grepl("error", colnames(error.fraction))]
-fraction <- 100 * error.fraction[,grepl("frac", colnames(error.fraction))]
-
-getJitter <- function(x){
-	jitter <- 0.15
-	runif(6, x-jitter, x+jitter)
-}
-
-l <- layout(matrix(c(1,2,3), nrow=3), heights=c(1,1,0.6))
-
-par(mar=c(0.5, 7, 0.5, 0.5))
-plot(1, type="n", xlim=c(1, 14), ylim=c(0,2.5), axes=F, xlab="", ylab="",
-		 cex.lab=1.5)
-text(x=getJitter(1.0), y=error[,1], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(2.0), y=error[,2], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(3.0), y=error[,3], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(4.0), y=error[,4], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(5.0), y=error[,5], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(6.0), y=error[,6], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(7.0), y=error[,7], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(8.0), y=error[,8], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(9.0), y=error[,9], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(10.0), y=error[,10], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(11.0), y=error[,11], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(12.0), y=error[,12], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(13.0), y=error[,13], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(14.0), y=error[,15], label=rownames(error.fraction), cex=1.25)
-text(x=10, y=2, label="*", cex=3, font=2)
-
-mtext(text="Error rate (%)", side=2, line=4, cex.lab=1.2)
-box()
-axis(2, las=2, at=seq(0,2.5, 0.5), label=format(seq(0, 2.5, 0.5),nsmall=1),
-		 cex.axis=1.5)
-
-abline(v=seq(3.5,13.5, 1), col="grey")
-abline(v=c(1.5, 2.5, 13.5), col="black", lwd=2)
-
-mtext(side=2, line=4.5, at=2.44, text="A", cex=2, font=2, las=1)
-
-par(mar=c(0.5, 7, 0.5, 0.5))
-plot(1, type="n", xlim=c(1, 14), ylim=c(0,100), axes=F,
-		 ylab="Reads remaining\nfrom basic criteria (%)", xlab="", cex.lab=1.5)
-text(x=getJitter(3.0), y=fraction[,1], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(4.0), y=fraction[,2], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(5.0), y=fraction[,3], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(6.0), y=fraction[,4], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(7.0), y=fraction[,5], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(8.0), y=fraction[,6], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(9.0), y=fraction[,7], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(10.0), y=fraction[,8], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(11.0), y=fraction[,9], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(12.0), y=fraction[,10], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(13.0), y=fraction[,11], label=rownames(error.fraction), cex=1.25)
-text(x=getJitter(14.0), y=fraction[,8], label=rownames(error.fraction), cex=1.25)
-
-box()
-abline(v=seq(3.5,13.5, 1), col="grey")
-abline(v=c(1.5, 2.5, 13.5), col="black", lwd=2)
-
-mtext(side=2, line=4.5, at=98, text="B", cex=2, font=2, las=1)
-axis(2, las=2, at=seq(0,100, 25), label=seq(0, 100, 25), cex.axis=1.5)
-text(1,0, label="NA")
-text(2,0, label="NA")
-
-plot(1, type="n", xlim=c(1, 14), ylim=c(0,1), axes=F, ylab="", xlab="")
-
-text(x=c(1:14)+0.2, y=rep(1, 11), label=c("Initial data", "Basic criteria",
-		"No Mismatches", "\u22641 Mismatch", "Coverage", "Quality Score",
-		"No Mismatches\n& Coverage", "\u22641 Mismatch\n& Coverage",
-		"No Mismatches\n& Quality", "\u22641 Mismatch\n& Quality",
-		"Coverage & Quality", "No Mismatches,\nCoverage & Quality",
-		"\u22641 Mismatch,\nCoverage & Quality", "After Pre-clustering"),
-		srt=90, pos=2, cex=1)
-
-layout(1)
-```
-
 ### Figure 3

### 2016-02-03 code/build_figure2.R

@@ -1,0 +1,72 @@
+error_summary <- read.table(file="data/process/error_summary.tsv", header=T)
+
+getJitter <- function(x){
+	jitter <- 0.15
+	runif(6, x-jitter, x+jitter)
+}
+
+
+methods <- c("raw", "basic", "pb", "pred", "coverage", "pb_coverage", "pb_pred",
+				"coverage_pred", "all", "precluster")
+
+
+
+cairo_pdf(file="results/figures/figure_2.pdf", width=6.5, height=6.0)
+
+l <- layout(matrix(c(1,2,3), nrow=3), heights=c(1,1,0.6))
+
+par(mar=c(0.5, 7, 0.5, 0.5))
+plot(1, type="n", xlim=c(1, length(methods)), ylim=c(0,1), axes=F, xlab="", ylab="")
+
+plotRegionMethod <- function(index, data){
+	method <- methods[index]
+	subset <- error_summary[error_summary$method==method,]
+	regions <- gsub("V", "", subset$region)
+	text(x=getJitter(index), y=subset[,data], label=regions, cex=1.25)
+}
+
+lapply(1:length(methods), plotRegionMethod, "error")
+text(x=4, y=0.8, label="*", cex=3, font=2)
+
+mtext(text="Error rate (%)", side=2, line=4, cex.lab=1.2)
+box()
+axis(2, las=2, at=seq(0,1, 0.2), label=format(seq(0, 1, 0.2),nsmall=1),
+		 cex.axis=1.5)
+
+abline(v=seq(3.5,9.5, 1), col="grey")
+abline(v=c(1.5, 2.5, 9.5), col="black", lwd=2)
+
+text(x=1, y=0.95, label="A", cex=2, font=2)
+
+par(mar=c(0.5, 7, 0.5, 0.5))
+plot(1, type="n", xlim=c(1, 10), ylim=c(0,100), axes=F,
+		 ylab="Reads remaining\nfrom basic criteria (%)", xlab="", cex.lab=1.5)
+
+lapply(1:length(methods), plotRegionMethod, "nseqs_red")
+
+
+box()
+abline(v=seq(3.5,9.5, 1), col="grey")
+abline(v=c(1.5, 2.5, 9.5), col="black", lwd=2)
+
+text(x=1, y=95, label="B", cex=2, font=2, las=1)
+axis(2, las=2, at=seq(0,100, 25), label=seq(0, 100, 25), cex.axis=1.5)
[…]

### 2016-02-03 code/build_figure3.R
Build Figure 3

@@ -1,0 +1,52 @@
+data <- read.table(file="data/process/taxonomy_depth_analysis.tsv", header=T)
+
+rdp <- 100 * cbind("Mock"=data[data$db=="pds" & data$dataset=="mock", "genus_species"],
+				 "Human"=data[data$db=="pds" & data$dataset=="human", "genus_species"],
+				 "Mouse"=data[data$db=="pds" & data$dataset=="mouse", "genus_species"],
+				 "Soil"=data[data$db=="pds" & data$dataset=="soil", "genus_species"])
+rownames(rdp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+
+gg <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "genus_species"],
+				"Human"=data[data$db=="gg" & data$dataset=="human", "genus_species"],
+				"Mouse"=data[data$db=="gg" & data$dataset=="mouse", "genus_species"],
+				"Soil"=data[data$db=="gg" & data$dataset=="soil", "genus_species"])
+rownames(gg) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+
+silva <- 100 * cbind("Mock"=data[data$db=="nr" & data$dataset=="mock", "genus_species"],
+					 "Human"=data[data$db=="nr" & data$dataset=="human", "genus_species"],
+					 "Mouse"=data[data$db=="nr" & data$dataset=="mouse", "genus_species"],
+					 "Soil"=data[data$db=="nr" & data$dataset=="soil", "genus_species"])
+rownames(silva) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+
+gg.sp <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "X7"],
+					 "Human"=data[data$db=="gg" & data$dataset=="human", "X7"],
+					 "Mouse"=data[data$db=="gg" & data$dataset=="mouse", "X7"],
+					 "Soil"=data[data$db=="gg" & data$dataset=="soil", "X7"])
+rownames(gg.sp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
+
+
+
+pdf(file="results/figures/figure_3.pdf", width=5, height=7)
+
+
[…]

### 2016-02-03 Schloss_PacBio16S_PeerJ_2016.Rmd
Build Figure 4

@@
 
-### Figure 3
-
-```{r, Figure_4, fig.width=7, fig.height=7}
-rdp <- 100 * cbind("Mock"=data[data$db=="pds" & data$dataset=="mock", "genus_species"],
-				 "Human"=data[data$db=="pds" & data$dataset=="human", "genus_species"],
-				 "Mouse"=data[data$db=="pds" & data$dataset=="mouse", "genus_species"],
-				 "Soil"=data[data$db=="pds" & data$dataset=="soil", "genus_species"])
-rownames(rdp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
-
-gg <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "genus_species"],
-				"Human"=data[data$db=="gg" & data$dataset=="human", "genus_species"],
-				"Mouse"=data[data$db=="gg" & data$dataset=="mouse", "genus_species"],
-				"Soil"=data[data$db=="gg" & data$dataset=="soil", "genus_species"])
-rownames(gg) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
-
-silva <- 100 * cbind("Mock"=data[data$db=="nr" & data$dataset=="mock", "genus_species"],
-					 "Human"=data[data$db=="nr" & data$dataset=="human", "genus_species"],
-					 "Mouse"=data[data$db=="nr" & data$dataset=="mouse", "genus_species"],
-					 "Soil"=data[data$db=="nr" & data$dataset=="soil", "genus_species"])
-rownames(silva) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
-
-gg.sp <- 100 * cbind("Mock"=data[data$db=="gg" & data$dataset=="mock", "X7"],
-					 "Human"=data[data$db=="gg" & data$dataset=="human", "X7"],
-					 "Mouse"=data[data$db=="gg" & data$dataset=="mouse", "X7"],
-					 "Soil"=data[data$db=="gg" & data$dataset=="soil", "X7"])
-rownames(gg.sp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
-
-par(mar=c(5, 5, 0.5, 1))
-dotchart(gg, xlim=c(0,100), col="black", xlab="Unique reads that classified\nto genus or species level (%)", pch=19)
-
-points(x=rdp[,"Mock"], y=25:30, pch=15, col="black")
-points(x=rdp[,"Human"], y=17:22, pch=15, col="black")
-points(x=rdp[,"Mouse"], y=9:14, pch=15, col="black")
-points(x=rdp[,"Soil"], y=1:6, pch=15, col="black")
-
-points(x=silva[,"Mock"], y=25:30, pch=17, col="black")
-points(x=silva[,"Human"], y=17:22, pch=17, col="black")
-points(x=silva[,"Mouse"], y=9:14, pch=17, col="black")
-points(x=silva[,"Soil"], y=1:6, pch=17, col="black")
-
-points(x=gg.sp[,"Mock"], y=25:30, pch=21, bg="gray")
-points(x=gg.sp[,"Human"], y=17:22, pch=21, bg="gray")
-points(x=gg.sp[,"Mouse"], y=9:14, pch=21, bg="gray")
-points(x=gg.sp[,"Soil"], y=1:6, pch=21, bg="gray")
-
-legend(x=63, y=12, legend=c("RDP (gen.)", "SILVA (gen.)", "greengenes (gen.+sp.)", "greengenes (sp.)"), pch=c(15, 17, 19, 21), col="black", pt.bg=c("black", "black", "black", "gray"), bg="white")
-```
-
-#plot one_offs
-```{r}
-one_offs <- read.table(file="data/process/non_random_analysis.tsv", header=T)
-
-regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
-
-z <- prop.table(as.matrix(one_offs), 2)[1:10,]
-plot(NA, xlim=c(1,10), ylim=c(0,1), xlab="", ylab="", axes=F)
-for(r in regions){
-	points(1:10, z[,r], type="l")
-}
-axis(1, at=1:10, label=1:10)
-axis(2)
-```
+**Figure 4.
+Percentage of 1-nt variants that occured up to ten times.** Sequences that were 1 nt different from the mock community reference sequences were counted to determine the number of times each variant appeared by region within the 16S rRNA gene.
 

### 2016-02-03 code/build_figure1.R

@@ -1,5 +1,5 @@
-regions <- c("V1V3", "V1V5", "V1V6", "V1V9", "V3V5", "V4")
-pretty_region <- c("V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9", "V3V5"="V3-V5", "V4"="V4")
+pretty_region <- c("V4"="V4", "V3V5"="V3-V5", "V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9")
+regions <- names(pretty_region)
 
 clrs <- rainbow(length(regions))
 names(clrs) <- regions

### 2016-02-03 code/build_figure4.R

@@ -1,0 +1,21 @@
+pretty_region <- c("V4"="V4", "V3V5"="V3-V5", "V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9")
+regions <- names(pretty_region)
+
+clrs <- rainbow(length(regions))
+names(clrs) <- regions
+
+#plot one_offs
+one_offs <- read.table(file="data/process/non_random_analysis.tsv", header=T)
+z <- prop.table(as.matrix(one_offs), 2)[1:10,]
+
+pdf(file="results/figures/figure_4.pdf", width=4, height=4)
+par(mar=c(4,4,0.5,0.5))
+plot(NA, xlim=c(1,10), ylim=c(0,1), xlab="Number of times 1-nt\nvariants were observed", ylab="Percentage of 1-nt variants", axes=F)
+for(r in regions){
+	points(1:10, z[,r], type="l", col=clrs[r], lwd=3)
+}
+axis(1, at=1:10, label=1:10)
+axis(2, at=seq(0,1,0.2), label=seq(0, 100, 20), las=2)
+box()
+legend(x=6, y=0.9, legend=pretty_region, lwd=3, col=clrs, cex=0.8)
+dev.off()

### 2016-02-03 Schloss_PacBio16S_PeerJ_2016.Rmd
Build Table 1

@@
 
-## Tables
-
-**Table 1. Summary of the primer pairs used to generate the 16S rRNA gene fragment fragments and the characteristics of each region.**
-
-```{r Table_1, warning=TRUE, results='asis'}
-
-getSloppyRevComp <- function(sequence){
-	rev.sequence <- paste(rev(unlist(strsplit(sequence, ""))), collapse="")
-	revcomp.sequence <- chartr("ATGCRYSWKMBDHVN", "TACG...........", rev.sequence)
-	return(revcomp.sequence)
-}
-
-getPos <- function(primer, sequence=ecoli, forward=TRUE){
-	if(forward==FALSE){
-		primer <- getSloppyRevComp(primer)
-	}
-	primer <- gsub("[^ATGC]", ".", primer)
-
-	pos <- regexpr(primer, sequence)
-
-	if(forward==FALSE){
-		pos <- pos + nchar(primer) -1
-	}
-	return(pos)
-
-}
-
-#taken from... http://www.ncbi.nlm.nih.gov/nuccore/NC_022648.1?report=fasta&from=4564120&to=4565661
-ecoli <- "AAATTGAAGAGTTTGATCATGGCTCAGATTGAACGCTGGCGGCAGGCCTAACACATGCAAGTCGAACGGTAACAGGAAGCAGCTTGCTGCTTTGCTGACGAGTGGCGGACGGGTGAGTAATGTCTGGGAAACTGCCTGATGGAGGGGGATAACTACTGGAAACGGTAGCTAATACCGCATAACGTCGCAAGACCAAAGAGGGGGACCTTCGGGCCTCTTGCCATCGGATGTGCCCAGATGGGATTAGCTAGTAGGTGGGGTAACGGCTCACCTAGGCGACGATCCCTAGCTGGTCTGAGAGGATGACCAGCCACACTGGAACTGAGACACGGTCCAGACTCCTACGGGAGGCAGCAGTGGGGAATATTGCACAATGGGCGCAAGCCTGATGCAGCCATGCCGCGTGTATGAAGAAGGCCTTCGGGTTGTAAAGTACTTTCAGCGGGGAGGAAGGGAGTAAAGTTAATACCTTTGCTCATTGACGTTACCCGCAGAAGAAGCACCGGCTAACTCCGTGCCAGCAGCCGCGGTAATACGGAGGGTGCAAGCGTTAATCGGAATTACTGGGCGTAAAGCGCACGCAGGCGGTTTGTTAAGTCAGATGTGAAATCCCCGGGCTCAACCTGGGAACTGCATCTGATACTGGCAAGCTTGAGTCTCGTAGAGGGGGGTAGAATTCCAGGTGTAGCGGTGAAATGCGTAGAGATCTGGAGGAATACCGGTGGCGAAGGCGGCCCCCTGGACGAAGACTGACGCTCAGGTGCGAAAGCGTGGGGAGCAAACAGGATTAGATACCCTGGTAGTCCACGCCGTAAACGATGTCGACTTGGAGGTTGTGCCCTTGAGGCGTGGCTTCCGGAGCTAACGCGTTAAGTCGACCGCCTGGGGAGTACGGCCGCAAGGTTAAAACTCAAATGAATTGACGGGGGCCCGCACAAGCGGTGGAGCATGTGGTTTAATTCGATGCAACGCGAAGAACCTTACCTGGTCTTGACATCCACGGAAGTTTTCAGAGATGAGAATGTGCCTTCGGGAACCGTGAGACAGGTGCTGCATGGCTGTCGTCAGCTCGTGTTGTGAAATGTTGGGTTAAGTCCCGCAACGAGCGCAACCCTTATCCTTTGTTGCCAGCGGTCCGGCCGGGAACTCAAAGGAGACTGCCAGTGATAAACTGGAGGAAGGTGGGGATGACGTCAAGTCATCATGGCCCTTACGACCAGGGCTACACACGTGCTACAATGGCGCATACAAAGAGAAGCGACCTCGCGAGAGCAAGCGGACCTCATAAAGTGCGTCGTAGTCCGGATTGGAGTCTGCAACTCGACTCCATGAAGTCGGAATCGCTAGTAATCGTGGATCAGAATGCCACGGTGAATACGTTCCCGGGCCTTGTACACACCGCCCGTCACACCATGGGAGTGGGTTGCAAAAGAAGTAGGTAGCTTAACCTTCGGGAGGGCGCTTACCACTTTGTGATTCATGACTGGGGTGAAGTCGTAACAAGGTAACCGTAGGGGAACCTGCGGTTGGATCACCTCCTTA"
-
-oligos <- read.table(file="data/references/pacbio.oligos", stringsAsFactors=F)[1:6,]
-rownames(oligos) <- pretty_region[oligos$V4]
-oligos <- oligos[,c("V2", "V3")]
-colnames(oligos) <- c("Forward", "Reverse")
-
-start <- sapply(oligos[,1], getPos)
-end <- sapply(oligos[,2], getPos, forward=FALSE)
-
-coordinates <- paste(start, end, sep="-")
-length <- end - start - nchar(oligos[,1]) - nchar(oligos[,2]) + 1
-t1 <- cbind(oligos, coordinates, length)
-knitr::kable(t1, digits=1, row.names=TRUE, col.names=c("Forward", "Reverse", "*E. coli* coordinates^a^", "Length (bp)^b^"), align=c("l", "l", "c", "c"))
-```
-
-^a^	The coordinates where the start and end of the forward and reverse primers anneal, respectively.
-
-^b^	The number of bases between the primers.
-
-**Table 2. Summary of the reasons that sequences were excluded because of the basic sequence curation steps**
-
-```{r Table_2, warning=TRUE, results='asis'}
-rownames(reasons.lost) <- pretty_region[rownames(reasons.lost)]
-reasons.lost <- cbind(orig.nseqs, 100*reasons.lost, basic.nseqs)
-knitr::kable(reasons.lost, digits=1, row.names=TRUE,
-						 col.names=c("Initial sequences (N)", "Good reads (%)",
-						 						"Wrong start/end position (%)",
-						 						"Excessively long homopolymers (%)",
-						 						"Ambiguous base calls (%)",
-						 						"Sequences remaining (N\\)"),
-						 align="c")
-```
-
 ## Figures

### 2016-02-03 results/tables/build_table1.Rmd
Build Table 1

---
output:
  pdf_document:
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=1.0in
---

\captionsetup{labelformat=empty}

```{r table1, results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
require(xtable, quietly=TRUE)
options(xtable.sanitize.text.function=identity)
options(xtable.sanitize.colnames.function=identity)

pretty_region <- c("V4"="V4", "V1V3"="V1-V3", "V3V5"="V3-V5", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9")
regions <- names(pretty_region)

mock_error <- read.table(file="data/process/mock.error.report")
mock_error_nochim <- mock_error[mock_error$numparents==1,]
mock_error_good <- mock_error_nochim[mock_error_nochim$error <= 0.1,]
n_seqs <- aggregate(mock_error_good$error, by=list(mock_error_good$region), length)
rownames(n_seqs) <- n_seqs$Group.1
n_seqs <- n_seqs[regions,]

getSloppyRevComp <- function(sequence){
	rev.sequence <- paste(rev(unlist(strsplit(sequence, ""))), collapse="")
	revcomp.sequence <- chartr("ATGCRYSWKMBDHVN", "TACG...........", rev.sequence)
	return(revcomp.sequence)
}

getPos <- function(primer, sequence=ecoli, forward=TRUE){
	if(forward==FALSE){
		primer <- getSloppyRevComp(primer)
	}
	primer <- gsub("[^ATGC]", ".", primer)

	pos <- regexpr(primer, sequence)

	if(forward==FALSE){
		pos <- pos + nchar(primer) -1
	}
	return(pos)

}

#taken from... http://www.ncbi.nlm.nih.gov/nuccore/NC_022648.1?report=fasta&from=4564120&to=4565661
ecoli <- "AAATTGAAGAGTTTGATCATGGCTCAGATTGAACGCTGGCGGCAGGCCTAACACATGCAAGTCGAACGGTAACAGGAAGCAGCTTGCTGCTTTGCTGACGAGTGGCGGACGGGTGAGTAATGTCTGGGAAACTGCCTGATGGAGGGGGATAACTACTGGAAACGGTAGCTAATACCGCATAACGTCGCAAGACCAAAGAGGGGGACCTTCGGGCCTCTTGCCATCGGATGTGCCCAGATGGGATTAGCTAGTAGGTGGGGTAACGGCTCACCTAGGCGACGATCCCTAGCTGGTCTGAGAGGATGACCAGCCACACTGGAACTGAGACACGGTCCAGACTCCTACGGGAGGCAGCAGTGGGGAATATTGCACAATGGGCGCAAGCCTGATGCAGCCATGCCGCGTGTATGAAGAAGGCCTTCGGGTTGTAAAGTACTTTCAGCGGGGAGGAAGGGAGTAAAGTTAATACCTTTGCTCATTGACGTTACCCGCAGAAGAAGCACCGGCTAACTCCGTGCCAGCAGCCGCGGTAATACGGAGGGTGCAAGCGTTAATCGGAATTACTGGGCGTAAAGCGCACGCAGGCGGTTTGTTAAGTCAGATGTGAAATCCCCGGGCTCAACCTGGGAACTGCATCTGATACTGGCAAGCTTGAGTCTCGTAGAGGGGGGTAGAATTCCAGGTGTAGCGGTGAAATGCGTAGAGATCTGGAGGAATACCGGTGGCGAAGGCGGCCCCCTGGACGAAGACTGACGCTCAGGTGCGAAAGCGTGGGGAGCAAACAGGATTAGATACCCTGGTAGTCCACGCCGTAAACGATGTCGACTTGGAGGTTGTGCCCTTGAGGCGTGGCTTCCGGAGCTAACGCGTTAAGTCGACCGCCTGGGGAGTACGGCCGCAAGGTTAAAACTCAAATGAATTGACGGGGGCCCGCACAAGCGGTGGAGCATGTGGTTTAATTCGATGCAACGCGAAGAACCTTACCTGGTCTTGACATCCACGGAAGTTTTCAGAGATGAGAATGTGCCTTCGGGAACCGTGAGACAGGTGCTGCATGGCTGTCGTCAGCTCGTGTTGTGAAATGTTGGGTTAAGTCCCGCAACGAGCGCAACCCTTATCCTTTGTTGCCAGCGGTCCGGCCGGGAACTCAAAGGAGACTGCCAGTGATAAACTGGAGGAAGGTGGGGATGACGTCAAGTCATCATGGCCCTTACGACCAGGGCTACACACGTGCTACAATGGCGCATACAAAGAGAAGCGACCTCGCGAGAGCAAGCGGACCTCATAAAGTGCGTCGTAGTCCGGATTGGAGTCTGCAACTCGACTCCATGAAGTCGGAATCGCTAGTAATCGTGGATCAGAATGCCACGGTGAATACGTTCCCGGGCCTTGTACACACCGCCCGTCACACCATGGGAGTGGGTTGCAAAAGAAGTAGGTAGCTTAACCTTCGGGAGGGCGCTTACCACTTTGTGATTCATGACTGGGGTGAAGTCGTAACAAGGTAACCGTAGGGGAACCTGCGGTTGGATCACCTCCTTA"

oligos <- read.table(file="data/references/pacbio.oligos", stringsAsFactors=F)[1:6,]
rownames(oligos) <- oligos$V4
oligos <- oligos[,c("V2", "V3")]
colnames(oligos) <- c("forward", "reverse")
oligos <- oligos[regions,]

start <- sapply(oligos[,1], getPos)
end <- sapply(oligos[,2], getPos, forward=FALSE)

coordinates <- paste(start, end, sep="-")
length <- end - start - nchar(oligos[,1]) - nchar(oligos[,2]) + 1
composite <- cbind(region = pretty_region, oligos, coorinates=coordinates, length=format(length, digits=0), nseqs=format(n_seqs$x, big.mark=','))

header <- c("Region", "\\bigcell{c}{ Forward }", "\\bigcell{c}{ Reverse }", "\\bigcell{c}{\\textit{E. coli} \\\\ Coordinates}", "\\bigcell{c}{Amplicon \\\\ Length}", "\\bigcell{c}{Sequences \\\\ (N)}")

header <- paste0("\\textbf{", header, "}")

addtorow <- list(pos=list(0),
                command=paste0(paste(header, collapse=" & "), "\\\\"))

composite <- xtable(composite)

caption <- "\\textbf{Table 1. Summary of the primer pairs used to generate the 16S rRNA gene fragment fragments and the characteristics of each region.}"

composite <- xtable(composite, caption = caption, label="")

align(composite) <- "ccllccc"

print(composite,
    add.to.row=addtorow,
	include.rownames=FALSE,
	include.colnames=FALSE,
	hline.after = c(-1, 0, nrow(composite)),
    comment=FALSE,
    caption.placement = "top",
    table.placement="!ht",
	size="small"
)
```

### 2016-02-04 Schloss_PacBio16S_PeerJ_2016.Rmd
Insert bibtex references and get to render w/o errors

@@
 sequencing of the 16S rRNA gene has largely been replaced by various platforms
-produced by 454/Roche {e.g. Sogin, 2006 #1510}, Illumina
-{e.g. Gloor, 2010 #2611}, and IonTorrent {e.g. Junemann, 2012 #2630}. It was
+produced by 454/Roche (e.g. @Sogin2006), Illumina
+(e.g. @Gloor2010), and IonTorrent (e.g. @Junemann2012). It was
 once common to sequence fewer than 100 16S rRNA gene sequences from several
-samples using the Sanger approach {e.g. McCaig, 1999 #592}. Now it is common to
+samples using the Sanger approach (e.g. @Mccaig1999). Now it is common to
 generate thousands of sequences from each of several hundred samples
-{The Human Microbiome Consortium, 2012 #2617}. The advance in throughput has
+[@HMP2012]. The advance in throughput has
 come at the cost of read length. Sanger sequencing regularly generated 800 nt
@@
 it has become imperative to develop stringent sequence curation and denoising
-algorithms {Kozich, 2013 #2719; Schloss, 2011 #2466}. There has been a tradeoff
+algorithms [@Kozich2013; @Schloss2011a]. There has been a tradeoff
 between read length, number of reads per sample, and the error rate. For
 instance, we recently demonstrated that using the Illumina MiSeq and the 454
-Titanium platforms the raw error rate varies between 1 and 2% {Kozich, 2013
-#2719; Schloss, 2011 #2466}. Yet, it was possible to obtain error rates below
+Titanium platforms the raw error rate varies between 1 and 2%
+[@Kozich2013; @Schloss2011a]. Yet, it was possible to obtain error rates below
 0.02% by adopting various denoising algorithms. However, the resulting fragments
@@
 Inadequate denoising of sequencing reads can have many negative effects
-including limited ability to identify chimeras {Haas, 2011 #2404;Edgar, 2011 #2406}
-and inflation of alpha- and beta-diversity metrics
-{Huse, 2010 #2034; Kunin, 2010 #2259; Kozich, 2013 #2719; Schloss, 2011 #2466}.
+including limited ability to identify chimeras [@Haas2011; @Edgar2011]
+and inflation of alpha- and beta-diversity metrics [@Huse2010; @Kunin2010; @Kozich2013; @Schloss2011a].
 Although MiSeq and 454 enjoy widespread use in the field, the MiSeq
@@
 sensitivity of classification against reference databases
-{Liu, 2008 #2188;Werner, 2012 #2474;Wang, 2007 #1809}. There is
-clearly a need to develop sequencing technologies that will allow researchers to
-generate high quality full-length 16S rRNA gene sequences in a high throughput
-manner.
+[@Liu2008; @Werner2011; @Wang2007]. There is clearly a need to develop
+sequencing technologies that will allow researchers to generate high quality
+full-length 16S rRNA gene sequences in a high throughput manner.
 
@@
 DNA Sequencing System has received limited application in the microbial ecology
-research domain {Mosher, 2014 #3376; Fichot, 2013 #3453; Mosher, 2013 #3454}.
+research domain @Mosher2013; @Mosher2014; @Fichot2013].
 The SMRT system ligates hairpin adapters (i.e. SMRTbells) to the ends of
@@
 circularized allowing the sequencing polymerase to process around the molecule
-multiple times {Au, 2012 #3455}. According to Pacific Biosciences the platform
+multiple times [@Au2012]. According to Pacific Biosciences the platform
 is able to generate median read lengths longer than 8 kb with the P4-C2
@@
 sequences, the Pacific Biosciences platform has not been widely adopted for
-sequencing 16S rRNA genes {Fichot, 2013 #3453; Mosher, 2014 #3376; Mosher, 2014 #3376}.
+sequencing 16S rRNA genes [@Fichot2013; @Mosher2014; @Mosher2013].
 Previous studies utilizing the technology have removed reads with mismatched
-primers and barcodes, ambiguous base  calls, and low quality scores {Fichot, 2013 #3453}.
+primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013].
 Others have utilized the platform without describing the bioinformatic pipeline
-that was utilized {Mosher, 2014 #3376; Mosher, 2014 #3376}. Regardless of the
+that was utilized [@Mosher2014; @Mosher2013]. Regardless of the
 curation methods, the error rates associated with sequencing the 16S rRNA gene
@@
 These same DNA extracts were previously used to develop an Illumina MiSeq-based
-sequencing strategy {Kozich, 2013 #2719}.  Briefly, we used a “Mock Community”
+sequencing strategy [@Kozich2013].  Briefly, we used a “Mock Community”
 composed of genomic DNA from 21 bacterial strains: *Acinetobacter baumannii*
@@
 within a single sequencing run. Methods describing PCR, amplicon cleanup, and
-pooling were described previously {Kozich, 2013 #2719}. The SMRTbell adapters
+pooling were described previously [@Kozich2013]. The SMRTbell adapters
 were ligated onto  the PCR products and the libraries were sequenced by Pacific
@@
 ***Data analysis.***  [ Revise ]
-All sequencing data were curated using mothur {Schloss, 2009 #1816} and analyzed
-using the R programming language {R Core Team, 2014 #3456}. The raw data can be
+All sequencing data were curated using mothur [@Schloss2009a] and analyzed
+using the R programming language [@R2016]. The raw data can be
 obtained from the Sequence Read Archive at NCBI under accession SRP051686, which
@@
 Identification of *in silico* chimeras and calculation of sequencing error rates
-was performed using the seq.error command in mothur {Schloss, 2011 #2466}. *De
+was performed using the seq.error command in mothur [@Schloss2011a]. *De
 novo* chimera detection was also performed on the mock and other sequence data
-using the abundance-based algorithm implemented in UCHIME {Edgar, 2011 #2406}.
+using the abundance-based algorithm implemented in UCHIME [@Edgar2011].
 Sequences sequences were aligned against a SILVA-based reference alignment
-{Pruesse, 2007 #1735} using a profile-based aligner {Schloss, 2009 #1825} and
-were classified against the SILVA {Pruesse, 2007 #1735}, RDP {Cole, 2014 #3460},
-and greengenes {Werner, 2012 #2474} reference taxonomies  using a naive Bayesian
-classifier {Wang, 2007 #1809}. Sequences were assigned to operational taxonomic
+[@Pruesse2007] using a profile-based aligner [@Schloss2009b] and
+were classified against the SILVA [@Pruesse2007], RDP [@Cole2013],
+and greengenes [@Werner2011] reference taxonomies  using a naive Bayesian
+classifier [@Wang2007]. Sequences were assigned to operational taxonomic
 units using the average neighbor clustering algorithm with a 3% distance
-threshold {Schloss, 2011 #3156}.  Detailed methods including this paper as an R
+threshold [@Schloss2011b].  Detailed methods including this paper as an R
 markdown file are available as a public online repository
-(http://github.com/SchlossLab/ Schloss_PacBio16S_PeerJ_2015).
+(http://github.com/SchlossLab/Schloss_PacBio16S_PeerJ_2016).
 
@@
 The substitution errors were equally likely and all four bases were equally likely to cause insertion errors.
-Interestingly, guanines (`r format(error_profile$del_bias$dG, digits=3, nsmall=1)`%) and adenines (`r format(error_profile$del_bias$dA, digits=3, nsmall=1)`%) were more likely to be deleted than cytosines (`r error_profile$del_bias$dC, digits=3, nsmall=1)`%) or thymidines (`r error_profile$del_bias$dT, digits=2, nsmall=1)`%).
+Interestingly, guanines (`r format(error_profile$del_bias$dG, digits=3, nsmall=1)`%) and adenines (`r format(error_profile$del_bias$dA, digits=3, nsmall=1)`%) were more likely to be deleted than cytosines (`r format(error_profile$del_bias$dC, digits=3, nsmall=1)`%) or thymidines (`r format(error_profile$del_bias$dT, digits=2, nsmall=1)`%).
 The PacBio quality values varied between `r paste(error_profile$qual_range, collapse=' and ')`.
@@
 ***Pre-clustering sequences to further reduce sequencing noise.***
-Previously, we implemented a pre-clustering algorithm where sequences were sorted by their abundance in decreasing order and rare sequences are clustered with a more abundant sequence if the rare sequences have fewer mismatches than a defined threshold when compared to the more abundant sequence [@refs].
+Previously, we implemented a pre-clustering algorithm where sequences were sorted by their abundance in decreasing order and rare sequences are clustered with a more abundant sequence if the rare sequences have fewer mismatches than a defined threshold when compared to the more abundant sequence [@Huse2010; @Schloss2011a].
 The recommended threshold was a 1-nt difference per 100-nt of sequence data.
@@
 The full-length (i.e.
-V1-V9) fragments had an error rate of `r round(error_summary[error_summary$method == "precluster" & error_summary$region=="V1V9","error"], digits=2)`% (Figure 2B), this is similar to what we have previously observed using the 454 and Illumina MiSeq platforms (0.02%){Schloss, 2011 #2466; Kozich, 2013 #2719}
+V1-V9) fragments had an error rate of `r round(error_summary[error_summary$method == "precluster" & error_summary$region=="V1V9","error"], digits=2)`% (Figure 2B), this is similar to what we have previously observed using the 454 and Illumina MiSeq platforms (0.02%)[@Schloss2011a; @Kozich2013].
 
@@
 sobs_data <- read.table(file="data/process/sobs_table.tsv", header=T)
+perfect_otu <- 19
 ```
@@
 ***Effects of error rates on OTU assignments.***
-The sequencing error rate is known to affect the number of OTUs that are observed {Schloss, 2011 #2466}.
-For each region, we determined that if there were no chimeras or PCR or sequencing errors, then we would expect to find `r perfect.otu[[1]]` OTUs.
-When achieved perfect chimera removal, but allowed for PCR and sequencing errors, we observed between `r diff<-otu.table[,2]-perfect.otu; round(min(diff), digits=1)` (`r names(which.min(diff))`) and `r round(max(diff), digits=1)` (`r names(which.max(diff))`) extra OTUs.
-The range in the number of extra OTUs was largely explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, pooled_error[,"precluster"])$estimate, 2)`).
+The sequencing error rate is known to affect the number of OTUs that are observed [Schloss2011a].
+For each region, we determined that if there were no chimeras or PCR or sequencing errors, then we would expect to find `r perfect_otu` OTUs.
+When achieved perfect chimera removal, but allowed for PCR and sequencing errors, we observed between `r diff<-sobs_data[,2]-perfect_otu; round(min(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.min(diff),])]`) and `r round(max(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.max(diff),])]`) extra OTUs.
+The range in the number of extra OTUs was largely explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, error_summary[error_summary$method=="precluster", "error"])$estimate, 2)`).
 Next, we determined the number of OTUs that were observed when we used UCHIME to identify chimeric sequence.
-Under these more realistic conditions, we observed between `r diff<-otu.table[,3]-perfect.otu; round(min(diff), digits=1)` (`r names(which.min(diff))`) and `r round(max(diff), digits=1)` (`r names(which.max(diff))`) extra OTUs.
+Under these more realistic conditions, we observed between `r diff<-sobs_data[,3]-perfect_otu; round(min(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.min(diff),])]`) and `r round(max(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.max(diff),])]`) extra OTUs.
 Finally, we calculated the number of OTUs in the soil, mouse, and human samples using the same pipeline with chimera detection and removal based on the UCHIME algorithm.
-Surpisingly, there was not a clear relationship across sample type and region.
-Again, we found that there was a strong correlation between the number of observed OTUs and the error rate for the mouse (R=`r format(round(cor.test(otu.table[,"Mouse"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`) and human samples (R=`r format(round(cor.test(otu.table[,"Human"], pooled_error[,"precluster"])$estimate, 2), nsmall=2)`).
+Surprisingly, there was not a clear relationship across sample type and region.
+Again, we found that there was a strong correlation between the number of observed OTUs and the error rate for the mouse (R=`r format(round(cor.test(sobs_data[,"mouse"], error_summary[error_summary$method=="precluster", "error"])$estimate, 2), nsmall=2)`) and human samples (R=`r format(round(cor.test(sobs_data[,"human"], error_summary[error_summary$method=="precluster", "error"])$estimate, 2), nsmall=2)`).
 These results underscore the effect of sequencing error on the inflation of the number of observed OTUs.
@@
 Surprisingly, the same 1-nt error appeared `r max_one_off["V1V6"]` times (`r format(round(percent_one_off[max_one_off["V1V6"], "V1V6"], digits=2), nsmall=2)`%) in the V1-V6 mock data and another 1-nt error appeared `r max_one_off["V1V9"]` times (`r format(round(percent_one_off[max_one_off["V1V9"], "V1V9"], digits=2), nsmall=2)`%) in the V1-V9 mock data.
-Contrary to previous reports {Carneiro, 2012 #3458;Koren, 2012 #3459}, these results indicate that reproducible errors occur with the PacBio sequencing platform and that they can be quite abundant.
+Contrary to previous reports [@Carneiro2012; @Koren2012], these results indicate that reproducible errors occur with the PacBio sequencing platform and that they can be quite abundant.
 Through the use of the pre-clustering step described above these 1-nt errors would be ameliorated; however, this result indicates that caution should be used when attempting to use fine-scale OTU definitions.
@@
 The various sequencing platforms that are available to microbial ecologists are able to fill unique needs and have their own strengths and weaknesses.
-For sequencing the 16S rRNA gene, the 454 platform is able to generate a moderate number of high-quality 500-nt sequence fragments (error rates below 0.02%) {Schloss, 2011 #2466} and the MiSeq platform is able to generate a large number of high-quality 250-nt sequence fragments (error rates below 0.02%) {Kozich, 2013 #2719}.
+For sequencing the 16S rRNA gene, the 454 platform is able to generate a moderate number of high-quality 500-nt sequence fragments (error rates below 0.02%) [@Schloss2011a] and the MiSeq platform is able to generate a large number of high-quality 250-nt sequence fragments (error rates below 0.02%) [@Kozich2013].
 The promise of the PacBio sequencing platform was the generation of high-quality near full-length sequence fragments.
@@
 
+\newpage
+
 ## Figures
@@
 
+\newpage
+
 ## References

### 2016-02-04 code/build_figure1.R
Point tables and figures to submission/

@@ -15,7 +15,7 @@ mock_error_basic <- mock_error_good[mock_error_good$good_coords & mock_error_goo
 subset <- mock_error_basic[sample(1:nrow(mock_error_good), 0.05*nrow(mock_error_good)),]
 
 
-pdf(file="results/figures/figure_1.pdf", width=6.5, height=3.0)
+pdf(file="submission/figure_1.pdf", width=6.5, height=3.0)
 layout(matrix(c(1,2,3), nrow=1), heights=1, widths=c(1,1,1))
 
 #Panel A depicting relationship between predicted and observed error rates

### 2016-02-04 code/build_figure2.R

@@ -11,7 +11,7 @@ methods <- c("raw", "basic", "pb", "pred", "coverage", "pb_coverage", "pb_pred",
 
 
 
-cairo_pdf(file="results/figures/figure_2.pdf", width=6.5, height=6.0)
+cairo_pdf(file="submission/figure_2.pdf", width=6.5, height=6.0)
 
 l <- layout(matrix(c(1,2,3), nrow=3), heights=c(1,1,0.6))
 

### 2016-02-04 code/build_figure3.R

@@ -26,7 +26,7 @@ rownames(gg.sp) <- c("V4", "V3-V5", "V1-V3", "V1-V5", "V1-V6", "V1-V9")
 
 
 
-pdf(file="results/figures/figure_3.pdf", width=5, height=7)
+pdf(file="submission/figure_3.pdf", width=5, height=7)
 
 
 par(mar=c(5, 3, 0.5, 0.5))

### 2016-02-04 code/build_figure4.R

@@ -8,7 +8,7 @@ names(clrs) <- regions
 one_offs <- read.table(file="data/process/non_random_analysis.tsv", header=T)
 z <- prop.table(as.matrix(one_offs), 2)[1:10,]
 
-pdf(file="results/figures/figure_4.pdf", width=4, height=4)
+pdf(file="submission/figure_4.pdf", width=4, height=4)
 par(mar=c(4,4,0.5,0.5))
 plot(NA, xlim=c(1,10), ylim=c(0,1), xlab="Number of times 1-nt\nvariants were observed", ylab="Percentage of 1-nt variants", axes=F)
 for(r in regions){

### 2016-02-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Edit Abstract section

@@
 ```{r knitr_settings, eval=TRUE, echo=FALSE, cache=FALSE}
+library(rjson)
+
 opts_chunk$set("tidy" = TRUE)
@@
 
-clrs <- rainbow(length(regions))
-names(clrs) <- regions
-
-library(rjson)
+error_profile <- fromJSON(scan("data/process/error_profile.json", quiet=T, what="character"))
+error_summary <- read.table(file="data/process/error_summary.tsv", header=T)
+sobs_data <- read.table(file="data/process/sobs_table.tsv", header=T)
+taxonomy_data <- read.table(file="data/process/taxonomy_depth_analysis.tsv", header=T)
+one_offs <- read.table(file="data/process/non_random_analysis.tsv", header=T)
 ```
@@
 
-## Abstract [ Revise ]
+## Abstract
 Over the past 10 years, microbial ecologists have largely abandoned sequencing
@@
 Illumina's MiSeq, have allowed researchers to obtain millions of high quality,
-but short sequences. These platforms have allowed researchers to significantly
-improve the design of their experiments. The tradeoff has been the decline in
+but short sequences. The result of the added sequencing depth has been significant
+improvements in experimental design. The tradeoff has been the decline in
 the number of full-length reference sequences that are deposited into databases.
@@
 Real-Time (SMRT) DNA sequencing platform to generate sequence reads from the 16S
-rRNA gene. We generated sequencing data from the V4, V3-V5, V1-V3, V1-V6, and
-V1-V9 variable regions from within the 16S rRNA gene from a synthetic mock
-community and natural samples collected from human feces, mouse feces, and soil.
-The synthetic mock community allowed us to assess the actual sequencing error
-rate and how that error rate changed when different curation methods were
+rRNA gene. We generated sequencing data from the V4, V3-V5, V1-V3, V1-V5, V1-V6,
+and V1-V9 variable regions from within the 16S rRNA gene using DNA from a synthetic mock community and natural samples collected from human feces, mouse feces, and soil.
+The mock community allowed us to assess the actual sequencing error rate and how that error rate changed when different curation methods were
 applied. We developed a simple method based on sequence characteristics and
-quality scores to reduce the observed error rate for the V1-V9 region from 2.16%
-to 0.32%. Unfortunately, this error rate was still 16-times higher than the
-error rate that has been observed for the shorter reads generated by 454 and
-Illumina's MiSeq sequencing platforms. Although the longer reads frequently
-provided better classification, the wider adoption of this approach for 16S rRNA
-gene sequencing is likely limited by its high sequencing error and low yield of
-sequencing data relative to the other available platforms.
+quality scores to reduce the observed error rate for the V1-V9 region from
+`r format(error_summary[error_summary$region == "V1V9" & error_summary$method == "raw", "error"], digits=2, nsmall=2)` to `r format(error_summary[error_summary$region == "V1V9" & error_summary$method == "precluster", "error"], digits=2, nsmall=2)`%.
+This error rate is comparable
+to what has been observed for the shorter reads generated by 454 and
+Illumina's MiSeq sequencing platforms. Although the per base sequencing cost is
+still significantly more than that of MiSeq, the prospect of supplementing reference databases with full-length sequences from organisms below the limit of detection from the Sanger approach is exciting
 
 ***Keywords:*** Microbial ecology, bioinformatics, sequencing error
+
+\newpage
 
@@
 
-```{r error_profile}
-error_profile <- fromJSON(scan("data/process/error_profile.json", quiet=T, what="character"))
-```
-
 ***The PacBio error profile and a basic sequence curation procedure.***
@@
 ```{r basic}
-error_summary <- read.table(file="data/process/error_summary.tsv", header=T)
-
 raw_to_basic_lost <- 100 * (1 - error_summary[error_summary$method == "basic","nseqs"]/error_summary[error_summary$method == "raw","nseqs"])
@@
 ```{r}
-sobs_data <- read.table(file="data/process/sobs_table.tsv", header=T)
 perfect_otu <- 19
@@
 
-```{r, classificationSummary}
-taxonomy_data <- read.table(file="data/process/taxonomy_depth_analysis.tsv", header=T)
-```
-
 ***Increasing sequence length improves classification.***
@@
 ```{r}
-one_offs <- read.table(file="data/process/non_random_analysis.tsv", header=T)
 percent_one_off <- 100 * prop.table(as.matrix(one_offs), 2)

### 2016-02-05 Schloss_PacBio16S_PeerJ_2016.Rmd
Editing Intro and Methods

@@
 
-**Authors:** Patrick D. Schloss^1#^, Sarah L. Westcott^1^, Matthew L. Jenior^1^,
-and Sarah K. Highlander^2^
+**Authors:** Patrick D. Schloss^1#^, Matthew L. Jenior^1^, Charles C. Koumpouras^1^, Sarah L. Westcott^1^, and Sarah K. Highlander^2^
 
@@
 ## Abstract
-Over the past 10 years, microbial ecologists have largely abandoned sequencing
-16S rRNA genes by the Sanger sequencing method and have instead adopted highly
-parallelized sequencing platforms. These new platforms, such as 454 and
-Illumina's MiSeq, have allowed researchers to obtain millions of high quality,
-but short sequences. The result of the added sequencing depth has been significant
-improvements in experimental design. The tradeoff has been the decline in
-the number of full-length reference sequences that are deposited into databases.
-To overcome this problem, we tested the ability of the PacBio Single Molecule,
-Real-Time (SMRT) DNA sequencing platform to generate sequence reads from the 16S
-rRNA gene. We generated sequencing data from the V4, V3-V5, V1-V3, V1-V5, V1-V6,
-and V1-V9 variable regions from within the 16S rRNA gene using DNA from a synthetic mock community and natural samples collected from human feces, mouse feces, and soil.
-The mock community allowed us to assess the actual sequencing error rate and how that error rate changed when different curation methods were
-applied. We developed a simple method based on sequence characteristics and
-quality scores to reduce the observed error rate for the V1-V9 region from
-`r format(error_summary[error_summary$region == "V1V9" & error_summary$method == "raw", "error"], digits=2, nsmall=2)` to `r format(error_summary[error_summary$region == "V1V9" & error_summary$method == "precluster", "error"], digits=2, nsmall=2)`%.
-This error rate is comparable
-to what has been observed for the shorter reads generated by 454 and
-Illumina's MiSeq sequencing platforms. Although the per base sequencing cost is
-still significantly more than that of MiSeq, the prospect of supplementing reference databases with full-length sequences from organisms below the limit of detection from the Sanger approach is exciting
+
+Over the past 10 years, microbial ecologists have largely abandoned sequencing 16S rRNA genes by the Sanger sequencing method and have instead adopted highly parallelized sequencing platforms.
+These new platforms, such as 454 and Illumina's MiSeq, have allowed researchers to obtain millions of high quality, but short sequences.
+The result of the added sequencing depth has been significant improvements in experimental design.
+The tradeoff has been the decline in the number of full-length reference sequences that are deposited into databases.
+To overcome this problem, we tested the ability of the PacBio Single Molecule, Real-Time (SMRT) DNA sequencing platform to generate sequence reads from the 16S rRNA gene.
+We generated sequencing data from the V4, V3-V5, V1-V3, V1-V5, V1-V6, and V1-V9 variable regions from within the 16S rRNA gene using DNA from a synthetic mock community and natural samples collected from human feces, mouse feces, and soil.
+The mock community allowed us to assess the actual sequencing error rate and how that error rate changed when different curation methods were applied.
+We developed a simple method based on sequence characteristics and quality scores to reduce the observed error rate for the V1-V9 region from `r format(error_summary[error_summary$region == "V1V9" & error_summary$method == "raw", "error"], digits=2, nsmall=2)` to `r format(error_summary[error_summary$region == "V1V9" & error_summary$method == "precluster", "error"], digits=2, nsmall=2)`%.
+This error rate is comparable to what has been observed for the shorter reads generated by 454 and Illumina's MiSeq sequencing platforms.
+Although the per base sequencing cost is still significantly more than that of MiSeq, the prospect of supplementing reference databases with full-length sequences from organisms below the limit of detection from the Sanger approach is exciting.
 
@@
 
-## Introduction [ Revise ]
-Advances in sequencing technologies over the past 10 years have introduced
-considerable advances to the field of microbial ecology. Clone-based Sanger
-sequencing of the 16S rRNA gene has largely been replaced by various platforms
-produced by 454/Roche (e.g. @Sogin2006), Illumina
-(e.g. @Gloor2010), and IonTorrent (e.g. @Junemann2012). It was
-once common to sequence fewer than 100 16S rRNA gene sequences from several
-samples using the Sanger approach (e.g. @Mccaig1999). Now it is common to
-generate thousands of sequences from each of several hundred samples
-[@HMP2012]. The advance in throughput has
-come at the cost of read length. Sanger sequencing regularly generated 800 nt
-per read and because the DNA was cloned, it was possible to obtain multiple
-reads per fragment to yield a full-length sequence from a representative single
-molecule. At approximately $8 (US) per sequencing read, most
-researchers have effectively decided that full-length sequences are not worth
-the increased cost relative to the cost of more recently developed approaches.
-There is still a clear need to generate high-throughput full-length sequence
-reads that are of sufficient quality that they can be used as references for
-analyses based on obtaining short sequence reads.
+## Introduction
 
-Historically, all sequencing platforms were created to primarily perform genome
-sequencing. When sequencing a genome, it is assumed that the same base of DNA
-will be sequenced multiple times and the consensus of multiple sequence reads is
-used to generate contigs. Thus, although an individual base call may have a high
-error rate, the consensus sequence will have a low error rate. To sequence the
-16S rRNA gene researchers use conserved primers to amplify a sub-region from
-within the gene that is isolated from many organisms. Because the fragments
-are not cloned, it is not possible to obtain high sequence coverage from the
-same DNA molecule using these platforms. Thus, to reduce sequencing error rates
-it has become imperative to develop stringent sequence curation and denoising
-algorithms [@Kozich2013; @Schloss2011a]. There has been a tradeoff
-between read length, number of reads per sample, and the error rate. For
-instance, we recently demonstrated that using the Illumina MiSeq and the 454
-Titanium platforms the raw error rate varies between 1 and 2%
-[@Kozich2013; @Schloss2011a]. Yet, it was possible to obtain error rates below
-0.02% by adopting various denoising algorithms. However, the resulting fragments
-were only 250-nt long. In the case of 454 Titanium, extending the length of the
-fragment introduces length-based errors and in the case of the Illumina MiSeq,
-increasing the length of the fragment reduces the overlap between the read pairs
-reducing the ability of each read to mutually reduce the sequencing error.
-Inadequate denoising of sequencing reads can have many negative effects
-including limited ability to identify chimeras [@Haas2011; @Edgar2011]
-and inflation of alpha- and beta-diversity metrics [@Huse2010; @Kunin2010; @Kozich2013; @Schloss2011a].
-Although MiSeq and 454 enjoy widespread use in the field, the MiSeq
-platform is emerging as the leader because of the ability to sequence 15-20
-million fragments that can be distributed across hundreds of samples for less
-than $5000 (US).
+Advances in sequencing technologies over the past 10 years have introduced considerable advances to the field of microbial ecology.
+Clone-based Sanger sequencing of the 16S rRNA gene has largely been replaced by various platforms produced by 454/Roche (e.g.
+@Sogin2006), Illumina (e.g.
+@Gloor2010), and IonTorrent (e.g.
+@Junemann2012).
+It was once common to sequence fewer than 100 16S rRNA gene sequences from several samples using the Sanger approach (e.g.
+@Mccaig1999).
+Now it is common to generate thousands of sequences from each of several hundred samples [@HMP2012].
+The advance in throughput has come at the cost of read length.
+Sanger sequencing regularly generated 800 nt per read and because the DNA was cloned, it was possible to obtain multiple reads per fragment to yield a full-length sequence from a representative single molecule.
+At approximately $8 (US) per sequencing read, most researchers have effectively decided that full-length sequences are not worth the increased cost relative to the cost of more recently developed approaches.
+There is still a clear need to generate high-throughput full-length sequence reads that are of sufficient quality that they can be used as references for analyses based on obtaining short sequence reads.
 
-As these sequencing platforms have grown in popularity, there has been a decline
-in the number of full-length 16S rRNA genes being deposited into GenBank that
-could serve as references. This is particularly frustrating since the
-technologies have significantly improved our ability to detect and identify
-novel populations for which we lack full-length reference sequences. A related
-problem is the perceived limitation that the short reads generated by the 454
-and Illumina platforms cannot be reliably classified to the genus or
-species level. Previous investigators have utilized simulations to
-demonstrate that increased read lengths usually increase the accuracy and
-sensitivity of classification against reference databases
-[@Liu2008; @Werner2011; @Wang2007]. There is clearly a need to develop
-sequencing technologies that will allow researchers to generate high quality
-full-length 16S rRNA gene sequences in a high throughput manner.
+Historically, all sequencing platforms were created to primarily perform genome sequencing.
+When sequencing a genome, it is assumed that the same base of DNA will be sequenced multiple times and the consensus of multiple sequence reads is used to generate contigs.
+Thus, although an individual base call may have a high error rate, the consensus sequence will have a low error rate.
+To sequence the 16S rRNA gene researchers use conserved primers to amplify a sub-region from within the gene that is isolated from many organisms.
+Because the fragments are not cloned, it is not possible to obtain high sequence coverage from the same DNA molecule using these platforms.
+Thus, to reduce sequencing error rates it has become imperative to develop stringent sequence curation and denoising algorithms [@Kozich2013; @Schloss2011a].
+There has been a tradeoff between read length, number of reads per sample, and the error rate.
+For instance, we recently demonstrated that using the Illumina MiSeq and the 454 Titanium platforms the raw error rate varies between 1 and 2% [@Kozich2013; @Schloss2011a].
+Yet, it was possible to obtain error rates below 0.02% by adopting various denoising algorithms; however, the resulting fragments were only 250-nt long.
+In the case of 454 Titanium, extending the length of the fragment introduces length-based errors and in the case of the Illumina MiSeq, increasing the length of the fragment reduces the overlap between the read pairs reducing the ability of each read to mutually reduce the sequencing error.
+Inadequate denoising of sequencing reads can have many negative effects including limited ability to identify chimeras [@Haas2011; @Edgar2011] and inflation of alpha- and beta-diversity metrics [@Huse2010; @Kunin2010; @Kozich2013; @Schloss2011a].
+Illumina's MiSeq plaform enjoys widespread use in the field because of the ability to sequence 15-20 million fragments that can be distributed across hundreds of samples for less than $5000 (US).
 
-New advances in single molecule sequencing technologies, such as the platform
-produced by Pacific Biosciences (PacBio), offer the opportunity to once again
-obtain full-length sequence reads with a high depth of coverage from a large
-number of samples. To this point, the PacBio Single Molecule, Real-Time (SMRT)
-DNA Sequencing System has received limited application in the microbial ecology
-research domain @Mosher2013; @Mosher2014; @Fichot2013].
-The SMRT system ligates hairpin adapters (i.e. SMRTbells) to the ends of
-double-stranded DNA. Although the DNA molecule is linear, it is effectively
-circularized allowing the sequencing polymerase to process around the molecule
-multiple times [@Au2012]. According to Pacific Biosciences the platform
-is able to generate median read lengths longer than 8 kb with the P4-C2
-chemistry; however, the single pass error rate is approximately 15%. Given the
-circular nature of the DNA fragment, the full read length can be used to cover
-the DNA fragment multiple times resulting in a reduced error rate. Therefore,
-one should be able to obtain multiple coverage of the full 16S rRNA gene at a
-reduced error rate.
+As these sequencing platforms have grown in popularity, there has been a decline in the number of full-length 16S rRNA genes being deposited into GenBank that could serve as references for sequence classification, phylogenetic analyses, and primer and probe design.
+This is particularly frustrating since the technologies have significantly improved our ability to detect and identify novel populations for which we lack full-length reference sequences.
+A related problem is the perceived limitation that the short reads generated by the 454 and Illumina platforms cannot be reliably classified to the genus or species level.
+Previous investigators have utilized simulations to demonstrate that increased read lengths usually increase the accuracy and sensitivity of classification against reference databases [@Liu2008; @Werner2011; @Wang2007].
+There is clearly a need to develop sequencing technologies that will allow researchers to generate high quality full-length 16S rRNA gene sequences in a high throughput manner.
 
-Despite the opportunity to potentially generate high-quality full-length
-sequences, the Pacific Biosciences platform has not been widely adopted for
-sequencing 16S rRNA genes [@Fichot2013; @Mosher2014; @Mosher2013].
-Previous studies utilizing the technology have removed reads with mismatched
-primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013].
-Others have utilized the platform without describing the bioinformatic pipeline
-that was utilized [@Mosher2014; @Mosher2013]. Regardless of the
-curation methods, the error rates associated with sequencing the 16S rRNA gene
-on the platform have never been reported. In the current study, we assessed the
-quality of data generated by the Pacific Biosciences sequencer and whether it
-could fill the need for generating high-quality, full-length sequence data. We
-hypothesized that by modulating the 16S rRNA gene fragment length we could alter
-the read depth and obtain reads longer than are currently available by the 454
-and Illumina platforms but with the same quality. To test this hypothesis, we
-developed a sequence curation pipeline that was optimized by reducing the
-sequencing error rate of a mock bacterial community with known composition. The
-resulting pipeline was then applied to 16S rRNA gene fragments that were
-isolated from soil and human and mouse feces.
+New advances in single molecule sequencing technologies are being developed to address this problem.
+One approach uses a random barcoding strategy to fragment, sequence, and assemble full-length amplicons using Illumina's HiSeq platform [@Burke2014; @Miller2013].
+Although the algorithms appear to have minimized the risk of assembly chimeras, it is unclear what the sequencing error rate is by this approach.
+An alternative is the use of single molecule technologies that offer read lengths that are thousands of bases long.
+Although the Oxford Nanopore Technology has been used to sequence 16S rRNA genes [@BentezPez2016], the the platform produced by Pacific Biosciences (PacBio) has received wider attention for this application [@Mosher2013; @Mosher2014; @Fichot2013; @Schloss2015].
+The PacBio Single Molecule, Real-Time (SMRT) DNA Sequencing System ligates hairpin adapters (i.e.
+SMRTbells) to the ends of double-stranded DNA.
+Although the DNA molecule is linear, the adapters effectively circularize the DNA allowing the sequencing polymerase to process around the molecule multiple times [@Au2012].
+According to Pacific Biosciences the platform is able to generate median read lengths longer than 8 kb with the P6-C4 chemistry; however, the single pass error rate is approximately 15%.
+Given the circular nature of the DNA fragment, the full read length can be used to cover the DNA fragment multiple times resulting in a reduced error rate.
+Therefore, one should be able to obtain multiple coverage of the full 16S rRNA gene at a reduced error rate.
+
+Despite the opportunity to potentially generate high-quality full-length sequences, the Pacific Biosciences platform has not been widely adopted for sequencing 16S rRNA genes [@Fichot2013; @Mosher2014; @Mosher2013].
+Previous studies utilizing the technology have removed reads with mismatched primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013].
+Others have utilized the platform without describing the bioinformatic pipeline that was utilized [@Mosher2014; @Mosher2013].
+The only study to report the error rate of the platform for sequencing 16S rRNA genes with a mock community used the P4-C2 chemistry and obtained an error rate of 0.32%, which is 16-fold higher than has been observed using the MiSeq or 454 platforms [@Schloss2011a; @Kozich2013].
+In the current study, we assessed the quality of data generated by the PacBio sequencer usingthe improved P6-C4 chemistry and on-sequencer data processing.
+The goal was to determine whether this strategy could fill the need for generating high-quality, full-length sequence data on par with other platforms.
+We hypothesized that by modulating the 16S rRNA gene fragment length we could alter the read depth and obtain reads longer than are currently available by the 454 and Illumina platforms but with the same quality.
+To test this hypothesis, we developed a sequence curation pipeline that was optimized by reducing the sequencing error rate of a mock bacterial community with known composition.
+The resulting pipeline was then applied to 16S rRNA gene fragments that were isolated from soil and human and mouse feces.
 
@@
 ***Community DNA.*** We utilized genomic DNA isolated from four communities.
-These same DNA extracts were previously used to develop an Illumina MiSeq-based
-sequencing strategy [@Kozich2013].  Briefly, we used a “Mock Community”
-composed of genomic DNA from 21 bacterial strains: *Acinetobacter baumannii*
-ATCC 17978, *Actinomyces odontolyticus* ATCC 17982, *Bacillus cereus* ATCC
-10987, *Bacteroides vulgatus* ATCC 8482, *Clostridium beijerinckii* ATCC 51743,
-*Deinococcus radiodurans* ATCC 13939, *Enterococcus faecalis* ATCC 47077,
-*Escherichia coli* ATCC 70096, *Helicobacter pylori* ATCC 700392, *Lactobacillus
-gasseri* ATCC 33323, *Listeria monocytogenes* ATCC BAA-679, *Neisseria
-meningitidis* ATCC BAA-335, *Porphyromonas gingivalis* ATCC 33277,
-*Propionibacterium acnes* DSM 16379, *Pseudomonas aeruginosa* ATCC 47085,
-*Rhodobacter sphaeroides* ATCC 17023, *Staphylococcus aureus* ATCC BAA-1718,
-*Staphylococcus epidermidis* ATCC 12228, *Streptococcus agalactiae* ATCC
-BAA-611, *Streptococcus mutans* ATCC 700610, *Streptococcus pneumoniae* ATCC
-BAA-334. The mock community DNA is available through BEI resources (v3.1,
-HM-278D). Genomic DNAs from the three other communities were obtained using the
-MO BIO PowerSoil DNA extraction kit. The human and mouse fecal samples were
-obtained using protocols that were reviewed and approved by the University
-Committee on Use and Care of Animals (Protocol #PRO00004877) and the
-Institutional Review Board at the University of Michigan (Protocol
-#HUM00057066). The human stool donor provided informed consent.
+These same DNA extracts were previously used to develop an Illumina MiSeq-based sequencing strategy [@Kozich2013].
+Briefly, we used a “Mock Community” composed of genomic DNA from 21 bacterial strains: *Acinetobacter baumannii* ATCC 17978, *Actinomyces odontolyticus* ATCC 17982, *Bacillus cereus* ATCC 10987, *Bacteroides vulgatus* ATCC 8482, *Clostridium beijerinckii* ATCC 51743, *Deinococcus radiodurans* ATCC 13939, *Enterococcus faecalis* ATCC 47077, *Escherichia coli* ATCC 70096, *Helicobacter pylori* ATCC 700392, *Lactobacillus gasseri* ATCC 33323, *Listeria monocytogenes* ATCC BAA-679, *Neisseria meningitidis* ATCC BAA-335, *Porphyromonas gingivalis* ATCC 33277, *Propionibacterium acnes* DSM 16379, *Pseudomonas aeruginosa* ATCC 47085, *Rhodobacter sphaeroides* ATCC 17023, *Staphylococcus aureus* ATCC BAA-1718, *Staphylococcus epidermidis* ATCC 12228, *Streptococcus agalactiae* ATCC BAA-611, *Streptococcus mutans* ATCC 700610, *Streptococcus pneumoniae* ATCC BAA-334.
+The mock community DNA is available through BEI resources (v3.1, HM-278D).
+Genomic DNAs from the three other communities were obtained using the MO BIO PowerSoil DNA extraction kit.
+The human and mouse fecal samples were obtained using protocols that were reviewed and approved by the University Committee on Use and Care of Animals (Protocol #PRO00004877) and the Institutional Review Board at the University of Michigan (Protocol #HUM00057066).
+The human stool donor provided informed consent.
 
-***Library generation and sequencing.*** [ Revise ]
-The DNAs were each amplified in triplicate using barcoded primers targeting the
-V4, V1-V3, V3-V5, V1-V5, V1-V6, and V1-V9 variable regions (Table 1). The
-primers were synthesized so that the 5’ end of the forward and reverse primers
-were each tagged with a 5-nt barcode sequence to allow multiplexing of samples
-within a single sequencing run. Methods describing PCR, amplicon cleanup, and
-pooling were described previously [@Kozich2013]. The SMRTbell adapters
-were ligated onto  the PCR products and the libraries were sequenced by Pacific
-Biosciences using the P6-C24 chemistry on a PacBio RS II SMRT DNA Sequencing
-System.
+***Library generation and sequencing.***
+The DNAs were each amplified in triplicate using barcoded primers targeting the V4, V1-V3, V3-V5, V1-V5, V1-V6, and V1-V9 variable regions (Table 1).
+The primers were synthesized so that the 5’ end of the forward and reverse primers were each tagged with paired 16-nt symmetric barcodes (https://github.com/PacificBiosciences/Bioinformatics-Training/wiki/Barcoding-with-SMRT-Analysis-2.3) to allow multiplexing of samples within a single sequencing run.
+Methods describing PCR, amplicon cleanup, and pooling were described previously [@Kozich2013].
+The SMRTbell adapters were ligated onto  the PCR products and the libraries were sequenced by Pacific Biosciences using the P6-C4 chemistry on a PacBio RS II SMRT DNA Sequencing System.
+Diffusion Loading was used for regions V4, V1-V3, and V3-V5 and MagBead loading was used for regions V1-V5, V1-V6, and V1-V9.
+Each region was sequenced separately using movies ranging in length between 180 and 360 minutes.
+The sequences were processed using pbccs (v.3.0.1; https://github.com/PacificBiosciences/pbccs), which generates predicted error rates using a proprietary algorithm.
 
-***Data analysis.***  [ Revise ]
-All sequencing data were curated using mothur [@Schloss2009a] and analyzed
-using the R programming language [@R2016]. The raw data can be
-obtained from the Sequence Read Archive at NCBI under accession SRP051686, which
-are associated with BioProject PRJNA271568. Several specific features were
-incorporated into mothur to facilitate the analysis of PacBio  sequence data.
-First, because non-ambiguous base calls are assigned to Phred quality scores of
-zero, the consensus fastq files were parsed so that scores of zero were
-interpreted as corresponding to an ambiguous base call (i.e. N) in the
-fastq.info command using the pacbio=T option. Second, because the consensus
-sequence can be generated in the forward and reverse complement orientations, a
-checkorient option was added to the trim.seqs command in order to identify the
-proper orientation. These features were incorporated into mothur v.1.30. Because
-chimeric molecules can be generated during PCR and would artificially inflate
-the sequencing error, it was necessary to remove these data prior to assessing
-the error rate. Because we knew the true sequences for the strains in the mock
-community we could calculate all possible chimeras between strains in the mock
-community (*in silico* chimeras). If a sequence read was 3 or more nucleotides
-more similar to an *in silico* chimera than it was to a non-chimeric reference
-sequence, it was classified as a chimera and removed from further consideration.
-Identification of *in silico* chimeras and calculation of sequencing error rates
-was performed using the seq.error command in mothur [@Schloss2011a]. *De
-novo* chimera detection was also performed on the mock and other sequence data
-using the abundance-based algorithm implemented in UCHIME [@Edgar2011].
-Sequences sequences were aligned against a SILVA-based reference alignment
-[@Pruesse2007] using a profile-based aligner [@Schloss2009b] and
-were classified against the SILVA [@Pruesse2007], RDP [@Cole2013],
-and greengenes [@Werner2011] reference taxonomies  using a naive Bayesian
-classifier [@Wang2007]. Sequences were assigned to operational taxonomic
-units using the average neighbor clustering algorithm with a 3% distance
-threshold [@Schloss2011b].  Detailed methods including this paper as an R
-markdown file are available as a public online repository
-(http://github.com/SchlossLab/Schloss_PacBio16S_PeerJ_2016).
+***Data analysis.***
+All sequencing data were curated using mothur (v1.36)[@Schloss2009a] and analyzed using the R programming language [@R2016].
+The raw data can be obtained from the Sequence Read Archive at NCBI under accession SRP051686, which are associated with BioProject PRJNXXXXX.
+Several specific features were incorporated into mothur to facilitate the analysis of PacBio sequence data.
+First, because non-ambiguous base calls are assigned to Phred quality scores of zero, the consensus fastq files were parsed so that scores of zero were interpreted as corresponding to an ambiguous base call (i.e.
+N) in the fastq.info command using the pacbio=T option.
+Second, because the consensus sequence can be generated in the forward and reverse complement orientations, a checkorient option was added to the trim.seqs command in order to identify the proper orientation.
+These features were incorporated into mothur v.1.30.
+Because chimeric molecules can be generated during PCR and would artificially inflate the sequencing error, it was necessary to remove these data prior to assessing the error rate.
+Because we knew the true sequences for the strains in the mock community we could calculate all possible chimeras between strains in the mock community (*in silico* chimeras).
+If a sequence read was 3 or more nucleotides more similar to an *in silico* chimera than it was to a non-chimeric reference sequence, it was classified as a chimera and removed from further consideration.
+Identification of *in silico* chimeras and calculation of sequencing error rates was performed using the seq.error command in mothur [@Schloss2011a].
+*De novo* chimera detection was also performed on the mock and other sequence data using the abundance-based algorithm implemented in UCHIME [@Edgar2011].
+Sequences sequences were aligned against a SILVA-based reference alignment [@Pruesse2007] using a profile-based aligner [@Schloss2009b] and were classified against the SILVA (v123) [@Pruesse2007], RDP (v10)[@Cole2013], and greengenes (v13_8_99)[@Werner2011] reference taxonomies  using a negative Bayesian classifier implemented within mothur [@Wang2007].
+Sequences were assigned to operational taxonomic units using the average neighbor clustering algorithm with a 3% distance threshold [@Schloss2011b].
+Detailed methods including this paper as an R markdown file are available as a public online repository (http://github.com/SchlossLab/Schloss_PacBio16S_PeerJ_2016).
 
@@
 
-**Figure 1. Summary of errors in data generated using PacBio sequencing platform
-to sequence various regions within the 16S rRNA gene.** The predicted error rate using PacBio's sequence analysis algorithm correlated well with the observed error rate (Pearson's R: `r format(error_profile$pred_error_corr, digits=2, nsmall=2)`; A).
+**Figure 1.
+Summary of errors in data generated using PacBio sequencing platform to sequence various regions within the 16S rRNA gene.** The predicted error rate using PacBio's sequence analysis algorithm correlated well with the observed error rate (Pearson's R: `r format(error_profile$pred_error_corr, digits=2, nsmall=2)`; A).
 Because of the large number of sequences, we randomly selected 5% of the data to show in panel A.
-The sequencing error rate of the amplified gene fragments
-increased with mismatches to the barcodes and primers (B). The sequencing error
-rate declined with increased sequencing coverage; however, increasing the
-sequencing depth beyond 10-fold coverage had no meaningful effect on the
-sequencing error rate (C).
+The sequencing error rate of the amplified gene fragments increased with mismatches to the barcodes and primers (B).
+The sequencing error rate declined with increased sequencing coverage; however, increasing the sequencing depth beyond 10-fold coverage had no meaningful effect on the sequencing error rate (C).
 
-**Figure 2. Change in error rate (A) and the percentage of sequences that were
-retained (B) when using various sequence curation methods.** The condition that
-was used for downstream analyses is indicated by the star. The plotted numbers
-represent the region that was sequenced. For example "15" represents the data
-for the V1-V5 region.
+**Figure 2.
+Change in error rate (A) and the percentage of sequences that were retained (B) when using various sequence curation methods.** The condition that was used for downstream analyses is indicated by the star.
+The plotted numbers represent the region that was sequenced.
+For example "19" represents the data for the V1-V9 region.
 
-**Figure 3. Percentage of unique sequences that could be classified.**
-Classifications were performed using taxonomy references curated from the RDP,
-SILVA, or greengenes databases for the four types of samples that were sequenced
-across the six regions from the 16S rRNA gene. Only the greengenes taxonomy
-reference provided species-level information.
+**Figure 3.
+Percentage of unique sequences that could be classified.** Classifications were performed using taxonomy references curated from the RDP, SILVA, or greengenes databases for the four types of samples that were sequenced across the six regions from the 16S rRNA gene.
+Only the greengenes taxonomy reference provided species-level information.
 
 **Figure 4.
-Percentage of 1-nt variants that occured up to ten times.** Sequences that were 1 nt different from the mock community reference sequences were counted to determine the number of times each variant appeared by region within the 16S rRNA gene.
+Percentage of 1-nt variants that occurred up to ten times.** Sequences that were 1 nt different from the mock community reference sequences were counted to determine the number of times each variant appeared by region within the 16S rRNA gene.
 

### 2016-02-05 code/build_figure1.R
Fix color scheme

@@ -1,7 +1,8 @@
 pretty_region <- c("V4"="V4", "V3V5"="V3-V5", "V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9")
 regions <- names(pretty_region)
 
-clrs <- rainbow(length(regions))
+library("wesanderson")
+clrs <- c(wes_palette("Darjeeling"), "black")
 names(clrs) <- regions
 
 

### 2016-02-05 code/build_figure4.R

@@ -1,7 +1,9 @@
 pretty_region <- c("V4"="V4", "V3V5"="V3-V5", "V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9")
 regions <- names(pretty_region)
 
-clrs <- rainbow(length(regions))
+library("wesanderson")
+clrs <- c(wes_palette("Darjeeling"), "black")
+
 names(clrs) <- regions
 
 #plot one_offs

### 2016-02-12 Schloss_PacBio16S_PeerJ_2016.Rmd
Incorporate Jenior's comments

@@
 When sequencing a genome, it is assumed that the same base of DNA will be sequenced multiple times and the consensus of multiple sequence reads is used to generate contigs.
-Thus, although an individual base call may have a high error rate, the consensus sequence will have a low error rate.
+Thus, an individual base call may have a high error rate, but the consensus sequence will have a low error rate.
 To sequence the 16S rRNA gene researchers use conserved primers to amplify a sub-region from within the gene that is isolated from many organisms.
 Because the fragments are not cloned, it is not possible to obtain high sequence coverage from the same DNA molecule using these platforms.
-Thus, to reduce sequencing error rates it has become imperative to develop stringent sequence curation and denoising algorithms [@Kozich2013; @Schloss2011a].
+To reduce sequencing error rates it has become imperative to develop stringent sequence curation and denoising algorithms [@Kozich2013; @Schloss2011a].
 There has been a tradeoff between read length, number of reads per sample, and the error rate.
@@
 The only study to report the error rate of the platform for sequencing 16S rRNA genes with a mock community used the P4-C2 chemistry and obtained an error rate of 0.32%, which is 16-fold higher than has been observed using the MiSeq or 454 platforms [@Schloss2011a; @Kozich2013].
-In the current study, we assessed the quality of data generated by the PacBio sequencer usingthe improved P6-C4 chemistry and on-sequencer data processing.
+In the current study, we assessed the quality of data generated by the PacBio sequencer using the improved P6-C4 chemistry and on-sequencer data processing.
 The goal was to determine whether this strategy could fill the need for generating high-quality, full-length sequence data on par with other platforms.
@@
 raw_to_basic_lost <- 100 * (1 - error_summary[error_summary$method == "basic","nseqs"]/error_summary[error_summary$method == "raw","nseqs"])
+names(raw_to_basic_lost) <- error_summary[error_summary$method == "basic","region"]
 ```
@@
 We established a simple curation procedure by culling any sequence that had a string of the same base repeated more than 8 times  or did not start and end at the expected alignment coordinates for that region of the 16S rRNA gene.
-This reduced the experiment-wide error rate from `r format(mean(error_summary[error_summary$method=="raw","error"]), digits=3, nsmall=2)` to `r format(mean(error_summary[error_summary$method=="basic","error"]), digits=2, nsmall=2)`%.
-This basic procedure resulted in the removal of between `r format(min(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(which.min(raw_to_basic_lost))]`) and `r format(max(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(which.max(raw_to_basic_lost))]`)% of the reads (Table 2).
+This reduced the experiment-wide error rate from `r format(mean(error_summary[error_summary$method=="raw","error"]), digits=2, nsmall=2)` to `r format(mean(error_summary[error_summary$method=="basic","error"]), digits=2, nsmall=2)`%.
+This basic procedure resulted in the removal of between `r format(min(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(raw_to_basic_lost[which.min(raw_to_basic_lost)])]`) and `r format(max(raw_to_basic_lost), digits=3, nsmall=1)` (`r pretty_region[names(raw_to_basic_lost[which.max(raw_to_basic_lost)])]`)% of the reads (Table 2).
 Although the percentage of reads removed increased with the length of the fragment, there was no obvious relationship between fragment length and error rate (Figure 2).
@@
 In contrast, classification of the human libraries resulted in more than 40% of the sequences being classified to a genus, regardless of the region.
-That the variation in species-level classification for the human libraries was less than 10% suggests that the benefit of added length is minimal considering the lower sequencing yield.
+These data demonstrate that for the samples we analyzed, the length of the sequence fragment was not as significant a factor in classification as the choice of database.
 

### 2016-02-12 Schloss_PacBio16S_PeerJ_2016.Rmd
Highlander comments

@@
 Third, based on the observed correlation between the predicted and observed error rates, we sought to identify a minimum predicted error rate that would allow us to reduce the observed error rate.
-The average observed error rate for sequences with predicted error rates between 0.01 and 0.10% was linear (Figure 1D).
+The average observed error rate for sequences with predicted error rates between 0.01 and 0.10% was linear.
 We decided to use a threshold of 0.01% because a large number of sequence reads were lost when we used a smaller threshold.
-When we used this threshold, we were able to reduce the error rate by `r paste(format(range(error_summary[error_summary$method == "pred","error_red"]), digits=3, nsmall=1), collapse=" to ")`% (Figure 2A).
+When we used this threshold, we were able to reduce the error rate by `r paste(format(range(error_summary[error_summary$method == "pred","error_red"]), digits=3, nsmall=1), collapse=" to ")`% (Figure 2).
 Finally, we quantified the effect of combining filters.
@@
 The full-length (i.e.
-V1-V9) fragments had an error rate of `r round(error_summary[error_summary$method == "precluster" & error_summary$region=="V1V9","error"], digits=2)`% (Figure 2B), this is similar to what we have previously observed using the 454 and Illumina MiSeq platforms (0.02%)[@Schloss2011a; @Kozich2013].
+V1-V9) fragments had an error rate of `r round(error_summary[error_summary$method == "precluster" & error_summary$region=="V1V9","error"], digits=2)`% (Figure 2; Table 2), this is similar to what we have previously observed using the 454 and Illumina MiSeq platforms (0.02%)[@Schloss2011a; @Kozich2013].
 
@@
 ***Effects of error rates on OTU assignments.***
-The sequencing error rate is known to affect the number of OTUs that are observed [Schloss2011a].
+The sequencing error rate is known to affect the number of OTUs that are observed [@Schloss2011a].
 For each region, we determined that if there were no chimeras or PCR or sequencing errors, then we would expect to find `r perfect_otu` OTUs.
-When achieved perfect chimera removal, but allowed for PCR and sequencing errors, we observed between `r diff<-sobs_data[,2]-perfect_otu; round(min(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.min(diff),])]`) and `r round(max(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.max(diff),])]`) extra OTUs.
+When achieved perfect chimera removal, but allowed for PCR and sequencing errors, we observed between `r diff<-sobs_data[,2]-perfect_otu; round(min(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.min(diff),])]`) and `r round(max(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.max(diff),])]`) extra OTUs (Table 2).
 The range in the number of extra OTUs was largely explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, error_summary[error_summary$method=="precluster", "error"])$estimate, 2)`).
@@
 
-***Increasing sequence length improves classification.***
+***Classification varies by region, environment, and database.***
 We classified all of the sequence data we generated using the naïve Bayesian classifier using the RDP, SILVA, and greengenes reference taxonomies (Figure 3).
@@
 This would manifest itself by the creation of duplicate sequences with the same error.
-We identified all of the mock community sequences that had a 1-nt difference to the true sequence.
+We identified all of the mock community sequences that had a 1-nt difference to the true sequence (Figure 4).
 For these three regions, between `r paste0(format(round(range(percent_one_off[1,]), digits=1), nsmall=1), collapse=" and ")` of the sequences with 1-nt errors were only observed once.
@@
 It remains to be determined whether the error rates we observed for full-length sequences are prohibitive for these applications.
-We can estimate the distribution of errors assuming that the errors follow a binomial distribution along the length of the 1,500 nt gene with the error rate that we achieved from the V1-V9 mock community data prior to pre-clustering the sequences, which was `r round(100*unique,2)`% (Figure 5).
+We can estimate the distribution of errors assuming that the errors follow a binomial distribution along the length of the 1,500 nt gene with the error rate that we achieved from the V1-V9 mock community data prior to pre-clustering the sequences, which was `r round(100*unique,2)`%.
 Under these conditions one would expect `r round(100*dbinom(0, 1500, unique), 2)`% of the sequences to have no errors and 50% of the sequences would have at least `r qbinom(0.50, 1500, unique)` errors.

### 2016-02-12 results/tables/build_table2.Rmd
Highlander comments

---
output:
  pdf_document:
    includes:
      in_header: table_header.tex
tables: true
geometry: margin=1.0in
---

\captionsetup{labelformat=empty}

```{r table1, results='asis', echo=FALSE, warning=FALSE, eval=TRUE}
require(xtable, quietly=TRUE)
options(xtable.sanitize.text.function=identity)
options(xtable.sanitize.colnames.function=identity)

pretty_region <- c("V4"="V4", "V1V3"="V1-V3", "V3V5"="V3-V5", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9")
region_order <- names(pretty_region)

mock_error <- read.table(file="data/process/error_summary.tsv", header=T, stringsAsFactors=F)
regions <- mock_error[mock_error$method=="basic","region"]
basic <- mock_error[mock_error$method=="basic","error"]
predicted <- mock_error[mock_error$method=="pred","error"]
precluster <- mock_error[mock_error$method=="precluster","error"]
percent_remaining <- 100-mock_error[mock_error$method=="precluster","nseqs_red"]
error_data <- data.frame(cbind(basic, predicted, precluster, percent_remaining))
rownames(error_data) <- regions
error_data <- error_data[region_order,]
error_data <- format(error_data, nsmall=2L, digits=2)

sobs_data <- read.table(file="data/process/sobs_table.tsv", header=T)
sobs_data <- sobs_data[region_order,-c(1,7)]
sobs_data <- format(sobs_data, nsmall=1L, digits=2)

composite <- cbind(error_data, sobs_data)
rownames(composite) <- pretty_region

#Error rate (%) for:
#% reads remaining from basic
#Average no. of OTUsa

header <- c("Basic", "Predicted", "Precluster", "", "Mock\\textsuperscript{b}", "Mock\\textsuperscript{c}", "Soil", "Mouse", "Human")

header <- paste0("\\textbf{", header, "}")
composite <- rbind(" " = header, composite)

addtorow <- list(pos=list(0,1),
                command=c('\\multirow{3}{*}{\\textbf{Region}} & \\multicolumn{3}{c}{\\textbf{Error rate following (\\%)}} & \\multirow{1}{*}{\\textbf{\\bigcell{c}{Reads \\\\ remaining (\\%)}}} & \\multicolumn{5}{c}{\\textbf{Average no. of OTUs\\textsuperscript{a}}} \\\\ \\cmidrule(r){2-4}	\\cmidrule(r){6-10}', '\\hline '))

caption <- "\\textbf{Table 2. Summary of the error rates and number of observed OTUs for each region.}"

composite <- xtable(composite, caption = caption, label="")

align(composite) <- "cccccccccc"
print(composite,
    add.to.row=addtorow,
	include.rownames=TRUE,
	include.colnames=FALSE,
	hline.after = c(-1, nrow(composite)),
    comment=FALSE,
    caption.placement = "top",
    table.placement="!ht",
	size="small",
	NA.string="ND"
)
```

^a^ The number of OTUs is based on rarefaction of each sample to 1,000 sequences per sample; cells labeled ND reflect samples that did not have at least 1,000 sequences.

^b^ Number of OTUs in the mock community when all chimeras were removed; in the absence of chimeras and sequencing errors, there should be 19 OTUs for all three regions.

^c^ Number of OTUs in the mock community when chimeras were removed using UCHIME.

### 2016-02-12 Schloss_PacBio16S_PeerJ_2016.Rmd
Add new PacBio reference

@@
 An alternative is the use of single molecule technologies that offer read lengths that are thousands of bases long.
-Although the Oxford Nanopore Technology has been used to sequence 16S rRNA genes [@BentezPez2016], the the platform produced by Pacific Biosciences (PacBio) has received wider attention for this application [@Mosher2013; @Mosher2014; @Fichot2013; @Schloss2015].
+Although the Oxford Nanopore Technology has been used to sequence 16S rRNA genes [@BentezPez2016], the the platform produced by Pacific Biosciences (PacBio) has received wider attention for this application [@Mosher2013; @Mosher2014; @Fichot2013; @Schloss2015; @Singer2016].
 The PacBio Single Molecule, Real-Time (SMRT) DNA Sequencing System ligates hairpin adapters (i.e.
@@
 
-Despite the opportunity to potentially generate high-quality full-length sequences, the Pacific Biosciences platform has not been widely adopted for sequencing 16S rRNA genes [@Fichot2013; @Mosher2014; @Mosher2013].
-Previous studies utilizing the technology have removed reads with mismatched primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013].
+Despite the opportunity to potentially generate high-quality full-length sequences, it is surprising that the Pacific Biosciences platform has not been more widely adopted for sequencing 16S rRNA genes.
+Previous studies utilizing the technology have removed reads with mismatched primers and barcodes, ambiguous base  calls, and low quality scores [@Fichot2013] or screened sequences based on the predicted error rate [@Singer2016].
 Others have utilized the platform without describing the bioinformatic pipeline that was utilized [@Mosher2014; @Mosher2013].

### 2016-02-15 doc/paper/rebuttal_v1.md
Incorporate previous reviewers' comments

## Reviewer 1

**The authors come to a sweeping conclusion about a technology using a version of the platform that is no longer available and has been superseded by two new generations of polymerase enzyme and sequencing chemistry. Since the paper focusses on a single technology, its conclusions using the P4-C2 chemistry are of very limited value to scientists currently working with the technology (current version: P6-C4). While it has to be appreciated that studies like these shoot at a moving target, the authors would at the very least have to make it very clear that their conclusions are only valid for the older version of the platform. The very strong statements of the current manuscript convey a message that ignores reported improvements.**

**Ideally, the authors should include additional analyses on the V1-V9 amplicon using the latest version of the PacBio platform. I believe this could be done without too much additional effort and greatly strengthen the impact of the paper. Alternatively, the authors need to place very visible disclaimers making it clear that all analyses are based on data generated with the P4-C2 chemistry, which is obsolete.
Error rates will also be influenced by the initial amplification. Were the exact same conditions used to generate the amplicons used for PacBio sequencing that were also used in the publication by Kozich et al., that serves as the reference for the error rates of the MiSeq platform.**

Reviewer 1 raised the important point that because the sequencing technology is a moving target, our conclusions were too strong and more importantly, out of date. In the last year we regenerated the data using the P6-C4 chemistry and the latest recommendations from PacBio on the appropriate methods for generating amplicon data. With that in mind we have performed a thorough rewrite of the paper that comes to a very different conclusion than the original manuscript. We appreciate the reviewer calling us on this point.

## Reviewer 2

**A general comment is, given the longer reads possible with this platform, perhaps there might not be need to focus on only one or a few commonly used hyper-variable region, but to focus on the nearly-full-length sequences. Considering the 3% divergence threshold commonly used for species definition, on the order of 40 errors could be permitted for a sequence length of 1500, and the goal of accurate genus classification and OTU clustering might be adequate. This may of course have other consequences when most analyses are done de novo, but I would suggest reference based analysis for chimera removal and OTU clustering. Although the authors have already done a very fine job with their assessment, perhaps they could address this.**

Given the cost of generating sequences by this platform, we feel that the primary use case for the technology is in generating reference sequences. As we discussed in the previous version, it is debatable whether sequences with 40 errors would be appropriate as a reference. We do not foresee the utility of this approach for broad scale sequencing surveys when there are cheaper methods for generating deep samples of datasets.

**Lines 140-141: Could the authors clarify the nature of the quality score from PacBio (Figure 1A does not appear to be comparable to that seen from other platforms)? How does it compare to that of Illumina/454 and how does mothur cope with the differences if there are any?**

We have added text towards the end of the first paragraph of the Results and Discussion section that points out that the PacBio quality scores are not the traditional Phred scores. In contrast to what we previously observed using the older chemistry, the new quality scores are not helpful in assessing the quality of the sequence data.

**Lines 279-282: “We found that the frequency of the most abundant 1-nt error paralleled the number of sequences. There were two sequences in the V4 dataset that occurred 76 times, one sequence in the V1-V5 dataset that occurred 30 times, and one sequence in the V3-V5 dataset that occurred 17 times.” Are there any general characteristics of the most abundant errors that could provide insight into why certain errors tend to occur? e.g. high AT content or high nucleotide diversity overall (or in a particular sliding window)?**

Unfortunately, we were not able to identify any signatures beyond the various insertion, deletion, and substitution biases in the overall data set.

**Figure 2 is bit difficult to follow with this “numbering” scheme, suggest to change to colored lines.**

Part of the design of this figure was that the regions generally overlap on top of each other indicating that the effectiveness of the methods is reproducible. We have colored the numbers and increased the horizontal jitter.

## Reviewer 3

**Comments: In my first few readings of the manuscript it was unclear how the additional sequencing depth was taken into account. In the beginning of the results it states the error rates were ~1.8% on average. But it was unclear whether this is the baseline error rate for a single pass, or whether it was the result of taking a consensus of multiple passes. Since the introduction states the single pass error is ~15%, it must be the latter which makes sense given the results, but I’d recommend explaining this at the beginning of the results section.**

**It would also be helpful to explain how the multiple reads are used to reduce error rates. For example, if the single pass error is 15%, then the chances of seeing an incorrect nucleotide on 5 or more out of 10 passes would be something like 0.006%, which seems pretty good. But I’m guessing that non-random errors invalidate this simple estimate, but I cant tell for sure because there’s no markings on the y-axis to figure 1C. If the high error rates stem from the fact that coverage rates are generally much lower than 10X, that would be good to know, because it means the platform could potentially be useful if read lengths are increased.**

We have added text to the beginning of the first paragraph of the Results and Discussion section to indicate that ever read we analyzed had at least 3-fold coverage. As we go on to demonstrate, at some point (>10-fold coverage) increased coverage does not improve the error rate. This was a surprising result and we suspect that the non-random nature of the errors is responsible for the departure from the expected error rate.

We have added text to the legend indicating that the y-axis scales are the same in Figure 1B and 1C.

**Finally, it was not clear if this study was based on a single sequencing run at a single facility. If so, it might be important to point out that results could vary from run to run, especially if read lengths/sequencing depth plays a big role.**

It was based on a series of sequencing runs at a single facility. We have added text to the Conclusion section to indicate the value of sequencing mock communities in parallel with samples to make sure that the error rate is maintained regardless of the facility.

**line 215: "whether" should be deleted.**

This was removed in the rewrite

**line 233: “one would unable…”**

This has been corrected.

**line 243: should say "when we achieved...".**

This has been corrected.

**lines 244ff.: in one place, they say they observed between 6 and 63.1 extra OTUs. in another place, between 7.4 and 86.8 extra OTUs. how do you have a fractional OTU?**

Because this is based on rarefying the data, these numbers represent an average number of extra OTUs.

**line 246: "sequence" should be plural.**

This has been corrected.

**Figure 2 is illegible. why not just have colored points or something? or add more left-right jitter?**

This has been altered as indicated above.

**line 280: if I understand correctly, it seems that some errors occurred once, but a small number occurred many times. That suggests the latter type may be errors during PCR or even cell division, which are not necessarily relevant to the sequencing methodology. If they are biases of the platform, however, that would be a major problem, so it would be good to have some clarification on this point.**

Given the expected error rates of the PCR polymerase is at least one or two orders of magnitude lower than the sequencing error rate, we suspect that these errors are due to the platform.

### 2016-02-15 Schloss_PacBio16S_PeerJ_2016.Rmd
Incorporate previous reviewers' comments

@@
 To build a sequence curation pipeline, we first needed to characterize the error rate associated with sequencing the 16S rRNA gene.
-We observed an average sequencing error rate of `r format(error_profile$overall_error, digits=2, nsmall=2)`%.
+Using the consensus sequence obtained from at least 3 sequencing passes of each fragment, we observed an average sequencing error rate of `r format(error_profile$overall_error, digits=2, nsmall=2)`%.
 Insertions, deletions, and substitutions accounted for `r format(error_profile$ins_rate, digits=3, nsmall=1)`, `r format(error_profile$del_rate, digits=3, nsmall=1)`, and `r format(error_profile$subs_rate, digits=3, nsmall=1)`% of the errors, respectively.
@@
 Surprisingly, the percentage of base calls that had the maximum quality value did not vary among correct base calls (`r format(error_profile$percent_over_93$m, digits=3, nsmall=1)`%), substitutions (`r format(error_profile$percent_over_93$s, digits=3, nsmall=1)`%), and insertions (`r format(error_profile$percent_over_93$i, digits=3, nsmall=1)`%).
-Although the individual base quality scores could not be used to screen sequence quality, we observed a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(error_profile$pred_error_corr, digits=2, nsmall=2)`; Figure 1A).
+It was not possible to use the individual base quality scores could not be used to screen sequence quality as has been possible in past studies using the Phred-based quality scores that accompany data generated using the 454 and Illumina technologies.
+We did observe a strong correlation between our observed error rate and the predicted error rate as calculated by the PacBio software (Pearson's R: `r format(error_profile$pred_error_corr, digits=2, nsmall=2)`; Figure 1A).
 
@@
 This approach removes residual PCR and sequencing errors while not overwhelming the resolution needed to identify OTUs that are based on a 3% distance threshold.
-The tradeoff of this approach is that one would unable to differentiate V1-V9 sequences that truly differed by less than 14 nt.
+The tradeoff of this approach is that one would be unable to differentiate V1-V9 sequences that truly differed by less than 14 nt.
 When we applied this approach to our PacBio data, we observed a reduction in the error rate between `r format(min(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)` (`r pretty_region[which.min(error_summary[error_summary$method == "precluster","nseqs_red"])]`) and `r format(max(error_summary[error_summary$method == "precluster","nseqs_red"]), digits=3, nsmall=1)`% (`r pretty_region[which.max(error_summary[error_summary$method == "precluster","nseqs_red"])]`).
@@
 For each region, we determined that if there were no chimeras or PCR or sequencing errors, then we would expect to find `r perfect_otu` OTUs.
-When achieved perfect chimera removal, but allowed for PCR and sequencing errors, we observed between `r diff<-sobs_data[,2]-perfect_otu; round(min(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.min(diff),])]`) and `r round(max(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.max(diff),])]`) extra OTUs (Table 2).
+When we achieved perfect chimera removal, but allowed for PCR and sequencing errors, we observed between `r diff<-sobs_data[,2]-perfect_otu; round(min(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.min(diff),])]`) and `r round(max(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.max(diff),])]`) extra OTUs (Table 2).
 The range in the number of extra OTUs was largely explained by the sequencing error rate (Pearson's R=`r round(cor.test(diff, error_summary[error_summary$method=="precluster", "error"])$estimate, 2)`).
-Next, we determined the number of OTUs that were observed when we used UCHIME to identify chimeric sequence.
+Next, we determined the number of OTUs that were observed when we used UCHIME to identify chimeric sequences.
 Under these more realistic conditions, we observed between `r diff<-sobs_data[,3]-perfect_otu; round(min(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.min(diff),])]`) and `r round(max(diff), digits=1)` (`r pretty_region[rownames(sobs_data[which.max(diff),])]`) extra OTUs.
@@
 The sequencing error rate declined with increased sequencing coverage; however, increasing the sequencing depth beyond 10-fold coverage had no meaningful effect on the sequencing error rate (C).
+The scale of they y-axis in B and C are the same.
 

### 2016-02-15 code/build_figure2.R
Change numbers to plotting symbols w/ color

@@ -1,10 +1,18 @@
-error_summary <- read.table(file="data/process/error_summary.tsv", header=T)
+error_summary <- read.table(file="data/process/error_summary.tsv", header=T, stringsAsFactors=F)
 
 getJitter <- function(x){
-	jitter <- 0.15
+	jitter <- 0.25
 	runif(6, x-jitter, x+jitter)
 }
 
+pretty_pch <- c("V4"=15, "V3V5"=23, "V1V3"=17, "V1V5"=24, "V1V6"=19, "V1V9"=25)
+pretty_region <- c("V4"="V4", "V3V5"="V3-V5", "V1V3"="V1-V3", "V1V5"="V1-V5", "V1V6"="V1-V6", "V1V9"="V1-V9")
+
+#regions <- names(pretty_region)
+
+library("wesanderson")
+clrs <- c(wes_palette("Darjeeling"), "black")
+names(clrs) <- names(pretty_pch)
 
 methods <- c("raw", "basic", "pb", "pred", "coverage", "pb_coverage", "pb_pred",
 				"coverage_pred", "all", "precluster")
@@ -13,16 +21,26 @@ methods <- c("raw", "basic", "pb", "pred", "coverage", "pb_coverage", "pb_pred",
 
 cairo_pdf(file="submission/figure_2.pdf", width=6.5, height=6.0)
 
-l <- layout(matrix(c(1,2,3), nrow=3), heights=c(1,1,0.6))
+l <- layout(matrix(c(1,2,3), nrow=3), heights=c(1,1,0.65))
 
 par(mar=c(0.5, 7, 0.5, 0.5))
 plot(1, type="n", xlim=c(1, length(methods)), ylim=c(0,1), axes=F, xlab="", ylab="")
 
 plotRegionMethod <- function(index, data){
 	method <- methods[index]
+	if(method == "precluster" && data == "nseqs_red"){
+		method <- "pred"
+	}
+
 	subset <- error_summary[error_summary$method==method,]
-	regions <- gsub("V", "", subset$region)
-	text(x=getJitter(index), y=subset[,data], label=regions, cex=1.25)
+
+
+	numbers <- subset[,data]
+	if(data == "nseqs_red") {
+		numbers <- 100 - numbers
+	}
+
[…]

### 2016-02-18 doc/paper/rebuttal_v1.md
Add introductory paragraph

@@
+Because the manuscript has been nearly entirely rewritten, we did not think it was appropriate to submit a version of the new manuscript that indicated the changes - the entire manuscript was changed.
+We appreciate the insights and suggestions from the original reviewers and apologize for taking so long to resubmit the manuscript.
+We took their advice to heart and regenerated the data using the P6C4 PacBio chemistry.
+
 ## Reviewer 1
@@
 Part of the design of this figure was that the regions generally overlap on top of each other indicating that the effectiveness of the methods is reproducible.
-We have colored the numbers and increased the horizontal jitter.
+We have replaced the numbers with colored symbols and increased the horizontal jitter.
 

### 2016-02-18 Schloss_PacBio16S_PeerJ_2016.Rmd
Add introductory paragraph

@@
 All sequencing data were curated using mothur (v1.36)[@Schloss2009a] and analyzed using the R programming language [@R2016].
-The raw data can be obtained from the Sequence Read Archive at NCBI under accession SRP051686, which are associated with BioProject PRJNXXXXX.
+The raw data can be obtained from the Sequence Read Archive at NCBI under accession SRP051686, which are associated with BioProject PRJNA271568.
+This accession and bioproject also contain data from the same samples sequenced using the P4-C2 chemistry.
 Several specific features were incorporated into mothur to facilitate the analysis of PacBio sequence data.

### 2016-02-18 Schloss_PacBio16S_PeerJ_2016.Rmd
Create TeX intermediate files

@@
   pdf_document:
+    keep_tex: true
     includes:
@@
 
-## Funding statement
-This study was supported by grants from the NIH (R01HG005975, R01GM099514 and P30DK034933 to PDS and U54HG004973 to SKH).
-
 \newpage

### 2016-02-18 submission/Schloss_PacBio16S_PeerJ_2016.tex
Create TeX intermediate files

\documentclass[11pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1.0in]{geometry}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system},
            colorlinks=true,
            citecolor=blue,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{0}

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\newcommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}
  \title{\textbf{Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA
sequencing system}}
  \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
  \author{}
  \preauthor{}\postauthor{}
  \date{}
  \predate{}\postdate{}

\usepackage{helvet} % Helvetica font
\renewcommand*\familydefault{\sfdefault} % Use the sans serif version of the font
\usepackage[T1]{fontenc}

\usepackage[none]{hyphenat}

\usepackage{setspace}
\doublespacing
\setlength{\parskip}{1em}

\usepackage{lineno}

\usepackage{pdfpages}

% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\begin{document}
\maketitle

\vspace{35mm}

\textbf{Running title:} 16S rRNA genes sequencing with PacBio

\vspace{15mm}

\textbf{Authors:} Patrick D. Schloss\textsuperscript{1\#}, Matthew L.
Jenior\textsuperscript{1}, Charles C. Koumpouras\textsuperscript{1},
Sarah L. Westcott\textsuperscript{1}, and Sarah K.
Highlander\textsuperscript{2}

\vspace{40mm}

\(\dagger\) To whom correspondence should be addressed:
\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}

1. Department of Microbiology and Immunology, 1500 W. Medical Center,
University of Michigan, Ann Arbor, MI 48109

2. Genomic Medicine, J. Craig Venter Institute, 4120 Capricorn Lane, La
Jolla, CA 92307

\newpage

\linenumbers

\subsection{Abstract}\label{abstract}

Over the past 10 years, microbial ecologists have largely abandoned
sequencing 16S rRNA genes by the Sanger sequencing method and have
instead adopted highly parallelized sequencing platforms. These new
platforms, such as 454 and Illumina's MiSeq, have allowed researchers to
obtain millions of high quality, but short sequences. The result of the
added sequencing depth has been significant improvements in experimental
design. The tradeoff has been the decline in the number of full-length
reference sequences that are deposited into databases. To overcome this
problem, we tested the ability of the PacBio Single Molecule, Real-Time
(SMRT) DNA sequencing platform to generate sequence reads from the 16S
rRNA gene. We generated sequencing data from the V4, V3-V5, V1-V3,
V1-V5, V1-V6, and V1-V9 variable regions from within the 16S rRNA gene
using DNA from a synthetic mock community and natural samples collected
from human feces, mouse feces, and soil. The mock community allowed us
to assess the actual sequencing error rate and how that error rate
changed when different curation methods were applied. We developed a
simple method based on sequence characteristics and quality scores to
reduce the observed error rate for the V1-V9 region from 0.69 to
0.027\%. This error rate is comparable to what has been observed for the
shorter reads generated by 454 and Illumina's MiSeq sequencing
platforms. Although the per base sequencing cost is still significantly
more than that of MiSeq, the prospect of supplementing reference
databases with full-length sequences from organisms below the limit of
detection from the Sanger approach is exciting.

\textbf{\emph{Keywords:}} Microbial ecology, bioinformatics, sequencing
error

\newpage

\subsection{Introduction}\label{introduction}

Advances in sequencing technologies over the past 10 years have
introduced considerable advances to the field of microbial ecology.
Clone-based Sanger sequencing of the 16S rRNA gene has largely been
replaced by various platforms produced by 454/Roche (e.g. Sogin et al.
(2006)), Illumina (e.g. Gloor et al. (2010)), and IonTorrent (e.g.
Jünemann et al. (2012)). It was once common to sequence fewer than 100
16S rRNA gene sequences from several samples using the Sanger approach
(e.g. Mccaig et al. (1999)). Now it is common to generate thousands of
sequences from each of several hundred samples (Consortium, 2012). The
advance in throughput has come at the cost of read length. Sanger
sequencing regularly generated 800 nt per read and because the DNA was
cloned, it was possible to obtain multiple reads per fragment to yield a
full-length sequence from a representative single molecule. At
approximately \$8 (US) per sequencing read, most researchers have
effectively decided that full-length sequences are not worth the
increased cost relative to the cost of more recently developed
approaches. There is still a clear need to generate high-throughput
full-length sequence reads that are of sufficient quality that they can
be used as references for analyses based on obtaining short sequence
reads.

Historically, all sequencing platforms were created to primarily perform
genome sequencing. When sequencing a genome, it is assumed that the same
base of DNA will be sequenced multiple times and the consensus of
multiple sequence reads is used to generate contigs. Thus, an individual
base call may have a high error rate, but the consensus sequence will
have a low error rate. To sequence the 16S rRNA gene researchers use
conserved primers to amplify a sub-region from within the gene that is
isolated from many organisms. Because the fragments are not cloned, it
is not possible to obtain high sequence coverage from the same DNA
molecule using these platforms. To reduce sequencing error rates it has
become imperative to develop stringent sequence curation and denoising
algorithms (Schloss, Gevers \& Westcott, 2011; Kozich et al., 2013).
There has been a tradeoff between read length, number of reads per
sample, and the error rate. For instance, we recently demonstrated that
using the Illumina MiSeq and the 454 Titanium platforms the raw error
rate varies between 1 and 2\% (Schloss, Gevers \& Westcott, 2011; Kozich
et al., 2013). Yet, it was possible to obtain error rates below 0.02\%
by adopting various denoising algorithms; however, the resulting
fragments were only 250-nt long. In the case of 454 Titanium, extending
the length of the fragment introduces length-based errors and in the
case of the Illumina MiSeq, increasing the length of the fragment
reduces the overlap between the read pairs reducing the ability of each
read to mutually reduce the sequencing error. Inadequate denoising of
sequencing reads can have many negative effects including limited
ability to identify chimeras (Haas et al., 2011; Edgar et al., 2011) and
inflation of alpha- and beta-diversity metrics (Kunin et al., 2010; Huse
et al., 2010; Schloss, Gevers \& Westcott, 2011; Kozich et al., 2013).
Illumina's MiSeq plaform enjoys widespread use in the field because of
the ability to sequence 15-20 million fragments that can be distributed
across hundreds of samples for less than \$5000 (US).

As these sequencing platforms have grown in popularity, there has been a
decline in the number of full-length 16S rRNA genes being deposited into
GenBank that could serve as references for sequence classification,
phylogenetic analyses, and primer and probe design. This is particularly
frustrating since the technologies have significantly improved our
ability to detect and identify novel populations for which we lack
full-length reference sequences. A related problem is the perceived
limitation that the short reads generated by the 454 and Illumina
platforms cannot be reliably classified to the genus or species level.
Previous investigators have utilized simulations to demonstrate that
increased read lengths usually increase the accuracy and sensitivity of
classification against reference databases (Wang et al., 2007; Liu et
al., 2008; Werner et al., 2011). There is clearly a need to develop
sequencing technologies that will allow researchers to generate high
quality full-length 16S rRNA gene sequences in a high throughput manner.

New advances in single molecule sequencing technologies are being
developed to address this problem. One approach uses a random barcoding
strategy to fragment, sequence, and assemble full-length amplicons using
Illumina's HiSeq platform (Miller et al., 2013; Burke \& Darling, 2014).
Although the algorithms appear to have minimized the risk of assembly
chimeras, it is unclear what the sequencing error rate is by this
approach. An alternative is the use of single molecule technologies that
offer read lengths that are thousands of bases long. Although the Oxford
Nanopore Technology has been used to sequence 16S rRNA genes
(Benítez-Páez, Portune \& Sanz, 2016), the the platform produced by
Pacific Biosciences (PacBio) has received wider attention for this
application (Fichot \& Norman, 2013; Mosher et al., 2013, 2014; Schloss
et al., 2015; Singer et al., 2016). The PacBio Single Molecule,
Real-Time (SMRT) DNA Sequencing System ligates hairpin adapters
(i.e.~SMRTbells) to the ends of double-stranded DNA. Although the DNA
molecule is linear, the adapters effectively circularize the DNA
allowing the sequencing polymerase to process around the molecule
multiple times (Au et al., 2012). According to Pacific Biosciences the
platform is able to generate median read lengths longer than 8 kb with
the P6-C4 chemistry; however, the single pass error rate is
approximately 15\%. Given the circular nature of the DNA fragment, the
full read length can be used to cover the DNA fragment multiple times
resulting in a reduced error rate. Therefore, one should be able to
obtain multiple coverage of the full 16S rRNA gene at a reduced error
rate.

Despite the opportunity to potentially generate high-quality full-length
sequences, it is surprising that the Pacific Biosciences platform has
not been more widely adopted for sequencing 16S rRNA genes. Previous
studies utilizing the technology have removed reads with mismatched
primers and barcodes, ambiguous base calls, and low quality scores
(Fichot \& Norman, 2013) or screened sequences based on the predicted
error rate (Singer et al., 2016). Others have utilized the platform
without describing the bioinformatic pipeline that was utilized (Mosher
et al., 2013, 2014). The only study to report the error rate of the
platform for sequencing 16S rRNA genes with a mock community used the
P4-C2 chemistry and obtained an error rate of 0.32\%, which is 16-fold
higher than has been observed using the MiSeq or 454 platforms (Schloss,
Gevers \& Westcott, 2011; Kozich et al., 2013). In the current study, we
assessed the quality of data generated by the PacBio sequencer using the
improved P6-C4 chemistry and on-sequencer data processing. The goal was
to determine whether this strategy could fill the need for generating
high-quality, full-length sequence data on par with other platforms. We
hypothesized that by modulating the 16S rRNA gene fragment length we
could alter the read depth and obtain reads longer than are currently
available by the 454 and Illumina platforms but with the same quality.
To test this hypothesis, we developed a sequence curation pipeline that
was optimized by reducing the sequencing error rate of a mock bacterial
community with known composition. The resulting pipeline was then
applied to 16S rRNA gene fragments that were isolated from soil and
human and mouse feces.

\subsection{Materials and Methods}\label{materials-and-methods}

\textbf{\emph{Community DNA.}} We utilized genomic DNA isolated from
four communities. These same DNA extracts were previously used to
develop an Illumina MiSeq-based sequencing strategy (Kozich et al.,
2013). Briefly, we used a ``Mock Community'' composed of genomic DNA
from 21 bacterial strains: \emph{Acinetobacter baumannii} ATCC 17978,
\emph{Actinomyces odontolyticus} ATCC 17982, \emph{Bacillus cereus} ATCC
10987, \emph{Bacteroides vulgatus} ATCC 8482, \emph{Clostridium
beijerinckii} ATCC 51743, \emph{Deinococcus radiodurans} ATCC 13939,
\emph{Enterococcus faecalis} ATCC 47077, \emph{Escherichia coli} ATCC
70096, \emph{Helicobacter pylori} ATCC 700392, \emph{Lactobacillus
gasseri} ATCC 33323, \emph{Listeria monocytogenes} ATCC BAA-679,
\emph{Neisseria meningitidis} ATCC BAA-335, \emph{Porphyromonas
gingivalis} ATCC 33277, \emph{Propionibacterium acnes} DSM 16379,
\emph{Pseudomonas aeruginosa} ATCC 47085, \emph{Rhodobacter sphaeroides}
ATCC 17023, \emph{Staphylococcus aureus} ATCC BAA-1718,
\emph{Staphylococcus epidermidis} ATCC 12228, \emph{Streptococcus
agalactiae} ATCC BAA-611, \emph{Streptococcus mutans} ATCC 700610,
\emph{Streptococcus pneumoniae} ATCC BAA-334. The mock community DNA is
available through BEI resources (v3.1, HM-278D). Genomic DNAs from the
three other communities were obtained using the MO BIO PowerSoil DNA
extraction kit. The human and mouse fecal samples were obtained using
protocols that were reviewed and approved by the University Committee on
Use and Care of Animals (Protocol \#PRO00004877) and the Institutional
Review Board at the University of Michigan (Protocol \#HUM00057066). The
human stool donor provided informed consent.

\textbf{\emph{Library generation and sequencing.}} The DNAs were each
amplified in triplicate using barcoded primers targeting the V4, V1-V3,
V3-V5, V1-V5, V1-V6, and V1-V9 variable regions (Table 1). The primers
were synthesized so that the 5' end of the forward and reverse primers
were each tagged with paired 16-nt symmetric barcodes
(\url{https://github.com/PacificBiosciences/Bioinformatics-Training/wiki/Barcoding-with-SMRT-Analysis-2.3})
to allow multiplexing of samples within a single sequencing run. Methods
describing PCR, amplicon cleanup, and pooling were described previously
(Kozich et al., 2013). The SMRTbell adapters were ligated onto the PCR
products and the libraries were sequenced by Pacific Biosciences using
the P6-C4 chemistry on a PacBio RS II SMRT DNA Sequencing System.
Diffusion Loading was used for regions V4, V1-V3, and V3-V5 and MagBead
loading was used for regions V1-V5, V1-V6, and V1-V9. Each region was
sequenced separately using movies ranging in length between 180 and 360
minutes. The sequences were processed using pbccs (v.3.0.1;
\url{https://github.com/PacificBiosciences/pbccs}), which generates
predicted error rates using a proprietary algorithm.

\textbf{\emph{Data analysis.}} All sequencing data were curated using
mothur (v1.36)(Schloss et al., 2009) and analyzed using the R
programming language (R Core Team, 2016). The raw data can be obtained
from the Sequence Read Archive at NCBI under accession SRP051686, which
are associated with BioProject PRJNA271568. This accession and
bioproject also contain data from the same samples sequenced using the
P4-C2 chemistry. Several specific features were incorporated into mothur
to facilitate the analysis of PacBio sequence data. First, because
non-ambiguous base calls are assigned to Phred quality scores of zero,
the consensus fastq files were parsed so that scores of zero were
interpreted as corresponding to an ambiguous base call (i.e.~N) in the
fastq.info command using the pacbio=T option. Second, because the
consensus sequence can be generated in the forward and reverse
complement orientations, a checkorient option was added to the trim.seqs
command in order to identify the proper orientation. These features were
incorporated into mothur v.1.30. Because chimeric molecules can be
generated during PCR and would artificially inflate the sequencing
error, it was necessary to remove these data prior to assessing the
error rate. Because we knew the true sequences for the strains in the
mock community we could calculate all possible chimeras between strains
in the mock community (\emph{in silico} chimeras). If a sequence read
was 3 or more nucleotides more similar to an \emph{in silico} chimera
than it was to a non-chimeric reference sequence, it was classified as a
chimera and removed from further consideration. Identification of
\emph{in silico} chimeras and calculation of sequencing error rates was
performed using the seq.error command in mothur (Schloss, Gevers \&
Westcott, 2011). \emph{De novo} chimera detection was also performed on
the mock and other sequence data using the abundance-based algorithm
implemented in UCHIME (Edgar et al., 2011). Sequences sequences were
aligned against a SILVA-based reference alignment (Pruesse et al., 2007)
using a profile-based aligner (Schloss, 2009) and were classified
against the SILVA (v123) (Pruesse et al., 2007), RDP (v10)(Cole et al.,
2013), and greengenes (v13\_8\_99)(Werner et al., 2011) reference
taxonomies using a negative Bayesian classifier implemented within
mothur (Wang et al., 2007). Sequences were assigned to operational
taxonomic units using the average neighbor clustering algorithm with a
3\% distance threshold (Schloss \& Westcott, 2011). Detailed methods
including this paper as an R markdown file are available as a public
online repository
(\url{http://github.com/SchlossLab/Schloss_PacBio16S_PeerJ_2016}).

\subsection{Results and Discussion}\label{results-and-discussion}

\textbf{\emph{The PacBio error profile and a basic sequence curation
procedure.}} To build a sequence curation pipeline, we first needed to
characterize the error rate associated with sequencing the 16S rRNA
gene. Using the consensus sequence obtained from at least 3 sequencing
passes of each fragment, we observed an average sequencing error rate of
0.65\%. Insertions, deletions, and substitutions accounted for 31.2,
17.9, and 50.9\% of the errors, respectively. The substitution errors
were equally likely and all four bases were equally likely to cause
insertion errors. Interestingly, guanines (39.4\%) and adenines (24.3\%)
were more likely to be deleted than cytosines (18.3\%) or thymidines
(18.0\%). The PacBio quality values varied between 2 and 93.
Surprisingly, the percentage of base calls that had the maximum quality
value did not vary among correct base calls (80.5\%), substitutions
(80.0\%), and insertions (80.4\%). It was not possible to use the
individual base quality scores could not be used to screen sequence
quality as has been possible in past studies using the Phred-based
quality scores that accompany data generated using the 454 and Illumina
technologies. We did observe a strong correlation between our observed
error rate and the predicted error rate as calculated by the PacBio
software (Pearson's R: -0.67; Figure 1A).

We established a simple curation procedure by culling any sequence that
had a string of the same base repeated more than 8 times or did not
start and end at the expected alignment coordinates for that region of
the 16S rRNA gene. This reduced the experiment-wide error rate from 0.68
to 0.65\%. This basic procedure resulted in the removal of between 0.714
(V1-V3) and 9.47 (V1-V9)\% of the reads (Table 2). Although the
percentage of reads removed increased with the length of the fragment,
there was no obvious relationship between fragment length and error rate
(Figure 2).

\textbf{\emph{Identifying correlates of increased sequencing error.}} In
contrast to the 454 and Illumina-based platforms where the sequencing
quality decays with length, the consensus sequencing approach employed
by the PacBio sequencer is thought to generate a uniform distribution of
errors. This makes it impossible to simply trim sequences to high
quality regions. Therefore, we sought to identify characteristics within
sequences that would allow us to identify and remove those sequences
with errors using three different approaches. First, we hypothesized
that errors in the barcode and primer would be correlated with the error
rate for the entire sequence. We observed a strong relationship between
the number of mismatches to the barcodes and primers and the error rate
of the rest of the sequence fragment (Figure 1B). Although allowing no
mismatches to the barcodes and primers yielded the lowest error rate,
that stringent criterion removed a large fraction of the reads from the
dataset. Allowing at most one mismatch only marginally increased the
error rate while retaining more sequences in the dataset (Figure 2).
Second, we hypothesized that increased sequencing coverage should yield
lower error rates. We found that once we had obtained 10-fold coverage
of the fragments, the error rate did not change appreciably (Figure 1C).
When we compared the error rates of reads with at least 10-fold coverage
to those with less coverage, we reduced the error rate by 8.48 to
37.08\% (Figure 2). Third, based on the observed correlation between the
predicted and observed error rates, we sought to identify a minimum
predicted error rate that would allow us to reduce the observed error
rate. The average observed error rate for sequences with predicted error
rates between 0.01 and 0.10\% was linear. We decided to use a threshold
of 0.01\% because a large number of sequence reads were lost when we
used a smaller threshold. When we used this threshold, we were able to
reduce the error rate by 51.4 to 70.0\% (Figure 2). Finally, we
quantified the effect of combining filters. We found that any
combination of filters that included the predicted error rate threshold
had the most significant impact on reducing the observed error rate.
Furthermore, the inclusion of the mismatch and coverage filters had a
negligible impact on error rates, but had a significant impact on the
number of sequences included in the analysis. For instance, among the
V1-V9 data, requiring sequences to have a predicted error rate less than
0.01\% resulted in a 69.2\% reduction in error and resulted in the
removal of 53.5\% of the sequences. Adding the mismatch or coverage
filter had no effect on the reduction of error, but resulted in the
removal of 56.2 and 56.6 \% of the sequences, respectively. Use of all
filters had no impact on the reduction in the observed error rate, but
resulted in the removal of 59.1\% of the sequences. The remainder of
this paper only uses sequences with a predicted error rate less than
0.01\%.

\textbf{\emph{Pre-clustering sequences to further reduce sequencing
noise.}} Previously, we implemented a pre-clustering algorithm where
sequences were sorted by their abundance in decreasing order and rare
sequences are clustered with a more abundant sequence if the rare
sequences have fewer mismatches than a defined threshold when compared
to the more abundant sequence (Huse et al., 2010; Schloss, Gevers \&
Westcott, 2011). The recommended threshold was a 1-nt difference per
100-nt of sequence data. For example, the threshold for 250 bp fragment
from the V4 region would be 2 nt or 14 for the 1458 bp V1-V9 fragments.
This approach removes residual PCR and sequencing errors while not
overwhelming the resolution needed to identify OTUs that are based on a
3\% distance threshold. The tradeoff of this approach is that one would
be unable to differentiate V1-V9 sequences that truly differed by less
than 14 nt. When we applied this approach to our PacBio data, we
observed a reduction in the error rate between 33.0 (V1-V3) and 48.7\%
(V1-V9). The final error rates varied between 0.02 (V1-V5) and 0.2\%
(V4). The full-length (i.e.~V1-V9) fragments had an error rate of 0.03\%
(Figure 2; Table 2), this is similar to what we have previously observed
using the 454 and Illumina MiSeq platforms (0.02\%)(Schloss, Gevers \&
Westcott, 2011; Kozich et al., 2013).

\textbf{\emph{Effects of error rates on OTU assignments.}} The
sequencing error rate is known to affect the number of OTUs that are
observed (Schloss, Gevers \& Westcott, 2011). For each region, we
determined that if there were no chimeras or PCR or sequencing errors,
then we would expect to find 19 OTUs. When we achieved perfect chimera
removal, but allowed for PCR and sequencing errors, we observed between
0.5 (V1-V9) and 14.4 (V4) extra OTUs (Table 2). The range in the number
of extra OTUs was largely explained by the sequencing error rate
(Pearson's R=1.0). Next, we determined the number of OTUs that were
observed when we used UCHIME to identify chimeric sequences. Under these
more realistic conditions, we observed between 8.2 (V1-V5) and 29.9 (V4)
extra OTUs. Finally, we calculated the number of OTUs in the soil,
mouse, and human samples using the same pipeline with chimera detection
and removal based on the UCHIME algorithm. Surprisingly, there was not a
clear relationship across sample type and region. Again, we found that
there was a strong correlation between the number of observed OTUs and
the error rate for the mouse (R=0.95) and human samples (R=0.60). These
results underscore the effect of sequencing error on the inflation of
the number of observed OTUs.

\textbf{\emph{Classification varies by region, environment, and
database.}} We classified all of the sequence data we generated using
the naïve Bayesian classifier using the RDP, SILVA, and greengenes
reference taxonomies (Figure 3). In general, increasing the length of
the region improved the ability to assign the sequence to a genus or
species. Interestingly, each of the samples we analyzed varied in the
ability to assign its sequences to the depth of genus or species.
Furthermore, the reference database that did the best job of classifying
the sequences varied by sample type. For example, the SILVA reference
did the best for the human feces and soil samples and the RDP did the
best for the mouse feces samples. An advantage of the greengenes
database is that it contains information for 2,514 species-level
lineages for 11\% of the reference sequences; the other databases only
provided taxonomic data to the genus level. There was a modest
association between the length of the fragment and the ability to
classify sequences to the species-level for the human samples; there was
no such association for the mouse and soil samples. In fact, at most
6.2\% of the soil sequences and 4.3\% of the mouse sequences could be
classified to a species. These results indicate that the ability to
classify sequences to the genus or species level is a function of read
length, sample type, and the reference database.

\textbf{\emph{Sequencing errors are not random.}} Above, we described
that although there was no obvious bias in the substitution or insertion
rate, we did observe that guanines and adenines were more likely to be
deleted than cytosines or thymidines. This lack of randomness in the
error profile suggested that there might be a systematic non-random
distribution of the errors across the sequences. This would manifest
itself by the creation of duplicate sequences with the same error. We
identified all of the mock community sequences that had a 1-nt
difference to the true sequence (Figure 4). For these three regions,
between 70.7 and 88.9 of the sequences with 1-nt errors were only
observed once. We found that the frequency of the most abundant 1-nt
error paralleled the number of sequences. Surprisingly, the same 1-nt
error appeared 1,954 times (0.02\%) in the V1-V6 mock data and another
1-nt error appeared 1,070 times (0.03\%) in the V1-V9 mock data.
Contrary to previous reports (Carneiro et al., 2012; Koren et al.,
2012), these results indicate that reproducible errors occur with the
PacBio sequencing platform and that they can be quite abundant. Through
the use of the pre-clustering step described above these 1-nt errors
would be ameliorated; however, this result indicates that caution should
be used when attempting to use fine-scale OTU definitions.

\subsection{Conclusions}\label{conclusions}

The various sequencing platforms that are available to microbial
ecologists are able to fill unique needs and have their own strengths
and weaknesses. For sequencing the 16S rRNA gene, the 454 platform is
able to generate a moderate number of high-quality 500-nt sequence
fragments (error rates below 0.02\%) (Schloss, Gevers \& Westcott, 2011)
and the MiSeq platform is able to generate a large number of
high-quality 250-nt sequence fragments (error rates below 0.02\%)
(Kozich et al., 2013). The promise of the PacBio sequencing platform was
the generation of high-quality near full-length sequence fragments. As
we have shown in this study, it is possible to generate near full-length
sequences with error rates that are slightly higher, but comparable to
the other platforms (i.e.~0.03\%). With the exception of the V4 region
(0.2\%), the error rates were less than 0.07\%. When we considered the
shorter V4 region, which is similar in length to what is sequenced by
the MiSeq platform, the error rates we observed with the PacBio platform
were nearly 8-fold higher than what has previously been reported on the
other platforms. It was unclear why these shorter reads had such a high
error rate relative to the other regions. At this point, the primary
limitation of generating full-length sequences on the PacBio platform is
the cost of generating the data and accessibility to the sequencers.

The widespread adoption of the 454 and MiSeq platforms and decrease in
the use of Sanger sequencing for the 16S rRNA gene has resulted in a
decrease in the generation of the full-length reference sequences that
are needed for performing phylogenetic analyses and designing lineage
specific PCR primers and fluorescent \emph{in situ} hybridization (FISH)
probes. It remains to be determined whether the error rates we observed
for full-length sequences are prohibitive for these applications. We can
estimate the distribution of errors assuming that the errors follow a
binomial distribution along the length of the 1,500 nt gene with the
error rate that we achieved from the V1-V9 mock community data prior to
pre-clustering the sequences, which was 0.2\%. Under these conditions
one would expect 4.3\% of the sequences to have no errors and 50\% of
the sequences would have at least 3 errors. After applying the
pre-clustering denoising step, the error rate drops by 7.7-fold to
0.03\%. With this error rate, we would expect 66.3\% of the sequences to
have no sequencing errors. The cost of the reduced error rate is the
loss of resolution among closely related sequences.

Full-length sequences are frequently seen as a panacea to overcome the
limitations of taxonomic classifications. The ability to classify each
of our sample types benefited from the generation of full-length
sequences. It was interesting that the benefit varied by sample type and
database. For example, using the mouse libraries, the ability to
classify each of the regions differed by less than 5\% when classifying
against the SILVA and greengenes databases. The effect of the database
that was used was also interesting. The RDP database outperformed the
other databases for the mouse samples and the SILVA database
outperformed the others for the human and soil samples. The three
databases were equally effective for classifying the mock community.
Finally, since only the greengenes database provided species-level
information for its reference sequences it was the only database that
allowed for resolution of species-level classification. The sequences
from the mouse and soil libraries were not effectively classified to the
species level (all less than 10\%). In contrast, classification of the
human libraries resulted in more than 40\% of the sequences being
classified to a genus, regardless of the region. These data demonstrate
that for the samples we analyzed, the length of the sequence fragment
was not as significant a factor in classification as the choice of
database.

The development of newer sequencing technologies continue to advance and
there is justifiable excitement to apply these technologies to sequence
the 16S rRNA gene. Although it is clearly possible to generate
sequencing data from these various platforms, it is critical that we
assess the platforms for their ability to generate high quality data and
the particular niche that the new approach will fill. With this in mind,
it is essential that researchers utilize mock communities as part of
their experimental design so that they can quantify their error rates.
The ability to generate near full-length 16S rRNA gene sequences is an
exciting advance that will hopefully expand our ability to improve the
characterization of microbial communities.

\subsection{Acknowledgements}\label{acknowledgements}

The Genomic DNA from Microbial Mock Community A (Even, Low
Concentration, v3.1, HM-278D) was obtained through the NIH Biodefense
and Emerging Infections Research Resources Repository, NIAID, NIH as
part of the Human Microbiome Project.

\newpage

\subsection{Figures}\label{figures}

\textbf{Figure 1. Summary of errors in data generated using PacBio
sequencing platform to sequence various regions within the 16S rRNA
gene.} The predicted error rate using PacBio's sequence analysis
algorithm correlated well with the observed error rate (Pearson's R:
-0.67; A). Because of the large number of sequences, we randomly
selected 5\% of the data to show in panel A. The sequencing error rate
of the amplified gene fragments increased with mismatches to the
barcodes and primers (B). The sequencing error rate declined with
increased sequencing coverage; however, increasing the sequencing depth
beyond 10-fold coverage had no meaningful effect on the sequencing error
rate (C). The scale of they y-axis in B and C are the same.

\textbf{Figure 2. Change in error rate (A) and the percentage of
sequences that were retained (B) when using various sequence curation
methods.} The condition that was used for downstream analyses is
indicated by the star. The plotted numbers represent the region that was
sequenced. For example ``19'' represents the data for the V1-V9 region.

\textbf{Figure 3. Percentage of unique sequences that could be
classified.} Classifications were performed using taxonomy references
curated from the RDP, SILVA, or greengenes databases for the four types
of samples that were sequenced across the six regions from the 16S rRNA
gene. Only the greengenes taxonomy reference provided species-level
information.

\textbf{Figure 4. Percentage of 1-nt variants that occurred up to ten
times.} Sequences that were 1 nt different from the mock community
reference sequences were counted to determine the number of times each
variant appeared by region within the 16S rRNA gene.

\newpage

\subsection*{References}\label{references}
\addcontentsline{toc}{subsection}{References}

\hypertarget{refs}{}
\hypertarget{ref-Au2012}{}
Au KF., Underwood JG., Lee L., Wong WH. 2012. Improving PacBio long read
accuracy by short read alignment. \emph{PLoS ONE} 7:e46679. DOI:
\href{https://doi.org/10.1371/journal.pone.0046679}{10.1371/journal.pone.0046679}.

\hypertarget{ref-BentezPez2016}{}
Benítez-Páez A., Portune KJ., Sanz Y. 2016. Species-level resolution of
16S rRNA gene amplicons sequenced through the MinION portable nanopore
sequencer. \emph{GigaScience} 5. DOI:
\href{https://doi.org/10.1186/s13742-016-0111-z}{10.1186/s13742-016-0111-z}.

\hypertarget{ref-Burke2014}{}
Burke C., Darling AE. 2014. Resolving microbial microdiversity with high
accuracy full length 16S rRNA illumina sequencing. DOI:
\href{https://doi.org/10.1101/010967}{10.1101/010967}.

\hypertarget{ref-Carneiro2012}{}
Carneiro MO., Russ C., Ross MG., Gabriel SB., Nusbaum C., DePristo MA.
2012. Pacific biosciences sequencing technology for genotyping and
variation discovery in human data. \emph{BMC Genomics} 13:375. DOI:
\href{https://doi.org/10.1186/1471-2164-13-375}{10.1186/1471-2164-13-375}.

\hypertarget{ref-Cole2013}{}
Cole JR., Wang Q., Fish JA., Chai B., McGarrell DM., Sun Y., Brown CT.,
Porras-Alfaro A., Kuske CR., Tiedje JM. 2013. Ribosomal database
project: Data and tools for high throughput rRNA analysis. \emph{Nucleic
Acids Research} 42:D633--D642. DOI:
\href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.

\hypertarget{ref-HMP2012}{}
Consortium THM. 2012. Structure, function and diversity of the healthy
human microbiome. \emph{Nature} 486:207--214. DOI:
\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.

\hypertarget{ref-Edgar2011}{}
Edgar RC., Haas BJ., Clemente JC., Quince C., Knight R. 2011. UCHIME
improves sensitivity and speed of chimera detection.
\emph{Bioinformatics} 27:2194--2200. DOI:
\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.

\hypertarget{ref-Fichot2013}{}
Fichot EB., Norman RS. 2013. Microbial phylogenetic profiling with the
pacific biosciences sequencing platform. \emph{Microbiome} 1:10. DOI:
\href{https://doi.org/10.1186/2049-2618-1-10}{10.1186/2049-2618-1-10}.

\hypertarget{ref-Gloor2010}{}
Gloor GB., Hummelen R., Macklaim JM., Dickson RJ., Fernandes AD.,
MacPhee R., Reid G. 2010. Microbiome profiling by illumina sequencing of
combinatorial sequence-tagged PCR products. \emph{PLoS ONE} 5:e15406.
DOI:
\href{https://doi.org/10.1371/journal.pone.0015406}{10.1371/journal.pone.0015406}.

\hypertarget{ref-Haas2011}{}
Haas BJ., Gevers D., Earl AM., Feldgarden M., Ward DV., Giannoukos G.,
Ciulla D., Tabbaa D., Highlander SK., Sodergren E., Methe B., DeSantis
TZ., Petrosino JF., Knight R., Birren BW. 2011. Chimeric 16S rRNA
sequence formation and detection in sanger and 454-pyrosequenced PCR
amplicons. \emph{Genome Research} 21:494--504. DOI:
\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.

\hypertarget{ref-Huse2010}{}
Huse SM., Welch DM., Morrison HG., Sogin ML. 2010. Ironing out the
wrinkles in the rare biosphere through improved OTU clustering.
\emph{Environmental Microbiology} 12:1889--1898. DOI:
\href{https://doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.

\hypertarget{ref-Junemann2012}{}
Jünemann S., Prior K., Szczepanowski R., Harks I., Ehmke B., Goesmann
A., Stoye J., Harmsen D. 2012. Bacterial community shift in treated
periodontitis patients revealed by ion torrent 16S rRNA gene amplicon
sequencing. \emph{PLoS ONE} 7:e41606. DOI:
\href{https://doi.org/10.1371/journal.pone.0041606}{10.1371/journal.pone.0041606}.

\hypertarget{ref-Koren2012}{}
Koren S., Schatz MC., Walenz BP., Martin J., Howard JT., Ganapathy G.,
Wang Z., Rasko DA., McCombie WR., Jarvis ED., Phillippy AM. 2012. Hybrid
error correction and de novo assembly of single-molecule sequencing
reads. \emph{Nat Biotechnol} 30:693--700. DOI:
\href{https://doi.org/10.1038/nbt.2280}{10.1038/nbt.2280}.

\hypertarget{ref-Kozich2013}{}
Kozich JJ., Westcott SL., Baxter NT., Highlander SK., Schloss PD. 2013.
Development of a dual-index sequencing strategy and curation pipeline
for analyzing amplicon sequence data on the MiSeq illumina sequencing
platform. \emph{Applied and Environmental Microbiology} 79:5112--5120.
DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.

\hypertarget{ref-Kunin2010}{}
Kunin V., Engelbrektson A., Ochman H., Hugenholtz P. 2010. Wrinkles in
the rare biosphere: Pyrosequencing errors can lead to artificial
inflation of diversity estimates. \emph{Environmental Microbiology}
12:118--123. DOI:
\href{https://doi.org/10.1111/j.1462-2920.2009.02051.x}{10.1111/j.1462-2920.2009.02051.x}.

\hypertarget{ref-Liu2008}{}
Liu Z., DeSantis TZ., Andersen GL., Knight R. 2008. Accurate taxonomy
assignments from 16S rRNA sequences produced by highly parallel
pyrosequencers. \emph{Nucleic Acids Research} 36:e120--e120. DOI:
\href{https://doi.org/10.1093/nar/gkn491}{10.1093/nar/gkn491}.

\hypertarget{ref-Mccaig1999}{}
Mccaig AE., Glover LA., James., Prosser I. 1999. Molecular analysis of
bacterial community structure and diversity in unimproved and improved
upland grass pastures. \emph{Appl Environ Microbiol} 65:1721--1730.

\hypertarget{ref-Miller2013}{}
Miller CS., Handley KM., Wrighton KC., Frischkorn KR., Thomas BC.,
Banfield JF. 2013. Short-read assembly of full-length 16S amplicons
reveals bacterial diversity in subsurface sediments. \emph{PLoS ONE}
8:e56018. DOI:
\href{https://doi.org/10.1371/journal.pone.0056018}{10.1371/journal.pone.0056018}.

\hypertarget{ref-Mosher2013}{}
Mosher JJ., Bernberg EL., Shevchenko O., Kan J., Kaplan LA. 2013.
Efficacy of a 3rd generation high-throughput sequencing platform for
analyses of 16S rRNA genes from environmental samples. \emph{Journal of
Microbiological Methods} 95:175--181. DOI:
\href{https://doi.org/10.1016/j.mimet.2013.08.009}{10.1016/j.mimet.2013.08.009}.

\hypertarget{ref-Mosher2014}{}
Mosher JJ., Bowman B., Bernberg EL., Shevchenko O., Kan J., Korlach J.,
Kaplan LA. 2014. Improved performance of the PacBio SMRT technology for
16S rDNA sequencing. \emph{Journal of Microbiological Methods}
104:59--60. DOI:
\href{https://doi.org/10.1016/j.mimet.2014.06.012}{10.1016/j.mimet.2014.06.012}.

\hypertarget{ref-Pruesse2007}{}
Pruesse E., Quast C., Knittel K., Fuchs BM., Ludwig W., Peplies J.,
Glockner FO. 2007. SILVA: A comprehensive online resource for quality
checked and aligned ribosomal RNA sequence data compatible with ARB.
\emph{Nucleic Acids Research} 35:7188--7196. DOI:
\href{https://doi.org/10.1093/nar/gkm864}{10.1093/nar/gkm864}.

\hypertarget{ref-R2016}{}
R Core Team. 2016. R: A language and environment for statistical
computing.

\hypertarget{ref-Schloss2009a}{}
Schloss PD., Westcott SL., Ryabin T., Hall JR., Hartmann M., Hollister
EB., Lesniewski RA., Oakley BB., Parks DH., Robinson CJ., Sahl JW.,
Stres B., Thallinger GG., Horn DJV., Weber CF. 2009. Introducing mothur:
Open-source, platform-independent, community-supported software for
describing and comparing microbial communities. \emph{Applied and
Environmental Microbiology} 75:7537--7541. DOI:
\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.

\hypertarget{ref-Schloss2009b}{}
Schloss PD. 2009. A high-throughput DNA sequence aligner for microbial
ecology studies. \emph{PLoS ONE} 4:e8230. DOI:
\href{https://doi.org/10.1371/journal.pone.0008230}{10.1371/journal.pone.0008230}.

\hypertarget{ref-Schloss2015}{}
Schloss PD., Westcott SL., Jenior ML., Highlander; SK. 2015. Sequencing
16S rRNA gene fragments using the pacBio sMRT dNA sequencing system.
DOI:
\href{https://doi.org/10.7287/peerj.preprints.778v1}{10.7287/peerj.preprints.778v1}.

\hypertarget{ref-Schloss2011b}{}
Schloss PD., Westcott SL. 2011. Assessing and improving methods used in
operational taxonomic unit-based approaches for 16S rRNA gene sequence
analysis. \emph{Applied and Environmental Microbiology} 77:3219--3226.
DOI: \href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.

\hypertarget{ref-Schloss2011a}{}
Schloss PD., Gevers D., Westcott SL. 2011. Reducing the effects of PCR
amplification and sequencing artifacts on 16S rRNA-based studies.
\emph{PLoS ONE} 6:e27310. DOI:
\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.

\hypertarget{ref-Singer2016}{}
Singer E., Bushnell B., Coleman-Derr D., Bowman B., Bowers RM., Levy A.,
Gies EA., Cheng J-F., Copeland A., Klenk H-P., Hallam SJ., Hugenholtz
P., Tringe SG., Woyke T. 2016. High-resolution phylogenetic microbial
community profiling. \emph{The ISME Journal}. DOI:
\href{https://doi.org/10.1038/ismej.2015.249}{10.1038/ismej.2015.249}.

\hypertarget{ref-Sogin2006}{}
Sogin ML., Morrison HG., Huber JA., Welch DM., Huse SM., Neal PR.,
Arrieta JM., Herndl GJ. 2006. Microbial diversity in the deep sea and
the underexplored ``rare biosphere''. \emph{Proceedings of the National
Academy of Sciences} 103:12115--12120. DOI:
\href{https://doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.

\hypertarget{ref-Wang2007}{}
Wang Q., Garrity GM., Tiedje JM., Cole JR. 2007. Naive bayesian
classifier for rapid assignment of rRNA sequences into the new bacterial
taxonomy. \emph{Applied and Environmental Microbiology} 73:5261--5267.
DOI: \href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.

\hypertarget{ref-Werner2011}{}
Werner JJ., Koren O., Hugenholtz P., DeSantis TZ., Walters WA., Caporaso
JG., Angenent LT., Knight R., Ley RE. 2011. Impact of training sets on
classification of high-throughput bacterial 16s rRNA gene surveys.
\emph{The ISME Journal} 6:94--103. DOI:
\href{https://doi.org/10.1038/ismej.2011.82}{10.1038/ismej.2011.82}.

\end{document}

### 2016-03-16 submission/table_2.tex
Correct rules generating TeX files for Tables 1 and 2

--- 2016-02-18 submission/Schloss_PacBio16S_PeerJ_2016.tex
@@
-\documentclass[11pt,]{article}
+\documentclass[]{article}
 \usepackage{lmodern}
@@
             pdfauthor={},
-            pdftitle={Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA sequencing system},
+            pdftitle={},
             colorlinks=true,
@@
 \urlstyle{same}  % don't use monospace font for urls
-\usepackage{graphicx,grffile}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+\usepackage{longtable,booktabs}
 \setlength{\parindent}{0pt}
@@
 \setlength{\droptitle}{-2em}
-  \title{\textbf{Sequencing 16S rRNA gene fragments using the PacBio SMRT DNA
-sequencing system}}
-  \pretitle{\vspace{\droptitle}\centering\huge}
-  \posttitle{\par}
+  \title{}
+  \pretitle{\vspace{\droptitle}}
+  \posttitle{}
   \author{}
@@
 
-\usepackage[none]{hyphenat}
+\usepackage{multirow}
 
-\usepackage{setspace}
-\doublespacing
-\setlength{\parskip}{1em}
+\pagenumbering{gobble}
 
-\usepackage{lineno}
+\usepackage{array}
 
-\usepackage{pdfpages}
+\newcommand{\bigcell}[2]{\begin{tabular}{@{}#1@{}}#2\end{tabular}}
+
+\usepackage{rotating}
+
+\usepackage{caption}
 
@@
 
-\vspace{35mm}
+\captionsetup{labelformat=empty}
 
-\textbf{Running title:} 16S rRNA genes sequencing with PacBio
+\begin{table}[!ht]
+\centering
+\caption{\textbf{Table 2. Summary of the error rates and number of observed OTUs for each region.}} 
+\label{}
+{\small
+\begin{tabular}{cccccccccc}
+  \hline
+  \multirow{3}{*}{\textbf{Region}} & \multicolumn{3}{c}{\textbf{Error rate following (\%)}} & \multirow{1}{*}{\textbf{\bigcell{c}{Reads \\ remaining (\%)}}} & \multicolumn{5}{c}{\textbf{Average no.
+of OTUs\textsuperscript{a}}} \\ \cmidrule(r){2-4} \cmidrule(r){6-10}  & \textbf{Basic} & \textbf{Predicted} & \textbf{Precluster} & \textbf{} & \textbf{Mock\textsuperscript{b}} & \textbf{Mock\textsuperscript{c}} & \textbf{Soil} & \textbf{Mouse} & \textbf{Human} \\
+   \hline V4 & 0.77 & 0.35 & 0.158 & 62.58 & 33.4 & 48.9 & 629.4 & 212.6 & 116.9 \\ 
+  V1-V3 & 0.56 & 0.27 & 0.068 & 67.01 & 20.5 & 27.7 & 726.0 & 107.3 & 106.4 \\ 
+  V3-V5 & 0.74 & 0.22 & 0.035 & 63.15 & 21.1 & 31.2 &    NA & 112.3 &  83.8 \\ 
+  V1-V5 & 0.57 & 0.21 & 0.019 & 59.29 & 19.8 & 27.2 & 694.4 &  65.9 &  84.9 \\ 
+  V1-V6 & 0.60 & 0.21 & 0.031 & 59.90 & 20.0 & 46.5 & 693.0 &  75.8 &  89.4 \\ 
+  V1-V9 & 0.68 & 0.21 & 0.027 & 51.33 & 19.5 & 40.4 &    NA & 102.5 & 117.3 \\ 
+   \hline
+\end{tabular}
+}
+\end{table}
 
-\vspace{15mm}
+\textsuperscript{a} The number of OTUs is based on rarefaction of each
+sample to 1,000 sequences per sample; cells labeled ND reflect samples
+that did not have at least 1,000 sequences.
 
-\textbf{Authors:} Patrick D. Schloss\textsuperscript{1\#}, Matthew L.
-Jenior\textsuperscript{1}, Charles C. Koumpouras\textsuperscript{1},
-Sarah L. Westcott\textsuperscript{1}, and Sarah K.
-Highlander\textsuperscript{2}
+\textsuperscript{b} Number of OTUs in the mock community when all
+chimeras were removed; in the absence of chimeras and sequencing errors,
+there should be 19 OTUs for all three regions.
 
-\vspace{40mm}
-
-\(\dagger\) To whom correspondence should be addressed:
-\href{mailto:pschloss@umich.edu}{\nolinkurl{pschloss@umich.edu}}
-
-1. Department of Microbiology and Immunology, 1500 W. Medical Center,
-University of Michigan, Ann Arbor, MI 48109
-
-2. Genomic Medicine, J. Craig Venter Institute, 4120 Capricorn Lane, La
-Jolla, CA 92307
-
-\newpage
-
-\linenumbers
-
-\subsection{Abstract}\label{abstract}
-
-Over the past 10 years, microbial ecologists have largely abandoned
-sequencing 16S rRNA genes by the Sanger sequencing method and have
-instead adopted highly parallelized sequencing platforms. These new
-platforms, such as 454 and Illumina's MiSeq, have allowed researchers to
-obtain millions of high quality, but short sequences. The result of the
-added sequencing depth has been significant improvements in experimental
-design. The tradeoff has been the decline in the number of full-length
-reference sequences that are deposited into databases. To overcome this
-problem, we tested the ability of the PacBio Single Molecule, Real-Time
-(SMRT) DNA sequencing platform to generate sequence reads from the 16S
-rRNA gene. We generated sequencing data from the V4, V3-V5, V1-V3,
-V1-V5, V1-V6, and V1-V9 variable regions from within the 16S rRNA gene
-using DNA from a synthetic mock community and natural samples collected
-from human feces, mouse feces, and soil. The mock community allowed us
-to assess the actual sequencing error rate and how that error rate
-changed when different curation methods were applied. We developed a
-simple method based on sequence characteristics and quality scores to
-reduce the observed error rate for the V1-V9 region from 0.69 to
-0.027\%. This error rate is comparable to what has been observed for the
-shorter reads generated by 454 and Illumina's MiSeq sequencing
-platforms. Although the per base sequencing cost is still significantly
-more than that of MiSeq, the prospect of supplementing reference
-databases with full-length sequences from organisms below the limit of
-detection from the Sanger approach is exciting.
-
-\textbf{\emph{Keywords:}} Microbial ecology, bioinformatics, sequencing
-error
-
-\newpage
-
-\subsection{Introduction}\label{introduction}
-
-Advances in sequencing technologies over the past 10 years have
-introduced considerable advances to the field of microbial ecology.
-Clone-based Sanger sequencing of the 16S rRNA gene has largely been
-replaced by various platforms produced by 454/Roche (e.g. Sogin et al.
-(2006)), Illumina (e.g. Gloor et al. (2010)), and IonTorrent (e.g.
-Jünemann et al. (2012)). It was once common to sequence fewer than 100
-16S rRNA gene sequences from several samples using the Sanger approach
-(e.g. Mccaig et al. (1999)). Now it is common to generate thousands of
-sequences from each of several hundred samples (Consortium, 2012). The
-advance in throughput has come at the cost of read length. Sanger
-sequencing regularly generated 800 nt per read and because the DNA was
-cloned, it was possible to obtain multiple reads per fragment to yield a
-full-length sequence from a representative single molecule. At
-approximately \$8 (US) per sequencing read, most researchers have
-effectively decided that full-length sequences are not worth the
-increased cost relative to the cost of more recently developed
-approaches. There is still a clear need to generate high-throughput
-full-length sequence reads that are of sufficient quality that they can
-be used as references for analyses based on obtaining short sequence
-reads.
-
-Historically, all sequencing platforms were created to primarily perform
-genome sequencing. When sequencing a genome, it is assumed that the same
-base of DNA will be sequenced multiple times and the consensus of
-multiple sequence reads is used to generate contigs. Thus, an individual
-base call may have a high error rate, but the consensus sequence will
-have a low error rate. To sequence the 16S rRNA gene researchers use
-conserved primers to amplify a sub-region from within the gene that is
-isolated from many organisms. Because the fragments are not cloned, it
-is not possible to obtain high sequence coverage from the same DNA
-molecule using these platforms. To reduce sequencing error rates it has
-become imperative to develop stringent sequence curation and denoising
-algorithms (Schloss, Gevers \& Westcott, 2011; Kozich et al., 2013).
-There has been a tradeoff between read length, number of reads per
-sample, and the error rate. For instance, we recently demonstrated that
-using the Illumina MiSeq and the 454 Titanium platforms the raw error
-rate varies between 1 and 2\% (Schloss, Gevers \& Westcott, 2011; Kozich
-et al., 2013). Yet, it was possible to obtain error rates below 0.02\%
-by adopting various denoising algorithms; however, the resulting
-fragments were only 250-nt long. In the case of 454 Titanium, extending
-the length of the fragment introduces length-based errors and in the
-case of the Illumina MiSeq, increasing the length of the fragment
-reduces the overlap between the read pairs reducing the ability of each
-read to mutually reduce the sequencing error. Inadequate denoising of
-sequencing reads can have many negative effects including limited
-ability to identify chimeras (Haas et al., 2011; Edgar et al., 2011) and
-inflation of alpha- and beta-diversity metrics (Kunin et al., 2010; Huse
-et al., 2010; Schloss, Gevers \& Westcott, 2011; Kozich et al., 2013).
-Illumina's MiSeq plaform enjoys widespread use in the field because of
-the ability to sequence 15-20 million fragments that can be distributed
-across hundreds of samples for less than \$5000 (US).
-
-As these sequencing platforms have grown in popularity, there has been a
-decline in the number of full-length 16S rRNA genes being deposited into
-GenBank that could serve as references for sequence classification,
-phylogenetic analyses, and primer and probe design. This is particularly
-frustrating since the technologies have significantly improved our
-ability to detect and identify novel populations for which we lack
-full-length reference sequences. A related problem is the perceived
-limitation that the short reads generated by the 454 and Illumina
-platforms cannot be reliably classified to the genus or species level.
-Previous investigators have utilized simulations to demonstrate that
-increased read lengths usually increase the accuracy and sensitivity of
-classification against reference databases (Wang et al., 2007; Liu et
-al., 2008; Werner et al., 2011). There is clearly a need to develop
-sequencing technologies that will allow researchers to generate high
-quality full-length 16S rRNA gene sequences in a high throughput manner.
-
-New advances in single molecule sequencing technologies are being
-developed to address this problem. One approach uses a random barcoding
-strategy to fragment, sequence, and assemble full-length amplicons using
-Illumina's HiSeq platform (Miller et al., 2013; Burke \& Darling, 2014).
-Although the algorithms appear to have minimized the risk of assembly
-chimeras, it is unclear what the sequencing error rate is by this
-approach. An alternative is the use of single molecule technologies that
-offer read lengths that are thousands of bases long. Although the Oxford
-Nanopore Technology has been used to sequence 16S rRNA genes
-(Benítez-Páez, Portune \& Sanz, 2016), the the platform produced by
-Pacific Biosciences (PacBio) has received wider attention for this
-application (Fichot \& Norman, 2013; Mosher et al., 2013, 2014; Schloss
-et al., 2015; Singer et al., 2016). The PacBio Single Molecule,
-Real-Time (SMRT) DNA Sequencing System ligates hairpin adapters
-(i.e.~SMRTbells) to the ends of double-stranded DNA. Although the DNA
-molecule is linear, the adapters effectively circularize the DNA
-allowing the sequencing polymerase to process around the molecule
-multiple times (Au et al., 2012). According to Pacific Biosciences the
-platform is able to generate median read lengths longer than 8 kb with
-the P6-C4 chemistry; however, the single pass error rate is
-approximately 15\%. Given the circular nature of the DNA fragment, the
-full read length can be used to cover the DNA fragment multiple times
-resulting in a reduced error rate. Therefore, one should be able to
-obtain multiple coverage of the full 16S rRNA gene at a reduced error
-rate.
-
-Despite the opportunity to potentially generate high-quality full-length
-sequences, it is surprising that the Pacific Biosciences platform has
-not been more widely adopted for sequencing 16S rRNA genes. Previous
-studies utilizing the technology have removed reads with mismatched
-primers and barcodes, ambiguous base calls, and low quality scores
-(Fichot \& Norman, 2013) or screened sequences based on the predicted
-error rate (Singer et al., 2016). Others have utilized the platform
-without describing the bioinformatic pipeline that was utilized (Mosher
-et al., 2013, 2014). The only study to report the error rate of the
-platform for sequencing 16S rRNA genes with a mock community used the
-P4-C2 chemistry and obtained an error rate of 0.32\%, which is 16-fold
-higher than has been observed using the MiSeq or 454 platforms (Schloss,
-Gevers \& Westcott, 2011; Kozich et al., 2013). In the current study, we
-assessed the quality of data generated by the PacBio sequencer using the
-improved P6-C4 chemistry and on-sequencer data processing. The goal was
-to determine whether this strategy could fill the need for generating
-high-quality, full-length sequence data on par with other platforms. We
-hypothesized that by modulating the 16S rRNA gene fragment length we
-could alter the read depth and obtain reads longer than are currently
-available by the 454 and Illumina platforms but with the same quality.
-To test this hypothesis, we developed a sequence curation pipeline that
-was optimized by reducing the sequencing error rate of a mock bacterial
-community with known composition. The resulting pipeline was then
-applied to 16S rRNA gene fragments that were isolated from soil and
-human and mouse feces.
-
-\subsection{Materials and Methods}\label{materials-and-methods}
-
-\textbf{\emph{Community DNA.}} We utilized genomic DNA isolated from
-four communities. These same DNA extracts were previously used to
-develop an Illumina MiSeq-based sequencing strategy (Kozich et al.,
-2013). Briefly, we used a ``Mock Community'' composed of genomic DNA
-from 21 bacterial strains: \emph{Acinetobacter baumannii} ATCC 17978,
-\emph{Actinomyces odontolyticus} ATCC 17982, \emph{Bacillus cereus} ATCC
-10987, \emph{Bacteroides vulgatus} ATCC 8482, \emph{Clostridium
-beijerinckii} ATCC 51743, \emph{Deinococcus radiodurans} ATCC 13939,
-\emph{Enterococcus faecalis} ATCC 47077, \emph{Escherichia coli} ATCC
-70096, \emph{Helicobacter pylori} ATCC 700392, \emph{Lactobacillus
-gasseri} ATCC 33323, \emph{Listeria monocytogenes} ATCC BAA-679,
-\emph{Neisseria meningitidis} ATCC BAA-335, \emph{Porphyromonas
-gingivalis} ATCC 33277, \emph{Propionibacterium acnes} DSM 16379,
-\emph{Pseudomonas aeruginosa} ATCC 47085, \emph{Rhodobacter sphaeroides}
-ATCC 17023, \emph{Staphylococcus aureus} ATCC BAA-1718,
-\emph{Staphylococcus epidermidis} ATCC 12228, \emph{Streptococcus
-agalactiae} ATCC BAA-611, \emph{Streptococcus mutans} ATCC 700610,
-\emph{Streptococcus pneumoniae} ATCC BAA-334. The mock community DNA is
-available through BEI resources (v3.1, HM-278D). Genomic DNAs from the
-three other communities were obtained using the MO BIO PowerSoil DNA
-extraction kit. The human and mouse fecal samples were obtained using
-protocols that were reviewed and approved by the University Committee on
-Use and Care of Animals (Protocol \#PRO00004877) and the Institutional
-Review Board at the University of Michigan (Protocol \#HUM00057066). The
-human stool donor provided informed consent.
-
-\textbf{\emph{Library generation and sequencing.}} The DNAs were each
-amplified in triplicate using barcoded primers targeting the V4, V1-V3,
-V3-V5, V1-V5, V1-V6, and V1-V9 variable regions (Table 1). The primers
-were synthesized so that the 5' end of the forward and reverse primers
-were each tagged with paired 16-nt symmetric barcodes
-(\url{https://github.com/PacificBiosciences/Bioinformatics-Training/wiki/Barcoding-with-SMRT-Analysis-2.3})
-to allow multiplexing of samples within a single sequencing run. Methods
-describing PCR, amplicon cleanup, and pooling were described previously
-(Kozich et al., 2013). The SMRTbell adapters were ligated onto the PCR
-products and the libraries were sequenced by Pacific Biosciences using
-the P6-C4 chemistry on a PacBio RS II SMRT DNA Sequencing System.
-Diffusion Loading was used for regions V4, V1-V3, and V3-V5 and MagBead
-loading was used for regions V1-V5, V1-V6, and V1-V9. Each region was
-sequenced separately using movies ranging in length between 180 and 360
-minutes. The sequences were processed using pbccs (v.3.0.1;
-\url{https://github.com/PacificBiosciences/pbccs}), which generates
-predicted error rates using a proprietary algorithm.
-
-\textbf{\emph{Data analysis.}} All sequencing data were curated using
-mothur (v1.36)(Schloss et al., 2009) and analyzed using the R
-programming language (R Core Team, 2016). The raw data can be obtained
-from the Sequence Read Archive at NCBI under accession SRP051686, which
-are associated with BioProject PRJNA271568. This accession and
-bioproject also contain data from the same samples sequenced using the
-P4-C2 chemistry. Several specific features were incorporated into mothur
-to facilitate the analysis of PacBio sequence data. First, because
-non-ambiguous base calls are assigned to Phred quality scores of zero,
-the consensus fastq files were parsed so that scores of zero were
-interpreted as corresponding to an ambiguous base call (i.e.~N) in the
-fastq.info command using the pacbio=T option. Second, because the
-consensus sequence can be generated in the forward and reverse
-complement orientations, a checkorient option was added to the trim.seqs
-command in order to identify the proper orientation. These features were
-incorporated into mothur v.1.30. Because chimeric molecules can be
-generated during PCR and would artificially inflate the sequencing
-error, it was necessary to remove these data prior to assessing the
-error rate. Because we knew the true sequences for the strains in the
-mock community we could calculate all possible chimeras between strains
-in the mock community (\emph{in silico} chimeras). If a sequence read
-was 3 or more nucleotides more similar to an \emph{in silico} chimera
-than it was to a non-chimeric reference sequence, it was classified as a
-chimera and removed from further consideration. Identification of
-\emph{in silico} chimeras and calculation of sequencing error rates was
-performed using the seq.error command in mothur (Schloss, Gevers \&
-Westcott, 2011). \emph{De novo} chimera detection was also performed on
-the mock and other sequence data using the abundance-based algorithm
-implemented in UCHIME (Edgar et al., 2011). Sequences sequences were
-aligned against a SILVA-based reference alignment (Pruesse et al., 2007)
-using a profile-based aligner (Schloss, 2009) and were classified
-against the SILVA (v123) (Pruesse et al., 2007), RDP (v10)(Cole et al.,
-2013), and greengenes (v13\_8\_99)(Werner et al., 2011) reference
-taxonomies using a negative Bayesian classifier implemented within
-mothur (Wang et al., 2007). Sequences were assigned to operational
-taxonomic units using the average neighbor clustering algorithm with a
-3\% distance threshold (Schloss \& Westcott, 2011). Detailed methods
-including this paper as an R markdown file are available as a public
-online repository
-(\url{http://github.com/SchlossLab/Schloss_PacBio16S_PeerJ_2016}).
-
-\subsection{Results and Discussion}\label{results-and-discussion}
-
-\textbf{\emph{The PacBio error profile and a basic sequence curation
-procedure.}} To build a sequence curation pipeline, we first needed to
-characterize the error rate associated with sequencing the 16S rRNA
-gene. Using the consensus sequence obtained from at least 3 sequencing
-passes of each fragment, we observed an average sequencing error rate of
-0.65\%. Insertions, deletions, and substitutions accounted for 31.2,
-17.9, and 50.9\% of the errors, respectively. The substitution errors
-were equally likely and all four bases were equally likely to cause
-insertion errors. Interestingly, guanines (39.4\%) and adenines (24.3\%)
-were more likely to be deleted than cytosines (18.3\%) or thymidines
-(18.0\%). The PacBio quality values varied between 2 and 93.
-Surprisingly, the percentage of base calls that had the maximum quality
-value did not vary among correct base calls (80.5\%), substitutions
-(80.0\%), and insertions (80.4\%). It was not possible to use the
-individual base quality scores could not be used to screen sequence
-quality as has been possible in past studies using the Phred-based
-quality scores that accompany data generated using the 454 and Illumina
-technologies. We did observe a strong correlation between our observed
-error rate and the predicted error rate as calculated by the PacBio
-software (Pearson's R: -0.67; Figure 1A).
-
-We established a simple curation procedure by culling any sequence that
-had a string of the same base repeated more than 8 times or did not
-start and end at the expected alignment coordinates for that region of
-the 16S rRNA gene. This reduced the experiment-wide error rate from 0.68
-to 0.65\%. This basic procedure resulted in the removal of between 0.714
-(V1-V3) and 9.47 (V1-V9)\% of the reads (Table 2). Although the
-percentage of reads removed increased with the length of the fragment,
-there was no obvious relationship between fragment length and error rate
-(Figure 2).
-
-\textbf{\emph{Identifying correlates of increased sequencing error.}} In
-contrast to the 454 and Illumina-based platforms where the sequencing
-quality decays with length, the consensus sequencing approach employed
-by the PacBio sequencer is thought to generate a uniform distribution of
-errors. This makes it impossible to simply trim sequences to high
-quality regions. Therefore, we sought to identify characteristics within
-sequences that would allow us to identify and remove those sequences
-with errors using three different approaches. First, we hypothesized
-that errors in the barcode and primer would be correlated with the error
-rate for the entire sequence. We observed a strong relationship between
-the number of mismatches to the barcodes and primers and the error rate
-of the rest of the sequence fragment (Figure 1B). Although allowing no
-mismatches to the barcodes and primers yielded the lowest error rate,
-that stringent criterion removed a large fraction of the reads from the
-dataset. Allowing at most one mismatch only marginally increased the
-error rate while retaining more sequences in the dataset (Figure 2).
-Second, we hypothesized that increased sequencing coverage should yield
-lower error rates. We found that once we had obtained 10-fold coverage
-of the fragments, the error rate did not change appreciably (Figure 1C).
-When we compared the error rates of reads with at least 10-fold coverage
-to those with less coverage, we reduced the error rate by 8.48 to
-37.08\% (Figure 2). Third, based on the observed correlation between the
-predicted and observed error rates, we sought to identify a minimum
-predicted error rate that would allow us to reduce the observed error
-rate. The average observed error rate for sequences with predicted error
-rates between 0.01 and 0.10\% was linear. We decided to use a threshold
-of 0.01\% because a large number of sequence reads were lost when we
-used a smaller threshold. When we used this threshold, we were able to
-reduce the error rate by 51.4 to 70.0\% (Figure 2). Finally, we
-quantified the effect of combining filters. We found that any
-combination of filters that included the predicted error rate threshold
-had the most significant impact on reducing the observed error rate.
-Furthermore, the inclusion of the mismatch and coverage filters had a
-negligible impact on error rates, but had a significant impact on the
-number of sequences included in the analysis. For instance, among the
-V1-V9 data, requiring sequences to have a predicted error rate less than
-0.01\% resulted in a 69.2\% reduction in error and resulted in the
-removal of 53.5\% of the sequences. Adding the mismatch or coverage
-filter had no effect on the reduction of error, but resulted in the
-removal of 56.2 and 56.6 \% of the sequences, respectively. Use of all
-filters had no impact on the reduction in the observed error rate, but
-resulted in the removal of 59.1\% of the sequences. The remainder of
-this paper only uses sequences with a predicted error rate less than
-0.01\%.
-
-\textbf{\emph{Pre-clustering sequences to further reduce sequencing
-noise.}} Previously, we implemented a pre-clustering algorithm where
-sequences were sorted by their abundance in decreasing order and rare
-sequences are clustered with a more abundant sequence if the rare
-sequences have fewer mismatches than a defined threshold when compared
-to the more abundant sequence (Huse et al., 2010; Schloss, Gevers \&
-Westcott, 2011). The recommended threshold was a 1-nt difference per
-100-nt of sequence data. For example, the threshold for 250 bp fragment
-from the V4 region would be 2 nt or 14 for the 1458 bp V1-V9 fragments.
-This approach removes residual PCR and sequencing errors while not
-overwhelming the resolution needed to identify OTUs that are based on a
-3\% distance threshold. The tradeoff of this approach is that one would
-be unable to differentiate V1-V9 sequences that truly differed by less
-than 14 nt. When we applied this approach to our PacBio data, we
-observed a reduction in the error rate between 33.0 (V1-V3) and 48.7\%
-(V1-V9). The final error rates varied between 0.02 (V1-V5) and 0.2\%
-(V4). The full-length (i.e.~V1-V9) fragments had an error rate of 0.03\%
-(Figure 2; Table 2), this is similar to what we have previously observed
-using the 454 and Illumina MiSeq platforms (0.02\%)(Schloss, Gevers \&
-Westcott, 2011; Kozich et al., 2013).
-
-\textbf{\emph{Effects of error rates on OTU assignments.}} The
-sequencing error rate is known to affect the number of OTUs that are
-observed (Schloss, Gevers \& Westcott, 2011). For each region, we
-determined that if there were no chimeras or PCR or sequencing errors,
-then we would expect to find 19 OTUs. When we achieved perfect chimera
-removal, but allowed for PCR and sequencing errors, we observed between
-0.5 (V1-V9) and 14.4 (V4) extra OTUs (Table 2). The range in the number
-of extra OTUs was largely explained by the sequencing error rate
-(Pearson's R=1.0). Next, we determined the number of OTUs that were
-observed when we used UCHIME to identify chimeric sequences. Under these
-more realistic conditions, we observed between 8.2 (V1-V5) and 29.9 (V4)
-extra OTUs. Finally, we calculated the number of OTUs in the soil,
-mouse, and human samples using the same pipeline with chimera detection
-and removal based on the UCHIME algorithm. Surprisingly, there was not a
-clear relationship across sample type and region. Again, we found that
-there was a strong correlation between the number of observed OTUs and
-the error rate for the mouse (R=0.95) and human samples (R=0.60). These
-results underscore the effect of sequencing error on the inflation of
-the number of observed OTUs.
-
-\textbf{\emph{Classification varies by region, environment, and
-database.}} We classified all of the sequence data we generated using
-the naïve Bayesian classifier using the RDP, SILVA, and greengenes
-reference taxonomies (Figure 3). In general, increasing the length of
-the region improved the ability to assign the sequence to a genus or
-species. Interestingly, each of the samples we analyzed varied in the
-ability to assign its sequences to the depth of genus or species.
-Furthermore, the reference database that did the best job of classifying
-the sequences varied by sample type. For example, the SILVA reference
-did the best for the human feces and soil samples and the RDP did the
-best for the mouse feces samples. An advantage of the greengenes
-database is that it contains information for 2,514 species-level
-lineages for 11\% of the reference sequences; the other databases only
-provided taxonomic data to the genus level. There was a modest
-association between the length of the fragment and the ability to
-classify sequences to the species-level for the human samples; there was
-no such association for the mouse and soil samples. In fact, at most
-6.2\% of the soil sequences and 4.3\% of the mouse sequences could be
-classified to a species. These results indicate that the ability to
-classify sequences to the genus or species level is a function of read
-length, sample type, and the reference database.
-
-\textbf{\emph{Sequencing errors are not random.}} Above, we described
-that although there was no obvious bias in the substitution or insertion
-rate, we did observe that guanines and adenines were more likely to be
-deleted than cytosines or thymidines. This lack of randomness in the
-error profile suggested that there might be a systematic non-random
-distribution of the errors across the sequences. This would manifest
-itself by the creation of duplicate sequences with the same error. We
-identified all of the mock community sequences that had a 1-nt
-difference to the true sequence (Figure 4). For these three regions,
-between 70.7 and 88.9 of the sequences with 1-nt errors were only
-observed once. We found that the frequency of the most abundant 1-nt
-error paralleled the number of sequences. Surprisingly, the same 1-nt
-error appeared 1,954 times (0.02\%) in the V1-V6 mock data and another
-1-nt error appeared 1,070 times (0.03\%) in the V1-V9 mock data.
-Contrary to previous reports (Carneiro et al., 2012; Koren et al.,
-2012), these results indicate that reproducible errors occur with the
-PacBio sequencing platform and that they can be quite abundant. Through
-the use of the pre-clustering step described above these 1-nt errors
-would be ameliorated; however, this result indicates that caution should
-be used when attempting to use fine-scale OTU definitions.
-
-\subsection{Conclusions}\label{conclusions}
-
-The various sequencing platforms that are available to microbial
-ecologists are able to fill unique needs and have their own strengths
-and weaknesses. For sequencing the 16S rRNA gene, the 454 platform is
-able to generate a moderate number of high-quality 500-nt sequence
-fragments (error rates below 0.02\%) (Schloss, Gevers \& Westcott, 2011)
-and the MiSeq platform is able to generate a large number of
-high-quality 250-nt sequence fragments (error rates below 0.02\%)
-(Kozich et al., 2013). The promise of the PacBio sequencing platform was
-the generation of high-quality near full-length sequence fragments. As
-we have shown in this study, it is possible to generate near full-length
-sequences with error rates that are slightly higher, but comparable to
-the other platforms (i.e.~0.03\%). With the exception of the V4 region
-(0.2\%), the error rates were less than 0.07\%. When we considered the
-shorter V4 region, which is similar in length to what is sequenced by
-the MiSeq platform, the error rates we observed with the PacBio platform
-were nearly 8-fold higher than what has previously been reported on the
-other platforms. It was unclear why these shorter reads had such a high
-error rate relative to the other regions. At this point, the primary
-limitation of generating full-length sequences on the PacBio platform is
-the cost of generating the data and accessibility to the sequencers.
-
-The widespread adoption of the 454 and MiSeq platforms and decrease in
-the use of Sanger sequencing for the 16S rRNA gene has resulted in a
-decrease in the generation of the full-length reference sequences that
-are needed for performing phylogenetic analyses and designing lineage
-specific PCR primers and fluorescent \emph{in situ} hybridization (FISH)
-probes. It remains to be determined whether the error rates we observed
-for full-length sequences are prohibitive for these applications. We can
-estimate the distribution of errors assuming that the errors follow a
-binomial distribution along the length of the 1,500 nt gene with the
-error rate that we achieved from the V1-V9 mock community data prior to
-pre-clustering the sequences, which was 0.2\%. Under these conditions
-one would expect 4.3\% of the sequences to have no errors and 50\% of
-the sequences would have at least 3 errors. After applying the
-pre-clustering denoising step, the error rate drops by 7.7-fold to
-0.03\%. With this error rate, we would expect 66.3\% of the sequences to
-have no sequencing errors. The cost of the reduced error rate is the
-loss of resolution among closely related sequences.
-
-Full-length sequences are frequently seen as a panacea to overcome the
-limitations of taxonomic classifications. The ability to classify each
-of our sample types benefited from the generation of full-length
-sequences. It was interesting that the benefit varied by sample type and
-database. For example, using the mouse libraries, the ability to
-classify each of the regions differed by less than 5\% when classifying
-against the SILVA and greengenes databases. The effect of the database
-that was used was also interesting. The RDP database outperformed the
-other databases for the mouse samples and the SILVA database
-outperformed the others for the human and soil samples. The three
-databases were equally effective for classifying the mock community.
-Finally, since only the greengenes database provided species-level
-information for its reference sequences it was the only database that
-allowed for resolution of species-level classification. The sequences
-from the mouse and soil libraries were not effectively classified to the
-species level (all less than 10\%). In contrast, classification of the
-human libraries resulted in more than 40\% of the sequences being
-classified to a genus, regardless of the region. These data demonstrate
-that for the samples we analyzed, the length of the sequence fragment
-was not as significant a factor in classification as the choice of
-database.
-
-The development of newer sequencing technologies continue to advance and
-there is justifiable excitement to apply these technologies to sequence
-the 16S rRNA gene. Although it is clearly possible to generate
-sequencing data from these various platforms, it is critical that we
-assess the platforms for their ability to generate high quality data and
-the particular niche that the new approach will fill. With this in mind,
-it is essential that researchers utilize mock communities as part of
-their experimental design so that they can quantify their error rates.
-The ability to generate near full-length 16S rRNA gene sequences is an
-exciting advance that will hopefully expand our ability to improve the
-characterization of microbial communities.
-
-\subsection{Acknowledgements}\label{acknowledgements}
-
-The Genomic DNA from Microbial Mock Community A (Even, Low
-Concentration, v3.1, HM-278D) was obtained through the NIH Biodefense
-and Emerging Infections Research Resources Repository, NIAID, NIH as
-part of the Human Microbiome Project.
-
-\newpage
-
-\subsection{Figures}\label{figures}
-
-\textbf{Figure 1. Summary of errors in data generated using PacBio
-sequencing platform to sequence various regions within the 16S rRNA
-gene.} The predicted error rate using PacBio's sequence analysis
-algorithm correlated well with the observed error rate (Pearson's R:
--0.67; A). Because of the large number of sequences, we randomly
-selected 5\% of the data to show in panel A. The sequencing error rate
-of the amplified gene fragments increased with mismatches to the
-barcodes and primers (B). The sequencing error rate declined with
-increased sequencing coverage; however, increasing the sequencing depth
-beyond 10-fold coverage had no meaningful effect on the sequencing error
-rate (C). The scale of they y-axis in B and C are the same.
-
-\textbf{Figure 2. Change in error rate (A) and the percentage of
-sequences that were retained (B) when using various sequence curation
-methods.} The condition that was used for downstream analyses is
-indicated by the star. The plotted numbers represent the region that was
-sequenced. For example ``19'' represents the data for the V1-V9 region.
-
-\textbf{Figure 3. Percentage of unique sequences that could be
-classified.} Classifications were performed using taxonomy references
-curated from the RDP, SILVA, or greengenes databases for the four types
-of samples that were sequenced across the six regions from the 16S rRNA
-gene. Only the greengenes taxonomy reference provided species-level
-information.
-
-\textbf{Figure 4. Percentage of 1-nt variants that occurred up to ten
-times.} Sequences that were 1 nt different from the mock community
-reference sequences were counted to determine the number of times each
-variant appeared by region within the 16S rRNA gene.
-
-\newpage
-
-\subsection*{References}\label{references}
-\addcontentsline{toc}{subsection}{References}
-
-\hypertarget{refs}{}
-\hypertarget{ref-Au2012}{}
-Au KF., Underwood JG., Lee L., Wong WH. 2012. Improving PacBio long read
-accuracy by short read alignment. \emph{PLoS ONE} 7:e46679. DOI:
-\href{https://doi.org/10.1371/journal.pone.0046679}{10.1371/journal.pone.0046679}.
-
-\hypertarget{ref-BentezPez2016}{}
-Benítez-Páez A., Portune KJ., Sanz Y. 2016. Species-level resolution of
-16S rRNA gene amplicons sequenced through the MinION portable nanopore
-sequencer. \emph{GigaScience} 5. DOI:
-\href{https://doi.org/10.1186/s13742-016-0111-z}{10.1186/s13742-016-0111-z}.
-
-\hypertarget{ref-Burke2014}{}
-Burke C., Darling AE. 2014. Resolving microbial microdiversity with high
-accuracy full length 16S rRNA illumina sequencing. DOI:
-\href{https://doi.org/10.1101/010967}{10.1101/010967}.
-
-\hypertarget{ref-Carneiro2012}{}
-Carneiro MO., Russ C., Ross MG., Gabriel SB., Nusbaum C., DePristo MA.
-2012. Pacific biosciences sequencing technology for genotyping and
-variation discovery in human data. \emph{BMC Genomics} 13:375. DOI:
-\href{https://doi.org/10.1186/1471-2164-13-375}{10.1186/1471-2164-13-375}.
-
-\hypertarget{ref-Cole2013}{}
-Cole JR., Wang Q., Fish JA., Chai B., McGarrell DM., Sun Y., Brown CT.,
-Porras-Alfaro A., Kuske CR., Tiedje JM. 2013. Ribosomal database
-project: Data and tools for high throughput rRNA analysis. \emph{Nucleic
-Acids Research} 42:D633--D642. DOI:
-\href{https://doi.org/10.1093/nar/gkt1244}{10.1093/nar/gkt1244}.
-
-\hypertarget{ref-HMP2012}{}
-Consortium THM. 2012. Structure, function and diversity of the healthy
-human microbiome. \emph{Nature} 486:207--214. DOI:
-\href{https://doi.org/10.1038/nature11234}{10.1038/nature11234}.
-
-\hypertarget{ref-Edgar2011}{}
-Edgar RC., Haas BJ., Clemente JC., Quince C., Knight R. 2011. UCHIME
-improves sensitivity and speed of chimera detection.
-\emph{Bioinformatics} 27:2194--2200. DOI:
-\href{https://doi.org/10.1093/bioinformatics/btr381}{10.1093/bioinformatics/btr381}.
-
-\hypertarget{ref-Fichot2013}{}
-Fichot EB., Norman RS. 2013. Microbial phylogenetic profiling with the
-pacific biosciences sequencing platform. \emph{Microbiome} 1:10. DOI:
-\href{https://doi.org/10.1186/2049-2618-1-10}{10.1186/2049-2618-1-10}.
-
-\hypertarget{ref-Gloor2010}{}
-Gloor GB., Hummelen R., Macklaim JM., Dickson RJ., Fernandes AD.,
-MacPhee R., Reid G. 2010. Microbiome profiling by illumina sequencing of
-combinatorial sequence-tagged PCR products. \emph{PLoS ONE} 5:e15406.
-DOI:
-\href{https://doi.org/10.1371/journal.pone.0015406}{10.1371/journal.pone.0015406}.
-
-\hypertarget{ref-Haas2011}{}
-Haas BJ., Gevers D., Earl AM., Feldgarden M., Ward DV., Giannoukos G.,
-Ciulla D., Tabbaa D., Highlander SK., Sodergren E., Methe B., DeSantis
-TZ., Petrosino JF., Knight R., Birren BW. 2011. Chimeric 16S rRNA
-sequence formation and detection in sanger and 454-pyrosequenced PCR
-amplicons. \emph{Genome Research} 21:494--504. DOI:
-\href{https://doi.org/10.1101/gr.112730.110}{10.1101/gr.112730.110}.
-
-\hypertarget{ref-Huse2010}{}
-Huse SM., Welch DM., Morrison HG., Sogin ML. 2010. Ironing out the
-wrinkles in the rare biosphere through improved OTU clustering.
-\emph{Environmental Microbiology} 12:1889--1898. DOI:
-\href{https://doi.org/10.1111/j.1462-2920.2010.02193.x}{10.1111/j.1462-2920.2010.02193.x}.
-
-\hypertarget{ref-Junemann2012}{}
-Jünemann S., Prior K., Szczepanowski R., Harks I., Ehmke B., Goesmann
-A., Stoye J., Harmsen D. 2012. Bacterial community shift in treated
-periodontitis patients revealed by ion torrent 16S rRNA gene amplicon
-sequencing. \emph{PLoS ONE} 7:e41606. DOI:
-\href{https://doi.org/10.1371/journal.pone.0041606}{10.1371/journal.pone.0041606}.
-
-\hypertarget{ref-Koren2012}{}
-Koren S., Schatz MC., Walenz BP., Martin J., Howard JT., Ganapathy G.,
-Wang Z., Rasko DA., McCombie WR., Jarvis ED., Phillippy AM. 2012. Hybrid
-error correction and de novo assembly of single-molecule sequencing
-reads. \emph{Nat Biotechnol} 30:693--700. DOI:
-\href{https://doi.org/10.1038/nbt.2280}{10.1038/nbt.2280}.
-
-\hypertarget{ref-Kozich2013}{}
-Kozich JJ., Westcott SL., Baxter NT., Highlander SK., Schloss PD. 2013.
-Development of a dual-index sequencing strategy and curation pipeline
-for analyzing amplicon sequence data on the MiSeq illumina sequencing
-platform. \emph{Applied and Environmental Microbiology} 79:5112--5120.
-DOI: \href{https://doi.org/10.1128/aem.01043-13}{10.1128/aem.01043-13}.
-
-\hypertarget{ref-Kunin2010}{}
-Kunin V., Engelbrektson A., Ochman H., Hugenholtz P. 2010. Wrinkles in
-the rare biosphere: Pyrosequencing errors can lead to artificial
-inflation of diversity estimates. \emph{Environmental Microbiology}
-12:118--123. DOI:
-\href{https://doi.org/10.1111/j.1462-2920.2009.02051.x}{10.1111/j.1462-2920.2009.02051.x}.
-
-\hypertarget{ref-Liu2008}{}
-Liu Z., DeSantis TZ., Andersen GL., Knight R. 2008. Accurate taxonomy
-assignments from 16S rRNA sequences produced by highly parallel
-pyrosequencers. \emph{Nucleic Acids Research} 36:e120--e120. DOI:
-\href{https://doi.org/10.1093/nar/gkn491}{10.1093/nar/gkn491}.
-
-\hypertarget{ref-Mccaig1999}{}
-Mccaig AE., Glover LA., James., Prosser I. 1999. Molecular analysis of
-bacterial community structure and diversity in unimproved and improved
-upland grass pastures. \emph{Appl Environ Microbiol} 65:1721--1730.
-
-\hypertarget{ref-Miller2013}{}
-Miller CS., Handley KM., Wrighton KC., Frischkorn KR., Thomas BC.,
-Banfield JF. 2013. Short-read assembly of full-length 16S amplicons
-reveals bacterial diversity in subsurface sediments. \emph{PLoS ONE}
-8:e56018. DOI:
-\href{https://doi.org/10.1371/journal.pone.0056018}{10.1371/journal.pone.0056018}.
-
-\hypertarget{ref-Mosher2013}{}
-Mosher JJ., Bernberg EL., Shevchenko O., Kan J., Kaplan LA. 2013.
-Efficacy of a 3rd generation high-throughput sequencing platform for
-analyses of 16S rRNA genes from environmental samples. \emph{Journal of
-Microbiological Methods} 95:175--181. DOI:
-\href{https://doi.org/10.1016/j.mimet.2013.08.009}{10.1016/j.mimet.2013.08.009}.
-
-\hypertarget{ref-Mosher2014}{}
-Mosher JJ., Bowman B., Bernberg EL., Shevchenko O., Kan J., Korlach J.,
-Kaplan LA. 2014. Improved performance of the PacBio SMRT technology for
-16S rDNA sequencing. \emph{Journal of Microbiological Methods}
-104:59--60. DOI:
-\href{https://doi.org/10.1016/j.mimet.2014.06.012}{10.1016/j.mimet.2014.06.012}.
-
-\hypertarget{ref-Pruesse2007}{}
-Pruesse E., Quast C., Knittel K., Fuchs BM., Ludwig W., Peplies J.,
-Glockner FO. 2007. SILVA: A comprehensive online resource for quality
-checked and aligned ribosomal RNA sequence data compatible with ARB.
-\emph{Nucleic Acids Research} 35:7188--7196. DOI:
-\href{https://doi.org/10.1093/nar/gkm864}{10.1093/nar/gkm864}.
-
-\hypertarget{ref-R2016}{}
-R Core Team. 2016. R: A language and environment for statistical
-computing.
-
-\hypertarget{ref-Schloss2009a}{}
-Schloss PD., Westcott SL., Ryabin T., Hall JR., Hartmann M., Hollister
-EB., Lesniewski RA., Oakley BB., Parks DH., Robinson CJ., Sahl JW.,
-Stres B., Thallinger GG., Horn DJV., Weber CF. 2009. Introducing mothur:
-Open-source, platform-independent, community-supported software for
-describing and comparing microbial communities. \emph{Applied and
-Environmental Microbiology} 75:7537--7541. DOI:
-\href{https://doi.org/10.1128/aem.01541-09}{10.1128/aem.01541-09}.
-
-\hypertarget{ref-Schloss2009b}{}
-Schloss PD. 2009. A high-throughput DNA sequence aligner for microbial
-ecology studies. \emph{PLoS ONE} 4:e8230. DOI:
-\href{https://doi.org/10.1371/journal.pone.0008230}{10.1371/journal.pone.0008230}.
-
-\hypertarget{ref-Schloss2015}{}
-Schloss PD., Westcott SL., Jenior ML., Highlander; SK. 2015. Sequencing
-16S rRNA gene fragments using the pacBio sMRT dNA sequencing system.
-DOI:
-\href{https://doi.org/10.7287/peerj.preprints.778v1}{10.7287/peerj.preprints.778v1}.
-
-\hypertarget{ref-Schloss2011b}{}
-Schloss PD., Westcott SL. 2011. Assessing and improving methods used in
-operational taxonomic unit-based approaches for 16S rRNA gene sequence
-analysis. \emph{Applied and Environmental Microbiology} 77:3219--3226.
-DOI: \href{https://doi.org/10.1128/aem.02810-10}{10.1128/aem.02810-10}.
-
-\hypertarget{ref-Schloss2011a}{}
-Schloss PD., Gevers D., Westcott SL. 2011. Reducing the effects of PCR
-amplification and sequencing artifacts on 16S rRNA-based studies.
-\emph{PLoS ONE} 6:e27310. DOI:
-\href{https://doi.org/10.1371/journal.pone.0027310}{10.1371/journal.pone.0027310}.
-
-\hypertarget{ref-Singer2016}{}
-Singer E., Bushnell B., Coleman-Derr D., Bowman B., Bowers RM., Levy A.,
-Gies EA., Cheng J-F., Copeland A., Klenk H-P., Hallam SJ., Hugenholtz
-P., Tringe SG., Woyke T. 2016. High-resolution phylogenetic microbial
-community profiling. \emph{The ISME Journal}. DOI:
-\href{https://doi.org/10.1038/ismej.2015.249}{10.1038/ismej.2015.249}.
-
-\hypertarget{ref-Sogin2006}{}
-Sogin ML., Morrison HG., Huber JA., Welch DM., Huse SM., Neal PR.,
-Arrieta JM., Herndl GJ. 2006. Microbial diversity in the deep sea and
-the underexplored ``rare biosphere''. \emph{Proceedings of the National
-Academy of Sciences} 103:12115--12120. DOI:
-\href{https://doi.org/10.1073/pnas.0605127103}{10.1073/pnas.0605127103}.
-
-\hypertarget{ref-Wang2007}{}
-Wang Q., Garrity GM., Tiedje JM., Cole JR. 2007. Naive bayesian
-classifier for rapid assignment of rRNA sequences into the new bacterial
-taxonomy. \emph{Applied and Environmental Microbiology} 73:5261--5267.
-DOI: \href{https://doi.org/10.1128/aem.00062-07}{10.1128/aem.00062-07}.
-
-\hypertarget{ref-Werner2011}{}
-Werner JJ., Koren O., Hugenholtz P., DeSantis TZ., Walters WA., Caporaso
-JG., Angenent LT., Knight R., Ley RE. 2011. Impact of training sets on
-classification of high-throughput bacterial 16s rRNA gene surveys.
-\emph{The ISME Journal} 6:94--103. DOI:
-\href{https://doi.org/10.1038/ismej.2011.82}{10.1038/ismej.2011.82}.
+\textsuperscript{c} Number of OTUs in the mock community when chimeras
+were removed using UCHIME.
 
